### 2020-02-11 notebooks/article_draft.Rmd
Adds repo structure and manuscript notebook (WIP)

---
title: "Ch. 2 Dissertation draft"
output: html_notebook
bibliography: ../docs/citations.bib
link-citations: yes
csl: ../docs/american-geophysical-union.csl
---

## README

This is a draft version of what will eventually become the paper I submit as the second chapter of my dissertation. 
Eventually it would be good for this to be a fully reproducible paper that I can submit to a journal, but that might require more $\LaTeX$ knowledge than I currently possess.

## Introduction

This is where the introduction will go.

## Methods

This is where the methods will go.

```{r, echo=FALSE, message=FALSE}
# Attach required packages
library(tidyverse)
library(here)
library(sf)
library(tcltk)
library(readxl)
library(broom)
library(rgdal)
library(raster)
library(tictoc)

tic("Total runtime")

# Function to import all .csv files of long format PAIPR results from a given directory, and
# concatenate the results into a single tibble
import_PAIPR = function(input_dir) {
  files = list.files(input_dir, pattern = "*.csv", full.names = TRUE)
  data = tibble()
  for (f in files) {
    data_f = read_csv(f)
    data = rbind(data, data_f)
  }
  return(data)
}

tic("PAIPR data import")

# Import and format PAIPR-generated SMB results
data = import_PAIPR( here("data/gamma_20111109"))
data$Year = as.integer(data$Year)

# Generate descriptive statistics based on imported gamma-fitted parameters
alpha = data$gamma_shape
alpha[which(alpha<1)] = 1
beta = 1/data$gamma_scale
mode.accum = (alpha-1)/beta
var.accum = alpha/beta^2

# Make df of relevant data in long format
accum.long = data %>% select(Lat, Lon, elev, Year) %>% 
  mutate(accum = mode.accum, sd = sqrt(var.accum))

# Find duplicate annual accum records and average into single record
accum.long = accum.long %>% group_by(Lat, Lon, elev, Year) %>%
  summarise(accum = mean(accum, na.rm = TRUE), sd = mean(sd, na.rm = TRUE)) %>% ungroup()

# This removes years prior to a cutoff, and clips the results to only time series that extend to cutoff
yr.first = 1985
yr.last = 2007
accum.long = accum.long %>% filter(Year>=yr.first) %>% filter(Year <= yr.last)
data.cutoff = accum.long %>% group_by(Lat, Lon) %>% summarise(yr.start = min(Year))

# Nest year/accum data into list-columns for each location  (insert # prior to "filter" pipe to avoid cutoff)
accum.loc = nest(accum.long, data = c(Year, accum, sd)) %>% filter(data.cutoff$yr.start <= yr.first)
rm(data, data.cutoff, alpha, beta, mode.accum, var.accum)

toc()
```

```{r, echo=FALSE, message=FALSE}
# Create sf point object, where each row has a unique XYZ coordinate in space, with associated list-columns for the geometry and data variables
accum.sf = st_as_sf(accum.loc, coords = c("Lon", "Lat"), dim = "XY")

# Define the coordinate reference system used in original data
st_crs(accum.sf) = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84"

# Import basin shapefiles (in Antarctic Polar Sterographic EPSG:3031)
basins = st_read(here("data/ANT_Basins_IMBIE2_v1.6/ANT_Basins_IMBIE2_v1.6.shp"))

# Transform SMB sf object into APS projection
accum.PS = accum.sf %>% st_transform(st_crs(basins))
rm(accum.sf)
```

```{r}
# Import Antarctic outline
Ant.outline = st_read(here("data/Ant_basemap/Coastline_medium_res_polygon.shp"))

# Import the SAMBA database from .xlsx file
samba = read_excel(here("data/DGK_SMB_compilation.xlsx"), 
                   sheet = "Accumulation", na = c("", "NAN", "NaN"))

# Remove columns with  no data
keep.idx = function(x) any(!is.na(x))
samba = samba %>% select_if(keep.idx)

# Create tbl of Site locations
core.locs = tibble(Site = colnames(samba[,2:ncol(samba)]), 
                   Latitude = as.numeric(samba[1,2:ncol(samba)]), 
                   Longitude = as.numeric(samba[2,2:ncol(samba)]), 
                   Elevation = as.numeric(samba[3,2:ncol(samba)]))

# Convert samba database to long format and rm NA values
core_long = samba[4:nrow(samba),] %>% rename(Year = Name) %>% 
  pivot_longer(-Year, names_to = "Site", values_to = "Accum") %>% 
  filter(!is.na(Accum))

# Recombine accum data in long format with location data and rearrange
cores.accum = left_join(core_long, core.locs, by = "Site")  %>% arrange(Site, Year) %>% 
  select(Site, Latitude, Longitude, Elevation, Year, Accum) %>% 
  mutate(Year = as.integer(Year))

# Keep only accumulation years within specified range
cores.accum = cores.accum %>% filter(Year >= yr.first) %>% filter(Year <= yr.last)

# This clips the results to only time series that extend to a selected year
cores.cutoff = cores.accum %>% group_by(Site) %>% 
  summarise(yr.start = min(Year))

# Nest year/accum data into list-columns for each location  (insert # prior to "filter" pipe to avoid cutoff)
tmp.nest = nest(cores.accum, Data = c(Year, Accum)) %>% filter(cores.cutoff$yr.start <= yr.first)

# Add column for the length of Data records and remove cores shorter than 75% of given duration
cores.nest = mutate(tmp.nest, data.length = unlist(lapply(tmp.nest$Data, nrow))) %>% 
  filter(data.length >= (0.75*(yr.last-yr.first)))

# Convert to simple feature object
cores.PS = cores.nest %>% 
  st_as_sf(coords = c("Longitude", "Latitude"), dim = "XY", 
           crs = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84") %>%
  st_transform(st_crs(Ant.outline))
rm(samba, keep.idx, core_long, tmp.nest)
```

```{r, fig_inset, echo=FALSE}
# Crop cores to chosen spatial extent
cores.PS = cores.PS %>% st_crop(xmin=-1.5e6, xmax=-8e5, ymin=-7e5, ymax=1e5)

# Plot of current Antarctic cores (blue) and soon-to-be-added (red)
cores.plt = ggplot(cores.PS) + geom_sf(data = Ant.outline, color = 'grey') +
  geom_sf(data = sample_n(accum.PS, size = 1000), color = 'red') + 
  geom_sf(aes(text=Site), color = 'blue') + coord_sf(datum = st_crs(cores.PS)) + 
  theme_void()
print(cores.plt)
```

## Results

This is where the results will go.

```{r, echo=FALSE, message=FALSE}
# Find the mean accumulation at each data location across all years (this method strips NA 
# values first, so the exact temporal range differs by location)
accum.PS = accum.PS %>% 
  mutate(accum_mu = simplify(map(data, ~ mean(.x$accum, na.rm = TRUE))))

# Create non-referenced tibble of accumulation by location
accum.loc_PS = as_tibble(st_coordinates(accum.PS)) %>%
  rename(Easting = X, Northing = Y) %>% bind_cols(st_set_geometry(accum.PS, NULL))

# Find the mean accumulation at each core location across all years (this method strips NA 
# values first, so the exact temporal range differs by location)
cores.PS = cores.PS %>% 
  mutate(accum_mu = simplify(map(Data, ~ mean(.x$Accum, na.rm = TRUE))))

# Create non-referenced tibble of core accumulation by location
cores.loc_PS = as_tibble(st_coordinates(cores.PS)) %>%
  rename(Easting = X, Northing = Y) %>% bind_cols(st_set_geometry(cores.PS, NULL))
```

```{r, accum_map, echo=FALSE}
# Plot mean accum on spatial grid
accum.plt1 = ggplot(accum.loc_PS, aes(x = Easting, y = Northing, color = accum_mu)) + 
  geom_point() + 
  geom_point(data = cores.loc_PS, aes(x = Easting, y = Northing, color = accum_mu), 
             shape = 12) + 
  scale_color_viridis_c() + 
  # ggtitle("Mean accum in space") + 
  xlab("Easting") + ylab("Northing") + theme_bw()
print(accum.plt1)
```

Tests with bootstrapping to better constrain regression parameters (but currently too computationally expensive for the data set).

```{r}

# library(car)
# test = accum.loc_PS %>% sample_n(size=1000)
# test$trend = NA
# test$bias = NA
# test$St.err = NA
# test$conf2.5 = NA
# test$conf97.5 = NA
# # test$r2 = NA
# tic("Bootstrap timer:")
# for (i in 1:nrow(test)) {
#   data.i = test$data[[i]]
#   lm.i = lm(accum ~ Year, data = data.i, weights = 1/sd^2)
#   # bs = function(formula, data, indices) {
#   #   data_smp = data[indices,]
#   #   fit = lm(accum ~ Year, data=data_smp, weights = 1/sd^2)
#   #   return(coef(fit))
#   # }
#   # boot.i = boot(data=data.i, statistic=bs, R=1000)
#   boot.i = Boot(lm.i, R=200, ncores = 4)
#   boot.sum = summary(boot.i)
#   boot.conf = confint(boot.i)
#   
#   test$trend[i] = boot.sum$original[2]
#   test$bias[i] = boot.sum$bootBias[2]
#   test$St.err[i] = boot.sum$bootSE[2]
#   test$conf2.5[i] = boot.conf[2,1]
#   test$conf97.5[i] = boot.conf[2,2]
# }
# toc()
# 
# # test.lm = lm(accum ~ Year, data = test$data[[1]], weights = 1/sd^2)
# # test.boot = Boot(test.lm)
# # summary(test.boot)
# # 
# # test.lm = map(test$data, ~ lm(accum ~ Year, data = .x, weights = 1/sd^2))
# # 
# # boot.fun = function(x)  Boot(x, R=500)
# # test.boot = lapply(test.lm, boot.fun)
# # 
# # test.boot = test.lm %>% map(function(x) Boot(x, R=500))

```

```{r}

tic("Time series regressions")
# Create linear models for time series at each location in data
accum.lm_loc = map(accum.loc_PS$data, ~ lm(accum ~ Year, data = .x, weights = 1/sd^2))

# I should correct significance values based on autocorrelation in the data
# It would be interesting to investigate the 2nd derivative in the data (the trend through time of the trend). 
# One approach to this would be to fit a line through each time series, but iteratively shorten the time series. In this way, I can see if the trend changes (e.g. accelerates) during recent years.
# Another approach could be to perform linear fits of each time series over a ?10?-year moving window

# Extract relevant model data and store in nested tibble
accum.loc_data = select(accum.loc_PS, -data, -accum_mu) %>% 
  mutate(data = map(accum.lm_loc, 
                    ~ select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))

# Extract model parameter estimates and p values and store as location-based tibble
tmp.tidy = map(accum.lm_loc, ~ tidy(.x))
trend.loc = accum.loc_PS %>% 
  select(Easting, Northing, elev, accum_mu) %>% 
  mutate(intrcpt = simplify(map(tmp.tidy, "estimate") %>% map(1))) %>%
  mutate(coeff_yr = simplify(map(tmp.tidy, "estimate") %>% map(2))) %>% 
  mutate(std.err = simplify(map(tmp.tidy, "std.error") %>% map(2))) %>% 
  mutate(p.val_yr = simplify(map(tmp.tidy, "p.value") %>% map(2)))

##### Same thing as above, but for the cores

# Create linear models for time series at each location in data
cores.lm_loc = map(cores.loc_PS$Data, ~ lm(Accum ~ Year, data = .x))

# Extract relevant model data and store in nested tibble
cores.loc_data = select(cores.loc_PS, -Data, -accum_mu, -data.length) %>% 
  mutate(data = map(cores.lm_loc, 
                    ~ select(augment(.), Year, Accum, .fitted, .se.fit, .resid, .std.resid)))

# Extract model parameter estimates and p values and store as location-based tibble
tmp.tidy = map(cores.lm_loc, ~ tidy(.x))
cores.trend = cores.loc_PS %>% 
  select(Site, Easting, Northing, Elevation, accum_mu) %>% 
  mutate(intrcpt = simplify(map(tmp.tidy, "estimate") %>% map(1))) %>%
  mutate(coeff_yr = simplify(map(tmp.tidy, "estimate") %>% map(2))) %>% 
  mutate(std.err = simplify(map(tmp.tidy, "std.error") %>% map(2))) %>% 
  mutate(p.val_yr = simplify(map(tmp.tidy, "p.value") %>% map(2)))

# Scale trend values by the mean in-situ accumulation rate
trend.scaled = trend.loc %>% mutate(coeff_yr = coeff_yr/accum_mu, 
                                    std.err = std.err/accum_mu)
cores.scaled = cores.trend %>% mutate(coeff_yr = coeff_yr/accum_mu, 
                                      std.err = std.err/accum_mu)

rm(accum.lm_loc, accum.loc_data, cores.lm_loc, cores.loc_data, tmp.tidy)
toc()
```

```{r, trend_map, echo=FALSE}
# Plot scaled trends significant at 95% CL
trend.map = ggplot(trend.scaled, aes(x = Easting, y = Northing)) + 
  geom_point(aes(color = 100*coeff_yr)) + 
  geom_point(data = filter(trend.scaled, p.val_yr < 0.05) %>% sample_n(2000), 
             aes(x=Easting, y=Northing), shape = 4, color = 'black', size = 0.05) + 
  geom_point(data = cores.scaled, 
             aes(x = Easting, y = Northing, color = 100*coeff_yr), shape = 8, size = 3) + 
  geom_point(data = filter(cores.scaled, p.val_yr < 0.05), 
             aes(x=Easting, y=Northing), shape = 4, color = 'black', size = 0.05) + 
  lims(x = c(-1.5e6,-9e5), y = c(-6e5,0.35e5)) +
  scale_colour_gradient2() +
  labs(x = "Easting", y = "Northing", color = "%/yr") + 
  theme_bw()
print(trend.map)
```

```{r}
test = filter(accum.loc_PS, Northing < -4.5E5 & Easting < -1.2E6)

test.smp = test %>% sample_n(size = 250)
test.lm = map(test.smp$data, ~ lm(accum ~ Year, data = .x, weights = 1/sd^2))

test.data = select(test.smp, -data, -accum_mu) %>% 
  mutate(data = map(test.lm, 
                    ~ select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))

tmp.tidy = map(test.lm, ~ tidy(.x))
test.smp = test.smp %>% 
  mutate(coeff_yr = simplify(map(tmp.tidy, "estimate") %>% map(2))) %>% 
  mutate(std.err = simplify(map(tmp.tidy, "std.error") %>% map(2))) %>% 
  mutate(p.val_yr = simplify(map(tmp.tidy, "p.value") %>% map(2)))

test2 = test.smp %>% sample_n(size = 10) %>% unnest(cols = c(data))

ggplot(test2, aes(Year, accum)) + geom_line(aes(group = Easting, color = factor(Easting)))

```

### Comparison of mean accumulation and slope

This section uses 8-meter mosaic tiles from the Reference Elevation Model of Antarctica (REMA) to compare mean accumulation rate to surface slope [@howat_reference_2019].

```{r}
# Find all mosaics that intersect the accumulation dataset
index.file = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_Tile_Index_Rel1.1/REMA_Tile_Index_Rel1.1.shp"
dem.index = st_read(index.file)
ggplot() + geom_sf(data = Ant.outline, color = 'grey') + geom_sf(data = dem.index)
keep.idx = unique(unlist(st_intersects(accum.PS, dem.index)))
mosaic.download = dem.index[keep.idx,] %>% mutate(fileurl = as.character(fileurl))

# output.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/WAIS_divide")
output.dir = tk_choose.dir(caption = "Select directory in which to save REMA mosaic tiles")

# Download files from REMA ftp server
for (file in 1:nrow(mosaic.download)) {
  tmp.file = tempfile()
  dir.name = basename(dirname(mosaic.download$fileurl[file]))
  # dir.create(file.path(output.dir, dir.name))

  # If not already downloaded, download tile and extract
  if (!dir.exists(file.path(output.dir, dir.name))) {
    download.file(mosaic.download$fileurl[file], tmp.file)
    untar(tmp.file, exdir = file.path(output.dir, dir.name))
  }
  unlink(tmp.file)
}
```

```{r}
# tic("Raster analysis")
# 
# dem.file = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/WAIS_divide/r.slope.tif"
# dem.slope = raster(dem.file)
# accum.loc_PS = accum.loc_PS %>% mutate(slope = extract(dem.slope, accum.PS))
# 
# toc()
```

```{r}
toc()
```

# REFERENCES

### 2020-02-12 notebooks/article_draft.Rmd
Adds chunks to download REMA DEMs and calculate/extract slope and aspect

@@
 # Make df of relevant data in long format
-accum.long = data %>% select(Lat, Lon, elev, Year) %>% 
+accum.long = data %>% dplyr::select(Lat, Lon, elev, Year) %>% 
   mutate(accum = mode.accum, sd = sqrt(var.accum))
@@
 cores.accum = left_join(core_long, core.locs, by = "Site")  %>% arrange(Site, Year) %>% 
-  select(Site, Latitude, Longitude, Elevation, Year, Accum) %>% 
+  dplyr::select(Site, Latitude, Longitude, Elevation, Year, Accum) %>% 
   mutate(Year = as.integer(Year))
@@
 This is where the results will go.
+
+```{r}
+## This code chunk serves to find which REMA DEM 8-meter tiles intersect our dataset 
+## and downloads any tiles not already present in the selected directory "output.dir"
+
+# REMA.dir = tk_choose.dir(caption = "Select directory in which to save REMA mosaic tiles")
+REMA.dir = file.path("G:/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
+# REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
+
+# Find all mosaics that intersect the accumulation dataset
+index.file = file.path(dirname(REMA.dir), "REMA_Tile_Index_Rel1.1/REMA_Tile_Index_Rel1.1.shp")
+dem.index = st_read(index.file)
+# ggplot() + geom_sf(data = Ant.outline, color = 'grey') + geom_sf(data = dem.index)
+keep.idx = unique(unlist(st_intersects(accum.PS, dem.index)))
+mosaic.download = dem.index[keep.idx,] %>% mutate(fileurl = as.character(fileurl))
+
+# Download files from REMA ftp server
+for (file in 1:nrow(mosaic.download)) {
+  tmp.file = tempfile()
+  dir.name = basename(dirname(mosaic.download$fileurl[file]))
+
+  # If not already downloaded, download tile and extract
+  if (!dir.exists(file.path(REMA.dir, dir.name))) {
+    download.file(mosaic.download$fileurl[file], tmp.file)
+    untar(tmp.file, exdir = file.path(REMA.dir, dir.name))
+  }
+  unlink(tmp.file)
+}
+```
+
+```{r}
+tic("Raster extraction")
+
+# Function to find which accum points are within a given REMA 8-meter mosaic tile 
+# and extract elevation, slope, and aspect from the REMA raster data
+get_topo_vals = function(tile, accum.pts, tile.dir) {
+
+  # Logical to filter accum data to those members within the tile extent
+  pts.log = st_intersects(accum.pts, tile, sparse = FALSE)[,1]
+  accum.i = accum.pts %>% filter(pts.log)
+
+  # Generate file paths to DEM, slope raster and aspect raster
+  dem.path = file.path(tile.dir, as.character(tile$tile), 
+                        paste(as.character(tile$tile), "_8m_dem.tif", sep = ""))
+  slope.path = file.path(tile.dir, as.character(tile$tile), 
+                        paste(as.character(tile$tile), "_8m_slope.tif", sep = ""))
+  aspect.path = file.path(tile.dir, as.character(tile$tile), 
+                        paste(as.character(tile$tile), "_8m_aspect.tif", sep = ""))
+
+  # Check if slope/aspect rasters exist. If they don't, calculate them from DEM 
+  # and save to REMA directory
+  if (!file.exists(slope.path)) {
+    tile.dem = raster(dem.path)
+    tile.slope = terrain(tile.dem, opt='slope', unit='degrees')
+    writeRaster(tile.slope, filename = slope.path, format = "GTiff")
+    tile.aspect = terrain(tile.dem, opt='aspect', unit='degrees')
+    writeRaster(tile.aspect, filename = aspect.path, format = "GTiff")
+  }
+
+  # Generate raster stack of DEM, slope, and aspect 
+  # and extract raster values at accum points
+  tile.stack = stack(list(dem.path, slope.path, aspect.path))
+  stack.vals = extract(tile.stack, accum.i)
+
+  # Get index values of accum data subset within the global accum dataset
+  pts.idx = (1:length(pts.log))[pts.log]
+
+  # Return index positions and raster stack values as list object
+  return(list(pts.idx, stack.vals))
+}
+
+# Add variables for REMA elevation, slope, and aspect
+accum.PS = accum.PS %>% mutate(elev.REMA = NA, slope = NA, aspect = NA)
+
+for (i in 1:nrow(mosaic.download)) {
+
+  # Get current tile with coverage in dataset
+  tile = mosaic.download[i,]
+
+  # Calculate and extract elevation, slope, and aspect
+  vals.list = get_topo_vals(tile, accum.PS, REMA.dir)
+  topo.idx = vals.list[[1]]
+  topo.vals = vals.list[[2]]
+
+  # Assign topo values to accum tbl
+  accum.PS$elev.REMA[topo.idx] = topo.vals[,1]
+  accum.PS$slope[topo.idx] = topo.vals[,2]
+  accum.PS$aspect[topo.idx] = topo.vals[,3]
+}
+
+toc()
+```
 
@@
 # Extract relevant model data and store in nested tibble
-accum.loc_data = select(accum.loc_PS, -data, -accum_mu) %>% 
+accum.loc_data = dplyr::select(accum.loc_PS, -data, -accum_mu) %>% 
   mutate(data = map(accum.lm_loc, 
-                    ~ select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))
+                    ~ dplyr::select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))
 
@@
 trend.loc = accum.loc_PS %>% 
-  select(Easting, Northing, elev, accum_mu) %>% 
+  dplyr::select(Easting, Northing, elev, accum_mu) %>% 
   mutate(intrcpt = simplify(map(tmp.tidy, "estimate") %>% map(1))) %>%
@@
 # Extract relevant model data and store in nested tibble
-cores.loc_data = select(cores.loc_PS, -Data, -accum_mu, -data.length) %>% 
+cores.loc_data = dplyr::select(cores.loc_PS, -Data, -accum_mu, -data.length) %>% 
   mutate(data = map(cores.lm_loc, 
-                    ~ select(augment(.), Year, Accum, .fitted, .se.fit, .resid, .std.resid)))
+                    ~ dplyr::select(augment(.), Year, Accum, .fitted, .se.fit, .resid, .std.resid)))
 
@@
 cores.trend = cores.loc_PS %>% 
-  select(Site, Easting, Northing, Elevation, accum_mu) %>% 
+  dplyr::select(Site, Easting, Northing, Elevation, accum_mu) %>% 
   mutate(intrcpt = simplify(map(tmp.tidy, "estimate") %>% map(1))) %>%
@@
 
-test.data = select(test.smp, -data, -accum_mu) %>% 
+test.data = dplyr::select(test.smp, -data, -accum_mu) %>% 
   mutate(data = map(test.lm, 
-                    ~ select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))
+                    ~ dplyr::select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))
 
@@
 ```{r}
-# Find all mosaics that intersect the accumulation dataset
-index.file = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_Tile_Index_Rel1.1/REMA_Tile_Index_Rel1.1.shp"
-dem.index = st_read(index.file)
-ggplot() + geom_sf(data = Ant.outline, color = 'grey') + geom_sf(data = dem.index)
-keep.idx = unique(unlist(st_intersects(accum.PS, dem.index)))
-mosaic.download = dem.index[keep.idx,] %>% mutate(fileurl = as.character(fileurl))
-
-# output.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/WAIS_divide")
-output.dir = tk_choose.dir(caption = "Select directory in which to save REMA mosaic tiles")
-
-# Download files from REMA ftp server
-for (file in 1:nrow(mosaic.download)) {
-  tmp.file = tempfile()
-  dir.name = basename(dirname(mosaic.download$fileurl[file]))
-  # dir.create(file.path(output.dir, dir.name))
-
-  # If not already downloaded, download tile and extract
-  if (!dir.exists(file.path(output.dir, dir.name))) {
-    download.file(mosaic.download$fileurl[file], tmp.file)
-    untar(tmp.file, exdir = file.path(output.dir, dir.name))
-  }
-  unlink(tmp.file)
-}
-```
-
-```{r}
-# tic("Raster analysis")
-# 
-# dem.file = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/WAIS_divide/r.slope.tif"
-# dem.slope = raster(dem.file)
-# accum.loc_PS = accum.loc_PS %>% mutate(slope = extract(dem.slope, accum.PS))
-# 
-# toc()
-```
-
-```{r}
 toc()

### 2021-06-08 docs/grsl-analysis.py
Adds function for variogram; generates random spatial variance

@@ -20,6 +20,7 @@ import panel as pn
 import seaborn as sns
 import xarray as xr
 from xrspatial import hillshade
+from scipy.spatial.distance import pdist
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -271,6 +272,131 @@ plt_res = gv.Points(
 
 plt_res = plt_res.redim.range(accum_res=(res_min,res_max))
 # plt_res
+
+
+
+
+# %% Spatial coherence and variability
+
+def semi_vario(
+    points_gdf, lag_size, d_metric='euclidean', 
+    vars='all', scale=True):
+    """A function to calculate the semivariogram for values associated with a geoDataFrame of points.
+
+    Args:
+        points_gdf (geopandas.geodataframe.GeoDataFrame): Location of points and values associated with those points to use for calculating distances and lagged semivariance.
+        lag_size (int): The size of the lagging interval to use for binning semivariance values.
+        d_metric (str, optional): The distance metric to use when calculating pairwise distances in the geoDataFrame. Defaults to 'euclidean'.
+        vars (list of str, optional): The names of variables to use when calculating semivariance. Defaults to 'all'.
+        scale (bool, optional): Whether to perform normalization (relative to max value) on semivariance values. Defaults to True.
+
+    Returns:
+        pandas.core.frame.DataFrame: The calculated semivariance values for each chosen input. Also includes the lag interval (index), the average separation distance (dist), and the number of paired points within each interval (cnt). 
+    """
+
+    # Get column names if 'all' is selected for "vars"
+    if vars == "all":
+        vars = points_gdf.drop(
+            columns='geometry').columns
+
+    # Extact trace coordinates and calculate pairwise distance
+    locs_arr = np.array(
+        [points_gdf.geometry.x, points_gdf.geometry.y]).T
+    dist_arr = pdist(locs_arr, metric=d_metric)
+
+    # Calculate the indices used for each pairwise calculation
+    i_idx = np.empty(dist_arr.shape)
+    j_idx = np.empty(dist_arr.shape)
[…]

### 2021-06-08 docs/grsl-analysis.py
Optimizing lag distance interval

@@ -341,8 +341,9 @@ def semi_vario(
     else:
         gamma_df = gamma_vals
 
-    # Add distance and count values to output
+    # Add distance, lag center, and count values to output
     gamma_df['dist'] = df_groups['dist'].mean()
+    gamma_df['lag_cent'] = lags[1::]-lag_size//2
     gamma_df['cnt'] = df_groups['dist'].count()
 
     return gamma_df
@@ -350,7 +351,7 @@ def semi_vario(
 # %%
 
 gamma_df = semi_vario(
-    gdf_PAIPR, lag_size=100, vars=['accum_mu', 'accum_res'], 
+    gdf_PAIPR, lag_size=250, vars=['accum_mu', 'accum_res'], 
     scale=True)
 
 
@@ -367,36 +368,36 @@ gdf_noise['accum_res'] = np.random.normal(
     size=gdf_noise.shape[0])
 
 gamma_noise = semi_vario(
-    gdf_noise, lag_size=100, vars=['accum_mu', 'accum_res'], 
+    gdf_noise, lag_size=250, vars=['accum_mu', 'accum_res'], 
     scale=True)
 
 # %% Plots/exploration of semivariograms
 
-threshold = 50
-data1 = gamma_df.query('cnt > @threshold')
-data2 = gamma_noise.query('cnt > @threshold')
+# According to the discussion [here](https://stats.stackexchange.com/questions/361220/how-can-i-understand-these-variograms), I should limit my empirical variogram to no more than 1/2 my total domain
+threshold = gamma_df['dist'].max()/2
+data1 = gamma_df.query('dist <= @threshold')
+data2 = gamma_noise.query('dist <= @threshold')
 
 fig1, ax1 = plt.subplots()
 
 data1.plot(
     kind='scatter', ax=ax1, color='blue', 
-    x='dist', y='accum_mu', label='Real data')
+    x='lag_cent', y='accum_mu', label='Real data')
 data2.plot(
     kind='scatter', ax=ax1, color='red', 
-    x='dist', y='accum_mu', label='Noise')
+    x='lag_cent', y='accum_mu', label='Noise')
 plt.title('Mean accumulation variogram')
 
 fig2, ax2 = plt.subplots()
 
 data1.plot(
     kind='scatter', ax=ax2, color='blue', 
-    x='dist', y='accum_res', label='Real data')
+    x='lag_cent', y='accum_res', label='Real data')
 data2.plot(
[…]

### 2021-06-09 src/my_functions.py
Updates function for extracting topo values from REMA

@@ -542,23 +542,30 @@ def get_REMA(tile_idx, output_dir):
             print(f"REMA tile {f_dir.name} already exists locally, moving to next download")
     print("All requested files downloaded")
     
-def topo_vals(tile_dir, locations):
-    """
-    Extracts elevation, slope, and aspect values at given locations.
-    
-    Parameters:
-    tile_dir {pathlib.PosixPath}: The relative or absolute path to a directory containing REMA tile DSM, slope and aspect geotiffs.
-    trace_locs {geopandas.geodataframe.GeoDataFrame}: A geodataframe containing the locations at which to extract raster data. These data should have the same coordinate reference system as the raster data, with the geometries stored in a column named "geometry".
-
+def topo_vals(tile_dir, locations, slope=False, aspect=False):
+    """Extracts elevation, slope, and aspect values at given locations.
     Dependencies: Requires the rasterio (as rio) module and, by extension, GDAL binaries.
     Requires the geopandas module.
+
+    Args:
+        tile_dir (pathlib.PosixPath): The relative or absolute path to a directory containing REMA tile DSM, slope and aspect geotiffs.
+        locations (geopandas.geodataframe.GeoDataFrame): A geodataframe containing the locations at which to extract raster data. These data should have the geometries stored in a column named "geometry" (the default for geopandas).
+        slope (bool, optional): Whether to also extract slope values at points of interest. Defaults to False.
+        aspect (bool, optional): Whether to also extract aspect values at points of interest. Defaults to False.
+
+    Returns:
+        geopandas.geodataframe.GeoDataFrame: The same input geodataframe with topographic values appended.
     """
+
     # Preallocate elevation variable in geodf
     if 'Elev' not in locations.columns:
         locations['Elev'] = None
     # locations = (
     #     locations.assign(elev=None)
     #     .assign(slope=None).assign(aspect=None))
+
+    # Get initial gdf crs
+    crs_init = locations.crs
 
     # Ensure locations are in same crs as REMA (EPSG:3031)
     locations.to_crs(epsg=3031, inplace=True)
@@ -579,25 +586,30 @@ def topo_vals(tile_dir, locations):
     locations.Elev[tile_mask] = tile_vals[tile_mask]
     src.close()
 
-    # # Extract slope values for all points within tile
-    # tile_path = [
[…]

### 2021-06-09 docs/grsl-analysis.py
Improves code readability

@@ -273,132 +273,6 @@ plt_res = gv.Points(
 plt_res = plt_res.redim.range(accum_res=(res_min,res_max))
 # plt_res
 
-
-
-
-# %% Spatial coherence and variability
-
-def semi_vario(
-    points_gdf, lag_size, d_metric='euclidean', 
-    vars='all', scale=True):
-    """A function to calculate the semivariogram for values associated with a geoDataFrame of points.
-
-    Args:
-        points_gdf (geopandas.geodataframe.GeoDataFrame): Location of points and values associated with those points to use for calculating distances and lagged semivariance.
-        lag_size (int): The size of the lagging interval to use for binning semivariance values.
-        d_metric (str, optional): The distance metric to use when calculating pairwise distances in the geoDataFrame. Defaults to 'euclidean'.
-        vars (list of str, optional): The names of variables to use when calculating semivariance. Defaults to 'all'.
-        scale (bool, optional): Whether to perform normalization (relative to max value) on semivariance values. Defaults to True.
-
-    Returns:
-        pandas.core.frame.DataFrame: The calculated semivariance values for each chosen input. Also includes the lag interval (index), the average separation distance (dist), and the number of paired points within each interval (cnt). 
-    """
-
-    # Get column names if 'all' is selected for "vars"
-    if vars == "all":
-        vars = points_gdf.drop(
-            columns='geometry').columns
-
-    # Extact trace coordinates and calculate pairwise distance
-    locs_arr = np.array(
-        [points_gdf.geometry.x, points_gdf.geometry.y]).T
-    dist_arr = pdist(locs_arr, metric=d_metric)
-
-    # Calculate the indices used for each pairwise calculation
-    i_idx = np.empty(dist_arr.shape)
-    j_idx = np.empty(dist_arr.shape)
-    m = locs_arr.shape[0]
-    for i in range(m):
-        for j in range(m):
-            if i < j < m:
-                i_idx[m*i + j - ((i + 2)*(i + 1))//2] = i
-                j_idx[m*i + j - ((i + 2)*(i + 1))//2] = j
-
-    # Create dfs for paired-point values
[…]

### 2021-06-09 src/my_functions.py
Adds function to calculate slope/aspect from REMA; addresses bugs in topo_vals function

@@ -12,6 +12,7 @@ import rasterio as rio
 from sklearn.neighbors import BallTree
 from scipy import signal
 import statsmodels.tsa.stattools as tsa
+import richdem as rd
 
 # Function to import and concatenate PAIPR .csv files
 def import_PAIPR(input_dir):
@@ -541,7 +542,53 @@ def get_REMA(tile_idx, output_dir):
         else:
             print(f"REMA tile {f_dir.name} already exists locally, moving to next download")
     print("All requested files downloaded")
+
+
+def calc_topo(dem_path):
+    """
+    Calculates slope and aspect from given DEM and saves output.
+    The function checks to see whether a slope/aspect file has already been created so as to avoid needless processing.
     
+    Parameters:
+    dem_path (pathlib.PosixPath): The relative or absolute path to an input DEM file.
+
+    Dependencies: 
+    richdem module
+    GDAL binaries
+    pathlib module
+    """
+    slope_path = Path(
+        str(dem_path).replace("dem", "slope"))
+    aspect_path = Path(
+        str(dem_path).replace("dem", "aspect"))
+
+    if ((not slope_path.is_file()) or 
+            (not aspect_path.is_file())):
+        
+        # Load DEM
+        dem = rd.LoadGDAL(str(dem_path))
+        
+        # Calculate slope values from DEM
+        if not slope_path.is_file():
+            print(f"Calculating slope values for REMA tile {dem_path.parent.name}...")
+            slope = rd.TerrainAttribute(
+                dem, attrib='slope_riserun')
+            rd.SaveGDAL(str(slope_path), slope)
+        else:
+                print(f"Slope data already exist locally for REMA tile {dem_path.parent.name}. Checking for aspect data...")
+        
+        # Calculate aspect values from DEM
+        if not aspect_path.is_file():
+            print(f"Calculating aspect values for REMA tile {dem_path.parent.name}...")
+            aspect = rd.TerrainAttribute(dem, attrib='aspect')
+            rd.SaveGDAL(str(aspect_path), aspect)
+        else:
+            print(f"Aspect data already exist locally for REMA tile {dem_path.parent.name}. Moving to next tile...")
+
+    else:
+        print(f"Slope/aspect geotifs already exist locally for REMA tile {dem_path.parent.name}. Moving to next tile...")
+
+
 def topo_vals(tile_dir, locations, slope=False, aspect=False):
[…]

### 2021-06-09 docs/grsl-analysis.py
Downloads/calculates/extracts topo data from REMA

@@ -26,7 +26,7 @@ from scipy.spatial.distance import pdist
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
 
 # Set project root directory
-ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
+ROOT_DIR = Path(__file__).parents[1]
 
 # Import custom project functions
 import sys
@@ -355,8 +355,9 @@ paipr_1to1_plt_comb = paipr_1to1_plt + site_res_plt
 # %% Spatial coherence and variability
 
 def vario(
-    points_gdf, lag_size, d_metric='euclidean', 
-    vars='all', scale=True):
+    points_gdf, lag_size, 
+    d_metric='euclidean', vars='all', 
+    stationarize=False, scale=True):
     """A function to calculate the semivariogram for values associated with a geoDataFrame of points.
 
     Args:
@@ -390,6 +391,10 @@ def vario(
                 i_idx[m*i + j - ((i + 2)*(i + 1))//2] = i
                 j_idx[m*i + j - ((i + 2)*(i + 1))//2] = j
 
+
+    if stationarize:
+        pass
+    
     # Create dfs for paired-point values
     i_vals = points_gdf[vars].iloc[i_idx].reset_index(
         drop=True)
@@ -428,7 +433,7 @@ def vario(
 
 # Calculate variogram for PAIPR results
 gamma_df = vario(
-    gdf_PAIPR, lag_size=250, vars=['accum_mu', 'accum_res'], 
+    gdf_PAIPR, lag_size=200, vars=['accum_mu', 'accum_res'], 
     scale=True)
 
 # Generate random noise with matched characteristics to PAIPR results
@@ -445,13 +450,13 @@ gdf_noise['accum_res'] = np.random.normal(
 
 # Calculate variogram for random noise results
 gamma_noise = vario(
-    gdf_noise, lag_size=250, vars=['accum_mu', 'accum_res'], 
+    gdf_noise, lag_size=200, vars=['accum_mu', 'accum_res'], 
     scale=True)
 
 # %% Plots/exploration of semivariograms
 
[…]

### 2021-06-10 docs/submission-docs/oib-repeat/response.md
Adds draft of reviewer response document

# Response to reviewers

We appreciate the time the reviewers took to offer suggestions for improvement, with their inclusion in the manuscript materially improving both the underlying science and its readability.

General response to editor and reviewers discussing the main thrusts of the edits.

In addition to these general changes, we also include specific responses to each of the comments provided by the reviewers, along with the locations in the updated manuscript of the implemented changes.

## Reviewer 1

> The amount of data used is extremely limited compared to the amount collected and available from OIB. It would be useful to significantly expand the amount of data used in this analysis and to discuss cross-region variability of the results.

We agree with the reviewer that it would be useful to expand the analysis to a wider region, with a discussion of cross-region variability.
We, however, feel such an investigation is outside the intended scope of the current study.
We have limited our analysis to regions where we have higher confidence in our depth-density modeling results (due to adequate *in-situ* density record coverage) and where repeat OIB flights with a number of years of separation between collection time exist.
Although research is ongoing to extend the density model used to wider regions of Antarctica, it is not possible to complete and implement this work in the short time frame required for revisions.
We have added additional text to the manuscript more explicitly stating the limits of this study, with hope to further expand it in the future.

> There should be some discussion and analysis of the effect of layer-echo signal to noise ratio in the precision of the layer measurements and the resulting accumulation estimates.

We were uncertain of the precise issue raised by this comment.
We understood it to mean...
The original paper outlining the PAIPR method discusses these concerns in greater detail.
*Discussion of uncertainty incorporation in PAIPR.*
We have included additional discussion in the manuscript to better highlight these points, and hope such discussion will satisfactorily resolve this concern.

## Reviewer 2

> The 2011 and 2016 transects were actually flown with slightly different radar parameters. 2011 had a bandwidth of 4 GHz (2-6 GHz sweep) and 2016 has a bandwidth of 6 GHz (2-8 GHz sweep). (If you’re using the CReSIS MATLAB files from their ftp site, you can check these parameters in the param_get_heights.radar.wfs variable that comes with the data.) This isn’t a huge difference in range resolution (~4 cm vs ~2.7 cm in snow), but I would expect some impact on the layer tracking results.This seems worth mentioning, both as a potential source of both the error and bias and because your results now actually suggest that the accumulation estimates are fairly robust to these kinds of small variations in the bandwidth/range resolution.

Blah blah blah.

> How did you choose the $S_Q = 0.50$ cutoff and the 10% quality cutoff for radargrams? Are those values you would expect to be reasonable for any data sets, or is there a logic to their choice that would be a useful for a reader to understand if they were interested in using these methods themself?

Blah blah blah.

> It was not totally clear to me why there is uncertainty/variability in the value of the logistic function parameters and how the ranges are chosen for the Monte Carlo implementation. If this can be clarified in a sentence or two, that  would be  nice  for readers  like  myself who are  not  fully familiar with the PAIPR method.

Blah blah blah.

> Reference of the in-situ accumulation estimates that are used for the bias distributions? It would be nice to have a quick sentence explaining where that data comes from and a pointerto any papers on it.

We believe poor wording on our part led to some confusion regarding this point, as more than one reviewer mentioned it.
We do not use other methods to compare to radar results (e.g. ice cores, climate model outputs, etc.) but instead focus exclusively on

> Is the mean in-situ accumulation rate one value over the entire study area?

Blah blah blah.

> I find the spatial correlation in bias shown in Fig. 3 very interesting and I think it’s quite an important result that suggests that mean error from sparse crossovers is not a robust way to quantify uncertainty in these types of radar measurements. I am curious whether you have looked at the spatial correlation of this bias pattern with either radar system parameters or physical conditions on the ice sheet beyond the accumulation  rate.  For  example,  I  could  imagine  that  variations  in  signal-to-noise/signal-to-clutter ratio/image quality rating, aircraft roll, or surface slope/surface roughness/topographic variability could play a role in the stability of the estimates. If so, these could be useful insights for future users of these methods when making some initial assessment of their study region and the available data.

We appreciate this suggest and agree some discussion of these correlations would be beneficial.
Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections.

> Consider briefly mentioning surface velocities for this region somewhere in the paper since horizontal advection might contribute to observed uncertainty in some of the faster flowing regions.  Although it doesn’t look off-hand like there is a correlation between ice velocity and bias in the study area, which is great.

Blah blah blah.

> Do you have any insights from this work on the length scale over which measurements can be considered repeatable? I’m assuming from the high RMSE that variability at the trace-by-trace level is quite high, but that averaged over some distance the agreement between seasons improved. If so that, would also be a useful note to add to the discussion/conclusions.

Blah blah blah.

> Fig. 1 - the color gradient on the flight lines is somewhat difficult to see. I would suggest making the flight line thicker.

We thickened the lines to better show the accumulation variability in this figure.

> Fig. 3 – consider shrinking the mean bias colorbar to the height of panel b, since it only applies to that panel.

We have shrunk the colorbar to the vertical limits of the lower panel.

> Fig. S2 –I assume  that  the dashed lines show some form of uncertainty bounds? Is this $\pm1 \sigma$, 95% confidence intervals, or something else? Please clarify in the caption.

We have addressed with oversight and explicit state in the figure caption that these lines represent $\pm1 \sigma$ errors.

## Reviewer 3

> The error analysis could be improved significantly by comparing the differences between each method with the error one would expect from repeat manual layer tracing.

Blah blah blah.

> Additionally, a comparison of accumulation results to regional climate model output would be a better way to determine “bias,” since those models are used throughout the literature and in policy recommendations.

Although we agree that comparisons to climate model outputs are valuable comparisons, they do not fit within the main focus and direction for this article.
This paper focuses on directly comparing radar results from 2011 and 2016, without attempting to assess whether the results are accurately representing annual accumulation rates (other studies focus more specifically on that question---Medley, Lewis, Keeler, etc.).
We instead investigate whether estimates derived from radar remain consistent and repeatable over a 5-year period, with the main goal assessing the suitability of comparing/combining accumulation estimates from differing radar collection times.
Additionally, the more coarse spatial footprint associated with climate modeling outputs would results in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.

> “Reconstructed from airborne radar imagery collected 5 years apart” could be “collected in 2011 and 2016” to reduce confusion.

We have incorporated the suggested wording (Line ).

> “both collection times” – refers to 2011 and 2016, but is not very clear in the abstract.

We have incorporated the suggested wording (Line ).

> Hyphen between “ice” and “penetrating” here, but not earlier in the abstract

We corrected this typo (Line ).

> Last sentence is a bit of a run-on. I suggest removing “in many cases” and splitting sentences after “by several years”

We removed the extraneous phrase and split into two sentences.

> Should “dominant control on sea level rise” be “dominant contributor to?”

We replaced "control on" with "contributor to" (Line ).

> SMB is not just an “important component of ice sheets,” but is now the single dominant component of ice sheet mass loss (van den Broeke et al., 2009, 2016).

Blah blah blah

> Change “this encompasses freshly-fallen...” to “this balance” or “this sum” or similar.

We have incorporated the suggested wording (Line ).

> Could cite a few ground-based ice penetrating radar papers to give validity to the airborne method and show how much more spatial coverage IceBridge can achieve in one season

Blah Blah blah.

> Give an end date for “NASA’s Operation IceBridge (OIB) campaign (commenced in 2009)”

Included the 2019 end of OIB.

> The Snow Radar does not “image” the subsurface, rather it collects geophysical data one can use to calculate SMB, or something similar

We have replaced "permit imaging" to "permit discernment" to better adhere to the precise definition in regards to radiostratigraphy.

> If the authors bring up “climate modeling” here then the reader would expect some comparison of Snow Radar accumulation rates to regional climate model output. See Koenig et al (2016 Figure 10 for example). RCM rates should be the accumulation rates you compare against, since they are the widely accepted “correct” values

Blah blah blah

> “...can change on shorter time scales” could be sub-annual time scales? Be more specific

Blah blah blah.

> Both [12 - Figure 8] and [13 – Figure 4] use OIB flights from different years to show “temporally stable” measurements at the cross-over points.

We include additional wording to better clarify the distinction between our work on prior studies.
Blah blah blah...

These crossover studies show internal consistency with increasing depth (temporal stability in the sense of repeatable reconstructions across the full time series of the record) but do not assess the effect of temporal stability in regards to time passing between data collection (i.e. they do not distinguish crossovers for the same flight, flights from the same year, flights spaced multiple years apart in time, etc.).
Koenig et al. (2016) includes additional snow deposition between flights as a potential source of error, but does not investigate the magnitude of this effect on estimate error or how the error changes as time between flights increases.
Lewis et al. (2017) similarly make no distinction in their crossover analysis for the timing of radar collection but are instead focused only on the spatial overlap between flights.
Additionally, Lewis et al. (2017) use flights only 2013--2014 OIB flights, limiting the potential effects of temporal instability that is the focus of our manuscript.

> “Most of these studies also focus on multi-decadal mean accumulation rates...” – can’t completely ignore Koenig’s snow radar work, which focused on single year accumulation rates from Snow Radar in Greenland. This manuscript is analyzing 1990-2010 accumulation rates, which is “multi-decadal mean accumulation rates.” This sentence seems to lead the reader to a different conclusion than where this manuscript is heading.

We have edited this section to increase clarity, and have explicitly distinguished Koenig et al. (2016, maybe others?) as also focusing on single year accumulation rates.

> “regardless of collection year” is a bit of a misnomer. You can’t trace annually resolved layers from 100+ years ago with the Snow Radar. So maybe “in recent decades” or similar

Blah

> “within 500 m of one another” – how much variability do you expect within those 500 m? Authors should reference some repeat ice core measurements, RCMs, or the Greenland Bamboo Forest to state how much variability/uncertainty in accumulation exists between these “overlapping” flights, which could then make your bias calculations look even more accurate.

Blah

>

### 2021-06-11 docs/grsl-analysis.py
Experiments with converting figures to matplotlib backend

@@ -11,11 +11,13 @@ import matplotlib.pyplot as plt
 import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
-from bokeh.io import output_notebook
 from shapely.geometry import Point
-output_notebook()
+# from bokeh.io import output_notebook
+# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
+# hv.extension('matplotlib')
+# gv.extension('matplotlib')
 import panel as pn
 import seaborn as sns
 import xarray as xr
@@ -192,11 +194,6 @@ plt_labels = hv.Labels(
         yoffset=20000, text_color='black', 
         text_font_size='28pt')
 
-# (
-#     plt_locCOMB * plt_manPTS 
-#     * (plt_accum2011 * plt_accum2016) * plt_labels
-# )
-
 # %%
 
 # Define Antarctic DEM file
@@ -288,7 +285,7 @@ for label in chunk_centers['Site']:
         crs=gdf_PAIPR.crs).distance(
             gdf_PAIPR.reset_index()) <= 30000).values
     
-    gdf_PAIPR['Site'][idx] = label
+    gdf_PAIPR.loc[idx,'Site'] = label
 
 # Create dataframes for scatter plots
 PAIPR_df = pd.DataFrame(
@@ -508,7 +505,7 @@ dem_list = [
     in REMA_DIR.joinpath("tiles_8m_v1.1").glob('**/*dem.tif') 
     if any(tile in str(path) for tile in dem_index.tile)]
 
-# Calculate slope and aspect for each DEM
+# Calculate slope and aspect for each DEM (only ones not already present)
 [calc_topo(dem) for dem in dem_list]
 
 # %%
@@ -1649,6 +1646,97 @@ cont_bar = cont_plt.opts(
     colorbar=True, fontscale=2.5, width=1200, height=1200)
 hv.save(cont_bar, ROOT_DIR.joinpath(
     'docs/Figures/oib-repeat/cont_bar.png'))
+
+
+
+
+# %% matplotlib versions
+
+# plt_accum2011 = gv.Points(
+#     gdf_2011, crs=ANT_proj, vdims=['accum']).opts(
+#         projection=ANT_proj, color='accum', 
+#         cmap='viridis', colorbar=True, s=50)
+# plt_accum2016 = gv.Points(
[…]

### 2021-06-11 docs/grsl-analysis.py
Adds functions to extract OIB flight data; function to calculate cummulative path distance

@@ -527,6 +527,106 @@ gdf_PAIPR.loc[gdf_PAIPR['slope']>60,'slope'] = np.nan
 gdf_PAIPR.loc[gdf_PAIPR['slope']<0,'elev'] = np.nan
 gdf_PAIPR.loc[gdf_PAIPR['slope']<0,'aspect'] = np.nan
 
+# %% Extract flight parameters
+
+def get_FlightData(flight_dir):
+    """Function to extract OIB flight parameter data from .nc files and convert to geodataframe.
+
+    Args:
+        flight_dir (pathlib.PosixPath): The directory containing the .nc OIB files to extract and convert.
+
+    Returns:
+        geopandas.geodataframe.GeoDataFrame: Table of OIB flight parameters (altitude, heading, lat, lon, pitch, and roll) with their corresponding surface location and collection time.
+    """
+    # List of files to extract from
+    nc_files = [file for file in flight_dir.glob('*.nc')]
+    
+    # Load as xarray dataset
+    xr_flight = xr.open_dataset(
+        nc_files.pop(0))[[
+            'altitude', 'heading', 'lat', 
+            'lon', 'pitch','roll']]
+
+    # Concatenate data from all flights in directory
+    for file in nc_files:
+
+        flight_i = xr.open_dataset(file)[[
+            'altitude', 'heading', 'lat', 
+            'lon', 'pitch','roll']]
+        xr_flight = xr.concat(
+            [xr_flight, flight_i], dim='time')
+
+    # Convert to dataframe and average values to ~200 m resolution
+    flight_data = xr_flight.to_dataframe()
+    flight_coarse = flight_data.rolling(
+        window=40, min_periods=1).mean().iloc[0::40]
+
+    # Convert to geodataframe in Antarctic coordinates
+    gdf_flight = gpd.GeoDataFrame(
+        data=flight_coarse.drop(columns=['lat', 'lon']), 
+        geometry=gpd.points_from_xy(
+            flight_coarse.lon, flight_coarse.lat), 
+        crs='EPSG:4326').reset_index().to_crs(epsg=3031)
+
+    return gdf_flight
+
+
+def path_dist(locs):
+    """Function to calculate the cummulative distance between points along a given path.
+
+    Args:
[…]

### 2021-06-11 docs/grsl-analysis.py
Saves 2011/2016 flight parameter data; rearranges for improved readability

@@ -347,287 +347,6 @@ paipr_1to1_plt_comb = paipr_1to1_plt + site_res_plt
 # paipr_1to1_plt_comb
 
 # %%[markdown]
-# ## Investigations of spatial variability
-#  
-# %% Spatial coherence and variability
-
-def vario(
-    points_gdf, lag_size, 
-    d_metric='euclidean', vars='all', 
-    stationarize=False, scale=True):
-    """A function to calculate the semivariogram for values associated with a geoDataFrame of points.
-
-    Args:
-        points_gdf (geopandas.geodataframe.GeoDataFrame): Location of points and values associated with those points to use for calculating distances and lagged semivariance.
-        lag_size (int): The size of the lagging interval to use for binning semivariance values.
-        d_metric (str, optional): The distance metric to use when calculating pairwise distances in the geoDataFrame. Defaults to 'euclidean'.
-        vars (list of str, optional): The names of variables to use when calculating semivariance. Defaults to 'all'.
-        scale (bool, optional): Whether to perform normalization (relative to max value) on semivariance values. Defaults to True.
-
-    Returns:
-        pandas.core.frame.DataFrame: The calculated semivariance values for each chosen input. Also includes the lag interval (index), the average separation distance (dist), and the number of paired points within each interval (cnt). 
-    """
-
-    # Get column names if 'all' is selected for "vars"
-    if vars == "all":
-        vars = points_gdf.drop(
-            columns='geometry').columns
-
-    # Extact trace coordinates and calculate pairwise distance
-    locs_arr = np.array(
-        [points_gdf.geometry.x, points_gdf.geometry.y]).T
-    dist_arr = pdist(locs_arr, metric=d_metric)
-
-    # Calculate the indices used for each pairwise calculation
-    i_idx = np.empty(dist_arr.shape)
-    j_idx = np.empty(dist_arr.shape)
-    m = locs_arr.shape[0]
-    for i in range(m):
-        for j in range(m):
-            if i < j < m:
-                i_idx[m*i + j - ((i + 2)*(i + 1))//2] = i
[…]

### 2021-06-15 docs/grsl-analysis.py
Investigates effect of topo/flight parameters; updates data directories

@@ -42,7 +42,7 @@ from my_functions import *
 # %% Import PAIPR-generated data
 
 # Import 20111109 results
-dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
+dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
 data_raw = import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
@@ -57,7 +57,7 @@ std2011_ALL = data_2011.pivot(
     index='Year', columns='trace_ID', values='std')
 
 # Import 20161109 results
-dir2 = ROOT_DIR.joinpath('data/PAIPR-outputs/20161109/')
+dir2 = ROOT_DIR.joinpath('data/PAIPR-repeat/20161109/smb/')
 data_raw = import_PAIPR(dir2)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
@@ -101,6 +101,7 @@ std_2016 = std2016_ALL.iloc[
 gdf_PAIPR = gpd.GeoDataFrame(
     {'ID_2011': dist_overlap.index.values, 
     'ID_2016': dist_overlap['trace_ID'].values, 
+    'trace_dist': dist_overlap['distance'].values,
     'accum_2011': 
         accum_2011.mean(axis=0).values, 
     'accum_2016': 
@@ -352,7 +353,7 @@ paipr_1to1_plt_comb = paipr_1to1_plt + site_res_plt
 # %%
 
 # Import and format PAIPR results
-dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
+dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
 data_raw = import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
@@ -367,7 +368,8 @@ std2011_ALL = data_2011.pivot(
     index='Year', columns='trace_ID', values='std')
 
 # Import and format manual results
-dir_0 = ROOT_DIR.joinpath('data/smb_manual/20111109/')
+dir_0 = ROOT_DIR.joinpath(
+    'data/PAIPR-repeat/20111109/smb-manual/')
 data_0 = import_PAIPR(dir_0)
 man_2011 = format_PAIPR(
     data_0, start_yr=1990, end_yr=2010).drop(
[…]

### 2021-06-15 docs/grsl-analysis.py
Investigates differences in horizon depth 2011 vs 2016

@@ -1810,6 +1810,125 @@ hv.save(cont_bar, ROOT_DIR.joinpath(
     'docs/Figures/oib-repeat/cont_bar.png'))
 
 
+# %%[markdown]
+# ## Investigations of depth variability and QC effect
+# 
+# %%
+
+# Import current 20111109 results
+dir_ = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
+data_raw = import_PAIPR(dir_)
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+smb_2011 = format_PAIPR(
+    data_0, start_yr=1990, end_yr=2010).drop(
+    'elev', axis=1).set_index(['collect_time', 'Year'])
+
+# Import 20111109 depth results
+dir_ = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/depth/')
+depth_2011 = import_PAIPR(dir_).rename(
+    columns={"IHR_year":"Year"}).set_index(
+        ['collect_time', 'Year'])
+
+# Join 2011 dataframes and extract gdf locations
+df_depths2011 = smb_2011.join(
+    depth_2011.drop(columns=['Lat', 'Lon']), 
+    how='inner').reset_index().groupby(
+        ['trace_ID', 'Year']).mean()
+df_grouped2011 = df_depths2011[['Lat', 'Lon']].groupby(
+        'trace_ID').mean().reset_index()
+gdf_depth2011 = gpd.GeoDataFrame(
+    data=df_grouped2011.drop(columns=['Lat', 'Lon']), 
+    geometry=gpd.points_from_xy(
+        x=df_grouped2011.Lon, y=df_grouped2011.Lat), 
+    crs="EPSG:4326")
+gdf_depth2011.to_crs(epsg=3031, inplace=True)
+
+# ## Repeat for 2016 data
+
+# Import current 20161109 results
+dir_ = ROOT_DIR.joinpath('data/PAIPR-repeat/20161109/smb/')
+data_raw = import_PAIPR(dir_)
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+smb_2016 = format_PAIPR(
[…]

### 2021-06-15 docs/grsl-analysis.py
Investigates effect of different random seeds on PAIPR results

@@ -1930,7 +1930,162 @@ df_depths['depth_res'] = (
 hv.Distribution(df_depths, kdims=['depth_res'], vdims=['Year']).groupby('Year')
 
 
+# %% Compare PAIPR results from different random seeds
 
+# Import old 20111109 results
+dir_old = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109')
+data_old = import_PAIPR(dir_old)
+data_old.query('QC_flag != 2', inplace=True)
+smb_old = data_old.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+data_test = format_PAIPR(
+    smb_old, start_yr=1990, end_yr=2010).drop(
+    'elev', axis=1)
+aTest_ALL = data_test.pivot(
+    index='Year', columns='trace_ID', values='accum')
+stdTest_ALL = data_test.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Import current 20111109 results
+dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
+data_raw = import_PAIPR(dir1)
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+data_2011 = format_PAIPR(
+    data_0, start_yr=1990, end_yr=2010).drop(
+    'elev', axis=1)
+a2011_ALL = data_2011.pivot(
+    index='Year', columns='trace_ID', values='accum')
+std2011_ALL = data_2011.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Create gdf of mean results for each trace and 
+# transform to Antarctic Polar Stereographic
+gdf_TestData = long2gdf(data_test)
+gdf_TestData.to_crs(epsg=3031, inplace=True)
+gdf_2011 = long2gdf(data_2011)
+gdf_2011.to_crs(epsg=3031, inplace=True)
+
+df_dist = nearest_neighbor(
+    gdf_2011, gdf_TestData, return_dist=True)
+idx_paipr = df_dist['distance'] <= 100
+dist_overlap = df_dist[idx_paipr]
+
+# Create numpy arrays for relevant results
+accum_2011_test = a2011_ALL.iloc[
+    :,dist_overlap.index]
+std_2011_test = std2011_ALL.iloc[
[…]

### 2021-06-16 docs/grsl-analysis.py
Modifies plots to improve visibility/readability

@@ -144,14 +144,14 @@ chunk_centers = gpd.GeoDataFrame({
 plt_accum2011 = gv.Points(
     gdf_2011, crs=ANT_proj, vdims=['accum']).opts(
         projection=ANT_proj, color='accum', 
-        cmap='viridis', colorbar=True, size=10, 
+        cmap='viridis', colorbar=True, size=12, 
         # bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
 plt_accum2016 = gv.Points(
     gdf_2016, crs=ANT_proj, vdims=['accum']).opts(
         projection=ANT_proj, color='accum', 
-        cmap='viridis', colorbar=True, size=10,
+        cmap='viridis', colorbar=True, size=12,
         # bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
@@ -173,8 +173,8 @@ plt_loc2016 = gv.Points(
 plt_locCOMB = gv.Points(
     gdf_PAIPR, crs=ANT_proj, 
     vdims=['accum_2011','accum_2016']).opts(
-        projection=ANT_proj, color='orange', 
-        size=18, 
+        projection=ANT_proj, color='cyan', 
+        size=24, 
         # bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
@@ -185,7 +185,7 @@ plt_manPTS = gv.Points(
     chunk_centers, crs=ANT_proj, 
     vdims='Site').opts(
         projection=ANT_proj, color='black', 
-        size=30, marker='square')
+        size=32, marker='square')
 
 plt_labels = hv.Labels(
     {'x': chunk_centers.geometry.x.values, 
@@ -222,7 +222,7 @@ elev_plt = hv.Image(xr_DEM.values, bounds=tpl_bnds).opts(
 
 # Generate contour plot
 cont_plt = hv.operation.contours(elev_plt, levels=15).opts(
-    cmap='cividis', show_legend=False, 
+    cmap='magma', show_legend=False, 
     colorbar=True, line_width=2)
 
 # Generate elevation hillshade
@@ -261,11 +261,13 @@ plt_accum = gv.Points(
 
 plt_res = gv.Points(
     data=gdf_PAIPR, crs=ANT_proj, vdims=['accum_res']).opts(
-        projection=ANT_proj, color='accum_res', size=10,
+        projection=ANT_proj, color='accum_res', size=12,
         # bgcolor='silver', 
-        colorbar=True, cmap='coolwarm_r', 
+        colorbar=True, 
[…]

### 2021-06-16 docs/submission-docs/oib-repeat/response.md
WIP: Responses to reviewer comments

@@
 In addition to these general changes, we also include specific responses to each of the comments provided by the reviewers, along with the locations in the updated manuscript of the implemented changes.
+
+Although we feel all the suggested edits are valuable, we believe the sum total of all these edits exceeds what can adequately be covered in such a short and targeted study.
+We have therefore incorporated all the suggestions we reasonably can while judiciously deferring less applicable suggestions to future studies.
+For each instance where we elected not to incorporate the suggested edit, we have detailed our reasoning and thoughts as to why such suggestions, while objectively meritorious, are best pursued in future work.
 
@@
 Although research is ongoing to extend the density model used to wider regions of Antarctica, it is not possible to complete and implement this work in the short time frame required for revisions.
-We have added additional text to the manuscript more explicitly stating the limits of this study, with hope to further expand it in the future.
+**We have, however, added additional text to the manuscript more explicitly stating the limits of this study, with hope to further expand it in the future (Lines ).**
 
@@
 *Discussion of uncertainty incorporation in PAIPR.*
-We have included additional discussion in the manuscript to better highlight these points, and hope such discussion will satisfactorily resolve this concern.
+**We have included additional discussion in the manuscript to better highlight these points, and hope such discussion will satisfactorily resolve this concern.**
 
@@
 
-Blah blah blah.
+We edited this to explicitly note this difference in bandwidth between the two flights and its impact on range resolution (Lines ).
+**We also now mention this when discussing potential sources of error in our results (Lines ) and indicate that our results suggest such variations have minimal impact on the reproducibility of results (Lines ).**
 
@@
 
-Blah blah blah.
+$S_Q = 0.50$ represents...
+Although somewhat arbitrary, the 10% cutoff was chosen to ensure more suspect data was flagged as questionable.
+Such results are still processed and generated, but were removed from the later analyses.
+Manual inspections/corrections are still possible with these flagged results.
+The lack of any trend in bias compared to image quality score support the use of these values for our analysis and can be viewed as lower bounds on the subset of data of adequate quality.
 
@@
 > Reference of the in-situ accumulation estimates that are used for the bias distributions?
-It would be nice to have a quick sentence explaining where that data comes from and a pointerto any papers on it.
+It would be nice to have a quick sentence explaining where that data comes from and a pointer to any papers on it.
 
@@
 We appreciate this suggest and agree some discussion of these correlations would be beneficial.
-Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections.
+Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections (Lines ).
+In brief, we find no obvious correlations between the strength of the bias and surface slope, surface aspect, plane altitude above the ice sheet surface, aircraft pitch, or aircraft roll.
+We see a positive correlation between surface elevation and the bias magnitude but believe this is an expression of lower accumulation rates at higher elevations that increases the error associated with SMB estimates (and therefore increases the potential for larger biases).
 
@@
 
-Blah blah blah.
+**We have added language discussing the range in ice velocities for the region and our rationale for considering horizontal advection to be negligible for most of our data in this study (Lines ).**
 
@@
 
-Blah blah blah.
+**Blah blah blah.**
 
@@
 We thickened the lines to better show the accumulation variability in this figure.
+We additionally made a number of other changes to this figure to improve the amount and form of the information conveyed.
+We have increased the size of all fonts to aid in visibility.
+We have also modified the color of the thicker line (denoting overlap) to better contrast with the accumulation color mappings.
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
 
@@
 We have shrunk the colorbar to the vertical limits of the lower panel.
+We have additionally increased the size of fonts to aid in readability.
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
 
@@
 Additionally, the more coarse spatial footprint associated with climate modeling outputs would results in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.
+
+**We have edited the manuscript so that we refer less explicitly to climate modeling (Lines ).**
 
@@
 
-We removed the extraneous phrase and split into two sentences.
+We removed the extraneous phrase and split the sentence into two sentences.
 
@@
 
-Blah blah blah
+We have added additional wording better highlighting the relevance of SMB, with particular reference to the work of van den Broeke et al.
 
@@
 
-Blah Blah blah.
+**Blah Blah blah.**
 
@@
 
-Included the 2019 end of OIB.
+We added the OIB end year of 2019.
 
@@
 
-Blah blah blah
+**Blah blah blah**
 
@@
 
-Blah blah blah.
+We have explicitly stated we are referring to time scales of a few years.
 
@@
 
-We include additional wording to better clarify the distinction between our work on prior studies.
-Blah blah blah...
+We include additional wording to better clarify the distinction between our work and prior studies (Lines ).
+We specifically highlight the differences between the crossover analyses performed in Lewis et al.
+(2017) and Koenig et al.
+(2016), and why our approach better targets this question of repeatability with disparate collection times specifically.
 
@@
 
-We have edited this section to increase clarity, and have explicitly distinguished Koenig et al. (2016, maybe others?) as also focusing on single year accumulation rates.
+We have edited this section to increase clarity, and have explicitly distinguished Koenig et al.
+(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodology.
 
@@
 
-Blah
+We have edited the sentence to reflect that we are specifically interested in understanding repeatability of radar-derived SMB collected over a period of years.
 
@@
 
+**Blah**
+
+> “additional details on this unit” do you mean “radar” or “radar unit”? A bit confusing as worded
+
+We have changed this to "Snow radar" to more clearly convey the unit in question.
+
+> “except we apply the model proposed in [20]...” Could you fit one sentence in this manuscript briefly describing that method?
+The depth-density relationship is critical for the rest of this analysis, so I feel it’s important to include
+
+We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ) as follows:
+
+This density model uses a positive-valued Bayesian semi-parametric model over space and depth below the ice sheet, informed by 57 \emph{in-situ} firn/ice cores.
+There are several advantages afforded by this method, most importantly its ability to leverage space-depth autocorrelation to interpolate the shape-constrained depth-density profiles to new locations with relatively few model assumptions.
+Moreover, because we fit the model in a Bayesian framework, our method provides rigorous uncertainty quantification of the depth-density profiles at arbitrary sites to propagate forward into uncertainties of final SMB estimates (see [] for further details).
+
+> What does “expert manual tracing” entail? One author tracing the lines once? Or multiple authors? Or multiple rounds of tracing? Etc.
+
+We have added context to explicitly show that these manual estimates are based on a single iteration of layer selection (Lines ).
+
+> “covering a range of accumulation rates and subsurface” – please provide the range here so it’s not just in the Figure 1 caption
+
+We have added the precise range in accumulation rates to this section of the text (Lines ).
+
+> “The error for the PAIPR method incorporates” implies that the “expert manual tracing” has zero error, when there certainly is some uncertainty
+
+We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method.
+Even using repeat iterations of manual tracing (either by the same individual or multiple individuals) would reflect psychological and physiological influences as much or more than the true error associated with manual tracing methods.
+We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty.
+
+> “and 1 (most suspect)” is not a proper scientific term
+
+We have removed the offending phrase (Line ).
+**We have also expanded our discussion of the image quality scoring metric to clarify what these values represent.**
+
+> IRH is defined after Equation 1, but also used in the paragraph before
+
+We have now properly defined this acronym prior to its first use (Line ).
+
+> What values of S_Q do you see throughout your data? One sentence describing the quality scores would be helpful
+
+**We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the variability in these values.**
+
+> Can you give a few examples of “regions where IRHs are discontinuous?” How much of Antarctica (or the region you studied) would fall into that category? Does SEAT2010-4 have a quality control flag?
+
+We have added a sentence stating such discontinuities are more common in coastal regions and areas that experience melt/refreeze (Lines ).
+We added a note specifying that 3.1% of the original data results were removed by the automated quality control flags (Lines ).
+**We added a note explicitly stating that issues in the SEAT2010-4 area were not associated with QC flagging but instead this region shows clear, well-delineated layers in both 2011 and 2016 results that do not agree with each other.**
+
+> “depths below the surface represented in the results will differ” – can you quantify how much deeper the IRHs are using the 2016 data? Just a range or mean and standard deviation would do
+
+We have added the mean difference in depth between 2016 and 2011 data for both the 2010 and 1990 horizons (Lines ).
+
+> The “IV. Results” is hidden above Figure 3, I suggest that the typesetting team make sure this section heading is visible
+
+We leave this to the typesetting team to resolve in the final version.
+
+> What are you considering “correct” when computing “the differences, biases, and errors?” Could be a good place to compare with RCMs or field measurements (if they exist).
+
+**Blah.**
+
+> “2016 estimates on average are 2.98...” – sentence starts with a number. Estimates of what?
+
+We have fixed the typos to ensure what is discussed is clearly evident (Line ).
+
+> How important is this 3% differences compared with tracing IRHs twice for the same flight? Or for two flights within 500 m of each other?
+
+**Blah.**
+
+> “when expressed as % bias relative to mean in-situ  annual accumulation” replace with “percent.” What are you using as the “correct” accumulation?
+
+We have fixed the typo (Line ) and added language to clarify that this % bias is relative to the average accumulation across all methods for a given location and year (Lines ).
+
+> “exhibit spatial coherence beyond what would be expected from random noise” can you quantify this somehow?
+
+We added a brief note that the biases are generally spatially correlated to a distance of ~75 km, based on an experimental variogram (Lines ).
+
+> “pronounced decrease in accumulation” – how large is this decrease? Is it statistically significant? It looks quite large from Figure S2. Can you see this trend in the firn core?
+
+**Blah**
+
+> “Site B in this study coincides with a firn core site” – replace with core “location”
+
+We incorporated the suggested edit (Line ).
+
+> Does SEAT2010-4 show any decrease in accumulation between 1990-2016, or is the difference entirely within the radar calculations?
+
 Blah
 
->
+> Can you briefly discuss “the limitations on the repeatability of radar imaging in these regions”
+
+Blah
+
+> “indicate that the automated PAIPR method is able to accurately replicate the manual approach” – can you quantify the difference in IRH depth between the two methods, not just the accumulation.
+Might be simple to add a column to one of the tables, since there is also uncertainty in density
+
+Blah
+
+> “when data collection is separated by a number of years” should be 5 years, since you can’t speculate if this would work with a longer separation
+
+Blah
+
+> “annual and sub-annual changes in surface conditions” if a regions begins to melt that had previously not experienced melt it would be harder for the snow radar to penetrate that region.
+So this statement is at least limited to the dry snow zone
+
+Blah
+
+> “implies the requisite collection frequency is likely much larger” – would you suggest collection every year?
+Two years?
+This is where you can make recommendations, especially now that OIB has finished
+
+Blah
+
+> Could bring in Koenig’s results to “broadly applicable to dry-firn regions of polar ice sheets in both hemispheres”
+
+Blah
+
+> Figure 1 – Fonts are all a bit small and difficult to read.
+“Mean accum” could be “Mean accumulation.” The thin line indicating mean accumulation is too small to detect variability, especially when on the white background.
+
+We have increased the size of all fonts and the size of the colored accumulation line to aid in visibility.
+We have also modified the color of the thicker line (denoting overlap) to better contrast with the accumulation color mappings.
+We have changed "Mean accum" to "Mean accumulation".
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
+
+> Figure 2 – I know this is straight from the NSIDC site, but you could Photoshop the text to be a bit larger and easier to read for those less familiar with radar echograms.
+Can you indicate on Figure 1 where this 5 km transect is?
+
+**Blah**
+
+> Figure 3 – The legend in Fig 3a.
+says “2016-2011” but the caption says “2011-2016.” Which is it?
+Colorbar extends beyond the height of Fig 3b.
+The colorbar is not white at 0 bias.
+Colored points are too small to see in Fig 3b.
+There seems to be a vertical line left of the word “Density”
+
+We have edited the caption to match the figure (2016-2011).
+We have modified the colorbar to match the bounds of the figure.
+We have ensured the colorbar is symmetric (although the center color is closer to gray than to white).
+We have increased the size of the colored points to increase visibility.
+We were unable to discern a vertical line left of the word "Density" in this figure.
+This may have been introduced during the conversion after uploading.
+If this continues to be an issue, please notify us and we will work to resolve it.
+
+> The link within the citation for [20] is just for the abstract, not the actual paper
+
+The full paper can be downloaded from the PDF link on the upper right-hand side of the website.
+
+> Figure S1 – All fonts are much too small to see.
+Nowhere in the figure or caption does it say that these numbers are accumulation in mm w.e.
+a^-1.
+I’d suggest changing the y axis for the histograms and only having one legend, otherwise it’s difficult to discern differences between the methods.
+
+We have increased the size of the fonts used, ensured axis labels are included, and made additional changes to increase the readability of the plots in this figure.
+
+> Figure S2 – Are the dashed lines uncertainty bounds?
+Need to say so, if true.
+How common are these missed/extra layers in the PAIPR method?
+Could you build in a manual correction to prevent this from happening elsewhere?
+
+We have clarified that the dashed lines are $1\sigma$ uncertainty bounds.
+We have additionally clarified and expounded on the missed/extra layers and how these are limited to the manually traced estimates.
+The probabilistic nature of PAIPR minimizes the impact of this effect for the automated results.

### 2021-06-17 docs/grsl-analysis.py
Additional analyses and figure edits for GRSL resubmission

@@ -102,6 +102,8 @@ gdf_PAIPR = gpd.GeoDataFrame(
     {'ID_2011': dist_overlap.index.values, 
     'ID_2016': dist_overlap['trace_ID'].values, 
     'trace_dist': dist_overlap['distance'].values,
+    'QC_2011': gdf_2011.loc[dist_overlap.index,'QC_med'].values,
+    'QC_2016': dist_overlap['QC_med'].values, 
     'accum_2011': 
         accum_2011.mean(axis=0).values, 
     'accum_2016': 
@@ -139,54 +141,57 @@ chunk_centers = gpd.GeoDataFrame({
 #             -4.639E5, -4.640E5, -4.668E5]), 
 #     crs="EPSG:3031")
 
-#%% Data location map
+#%% Data map components
 
+# Accumulation plots for both flights
 plt_accum2011 = gv.Points(
     gdf_2011, crs=ANT_proj, vdims=['accum']).opts(
         projection=ANT_proj, color='accum', 
         cmap='viridis', colorbar=True, size=12, 
-        # bgcolor='silver', 
+        bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
 plt_accum2016 = gv.Points(
     gdf_2016, crs=ANT_proj, vdims=['accum']).opts(
         projection=ANT_proj, color='accum', 
         cmap='viridis', colorbar=True, size=12,
-        # bgcolor='silver', 
+        bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
 
+# Location plots for flights (and combined overlap)
 plt_loc2011 = gv.Points(
     gdf_2011, crs=ANT_proj, vdims=['accum','std']).opts(
         projection=ANT_proj, color='blue', alpha=0.9, 
         size=5, 
-        # bgcolor='silver', 
+        bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
 plt_loc2016 = gv.Points(
     gdf_2016, crs=ANT_proj, vdims=['accum','std']).opts(
         projection=ANT_proj, color='red', alpha=0.9, 
         size=5, 
-        # bgcolor='silver', 
+        bgcolor='silver', 
         tools=['hover'], 
         width=700, height=700)
 plt_locCOMB = gv.Points(
[…]

### 2021-06-17 src/my_functions.py
Exports QC_med values when formatting PAIPR output

@@ -78,7 +78,7 @@ def format_PAIPR(data_df, start_yr=None, end_yr=None):
         # New df (in long format) with accum data assigned
         data_long = (
             data_df.filter(['trace_ID', 'collect_time', 
-            'QC_flag', 'Lat', 'Lon', 
+            'QC_flag', 'QC_med', 'Lat', 'Lon', 
             'elev', 'Year']).assign(
                 accum=data_df['accum_mu'], 
                 std=data_df['accum_std']).reset_index(drop=True))

### 2021-06-17 docs/submission-docs/oib-repeat/response.md
WIP: Adds additional responses to reviewer comments

@@
 
+We include two copies of the manuscript in this submission.
+The first is a direct update from the initial submission.
+The second highlights the specific changes we have made since the original submission.
+Line numbers in the following comments reference the location of edits in the second, change-highlighted document.
+
 Although we feel all the suggested edits are valuable, we believe the sum total of all these edits exceeds what can adequately be covered in such a short and targeted study.
@@
 For each instance where we elected not to incorporate the suggested edit, we have detailed our reasoning and thoughts as to why such suggestions, while objectively meritorious, are best pursued in future work.
+
+The edits incorporated into the manuscript also necessitated the moving of the discussion of updates to the PAIPR methodology to the supplement.
+We do not, however, feel this change materially impacts the readability of the manuscript or causes undue jumps in logic when trying to follow the methods, results and their implications.
+We have made edits throughout the manuscript to ensure no loss in...
 
@@
 Although research is ongoing to extend the density model used to wider regions of Antarctica, it is not possible to complete and implement this work in the short time frame required for revisions.
-**We have, however, added additional text to the manuscript more explicitly stating the limits of this study, with hope to further expand it in the future (Lines ).**
+**We have, however, added additional text to the manuscript more explicitly stating the limits of this study, with hopes to further expand it in the future (Lines ).**
 
@@
 
-We were uncertain of the precise issue raised by this comment.
-We understood it to mean...
-The original paper outlining the PAIPR method discusses these concerns in greater detail.
-*Discussion of uncertainty incorporation in PAIPR.*
-**We have included additional discussion in the manuscript to better highlight these points, and hope such discussion will satisfactorily resolve this concern.**
+SNR of echo layers are relevant to this research, so we are glad the reviewer raised this point.
+We have added a discussion of the impact of radar attenuation on SNR and its effect on layer selection (Lines ).
+We also briefly discuss how these impacts are limited in our study (Lines ).
 
@@
 2011 had a bandwidth of 4 GHz (2-6 GHz sweep) and 2016 has a bandwidth of 6 GHz (2-8 GHz sweep).
-(If you’re using the CReSIS MATLAB files from their ftp site, you can check these parameters in the param_get_heights.radar.wfs variable that comes with the data.) This isn’t a huge difference in range resolution (~4 cm vs ~2.7 cm in snow), but I would expect some impact on the layer tracking results.This seems worth mentioning, both as a potential source of both the error and bias and because your results now actually suggest that the accumulation estimates are fairly robust to these kinds of small variations in the bandwidth/range resolution.
+(If you’re using the CReSIS MATLAB files from their ftp site, you can check these parameters in the param_get_heights.radar.wfs variable that comes with the data.) This isn’t a huge difference in range resolution (~4 cm vs ~2.7 cm in snow), but I would expect some impact on the layer tracking results.
+This seems worth mentioning, both as a potential source of both the error and bias and because your results now actually suggest that the accumulation estimates are fairly robust to these kinds of small variations in the bandwidth/range resolution.
 
-We edited this to explicitly note this difference in bandwidth between the two flights and its impact on range resolution (Lines ).
-**We also now mention this when discussing potential sources of error in our results (Lines ) and indicate that our results suggest such variations have minimal impact on the reproducibility of results (Lines ).**
+We appreciate the reviewer bringing up this point and its significance to our results.
+We edited the manuscript to explicitly note this difference in bandwidth between the two flights and its impact on range resolution (Lines ).
+**We also now mention this difference when discussing potential sources of error in our results (Lines ) and indicate that our results suggest such variations have minimal impact on the reproducibility of SMB estimates from radar (Lines ).**
 
@@
 
-$S_Q = 0.50$ represents...
-Although somewhat arbitrary, the 10% cutoff was chosen to ensure more suspect data was flagged as questionable.
-Such results are still processed and generated, but were removed from the later analyses.
-Manual inspections/corrections are still possible with these flagged results.
-The lack of any trend in bias compared to image quality score support the use of these values for our analysis and can be viewed as lower bounds on the subset of data of adequate quality.
+This comment raises a good point regarding the selection of these thresholds.
+We have added additional discussion on the $S_Q$ parameter to better give intuition to its physical meaning (Lines ).
+We further include a brief discussion regarding the choice of parameter values, the rationale behind that choice, we indicate that a small amount of the overall dataset (~3%) were removed through this, and present some evidence suggesting our choice of parameter values is warranted (Lines ).
 
@@
 We believe poor wording on our part led to some confusion regarding this point, as more than one reviewer mentioned it.
-We do not use other methods to compare to radar results (e.g. ice cores, climate model outputs, etc.) but instead focus exclusively on
+We do not use other methods to compare to radar results (e.g.
+ice cores, climate model outputs, etc.) but instead focus exclusively on the biases and errors between radar results from different methods and flights.
+The target of this research is not to determine whether radar-derived accumulation is accurate compared to other independent methods, but rather whether radar-derived estimates are self-consistent over time.
+We therefore do not perform direct comparisons between radar results and independent methods (e.g. firn cores, RCM outputs, etc.).
+
+We have modified language throughout the manuscript to make this point more clear.
+We have specifically removed all references to *in-situ data* throughout the manuscript to avoid additional confusion.
 
@@
 
-Blah blah blah.
+This mean value refers to the location-specific mean accumulation across all four methods for a given year.
+It therefore is not a single consistent value for the whole region but instead varies by location.
+We elected to use a spatially-varying mean to scale our biases due to the large gradient in accumulation rates across the region (varying by a factor of ~3).
+We have removed the term *in-situ* and replaced it with wording that more clearly articulates this definition (Lines ).
 
@@
 Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections (Lines ).
-In brief, we find no obvious correlations between the strength of the bias and surface slope, surface aspect, plane altitude above the ice sheet surface, aircraft pitch, or aircraft roll.
+We find no obvious correlations between the strength of the bias and surface slope, surface aspect, plane altitude above the ice sheet surface, aircraft pitch, or aircraft roll.
 We see a positive correlation between surface elevation and the bias magnitude but believe this is an expression of lower accumulation rates at higher elevations that increases the error associated with SMB estimates (and therefore increases the potential for larger biases).
@@
 
-We have addressed with oversight and explicit state in the figure caption that these lines represent $\pm1 \sigma$ errors.
+We have addressed with oversight and explicitly state in the figure caption that these lines represent $\pm1 \sigma$ errors.
 
@@
 This paper focuses on directly comparing radar results from 2011 and 2016, without attempting to assess whether the results are accurately representing annual accumulation rates (other studies focus more specifically on that question---Medley, Lewis, Keeler, etc.).
+The target of this research is not to determine whether radar-derived accumulation is accurate compared to other independent methods, but rather whether radar-derived estimates are self-consistent over time.
+We therefore do not perform direct comparisons between radar results and independent methods (e.g. firn cores, RCM outputs, etc.).
 We instead investigate whether estimates derived from radar remain consistent and repeatable over a 5-year period, with the main goal assessing the suitability of comparing/combining accumulation estimates from differing radar collection times.
-Additionally, the more coarse spatial footprint associated with climate modeling outputs would results in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.
+Additionally, the more coarse spatial footprint associated with climate modeling outputs (10's of km vs ~200 m) specifically would result in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.
 
@@
 
-These crossover studies show internal consistency with increasing depth (temporal stability in the sense of repeatable reconstructions across the full time series of the record) but do not assess the effect of temporal stability in regards to time passing between data collection (i.e.
+**These crossover studies show internal consistency with increasing depth (temporal stability in the sense of repeatable reconstructions across the full time series of the record) but do not assess the effect of temporal stability in regards to time passing between data collection (i.e.
 they do not distinguish crossovers for the same flight, flights from the same year, flights spaced multiple years apart in time, etc.).
@@
 Lewis et al. (2017) similarly make no distinction in their crossover analysis for the timing of radar collection but are instead focused only on the spatial overlap between flights.
-Additionally, Lewis et al. (2017) use flights only 2013--2014 OIB flights, limiting the potential effects of temporal instability that is the focus of our manuscript.
+Additionally, Lewis et al. (2017) use flights only 2013--2014 OIB flights, limiting the potential effects of temporal instability that is the focus of our manuscript.**
 
@@
 We have edited this section to increase clarity, and have explicitly distinguished Koenig et al.
-(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodology.
+(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodological focus (Lines ).
 
@@
 
-We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ) as follows:
-
-This density model uses a positive-valued Bayesian semi-parametric model over space and depth below the ice sheet, informed by 57 \emph{in-situ} firn/ice cores.
-There are several advantages afforded by this method, most importantly its ability to leverage space-depth autocorrelation to interpolate the shape-constrained depth-density profiles to new locations with relatively few model assumptions.
-Moreover, because we fit the model in a Bayesian framework, our method provides rigorous uncertainty quantification of the depth-density profiles at arbitrary sites to propagate forward into uncertainties of final SMB estimates (see [] for further details).
+We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ).
 
@@
 
-We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method.
+We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method (Lines ).
 Even using repeat iterations of manual tracing (either by the same individual or multiple individuals) would reflect psychological and physiological influences as much or more than the true error associated with manual tracing methods.
-We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty.
+We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty (Lines ).
 
@@
 We have removed the offending phrase (Line ).
-**We have also expanded our discussion of the image quality scoring metric to clarify what these values represent.**
+We have also expanded our discussion of the image quality scoring metric to clarify what these values represent (Lines ).
 
@@
 
-**We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the variability in these values.**
+We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the choice parameter values $S_Q$ and the cutoff percentage threshold (Lines ).
 
@@
 
-**Blah.**
+We believe poor wording on our part led to some confusion regarding this point, as more than one reviewer mentioned it.
+We do not use other methods to compare to radar results (e.g.
+ice cores, climate model outputs, etc.) but instead focus exclusively on the biases and errors between radar results from different methods and flights.
+The target of this research is not to determine whether radar-derived accumulation is accurate compared to other independent methods, but rather whether radar-derived estimates are self-consistent over time.
+We therefore do not perform direct comparisons between radar results and independent methods (e.g. firn cores, RCM outputs, etc.).
+We instead use the mean accumulation value at a given location and year to determine percent differences between different methods and flight data.
+**We have modified wording referencing this throughout the manuscript to hopefully make this distinction more clear (Lines ).**
 
@@
 
-Blah
+**Blah**
 
@@
 
-Blah
+**Blah**
+Although we believe this could be useful information, we fear its inclusion could lead to more misdirection than context.
+Due to the probabilistic nature of PAIPR, the depths assigned to a given year incorporate a greater degree of uncertainty...
 

### 2021-06-17 docs/grsl-analysis.py
Adds cell to extract surface ice velocities

@@ -245,6 +245,30 @@ xr_HS = hillshade(xr_DEM)
 hill_plt = hv.Image(xr_HS.values, bounds=tpl_bnds).opts(
         alpha=0.25, cmap='gray', colorbar=False)
 
+# %% Get ice velocities of trace locations
+
+xr_vice = xr.open_dataset(ROOT_DIR.joinpath(
+    'data/ice-velocities/antarctica_ice_velocity_450m_v2.nc'))
+xr_clip = xr_vice.sel(
+    x=slice(
+        gdf_PAIPR.total_bounds[0], 
+        gdf_PAIPR.total_bounds[2]), 
+    y=slice(
+        gdf_PAIPR.total_bounds[3], 
+        gdf_PAIPR.total_bounds[1]))
+xr_pts = extract_at_pts(xr_clip, gdf_PAIPR, return_dist=True)
+vice_pts = gpd.GeoDataFrame(
+    data=xr_pts[['VX', 'VY']], crs='epsg:3031', 
+    geometry=xr_pts.geometry)
+vice_pts['Vxy'] = np.sqrt(
+    vice_pts['VX']**2 + vice_pts['VY']**2)
+
+gv.Points(vice_pts, crs=ANT_proj, 
+vdims=['Vxy', 'VX', 'VY']).opts(
+    projection=ANT_proj, color='Vxy', cmap='viridis', 
+    size=10, colorbar=True, logz=True,
+    width=800, height=800, tools=['hover'])
+
 # %% PAIPR residuals
 
 # Calculate residuals (as % bias of mean accumulation)

### 2021-06-18 docs/grsl-analysis.py
Edits/tweaks to GRSL figures

@@ -175,7 +175,7 @@ plt_loc2016 = gv.Points(
         tools=['hover'], 
         width=700, height=700)
 plt_locCOMB = gv.Points(
-    gdf_PAIPR, crs=ANT_proj, 
+    gdf_PAIPR.iloc[0::25], crs=ANT_proj, 
     vdims=['accum_2011','accum_2016']).opts(
         projection=ANT_proj, color='cyan', 
         size=24, alpha=0.5,
@@ -200,7 +200,7 @@ plt_labels = hv.Labels(
         yoffset=20000, text_color='black', 
         text_font_size='28pt')
 
-# Location of example echograms in Fig 2
+# Location of example echograms in Fig S1
 echo_pt = gpd.GeoDataFrame(
     gpd.GeoSeries(Point(-113.250, -79.215), crs='epsg:4326'))
 echo_pt = echo_pt.rename(
@@ -256,7 +256,8 @@ xr_clip = xr_vice.sel(
     y=slice(
         gdf_PAIPR.total_bounds[3], 
         gdf_PAIPR.total_bounds[1]))
-xr_pts = extract_at_pts(xr_clip, gdf_PAIPR, return_dist=True)
+xr_pts = extract_at_pts(
+    xr_clip, gdf_PAIPR.copy(), return_dist=True)
 vice_pts = gpd.GeoDataFrame(
     data=xr_pts[['VX', 'VY']], crs='epsg:3031', 
     geometry=xr_pts.geometry)
@@ -1818,8 +1819,10 @@ data_map = (
     # elev_plt.opts(colorbar=False)
     hill_plt
     * cont_plt.opts(colorbar=False)
-    * plt_locCOMB * plt_manPTS 
+    * plt_locCOMB 
+    * plt_manPTS 
     * (plt_accum2011 * plt_accum2016) 
+    * plt_echo
     * plt_labels.opts(text_font_size='36pt')
     ).opts(fontscale=3, width=1200, height=1200)
 
@@ -1865,11 +1868,13 @@ cont_bar = cont_plt.opts(
 hv.save(cont_bar, ROOT_DIR.joinpath(
     'docs/Figures/oib-repeat/cont_bar.png'))
 
-# %% Example echogram figures
+# %% Example 2011 echogram figure
 
-echo2011 = xr.load_dataset(ROOT_DIR.joinpath(
-    'data/oib-examples/IRSNO1B_20111109_02_255.nc'))[
[…]

### 2021-06-18 docs/submission-docs/oib-repeat/response.md
Comments included in draft revision of grsl

@@
+---
+geometry: margin=2.5cm
+---
+
 # Response to reviewers
@@
 We appreciate the time the reviewers took to offer suggestions for improvement, with their inclusion in the manuscript materially improving both the underlying science and its readability.
+We include specific responses to each of the comments provided by the reviewers, along with the locations in the updated manuscript of the implemented changes.
+Although we feel all the suggested edits were valuable, we believe the sum total of all these edits exceeds what can adequately be covered in such a short and targeted study.
+We have therefore incorporated all the suggestions we reasonably can while judiciously deferring less applicable suggestions to future endeavors.
+For each instance where we elected not to incorporate the suggested edit, we have detailed our reasoning and thoughts as to why such suggestions, while objectively meritorious, are best pursued in later work.
 
-General response to editor and reviewers discussing the main thrusts of the edits.
-
-In addition to these general changes, we also include specific responses to each of the comments provided by the reviewers, along with the locations in the updated manuscript of the implemented changes.
+The edits incorporated into the manuscript also necessitated the moving of the discussion of updates to the PAIPR methodology and the figure showing representative echograms to the supplement.
+We do not, however, feel these changes materially impact the readability of the manuscript or cause undue jumps in logic when trying to follow the methods, results, and implications of the research.
+We have made edits throughout the manuscript to ensure no loss in continuity due to these adjustments.
 
 We include two copies of the manuscript in this submission.
-The first is a direct update from the initial submission.
-The second highlights the specific changes we have made since the original submission.
-Line numbers in the following comments reference the location of edits in the second, change-highlighted document.
+The first is a direct update from the initial submission in the.
+The second highlights the specific changes we have made since the original submission with additions in blue and deletions in red.
+Line numbers in the following comments reference the location of edits in the second, color-highlighted document.
 
-Although we feel all the suggested edits are valuable, we believe the sum total of all these edits exceeds what can adequately be covered in such a short and targeted study.
-We have therefore incorporated all the suggestions we reasonably can while judiciously deferring less applicable suggestions to future studies.
-For each instance where we elected not to incorporate the suggested edit, we have detailed our reasoning and thoughts as to why such suggestions, while objectively meritorious, are best pursued in future work.
-
-The edits incorporated into the manuscript also necessitated the moving of the discussion of updates to the PAIPR methodology to the supplement.
-We do not, however, feel this change materially impacts the readability of the manuscript or causes undue jumps in logic when trying to follow the methods, results and their implications.
-We have made edits throughout the manuscript to ensure no loss in...
-
-## Reviewer 1
+## Responses to Reviewer 1
 
@@
 
-We agree with the reviewer that it would be useful to expand the analysis to a wider region, with a discussion of cross-region variability.
+*We agree with the reviewer that it would be useful to expand the analysis to a wider region, with a discussion of cross-region variability.
 We, however, feel such an investigation is outside the intended scope of the current study.
@@
 Although research is ongoing to extend the density model used to wider regions of Antarctica, it is not possible to complete and implement this work in the short time frame required for revisions.
-**We have, however, added additional text to the manuscript more explicitly stating the limits of this study, with hopes to further expand it in the future (Lines ).**
+We nonetheless believe the data included in the study is sufficiently spatially extensive to be useful and informative to the wider scientific community.*
 
@@
 
-SNR of echo layers are relevant to this research, so we are glad the reviewer raised this point.
+*SNR of echo layers are relevant to this research, so we are glad the reviewer raised this point.
 We have added a discussion of the impact of radar attenuation on SNR and its effect on layer selection (Lines ).
-We also briefly discuss how these impacts are limited in our study (Lines ).
+We also briefly discuss how these impacts are limited in our study (Lines ).*
 
-## Reviewer 2
+## Responses to Reviewer 2
 
@@
 
-We appreciate the reviewer bringing up this point and its significance to our results.
+*We appreciate the reviewer bringing up this point and its significance to our results.
 We edited the manuscript to explicitly note this difference in bandwidth between the two flights and its impact on range resolution (Lines ).
-**We also now mention this difference when discussing potential sources of error in our results (Lines ) and indicate that our results suggest such variations have minimal impact on the reproducibility of SMB estimates from radar (Lines ).**
+We also now mention this difference when discussing potential sources of error in our results and indicate that our results suggest such variations have minimal impact on the reproducibility of SMB estimates from radar (Lines ).*
 
@@
 
-This comment raises a good point regarding the selection of these thresholds.
+*This comment raises a good point regarding the selection of these thresholds.
 We have added additional discussion on the $S_Q$ parameter to better give intuition to its physical meaning (Lines ).
-We further include a brief discussion regarding the choice of parameter values, the rationale behind that choice, we indicate that a small amount of the overall dataset (~3%) were removed through this, and present some evidence suggesting our choice of parameter values is warranted (Lines ).
+We further include a brief discussion regarding the choice of parameter values, the rationale behind that choice, we indicate that a small amount of the overall dataset (~3%) were removed through this, and present some evidence supporting our choice of parameter values (Lines ).*
 
@@
 
-**Blah blah blah.**
+*We have added clarifying language to address this confusion (Lines ).
+Essentially we optimize the logistic function parameters using manual layer traces, which we repeat across all traces of multiple echograms.
+This leads to estimated parameter values from each trace, and a distribution of ~15,000 estimates across the full training dataset.
+We use this distribution to inform our expected value and variability of logistic parameters when performing our Monte Carlo simulations.*
 
@@
 
-We believe poor wording on our part led to some confusion regarding this point, as more than one reviewer mentioned it.
+*We believe poor wording on our part led to some confusion regarding this point, as more than one reviewer mentioned it.
 We do not use other methods to compare to radar results (e.g.
@@
 We therefore do not perform direct comparisons between radar results and independent methods (e.g. firn cores, RCM outputs, etc.).
-
 We have modified language throughout the manuscript to make this point more clear.
-We have specifically removed all references to *in-situ data* throughout the manuscript to avoid additional confusion.
+We have specifically removed all references to *in-situ data* throughout the manuscript to avoid additional confusion.*
 
@@
 
-This mean value refers to the location-specific mean accumulation across all four methods for a given year.
+*This mean value refers to the location-specific mean accumulation across all four methods for a given year.
 It therefore is not a single consistent value for the whole region but instead varies by location.
 We elected to use a spatially-varying mean to scale our biases due to the large gradient in accumulation rates across the region (varying by a factor of ~3).
-We have removed the term *in-situ* and replaced it with wording that more clearly articulates this definition (Lines ).
+We have removed the term *in-situ* and replaced it with wording that more clearly articulates this definition (Lines ).*
 
@@
 
-We appreciate this suggest and agree some discussion of these correlations would be beneficial.
+*We appreciate this suggestion and agree some discussion of these correlations would be beneficial.
 Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections (Lines ).
 We find no obvious correlations between the strength of the bias and surface slope, surface aspect, plane altitude above the ice sheet surface, aircraft pitch, or aircraft roll.
-We see a positive correlation between surface elevation and the bias magnitude but believe this is an expression of lower accumulation rates at higher elevations that increases the error associated with SMB estimates (and therefore increases the potential for larger biases).
+We see a positive correlation between surface elevation and the bias magnitude but believe this is an expression of lower accumulation rates at higher elevations that increases the error associated with SMB estimates (and therefore increases the potential for larger biases).*
 
@@
 
-**We have added language discussing the range in ice velocities for the region and our rationale for considering horizontal advection to be negligible for most of our data in this study (Lines ).**
+*We have added language discussing the range in ice velocities for the region and our rationale for considering horizontal advection to be negligible for most of our data in this study (Lines ).*
 
-> Do you have any insights from this work on the length scale over which measurements can be considered repeatable?
+> **Do you have any insights from this work on the length scale over which measurements can be considered repeatable?
 I’m assuming from the high RMSE that variability at the trace-by-trace level is quite high, but that averaged over some distance the agreement between seasons improved.
-If so that, would also be a useful note to add to the discussion/conclusions.
+If so that, would also be a useful note to add to the discussion/conclusions.**
 
-**Blah blah blah.**
+*I don't really have the space to go into this at all, so any suggestions on how best to respond to this comment?*
 
@@
 
-We thickened the lines to better show the accumulation variability in this figure.
+*We thickened the lines to better show the accumulation variability in this figure.
 We additionally made a number of other changes to this figure to improve the amount and form of the information conveyed.
@@
 We have also modified the color of the thicker line (denoting overlap) to better contrast with the accumulation color mappings.
-We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.*
 
@@
 
-We have shrunk the colorbar to the vertical limits of the lower panel.
+*We have shrunk the colorbar to the vertical limits of the lower panel.
 We have additionally increased the size of fonts to aid in readability.
-We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.*
 
@@
 
-We have addressed with oversight and explicitly state in the figure caption that these lines represent $\pm1 \sigma$ errors.
+*We have addressed with oversight and explicitly state in the figure caption that these lines represent $\pm1 \sigma$ errors.*
 
-## Reviewer 3
+## Responses to Reviewer 3
 
-> The error analysis could be improved significantly by comparing the differences between each method with the error one would expect from repeat manual layer tracing.
+> **The error analysis could be improved significantly by comparing the differences between each method with the error one would expect from repeat manual layer tracing.**
 
-**Blah blah blah.**
+*I'm still not really sure how to respond to this. It is not something that will be addressed in the revised manuscript so any recommendations on the best way to refute the point?*
 
@@
 
-Although we agree that comparisons to climate model outputs are valuable comparisons, they do not fit within the main focus and direction for this article.
-This paper focuses on directly comparing radar results from 2011 and 2016, without attempting to assess whether the results are accurately representing annual accumulation rates (other studies focus more specifically on that question---Medley, Lewis, Keeler, etc.).
+*Although we agree that comparisons to climate model outputs are valuable comparisons, they do not fit within the main focus and direction for this article.
+This paper focuses on directly comparing radar results from 2011 and 2016, without attempting to assess whether the results are accurately representing annual accumulation rates (other studies focus more specifically on that question---e.g.
+Medley et al.
+(2013), Lewis et al.
+(2017), Keeler et al.
+(2020), etc.
 The target of this research is not to determine whether radar-derived accumulation is accurate compared to other independent methods, but rather whether radar-derived estimates are self-consistent over time.
@@
 We instead investigate whether estimates derived from radar remain consistent and repeatable over a 5-year period, with the main goal assessing the suitability of comparing/combining accumulation estimates from differing radar collection times.
-Additionally, the more coarse spatial footprint associated with climate modeling outputs (10's of km vs ~200 m) specifically would result in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.
-
-**We have edited the manuscript so that we refer less explicitly to climate modeling (Lines ).**
+Additionally, the more coarse spatial footprint associated with climate modeling outputs (10's of km vs ~200 m) specifically would result in ambiguity in whether the differences with radar-derived estimates result from simple spatial variability in accumulation rates or from some limitation/inaccuracy in the radar-derived results.*
 
@@
 
-We have incorporated the suggested wording (Line ).
+*We have incorporated the suggested wording (Line ).*
 
@@
 
-We corrected this typo (Line ).
+*We corrected this typo (Line ).*
 
@@
 
-We removed the extraneous phrase and split the sentence into two sentences.
+*We removed the extraneous phrase and split the sentence into two sentences.*
 
@@
 
-We replaced "control on" with "contributor to" (Line ).
+*We replaced "control on" with "contributor to" (Line ).*
 
@@
 
-We have added additional wording better highlighting the relevance of SMB, with particular reference to the work of van den Broeke et al.
+*We have added additional wording better highlighting the relevance of SMB, with particular reference to the work of van den Broeke et al. (Lines )*
 
@@
 
-**Blah Blah blah.**
+*We have added additional citations of ground-based radar studies to better highlight the increased spatial coverage of Operation IceBridge.*
 
@@
 
-We added the OIB end year of 2019.
+*We added the OIB end year of 2019.*
 
@@
 
-We have replaced "permit imaging" to "permit discernment" to better adhere to the precise definition in regards to radiostratigraphy.
+*We have replaced "permit imaging" to "permit discernment" to better adhere to the precise definition in regards to radiostratigraphy.*
 
@@
 
-**Blah blah blah**
+*The reference to climate modeling here is merely summarizing the findings of the cited study used as an example of rapidly changing surface conditions.
+Nonetheless, we have removed this example as to avoid inadvertently leading the reader to conclusions of RCM comparisons (Lines ).*
 
@@
 
-We have explicitly stated we are referring to time scales of a few years.
+*We have explicitly stated we are referring to time scales of a few years.*
 
@@
 
-We include additional wording to better clarify the distinction between our work and prior studies (Lines ).
+*We include additional wording to better clarify the distinction between our work and prior studies (Lines ).
 We specifically highlight the differences between the crossover analyses performed in Lewis et al.
 (2017) and Koenig et al.
-(2016), and why our approach better targets this question of repeatability with disparate collection times specifically.
-
-**These crossover studies show internal consistency with increasing depth (temporal stability in the sense of repeatable reconstructions across the full time series of the record) but do not assess the effect of temporal stability in regards to time passing between data collection (i.e.
-they do not distinguish crossovers for the same flight, flights from the same year, flights spaced multiple years apart in time, etc.).
-Koenig et al.
-(2016) includes additional snow deposition between flights as a potential source of error, but does not investigate the magnitude of this effect on estimate error or how the error changes as time between flights increases.
-Lewis et al. (2017) similarly make no distinction in their crossover analysis for the timing of radar collection but are instead focused only on the spatial overlap between flights.
-Additionally, Lewis et al. (2017) use flights only 2013--2014 OIB flights, limiting the potential effects of temporal instability that is the focus of our manuscript.**
+(2016), and why our approach better targets the specific question of repeatability with disparate collection times specifically.*
 
@@
 
-We have edited this section to increase clarity, and have explicitly distinguished Koenig et al.
-(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodological focus (Lines ).
+*We have edited this section to increase clarity, and have explicitly distinguished Koenig et al.
+(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodological focus (Lines ).*
 
@@
 
-We have edited the sentence to reflect that we are specifically interested in understanding repeatability of radar-derived SMB collected over a period of years.
+*We have edited the sentence to reflect that we are specifically interested in understanding repeatability of radar-derived SMB collected over a period of years.*
 
@@
 
-**Blah**
+*This is an important point that we appreciate the reviewer raising.
+We have added two citations giving context to the magnitude of small-scale spatial variability in Antarctica (Lines ) and also again make reference to this variability in our conclusions (Lines ).*
 
@@
 
-We have changed this to "Snow radar" to more clearly convey the unit in question.
+*We have changed this to "Snow radar" to more clearly convey the unit in question.*
 
@@
 
-We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ).
+*We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ).*
 
@@
 
-We have added context to explicitly show that these manual estimates are based on a single iteration of layer selection (Lines ).
+*We have added context to explicitly show that these manual estimates are based on a single iteration of layer selection (Lines ).*
 
@@
 
-We have added the precise range in accumulation rates to this section of the text (Lines ).
+*We have added the precise range in accumulation rates to this section of the text (Lines ).*
 
@@
 
-We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method (Lines ).
+*We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method (Lines ).
 Even using repeat iterations of manual tracing (either by the same individual or multiple individuals) would reflect psychological and physiological influences as much or more than the true error associated with manual tracing methods.
-We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty (Lines ).
+We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty (Lines ).*
 
@@
 
-We have removed the offending phrase (Line ).
-We have also expanded our discussion of the image quality scoring metric to clarify what these values represent (Lines ).
+*We have removed the offending phrase (Line ).
+We have also expanded our discussion of the image quality scoring metric to clarify what these values represent (Lines ).*
 
@@
 
-We have now properly defined this acronym prior to its first use (Line ).
+*We now properly define this acronym prior to its first use (Line ).*
 
@@
 
-We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the choice parameter values $S_Q$ and the cutoff percentage threshold (Lines ).
+*We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the choice parameter values $S_Q$ and the cutoff percentage threshold (Lines ).*
 
@@
 
-We have added a sentence stating such discontinuities are more common in coastal regions and areas that experience melt/refreeze (Lines ).
+*We have added a sentence stating such discontinuities are more common in coastal regions and areas that experience melt/refreeze and are therefore less impactful on our specific study (Lines ).
 We added a note specifying that 3.1% of the original data results were removed by the automated quality control flags (Lines ).
-**We added a note explicitly stating that issues in the SEAT2010-4 area were not associated with QC flagging but instead this region shows clear, well-delineated layers in both 2011 and 2016 results that do not agree with each other.**
+We added a note stating that issues in the Site B vicinity were not associated with QC flagging but instead this region shows clear, well-delineated layers in both 2011 and 2016 results that do not agree with each other (Lines ).*
 
@@
 
-We have added the mean difference in depth between 2016 and 2011 data for both the 2010 and 1990 horizons (Lines ).
+*We have added the mean difference in depth between 2016 and 2011 data for both the 2010 and 1990 horizons (Lines ).*
 
@@
 
-We leave this to the typesetting team to resolve in the final version.
+*We leave this to the typesetting team to resolve in the final version.*
 
@@
 We instead use the mean accumulation value at a given location and year to determine percent differences between different methods and flight data.
-**We have modified wording referencing this throughout the manuscript to hopefully make this distinction more clear (Lines ).**
+We have modified the wording referencing this throughout the manuscript to hopefully make this distinction more clear.*
 
@@
 
-We have fixed the typos to ensure what is discussed is clearly evident (Line ).
+*We have fixed the typos to ensure what is discussed is clearly evident (Line ).*
 
@@
 
-**Blah.**
+*We added context for this difference compared to small-scale spatial variability (Lines ), reinforcing the conclusion that the measured bias between flight results is relatively small.*
 
@@
 
-We have fixed the typo (Line ) and added language to clarify that this % bias is relative to the average accumulation across all methods for a given location and year (Lines ).
+*We have fixed the typo (Line ) and added language to clarify that this % bias is relative to the average accumulation across all methods for a given location and year (Lines ).*
 
@@
 
-We added a brief note that the biases are generally spatially correlated to a distance of ~75 km, based on an experimental variogram (Lines ).
+*We added a brief note that the biases are generally spatially correlated to a distance of ~75 km, based on an experimental variogram (Lines ).*
 
-> “pronounced decrease in accumulation” – how large is this decrease? Is it statistically significant? It looks quite large from Figure S2. Can you see this trend in the firn core?
+> **“pronounced decrease in accumulation” – how large is this decrease? Is it statistically significant? It looks quite large from Figure S2. Can you see this trend in the firn core?**
 
-**Blah**
+*I feel, especially given the space constraints, discussing the significance and source of this decrease is counterproductive. Do you agree?*
 
@@
 
-We incorporated the suggested edit (Line ).
+*We incorporated the suggested edit (Line ).*
 
-> Does SEAT2010-4 show any decrease in accumulation between 1990-2016, or is the difference entirely within the radar calculations?
+> **Does SEAT2010-4 show any decrease in accumulation between 1990-2016, or is the difference entirely within the radar calculations?**
 
-**Blah**
+*Same as the earlier question regarding the decrease in accumulation...*
 
-> Can you briefly discuss “the limitations on the repeatability of radar imaging in these regions”
+> **Can you briefly discuss “the limitations on the repeatability of radar imaging in these regions”**
 
-**Blah**
+*Once again with the space constraints I don't think it is worth it to go into this, but I'm not sure how to state that to the reviewers. Any thoughts would be helpful.*
 
@@
 
-**Blah**
-Although we believe this could be useful information, we fear its inclusion could lead to more misdirection than context.
-Due to the probabilistic nature of PAIPR, the depths assigned to a given year incorporate a greater degree of uncertainty...
+*Although we agree this would be potentially useful information, its inclusion with our specific methodology would be more involved than it might first appear.
+The PAIPR method first picks as many potential layers as possible and then assigns each layer a likelihood of representing a true annual layer.
+It then performs Monte Carlo simulations based on these probabilities to generate age-depth profile distributions for each trace.
+No picked layer therefore is absolutely considered an annual layer, making direct comparisons of layer depths to manually picked annual layers rather complicated and non-intuitive.
+We feel the necessary context and nuance required to adequately explain this in such a short paper (particularly for an aspect not directly related to the research target) would detract from the overall manuscript as it would require the removal of more pertinent information.*
 
@@
 
-Blah
+*We have edited this sentence to explicitly refer only to the 5 year separation period.*
 
@@
 
-Blah
+*We have edited this sentence to explicitly note that these results are applicable to the dry firn zone of ice sheets specifically (Lines ).*
 
@@
 
-Blah
+*There appears to be a slight misunderstanding in our statement.
+We conclude that a collection interval of 5 years appears to be sufficient, but some of our evidence supports the notion that longer intervals between data collection are likely acceptable.
+We reworded the specific sentence to better reflect this meaning and to avoid confusion (Lines ).*
 
@@
 
-Blah
+*As this is a concluding summary statement, we do not believe this is the correct location to discuss/compare results from other studies.
+We have, however, added additional commentary on Koenig et al. (2016) earlier in the manuscript (Lines ).*
 
@@
 
-We have increased the size of all fonts and the size of the colored accumulation line to aid in visibility.
+*We have increased the size of all fonts and the size of the colored accumulation line to aid in visibility.
 We have also modified the color of the thicker line (denoting overlap) to better contrast with the accumulation color mappings.
 We have changed "Mean accum" to "Mean accumulation".
-We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.
+We have further incorporated additional modifications to the figure (addition of hillshade and contour lines) to provide better contextual information of the region of study.*
 
@@
 
-**Blah**
+*We have modified the figure to make the axes and labels easier to read.
+We have also expanded this figure to include comparisons between 2011 and 2016 flights for this 5 km transect.
+We denote the location of this transect in Fig. 1 with a white star.
+This figure now appears in the Supplement due to space constraints.*
 
@@
 
-We have edited the caption to match the figure (2016-2011).
+*We have edited the caption to match the figure (2016-2011).
 We have modified the colorbar to match the bounds of the figure.
@@
 This may have been introduced during the conversion after uploading.
-If this continues to be an issue, please notify us and we will work to resolve it.
+If this continues to be an issue, please notify us and we will work to resolve it.*
 
@@
 
-The full paper can be downloaded from the PDF link on the upper right-hand side of the website.
+*The full paper can be downloaded from the PDF link on the upper right-hand side of the website.
+Nevertheless, we have updated the citation link to point directly to the PDF to avoid confusion.*
 
@@
 
-We have increased the size of the fonts used, ensured axis labels are included, and made additional changes to increase the readability of the plots in this figure.
+*We have increased the size of the fonts used, ensured axis labels are included, and made additional changes to increase the readability of the plots in this figure.*
 
@@
 
-We have clarified that the dashed lines are $1\sigma$ uncertainty bounds.
+*We have clarified that the dashed lines are $1\sigma$ uncertainty bounds.
 We have additionally clarified and expounded on the missed/extra layers and how these are limited to the manually traced estimates.
-The probabilistic nature of PAIPR minimizes the impact of this effect for the automated results.
+The probabilistic nature of PAIPR minimizes the impact of this effect for the automated results.*

### 2021-06-22 docs/submission-docs/oib-repeat/response.md
Final response to requested revisions

@@
 We, however, feel such an investigation is outside the intended scope of the current study.
-We have limited our analysis to regions where we have higher confidence in our depth-density modeling results (due to adequate *in-situ* density record coverage) and where repeat OIB flights with a number of years of separation between collection time exist.
-Although research is ongoing to extend the density model used to wider regions of Antarctica, it is not possible to complete and implement this work in the short time frame required for revisions.
+We limited our analyses to regions where (1) we have more in-situ density data to inform the depth-density modeling, (2) repeat OIB flights with multiple years of separation between collection time exists, and (3) reside towards the interior to avoid issues where melt events can significantly impact the results.
+There may still be other sites that fit these criteria, but given the limited scope of the topic, limited space in the paper, and limited time given to address the reviews, we do not expand the analyses in this study to other regions.
 We nonetheless believe the data included in the study is sufficiently spatially extensive to be useful and informative to the wider scientific community.*
@@
 *SNR of echo layers are relevant to this research, so we are glad the reviewer raised this point.
-We have added a discussion of the impact of radar attenuation on SNR and its effect on layer selection (Lines ).
-We also briefly discuss how these impacts are limited in our study (Lines ).*
+We have added a discussion of the impact of radar attenuation on SNR, its effect on layer selection, and how these impacts are limited in our study **(Lines 192-201)**.*
 
@@
 *We appreciate the reviewer bringing up this point and its significance to our results.
-We edited the manuscript to explicitly note this difference in bandwidth between the two flights and its impact on range resolution (Lines ).
-We also now mention this difference when discussing potential sources of error in our results and indicate that our results suggest such variations have minimal impact on the reproducibility of SMB estimates from radar (Lines ).*
+We edited the manuscript to explicitly note this difference in bandwidth between the two flights and its impact on range resolution **(Lines 152-155)**.
+We also now mention this difference when discussing potential sources of error in our results and indicate that our results suggest such variations have minimal impact on the reproducibility of SMB estimates from radar **(Lines 457-464)**.*
 
@@
 *This comment raises a good point regarding the selection of these thresholds.
-We have added additional discussion on the $S_Q$ parameter to better give intuition to its physical meaning (Lines ).
-We further include a brief discussion regarding the choice of parameter values, the rationale behind that choice, we indicate that a small amount of the overall dataset (~3%) were removed through this, and present some evidence supporting our choice of parameter values (Lines ).*
+We have added additional discussion on the $S_Q$ parameter to better give intuition to its physical meaning **(Supplement Lines 27-30)**.
+We further include a brief discussion regarding the choice of parameter values, the rationale behind that choice, we indicate that a small amount of the overall dataset (~3%) were removed through this, and present some evidence supporting our choice of parameter values **(Lines 41-49)**.*
 
@@
 
-*We have added clarifying language to address this confusion (Lines ).
+*We have added clarifying language to address this confusion **(Lines 67-71)**.
 Essentially we optimize the logistic function parameters using manual layer traces, which we repeat across all traces of multiple echograms.
@@
 We have modified language throughout the manuscript to make this point more clear.
-We have specifically removed all references to *in-situ data* throughout the manuscript to avoid additional confusion.*
+We have specifically removed all references to in-situ data throughout the manuscript to avoid additional confusion.*
 
@@
 We elected to use a spatially-varying mean to scale our biases due to the large gradient in accumulation rates across the region (varying by a factor of ~3).
-We have removed the term *in-situ* and replaced it with wording that more clearly articulates this definition (Lines ).*
+We have removed the term in-situ and replaced it with wording that more clearly articulates this definition **(Lines 371-372, 377-379)**.*
 
@@
 *We appreciate this suggestion and agree some discussion of these correlations would be beneficial.
-Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections (Lines ).
+Although a full investigation of these relationships is beyond the scope of this short paper, we have included additional language briefly discussing this and our key findings regarding these connections **(Lines 393-403)**.
+We also include an additional figure in the Supplement that visually summarizes these findings.
 We find no obvious correlations between the strength of the bias and surface slope, surface aspect, plane altitude above the ice sheet surface, aircraft pitch, or aircraft roll.
@@
 
-*We have added language discussing the range in ice velocities for the region and our rationale for considering horizontal advection to be negligible for most of our data in this study (Lines ).*
+*We have added language discussing the range in ice velocities for the region and our rationale for considering horizontal advection to be negligible for most of our data in this study **(Lines 470-474)**.*
 
-> **Do you have any insights from this work on the length scale over which measurements can be considered repeatable?
+> Do you have any insights from this work on the length scale over which measurements can be considered repeatable?
 I’m assuming from the high RMSE that variability at the trace-by-trace level is quite high, but that averaged over some distance the agreement between seasons improved.
-If so that, would also be a useful note to add to the discussion/conclusions.**
+If so that, would also be a useful note to add to the discussion/conclusions.
 
-*I don't really have the space to go into this at all, so any suggestions on how best to respond to this comment?*
+*While we don't directly address this question in the text, we do add a brief discussion of the results of a semivariogram analysis that shows the spatial correlation length scale of biases in the region the implications of this **(Lines 389-393)**.
+Unfortunately, there was not enough space to add any additional discussion or analysis, and we prioritized other additions/edits here.*
 
@@
 
-*We have addressed with oversight and explicitly state in the figure caption that these lines represent $\pm1 \sigma$ errors.*
+*We have addressed this oversight and explicitly state in the figure caption that these lines represent $\pm1 \sigma$ errors.*
 
@@
 
-> **The error analysis could be improved significantly by comparing the differences between each method with the error one would expect from repeat manual layer tracing.**
+> The error analysis could be improved significantly by comparing the differences between each method with the error one would expect from repeat manual layer tracing.
 
-*I'm still not really sure how to respond to this. It is not something that will be addressed in the revised manuscript so any recommendations on the best way to refute the point?*
+*Although we agree such a comparison would be interesting, we feel the assessment of uncertainties associated with repeat manual tracing to the same quantitative level and rigor as PAIPR is beyond the constraints and scope of this paper.
+Since we are not able to produce uncertainties comparable to PAIPR, we instead discuss the difference in uncertainties between the two **(Lines 329-341)** to make it clear to the reader they are not perfect comparisons.
+We added clarifying language to the manuscript discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it **(Lines 335-338)**.
+We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty.
+From our overview of the published literature regarding the manual tracing of annual layers in radar, the use of repeat manual tracing to determine quantitative error bounds does not appear common.
+We therefore believe our exclusion of it here (especially given the focus for our study and the brevity of the publication) is acceptable.*
 
@@
 
-*We have incorporated the suggested wording (Line ).*
+*We incorporated the suggested wording **(Lines 9-10)**.*
 
@@
 
-*We have incorporated the suggested wording (Line ).*
+*We incorporated the suggested wording **(Line 11-12)**.*
 
@@
 
-*We corrected this typo (Line ).*
+*We corrected this typo **(Line 4)**.*
 
@@
 
-*We removed the extraneous phrase and split the sentence into two sentences.*
+*We removed the extraneous phrase and split the sentence into two sentences **(Lines 19-24)**.*
 
@@
 
-*We replaced "control on" with "contributor to" (Line ).*
+*We replaced "control on" with "contributor to" **(Line 34-35)**.*
 
@@
 
-*We have added additional wording better highlighting the relevance of SMB, with particular reference to the work of van den Broeke et al. (Lines )*
+*We added additional wording better highlighting the relevance of SMB, with particular reference to the work of van den Broeke et al. **(Lines 41-45)***
 
@@
 
-*We have incorporated the suggested wording (Line ).*
+*We incorporated the suggested wording **(Line 45)**.*
 
@@
 
-*We have added additional citations of ground-based radar studies to better highlight the increased spatial coverage of Operation IceBridge.*
+*We added additional citations of ground-based radar studies to better highlight the increased spatial coverage of Operation IceBridge **(Line 64)**.*
 
@@
 
-*We added the OIB end year of 2019.*
+*We added the OIB end year of 2019 **(Line 70)**.*
 
@@
 
-*We have replaced "permit imaging" to "permit discernment" to better adhere to the precise definition in regards to radiostratigraphy.*
+*We replaced "permit imaging" to "permit discernment" to better adhere to the precise definition in regards to radiostratigraphy **(Line 77)**.*
 
@@
 *The reference to climate modeling here is merely summarizing the findings of the cited study used as an example of rapidly changing surface conditions.
-Nonetheless, we have removed this example as to avoid inadvertently leading the reader to conclusions of RCM comparisons (Lines ).*
+Nonetheless, we removed this example as to avoid inadvertently leading the reader to conclusions of RCM comparisons **(Lines 95-100)**.*
 
@@
 
-*We have explicitly stated we are referring to time scales of a few years.*
+*We have explicitly stated we are referring to time scales of a few years **(Line 104)**.*
 
@@
 
-*We include additional wording to better clarify the distinction between our work and prior studies (Lines ).
+*We include additional wording to better clarify the distinction between our work and prior studies **(Lines 111-125)**.
 We specifically highlight the differences between the crossover analyses performed in Lewis et al.
@@
 
-*We have edited this section to increase clarity, and have explicitly distinguished Koenig et al.
-(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodological focus (Lines ).*
+*We edited this section to increase clarity, and have explicitly distinguished Koenig et al.
+(2016) as also focusing on single year accumulation rates while highlighting again the difference in and significance of our methodological focus **(Lines 113-117)**.*
 
@@
 
-*We have edited the sentence to reflect that we are specifically interested in understanding repeatability of radar-derived SMB collected over a period of years.*
+*We edited the sentence to reflect that we are specifically interested in understanding repeatability of radar-derived SMB collected over a period of years **(Lines 136-137)**.*
 
@@
 *This is an important point that we appreciate the reviewer raising.
-We have added two citations giving context to the magnitude of small-scale spatial variability in Antarctica (Lines ) and also again make reference to this variability in our conclusions (Lines ).*
+We added two citations giving context to the magnitude of small-scale spatial variability in Antarctica **(Lines 464-469)**.*
 
@@
 
-*We have changed this to "Snow radar" to more clearly convey the unit in question.*
+*We changed this to "Snow Radar" to more clearly convey the unit in question **(Line 159)**.*
 
@@
 
-*We added additional sentences to describe this method and the specific advantages/rationale for using it (Lines ).*
+*We added additional sentences to describe this method and the specific advantages/rationale for using it **(Lines 239-237)**.*
 
@@
 
-*We have added context to explicitly show that these manual estimates are based on a single iteration of layer selection (Lines ).*
+*We added context to explicitly show that these manual estimates are based on a single iteration of layer selection **(Lines 320-322)**.*
 
@@
 
-*We have added the precise range in accumulation rates to this section of the text (Lines ).*
+*We added the precise range in accumulation rates to this section of the text **(Lines 324)**.*
 
@@
 
-*We have added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method (Lines ).
-Even using repeat iterations of manual tracing (either by the same individual or multiple individuals) would reflect psychological and physiological influences as much or more than the true error associated with manual tracing methods.
-We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty (Lines ).*
+*We added clarifying language discussing that the manual method indeed entails additional uncertainty, but that we do not attempt to quantify it due to the subjective nature of the method **(Lines 335-338)**.
+<!-- Even using repeat iterations of manual tracing (either by the same individual or multiple individuals) would reflect psychological and physiological influences as much or more than the true error associated with manual tracing methods.
+-->
+We instead use this to highlight one of the key advantages of an automated method like PAIPR in that it allows for more rigorous quantification of this uncertainty.*
 
@@
 
-*We have removed the offending phrase (Line ).
-We have also expanded our discussion of the image quality scoring metric to clarify what these values represent (Lines ).*
+*We removed the offending phrase **(Supplement Line 17)**.
+We also expanded our discussion of the image quality scoring metric to clarify what these values represent **(Supplement Lines 27-30)**.*
 
@@
 
-*We now properly define this acronym prior to its first use (Line ).*
+*We now properly define this acronym prior to its first use **(Line 184)**.*
 
@@
 
-*We have expanded our discussion of the image quality scoring metric to clarify what these values represent, and the choice parameter values $S_Q$ and the cutoff percentage threshold (Lines ).*
+*We expanded our discussion of the image quality scoring metric to clarify what these values represent, and the choice parameter values $S_Q$ and the cutoff percentage threshold **(Supplement Lines 41-46)**.*
 
@@
 
-*We have added a sentence stating such discontinuities are more common in coastal regions and areas that experience melt/refreeze and are therefore less impactful on our specific study (Lines ).
-We added a note specifying that 3.1% of the original data results were removed by the automated quality control flags (Lines ).
-We added a note stating that issues in the Site B vicinity were not associated with QC flagging but instead this region shows clear, well-delineated layers in both 2011 and 2016 results that do not agree with each other (Lines ).*
+*We added a sentence stating such discontinuities are more common in coastal regions and areas that experience melt/refreeze and are therefore less impactful on our specific study **(Supplement Lines 62-65)**.
+We added a note specifying that 3.1% of the original data results were removed by the automated quality control flags **(Supplement Lines 46-47)**.
+We added a note stating that issues in the Site B vicinity were not associated with QC flagging but instead this region shows clear, well-delineated layers in both 2011 and 2016 results that do not agree with each other **(Lines 412-414)**.*
 
@@
 
-*We have added the mean difference in depth between 2016 and 2011 data for both the 2010 and 1990 horizons (Lines ).*
+*We added the mean difference in depth between 2016 and 2011 data for both the 2010 and 1990 horizons **(Lines 352-354)**.*
 
@@
 
-*We leave this to the typesetting team to resolve in the final version.*
+*This issue is resolved with the added revisions. We trust the typesetting team to resolve this if it occurs in final preparations for publication.*
 
@@
 
-*We have fixed the typos to ensure what is discussed is clearly evident (Line ).*
+*We fixed the typos to ensure what is discussed is clearly evident **(Lines 361-362)**.*
 
@@
 
-*We added context for this difference compared to small-scale spatial variability (Lines ), reinforcing the conclusion that the measured bias between flight results is relatively small.*
+*We added context for this difference compared to small-scale spatial variability **(Lines 464-469)**, reinforcing the conclusion that the measured bias between flight results is relatively small.*
 
@@
 
-*We have fixed the typo (Line ) and added language to clarify that this % bias is relative to the average accumulation across all methods for a given location and year (Lines ).*
+*We fixed the typo **(Line 370)** and added language to clarify that this % bias is relative to the average accumulation across all methods for a given location and year **(Lines 371-372)**.*
 
@@
 
-*We added a brief note that the biases are generally spatially correlated to a distance of ~75 km, based on an experimental variogram (Lines ).*
+*We added a brief note that the biases are generally spatially correlated to a distance of ~50 km, based on an experimental variogram **(Lines 389-393)**.*
 
-> **“pronounced decrease in accumulation” – how large is this decrease? Is it statistically significant? It looks quite large from Figure S2. Can you see this trend in the firn core?**
+> “pronounced decrease in accumulation” – how large is this decrease? Is it statistically significant? It looks quite large from Figure S2. Can you see this trend in the firn core?
 
-*I feel, especially given the space constraints, discussing the significance and source of this decrease is counterproductive. Do you agree?*
+*The limited space of this publication does not permit a broader discussion on the significance and source of this decrease.
+We merely mention it in the text as the reason for the discrepancy between 2016 and 2011 results without additional discussion or theorizing as to its cause.
+We make no direct comparisons in the text to the results from firn/ice cores and have added language to clarify this point.
+We reference the SEAT2010-4 only in the context that Burgener et al.
+expressed difficulties in the dating of this core as well, providing evidence of pervasive issues in the area from both radar imaging and geochemical analyses.*
 
@@
 
-*We incorporated the suggested edit (Line ).*
+*We incorporated the suggested edit **(Line 414)**.*
 
-> **Does SEAT2010-4 show any decrease in accumulation between 1990-2016, or is the difference entirely within the radar calculations?**
+> Does SEAT2010-4 show any decrease in accumulation between 1990-2016, or is the difference entirely within the radar calculations?
 
-*Same as the earlier question regarding the decrease in accumulation...*
+*Yes, the SEAT2010-4 core does show decreased accumulation between 1990-2016.
+Unfortunately, we do not include this discussion in the paper due to space constraints.
+While we agree with the reviewer that this is an interesting point, we felt this discussion did not add as much to the improvement of the manuscript as the other requested revisions.*
 
-> **Can you briefly discuss “the limitations on the repeatability of radar imaging in these regions”**
+> Can you briefly discuss “the limitations on the repeatability of radar imaging in these regions”
 
-*Once again with the space constraints I don't think it is worth it to go into this, but I'm not sure how to state that to the reviewers. Any thoughts would be helpful.*
+*We add additional discussion on the repeatability of radar in the region in the Conclusion, emphasizing the repeatability despite many potential sources of uncertainty **(Lines 457-476)**.
+We also discuss some of the challenges at local scales, where greater biases can be introduced, but that at more regional scales these biases are minimized.*
 
@@
 
-*We have edited this sentence to explicitly refer only to the 5 year separation period.*
+*We have edited this sentence to explicitly refer only to the 5 year separation period **(Lines 457)**.*
 
@@
 
-*We have edited this sentence to explicitly note that these results are applicable to the dry firn zone of ice sheets specifically (Lines ).*
+*We have edited this sentence to explicitly note that these results are applicable to the dry firn zone of ice sheets specifically **(Lines 481-482, 505)**.*
 
@@
 We conclude that a collection interval of 5 years appears to be sufficient, but some of our evidence supports the notion that longer intervals between data collection are likely acceptable.
-We reworded the specific sentence to better reflect this meaning and to avoid confusion (Lines ).*
+We reworded the specific sentence to better reflect this meaning and to avoid confusion **(Lines 484-486)**.*
 
@@
 *As this is a concluding summary statement, we do not believe this is the correct location to discuss/compare results from other studies.
-We have, however, added additional commentary on Koenig et al. (2016) earlier in the manuscript (Lines ).*
+We have, however, added additional commentary on Koenig et al. (2016) earlier in the manuscript **(Lines 113-117)**.*
 
@@
 
-*The full paper can be downloaded from the PDF link on the upper right-hand side of the website.
-Nevertheless, we have updated the citation link to point directly to the PDF to avoid confusion.*
+*We have modified the citation to point directly to the full article.*
 

### 2021-06-22 docs/grsl-analysis.py
Prettifies correlation plots for revised submission

@@ -1290,7 +1290,7 @@ dist_2016['plane_elev'] = dist_2016['altitude_res']-gdf_PAIPR['elev']
 #         columns=['time', 'geometry']).add_suffix('_2016')
 # gdf_PAIPR = gdf_PAIPR.join(dist_2016)
 
-plane_res = (dist_2016 - dist_2011).drop(
+plane_res = abs(dist_2016 - dist_2011).drop(
     columns=['altitude_res', 'distance_res'])
 
 # Add column for absolute value of accum res
@@ -1301,35 +1301,60 @@ gdf_PAIPR = gdf_PAIPR.join(plane_res)
 
 # %% Paired correlation plots for variables
 
-# sns.pairplot(
-#     data=gdf_PAIPR, 
-#     vars=[
-#         'accum_res', 'trace_dist', 'elev', 'slope', 'aspect', 'altitude_2011', 'heading_2011', 'pitch_2011', 'roll_2011', 
-#         'altitude_2016', 'heading_2016', 'pitch_2016', 'roll_2016'], 
-#     kind='kde', dropna=True)
+lab_replace = {
+    'res_abs': 'Accum % bias',
+    'trace_dist': 'Trace distance (m)', 
+    'elev': 'Elevation (m)', 
+    'slope': 'Slope', 'aspect':'Aspect (degrees)',
+    'plane_elev':'Altitude bias (m)', 
+    'heading_res':'Heading bias (degrees)', 
+    'pitch_res':'Pitch bias (degrees)', 
+    'roll_res':'Roll bias (degrees)'}
 
-sns.pairplot(
-    data=gdf_PAIPR, 
+pplt1 = sns.pairplot(
+    data=gdf_PAIPR, kind='scatter',
     x_vars=[
-        'accum_res', 'res_abs', 'accum_mu', 
-        'trace_dist', 'elev', 'slope', 'aspect'],
-    y_vars=['accum_res', 'res_abs', 'accum_mu'], 
-    diag_kind='kde', dropna=True)
+        'trace_dist', 'elev', 
+        'slope', 'aspect'], 
+    y_vars=['res_abs'], dropna=True, 
+    plot_kws={
+        # 'line_kws':{'color':'red'}, 
+        # 'scatter_kws': {'alpha': 0.1}}, 
+        'alpha':0.1},
+    height=5)
 
-# sns.pairplot(
-#     data=gdf_PAIPR, 
-#     x_vars=['accum_res', 'res_abs', 
[…]

### 2021-07-16 docs/submission-docs/oib-repeat/response.md
Adds final responses to GRSL edits

@@
 
-We include two copies of the manuscript in this submission.
-The first is a direct update from the initial submission in the.
-The second highlights the specific changes we have made since the original submission with additions in blue and deletions in red.
-Line numbers in the following comments reference the location of edits in the second, color-highlighted document.
+We include two copies of the manuscript (and two corresponding copies of the supplement) in this submission.
+The first copy (main-revised.pdf and supplement-revised.pdf) is a direct update from the initial submission.
+The second copy (main-revised_diff.pdf and supp-revised_diff.pdf) highlights the specific changes we have made since the original submission with additions in blue and deletions in red.
+Line numbers in the following comments reference the location of edits in the second, color-highlighted documents.
 

### 2021-07-16 docs/analysis.py
Grsl edits (#6)
* Post def (#5)

* Improves plotting assignment and formatting for publication figure generation

* Re-implements rio code after VSC fix; adds submission-docs to ignore file

* Updates variable names to better reflect true meaning

* Prints outputs at end of script; improved cut-off bounds for plotting; updates t_perc to actual percent values; improves final plot panel groupings

* Cleans codebase by removing legacy and deprecated code

* Updates figure file paths to reflect name changes

* Cleans up old deprecated files

* Cleans and updates script to improved processing steps

@@ -12,11 +12,14 @@ import matplotlib.gridspec as gridspec
 import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
+from cartopy import feature as cf
 from bokeh.io import output_notebook
 output_notebook()
 hv.extension('bokeh')
 gv.extension('bokeh')
 from shapely.geometry import Polygon, Point
+import xarray as xr
+from xrspatial import hillshade
 
 # hv.archive.auto()
 
@@ -36,25 +39,20 @@ from my_functions import *
 ANT_proj = ccrs.SouthPolarStereo(
     true_scale_latitude=-71)
 
-# Define Antarctic boundary file
-shp = str(ROOT_DIR.joinpath(
-    'data/Ant_basemap/Coastline_medium_res_polygon.shp'))
-gdf_ANT = gpd.read_file(shp)
+# # Define Antarctic boundary file
+# shp = str(ROOT_DIR.joinpath(
+#     'data/Ant_basemap/Coastline_medium_res_polygon.shp'))
 
-
-import xarray as xr
-# Define Antarctic DEM file
-xr_DEM = xr.open_rasterio(
-    ROOT_DIR.joinpath(
-        'data/Antarctica_Cryosat2_1km_DEMv1.0.tif')).squeeze()
+# Create Antarctic coastline polygon
+world = gpd.read_file(
+    gpd.datasets.get_path('naturalearth_lowres'))
+Ant = world.query('continent=="Antarctica"')
+Ant = Ant.to_crs(epsg=3031)
 
 #%% Import and format PAIPR-generated results
 
 # Import raw data
 data_list = [folder for folder in DATA_DIR.glob('*')]
-# data_list = [
-#     path for path in data_list 
-#     if "20141103" not in str(path)] #Removes 2014 results, as there are too few and are suspect
 data_raw = pd.DataFrame()
 for folder in data_list:
     data = import_PAIPR(folder)
@@ -67,8 +65,6 @@ data_0 = data_raw.query(
     ['collect_time', 'Year']).reset_index(drop=True)
 
 # Format and sort results for further processing
-# data_form = format_PAIPR(data_0).drop(
-#     'elev', axis=1)
 data_form = format_PAIPR(data_0)
 
[…]

### 2021-07-16 docs/grsl-analysis.py

@@ -11,20 +11,24 @@ import matplotlib.pyplot as plt
 import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
-from bokeh.io import output_notebook
 from shapely.geometry import Point
-output_notebook()
+# from bokeh.io import output_notebook
+# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
+# hv.extension('matplotlib')
+# gv.extension('matplotlib')
 import panel as pn
 import seaborn as sns
 import xarray as xr
+from xrspatial import hillshade
+from scipy.spatial.distance import pdist
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
 
 # Set project root directory
-ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
+ROOT_DIR = Path(__file__).parents[1]
 
 # Import custom project functions
 import sys
@@ -38,7 +42,7 @@ from my_functions import *
 # %% Import PAIPR-generated data
 
 # Import 20111109 results
-dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
+dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
 data_raw = import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
@@ -53,7 +57,7 @@ std2011_ALL = data_2011.pivot(
     index='Year', columns='trace_ID', values='std')
 
 # Import 20161109 results
-dir2 = ROOT_DIR.joinpath('data/PAIPR-outputs/20161109/')
+dir2 = ROOT_DIR.joinpath('data/PAIPR-repeat/20161109/smb/')
 data_raw = import_PAIPR(dir2)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
@@ -97,6 +101,9 @@ std_2016 = std2016_ALL.iloc[
 gdf_PAIPR = gpd.GeoDataFrame(
     {'ID_2011': dist_overlap.index.values, 
     'ID_2016': dist_overlap['trace_ID'].values, 
+    'trace_dist': dist_overlap['distance'].values,
+    'QC_2011': gdf_2011.loc[dist_overlap.index,'QC_med'].values,
+    'QC_2016': dist_overlap['QC_med'].values, 
     'accum_2011': 
[…]

### 2021-07-16 scratch/update-valid.py

@@ -1,0 +1,426 @@
+# An updated validation of the most recent PAIPR results using both firn cores and manual radar data.
+
+# %% Set up environment
+
+# Import requisite modules
+import numpy as np
+import pandas as pd
+import geopandas as gpd
+from pathlib import Path
+import matplotlib.pyplot as plt
+from cartopy import crs as ccrs
+
+# Define plotting projection to use
+ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
+
+# Set project root directory
+ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
+
+# Import custom project functions
+import sys
+SRC_DIR = ROOT_DIR.joinpath('src')
+sys.path.append(str(SRC_DIR))
+from my_functions import *
+
+# %% Import and format radar data
+
+# Import and format PAIPR results
+dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
+data_raw = import_PAIPR(dir1)
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+paipr_data = format_PAIPR(
+    data_0, start_yr=1980, end_yr=2010).drop(
+    'elev', axis=1)
+paipr_ALL = paipr_data.pivot(
+    index='Year', columns='trace_ID', values='accum')
+std_ALL = paipr_data.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Import and format manual results
+dir_0 = ROOT_DIR.joinpath('data/smb_manual/20111109/')
+data_0 = import_PAIPR(dir_0)
+man_data = format_PAIPR(
+    data_0, start_yr=1980, end_yr=2010).drop(
+    'elev', axis=1)
+man_ALL = man_data.pivot(
+    index='Year', columns='trace_ID', values='accum')
+manSTD_ALL = man_data.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Create gdf of mean results for each trace and 
+# transform to Antarctic Polar Stereographic
+gdf_paipr = long2gdf(paipr_data)
+gdf_paipr.to_crs(epsg=3031, inplace=True)
+gdf_man = long2gdf(man_data)
+gdf_man.to_crs(epsg=3031, inplace=True)
+
+# Drop unwanted columns
+gdf_paipr.drop(
[…]

### 2021-07-16 src/my_functions.py

@@ -8,10 +8,11 @@ from pathlib import Path
 import time
 import shutil
 import os
-# import rasterio as rio
+import rasterio as rio
 from sklearn.neighbors import BallTree
 from scipy import signal
 import statsmodels.tsa.stattools as tsa
+import richdem as rd
 
 # Function to import and concatenate PAIPR .csv files
 def import_PAIPR(input_dir):
@@ -77,7 +78,7 @@ def format_PAIPR(data_df, start_yr=None, end_yr=None):
         # New df (in long format) with accum data assigned
         data_long = (
             data_df.filter(['trace_ID', 'collect_time', 
-            'QC_flag', 'Lat', 'Lon', 
+            'QC_flag', 'QC_med', 'Lat', 'Lon', 
             'elev', 'Year']).assign(
                 accum=data_df['accum_mu'], 
                 std=data_df['accum_std']).reset_index(drop=True))
@@ -489,10 +490,10 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
 
     trend_mu = np.nanmean(trends_bs, axis=0)
     intercept_mu = np.nanmean(intercepts, axis=0)
-    trend_05 = np.nanpercentile(trends_bs, 2.5, axis=0)
-    trend_95 = np.nanpercentile(trends_bs, 97.5, axis=0)
+    trendCI_lb = np.nanpercentile(trends_bs, 2.5, axis=0)
+    trendCI_ub = np.nanpercentile(trends_bs, 97.5, axis=0)
 
-    return trend_mu, intercept_mu, trend_05, trend_95
+    return trend_mu, intercept_mu, trendCI_lb, trendCI_ub
 
 # Function to perform autocorrelation
 def acf(df):
@@ -513,94 +514,215 @@ def acf(df):
 
     return acf_df
 
-# import requests
-# def get_REMA(tile_idx, output_dir):
-#     """
-#     Downloads, unzips, and saves DEM tiles from the Reference Elevation Model of Antarctica (REMA) dataset.
-#     Required inputs:
-#     tile_idx {pandas.core.frame.DataFrame}: Dataframe with the names (name), tile IDs (tile), and file urls (fileurl) of the requested data for download, taken from the [REMA tile shapefile](http://data.pgc.umn.edu/elev/dem/setsm/REMA/indexes/).
-#     output_dir {pathlib.PosixPath}: Path of directory at which to download the requested file.
-#     """
+import requests
[…]

### 2021-07-16 docs/analysis.py
Updates custom functions to use proper importing of modules within the new 'my_mods' package

@@ -13,10 +13,10 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from cartopy import feature as cf
-from bokeh.io import output_notebook
-output_notebook()
-hv.extension('bokeh')
-gv.extension('bokeh')
+# from bokeh.io import output_notebook
+# output_notebook()
+# hv.extension('bokeh')
+# gv.extension('bokeh')
 from shapely.geometry import Polygon, Point
 import xarray as xr
 from xrspatial import hillshade
@@ -33,7 +33,8 @@ DATA_DIR = ROOT_DIR.joinpath('data/PAIPR-outputs')
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, stats
+import my_mods.spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(
@@ -55,7 +56,7 @@ Ant = Ant.to_crs(epsg=3031)
 data_list = [folder for folder in DATA_DIR.glob('*')]
 data_raw = pd.DataFrame()
 for folder in data_list:
-    data = import_PAIPR(folder)
+    data = paipr.import_PAIPR(folder)
     data_raw = data_raw.append(data)
 
 # Remove results for below QC data reliability
@@ -65,7 +66,7 @@ data_0 = data_raw.query(
     ['collect_time', 'Year']).reset_index(drop=True)
 
 # Format and sort results for further processing
-data_form = format_PAIPR(data_0)
+data_form = paipr.format_PAIPR(data_0)
 
 # Create time series arrays for annual accumulation 
 # and error
@@ -76,7 +77,7 @@ std_ALL = data_form.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_traces = long2gdf(data_form)
+gdf_traces = paipr.long2gdf(data_form)
 gdf_traces.to_crs(epsg=3031, inplace=True)
 
 
@@ -84,7 +85,7 @@ gdf_traces.to_crs(epsg=3031, inplace=True)
 
 # xr_vice = xr.open_dataset(ROOT_DIR.joinpath(
[…]

### 2021-07-16 docs/diss-ch5.py

@@ -13,10 +13,9 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from shapely.geometry import Point, Polygon
-# from bokeh.io import output_notebook
-# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
+from scipy import signal
 
 # Set project root directory
 ROOT_DIR = Path('/home/durbank/Documents/Research/Antarctica/WAIS-central/')
@@ -28,7 +27,8 @@ DATA_DIR = ROOT_DIR.joinpath('data')
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, stats
+import my_mods.spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -108,7 +108,7 @@ data_list = [
     DATA_DIR.joinpath('PAIPR-outputs').glob('*')]
 data_raw = pd.DataFrame()
 for folder in data_list:
-    data = import_PAIPR(folder)
+    data = paipr.import_PAIPR(folder)
     data_raw = data_raw.append(data)
 
 # Remove results for below QC data reliability
@@ -118,7 +118,7 @@ data_0 = data_raw.query(
     ['collect_time', 'Year']).reset_index(drop=True)
 
 # Format and sort results for further processing
-data_form = format_PAIPR(data_0)
+data_form = paipr.format_PAIPR(data_0)
 
 # Create time series arrays for annual accumulation 
 # and error
@@ -129,16 +129,16 @@ std_ALL = data_form.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_traces = long2gdf(data_form)
+gdf_traces = paipr.long2gdf(data_form)
 gdf_traces.to_crs(epsg=3031, inplace=True)
 
 # %% PAIPR data aggregation by grid
 
 # Combine trace time series based on grid cells
 grid_res = 1000
-tmp_grids = pts2grid(gdf_traces, resolution=grid_res)
+tmp_grids = so.pts2grid(gdf_traces, resolution=grid_res)
 (gdf_grid_ALL, accum_grid_ALL, 
-    MoE_grid_ALL, yr_count_ALL) = trace_combine(
[…]

### 2021-07-16 docs/grsl-analysis.py

@@ -12,17 +12,12 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from shapely.geometry import Point
-# from bokeh.io import output_notebook
-# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
-# hv.extension('matplotlib')
-# gv.extension('matplotlib')
 import panel as pn
 import seaborn as sns
 import xarray as xr
 from xrspatial import hillshade
-from scipy.spatial.distance import pdist
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -34,7 +29,8 @@ ROOT_DIR = Path(__file__).parents[1]
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, rema, viz, stats
+import my_mods.spat_ops as so
 
 # %%[markdown]
 # ## Data and Study Site
@@ -43,12 +39,12 @@ from my_functions import *
 
 # Import 20111109 results
 dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
-data_raw = import_PAIPR(dir1)
+data_raw = paipr.import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-data_2011 = format_PAIPR(
+data_2011 = paipr.format_PAIPR(
     data_0, start_yr=1990, end_yr=2010).drop(
     'elev', axis=1)
 a2011_ALL = data_2011.pivot(
@@ -58,12 +54,12 @@ std2011_ALL = data_2011.pivot(
 
 # Import 20161109 results
 dir2 = ROOT_DIR.joinpath('data/PAIPR-repeat/20161109/smb/')
-data_raw = import_PAIPR(dir2)
+data_raw = paipr.import_PAIPR(dir2)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-data_2016 = format_PAIPR(
+data_2016 = paipr.format_PAIPR(
[…]

### 2021-07-16 docs/update-valid.py

@@ -9,6 +9,8 @@ import geopandas as gpd
 from pathlib import Path
 import matplotlib.pyplot as plt
 from cartopy import crs as ccrs
+import holoviews as hv
+hv.extension('bokeh', 'matplotlib')
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -20,18 +22,19 @@ ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, viz
+import my_mods.spat_ops as so
 
 # %% Import and format radar data
 
 # Import and format PAIPR results
 dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
-data_raw = import_PAIPR(dir1)
+data_raw = paipr.import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-paipr_data = format_PAIPR(
+paipr_data = paipr.format_PAIPR(
     data_0, start_yr=1980, end_yr=2010).drop(
     'elev', axis=1)
 paipr_ALL = paipr_data.pivot(
@@ -41,8 +44,8 @@ std_ALL = paipr_data.pivot(
 
 # Import and format manual results
 dir_0 = ROOT_DIR.joinpath('data/smb_manual/20111109/')
-data_0 = import_PAIPR(dir_0)
-man_data = format_PAIPR(
+data_0 = paipr.import_PAIPR(dir_0)
+man_data = paipr.format_PAIPR(
     data_0, start_yr=1980, end_yr=2010).drop(
     'elev', axis=1)
 man_ALL = man_data.pivot(
@@ -52,9 +55,9 @@ manSTD_ALL = man_data.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_paipr = long2gdf(paipr_data)
+gdf_paipr = paipr.long2gdf(paipr_data)
 gdf_paipr.to_crs(epsg=3031, inplace=True)
-gdf_man = long2gdf(man_data)
+gdf_man = paipr.long2gdf(man_data)
[…]

### 2021-07-16 src/my_mods/stats.py

@@ -1,0 +1,170 @@
+# Module containing custon stats functions and operations
+
+# Requisite modules
+import pandas as pd
+import time
+import numpy as np
+from scipy import signal
+import statsmodels.tsa.stattools as tsa
+from scipy.spatial.distance import pdist
+
+def trend_bs(df, nsim, df_err=pd.DataFrame()):
+    """
+    Dpc string goes here.
+    """
+    tic = time.perf_counter()
+    
+    # Preallocate results arrays
+    trends_bs = pd.DataFrame(columns=df.columns)
+    intercepts = pd.DataFrame(columns=df.columns)
+
+    # In no errors exist, create neutral weights
+    if df_err.empty:
+        df_err = pd.DataFrame(
+            np.ones(df.shape), index=df.index, 
+            columns=df.columns)
+
+    # Peform bootstrapping
+    for _ in range(nsim):
+        # Randomly resample data
+        data_bs = df.sample(
+            len(df), replace=True).sort_index()
+        
+        # Generate mean weights
+        weights_bs = (
+                1/df_err.loc[data_bs.index]).mean(axis=1)
+
+        # Check for nans (requires for loop w/ 2D array)
+        if np.any(np.isnan(data_bs)):
+            
+            data_ma = np.ma.array(data_bs, mask=np.isnan(data_bs))
+            coeffs = np.zeros((2,data_ma.shape[1]))
+            for idx in range(coeffs.shape[1]):
+                data_i = data_ma[:,idx]
+                if (np.invert(data_i.mask)).sum() < 5:
+                    coeffs[:,idx] = np.nan
+                else:
+                    coeffs[:,idx] = np.ma.polyfit(
+                        data_bs.index, data_i, 1, w=weights_bs)
+                
+        
+        else:
+            # If no nan's present, perform vectorized fitting
+            coeffs = np.polyfit(
+                data_bs.index, data_bs, 1, w=weights_bs)
+
+        trends_bs = trends_bs.append(
+            pd.Series(coeffs[0], index=df.columns), 
+            ignore_index=True)
+        intercepts = intercepts.append(
+            pd.Series(coeffs[1], index=df.columns), 
+            ignore_index=True)
+
+    toc = time.perf_counter()
+    print(f"Execution time of bootstrapping: {toc-tic} s")
+
+    trend_mu = np.nanmean(trends_bs, axis=0)
[…]

### 2021-07-16 src/my_mods/viz.py

@@ -1,0 +1,340 @@
+# Module for project-specific visualizations and figure generation
+
+# Requisite modules
+import numpy as np
+import pandas as pd
+import geopandas as gpd
+import matplotlib.pyplot as plt
+import holoviews as hv
+
+
+def plot_TScomp(
+    ts_df1, ts_df2, gdf_combo, labels, ts_cores=None,
+    yaxis=True, xlims=None, ylims=None,
+    colors=['blue', 'red'], ts_err1=None, ts_err2=None):
+    """This is a function to generate matplotlib objects that compare spatially overlapping accumulation time series.
+
+    Args:
+        ts_df1 (pandas.DataFrame): Dataframe containing time series for the first dataset.
+        ts_df2 (pandas.DataFrame): Dataframe containing time series for the second dataset.
+        gdf_combo (geopandas.geoDataFrame): Geodataframe with entries corresponding to the paired time series locations. Also contains a column 'Site' that groups the different time series according to their manual tracing location.
+        labels (list of str): The labels used in the output plot to differentiate the time series dataframes.
+        colors (list, optional): The colors to use when plotting the time series. Defaults to ['blue', 'red'].
+        ts_err1 (pandas.DataFrame, optional): DataFrame containing time series errors corresponding to ts_df1. If "None" then the error is estimated from the standard deviations in annual results. Defaults to None.
+        ts_err2 (pandas.DataFrame, optional): DataFrame containing time series errors corresponding to ts_df2. If "None" then the error is estimated from the standard deviations in annual results. Defaults to None.
+
+    Returns:
+        matplotlib.pyplot.figure: Generated figure comparing the two overlapping time series.
+    """
+
+    # Remove observations without an assigned site
+    site_list = np.unique(gdf_combo['Site']).tolist()
+    if "Null" in site_list:
+        site_list.remove("Null")
+
+    # Generate figure with a row for each site
+    fig, axes = plt.subplots(
+        ncols=1, nrows=len(site_list), 
+        constrained_layout=True, 
+        figsize=(6,24))
+
+    for i, site in enumerate(site_list):
+        
+        # Subset results to specific site
+        idx = np.flatnonzero(gdf_combo['Site']==site)
+        df1 = ts_df1.iloc[:,idx]
+        df2 = ts_df2.iloc[:,idx]
+
[…]

### 2021-07-16 docs/analysis.py
Adds __init__ to improve import handling

@@ -34,7 +34,7 @@ import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
 from my_mods import paipr, stats
-import my_mods.spat_ops as so
+from my_mods import spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(

### 2021-07-16 docs/diss-ch5.py

@@ -28,7 +28,7 @@ import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
 from my_mods import paipr, stats
-import my_mods.spat_ops as so
+from my_mods import spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)

### 2021-07-16 docs/grsl-analysis.py

@@ -30,7 +30,7 @@ import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
 from my_mods import paipr, rema, viz, stats
-import my_mods.spat_ops as so
+from my_mods import spat_ops as so
 
 # %%[markdown]
 # ## Data and Study Site

### 2021-07-16 docs/update-valid.py

@@ -23,7 +23,7 @@ import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
 from my_mods import paipr, viz
-import my_mods.spat_ops as so
+from my_mods import spat_ops as so
 
 # %% Import and format radar data
 

### 2021-07-16 docs/analysis.py
Modules (#8)
* Updates custom functions to use proper importing of modules within the new 'my_mods' package

* Adds __init__ to improve import handling

@@ -13,10 +13,10 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from cartopy import feature as cf
-from bokeh.io import output_notebook
-output_notebook()
-hv.extension('bokeh')
-gv.extension('bokeh')
+# from bokeh.io import output_notebook
+# output_notebook()
+# hv.extension('bokeh')
+# gv.extension('bokeh')
 from shapely.geometry import Polygon, Point
 import xarray as xr
 from xrspatial import hillshade
@@ -33,7 +33,8 @@ DATA_DIR = ROOT_DIR.joinpath('data/PAIPR-outputs')
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, stats
+from my_mods import spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(
@@ -55,7 +56,7 @@ Ant = Ant.to_crs(epsg=3031)
 data_list = [folder for folder in DATA_DIR.glob('*')]
 data_raw = pd.DataFrame()
 for folder in data_list:
-    data = import_PAIPR(folder)
+    data = paipr.import_PAIPR(folder)
     data_raw = data_raw.append(data)
 
 # Remove results for below QC data reliability
@@ -65,7 +66,7 @@ data_0 = data_raw.query(
     ['collect_time', 'Year']).reset_index(drop=True)
 
 # Format and sort results for further processing
-data_form = format_PAIPR(data_0)
+data_form = paipr.format_PAIPR(data_0)
 
 # Create time series arrays for annual accumulation 
 # and error
@@ -76,7 +77,7 @@ std_ALL = data_form.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_traces = long2gdf(data_form)
+gdf_traces = paipr.long2gdf(data_form)
 gdf_traces.to_crs(epsg=3031, inplace=True)
 
 
@@ -84,7 +85,7 @@ gdf_traces.to_crs(epsg=3031, inplace=True)
 
 # xr_vice = xr.open_dataset(ROOT_DIR.joinpath(
[…]

### 2021-07-16 docs/diss-ch5.py

@@ -13,10 +13,9 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from shapely.geometry import Point, Polygon
-# from bokeh.io import output_notebook
-# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
+from scipy import signal
 
 # Set project root directory
 ROOT_DIR = Path('/home/durbank/Documents/Research/Antarctica/WAIS-central/')
@@ -28,7 +27,8 @@ DATA_DIR = ROOT_DIR.joinpath('data')
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, stats
+from my_mods import spat_ops as so
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -108,7 +108,7 @@ data_list = [
     DATA_DIR.joinpath('PAIPR-outputs').glob('*')]
 data_raw = pd.DataFrame()
 for folder in data_list:
-    data = import_PAIPR(folder)
+    data = paipr.import_PAIPR(folder)
     data_raw = data_raw.append(data)
 
 # Remove results for below QC data reliability
@@ -118,7 +118,7 @@ data_0 = data_raw.query(
     ['collect_time', 'Year']).reset_index(drop=True)
 
 # Format and sort results for further processing
-data_form = format_PAIPR(data_0)
+data_form = paipr.format_PAIPR(data_0)
 
 # Create time series arrays for annual accumulation 
 # and error
@@ -129,16 +129,16 @@ std_ALL = data_form.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_traces = long2gdf(data_form)
+gdf_traces = paipr.long2gdf(data_form)
 gdf_traces.to_crs(epsg=3031, inplace=True)
 
 # %% PAIPR data aggregation by grid
 
 # Combine trace time series based on grid cells
 grid_res = 1000
-tmp_grids = pts2grid(gdf_traces, resolution=grid_res)
+tmp_grids = so.pts2grid(gdf_traces, resolution=grid_res)
 (gdf_grid_ALL, accum_grid_ALL, 
-    MoE_grid_ALL, yr_count_ALL) = trace_combine(
[…]

### 2021-07-16 docs/grsl-analysis.py

@@ -12,17 +12,12 @@ import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
 from shapely.geometry import Point
-# from bokeh.io import output_notebook
-# output_notebook()
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
-# hv.extension('matplotlib')
-# gv.extension('matplotlib')
 import panel as pn
 import seaborn as sns
 import xarray as xr
 from xrspatial import hillshade
-from scipy.spatial.distance import pdist
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -34,7 +29,8 @@ ROOT_DIR = Path(__file__).parents[1]
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, rema, viz, stats
+from my_mods import spat_ops as so
 
 # %%[markdown]
 # ## Data and Study Site
@@ -43,12 +39,12 @@ from my_functions import *
 
 # Import 20111109 results
 dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20111109/smb/')
-data_raw = import_PAIPR(dir1)
+data_raw = paipr.import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-data_2011 = format_PAIPR(
+data_2011 = paipr.format_PAIPR(
     data_0, start_yr=1990, end_yr=2010).drop(
     'elev', axis=1)
 a2011_ALL = data_2011.pivot(
@@ -58,12 +54,12 @@ std2011_ALL = data_2011.pivot(
 
 # Import 20161109 results
 dir2 = ROOT_DIR.joinpath('data/PAIPR-repeat/20161109/smb/')
-data_raw = import_PAIPR(dir2)
+data_raw = paipr.import_PAIPR(dir2)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-data_2016 = format_PAIPR(
+data_2016 = paipr.format_PAIPR(
[…]

### 2021-07-16 notebooks/manual_comp.py

@@ -1,980 +1,0 @@
-# This script is for comparing manually-selected smb results vs PAIPR-generated SMB results, as well as comparing repeatability between 2011 and 2016, both with manual results and with PAIPR results
-
-# %%
-# Import requisite modules
-import numpy as np
-import pandas as pd
-import geopandas as gpd
-from pathlib import Path
-import matplotlib.pyplot as plt
-import geoviews as gv
-import holoviews as hv
-from cartopy import crs as ccrs
-from bokeh.io import output_notebook
-from shapely.geometry import Point
-output_notebook()
-hv.extension('bokeh')
-gv.extension('bokeh')
-import seaborn as sns
-
-# Define plotting projection to use
-ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
-
-# Set project root directory
-ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
-
-# Import custom project functions
-import sys
-SRC_DIR = ROOT_DIR.joinpath('src')
-sys.path.append(str(SRC_DIR))
-from my_functions import *
-
-# # Import Antarctic outline shapefile
-# ant_path = ROOT_DIR.joinpath(
-#     'data/Ant_basemap/Coastline_medium_res_polygon.shp')
-# ant_outline = gpd.read_file(ant_path)
-
-# %% Import manual results
-# Get list of 2011 manual files
-# list_2011 = [file for file in Path('data/smb_manual').glob('*20111109.csv')]
-dir_0 = ROOT_DIR.joinpath('data/smb_manual/20111109/')
-data_0 = import_PAIPR(dir_0)
-man_2011 = format_PAIPR(
-    data_0, start_yr=1990, end_yr=2010).drop(
-    'elev', axis=1)
-man2011_ALL = man_2011.pivot(
-    index='Year', columns='trace_ID', values='accum')
-manSTD_2011_ALL = man_2011.pivot(
-    index='Year', columns='trace_ID', values='std')
-
-# Create gdf of mean results for each trace and 
-# transform to Antarctic Polar Stereographic
-gdf_man2011 = long2gdf(man_2011)
-gdf_man2011.to_crs(epsg=3031, inplace=True)
-
[…]

### 2021-07-16 src/my_functions.py

@@ -1,391 +1,391 @@
 # Module script to store project-specific custom functions and variables
 
-# Import requisite modules
-import numpy as np
-import pandas as pd
-import geopandas as gpd
-from pathlib import Path
-import time
-import shutil
-import os
-import rasterio as rio
-from sklearn.neighbors import BallTree
-from scipy import signal
-import statsmodels.tsa.stattools as tsa
-import richdem as rd
+# # Import requisite modules
+# import numpy as np
+# import pandas as pd
+# import geopandas as gpd
+# from pathlib import Path
+# import time
+# import shutil
+# import os
+# import rasterio as rio
+# from sklearn.neighbors import BallTree
+# from scipy import signal
+# import statsmodels.tsa.stattools as tsa
+# import richdem as rd
 
-# Function to import and concatenate PAIPR .csv files
-def import_PAIPR(input_dir):
-    """
-    Function to import PAIPR-derived accumulation data.
-    This concatenates all files within the given directory into a single pandas dataframe.
+# # Function to import and concatenate PAIPR .csv files
+# def import_PAIPR(input_dir):
+#     """
+#     Function to import PAIPR-derived accumulation data.
+#     This concatenates all files within the given directory into a single pandas dataframe.
 
-    Parameters:
-    input_dir (pathlib.PosixPath): Absolute path of directory containing .csv files of PAIPR results (gamma-distributions fitted to accumulation curves). This directory can contain multiple files, which will be concatenated to a single dataframe.
-    """
+#     Parameters:
+#     input_dir (pathlib.PosixPath): Absolute path of directory containing .csv files of PAIPR results (gamma-distributions fitted to accumulation curves). This directory can contain multiple files, which will be concatenated to a single dataframe.
+#     """
 
-    data = pd.DataFrame()
-    for file in input_dir.glob("*.csv"):
-        data_f = pd.read_csv(file)
-        data = data.append(data_f)
-    data['collect_time'] = pd.to_datetime(
-        data['collect_time'])
-    return data
+#     data = pd.DataFrame()
+#     for file in input_dir.glob("*.csv"):
+#         data_f = pd.read_csv(file)
+#         data = data.append(data_f)
[…]

### 2021-07-16 src/my_mods/__init__.py

@@ -1,0 +1,12 @@
+
+# print(f'Invoking __init__.py for {__name__}')
+
+import my_mods.paipr, my_mods.rema, my_mods.spat_ops, my_mods.stats, my_mods.viz
+
+__all__ = [
+        'paipr',
+        'rema',
+        'spat_ops',
+        'stats',
+        'viz'
+        ]

### 2021-07-16 src/my_mods/paipr.py

@@ -1,0 +1,177 @@
+# Module of helper functions for formatting, splitting, and otherwise dealing with PAIPR-generated results
+
+from pathlib import Path
+import pandas as pd
+import geopandas as gpd
+import numpy as np
+import xarray as xr
+
+def import_PAIPR(input_dir):
+    """
+    Function to import PAIPR-derived accumulation data.
+    This concatenates all files within the given directory into a single pandas dataframe.
+
+    Parameters:
+    input_dir (pathlib.PosixPath): Absolute path of directory containing .csv files of PAIPR results. This directory can contain multiple files, which will be concatenated to a single dataframe.
+    """
+
+    data = pd.DataFrame()
+    for file in input_dir.glob("*.csv"):
+        data_f = pd.read_csv(file)
+        data = data.append(data_f)
+    data['collect_time'] = pd.to_datetime(
+        data['collect_time'])
+    return data
+
+def format_PAIPR(data_df, start_yr=None, end_yr=None):
+    """
+
+    """
+    # Create groups based on trace locations
+    traces = data_df.groupby(
+        ['collect_time', 'Lat', 'Lon'])
+    data_df = data_df.assign(trace_ID = traces.ngroup())
+    traces = data_df.groupby('trace_ID')
+
+    if start_yr != end_yr:
+        # Remove time series with data missing from period
+        # of interest (and clip to period of interest)
+        data_df = traces.filter(
+            lambda x: min(x['Year']) <= start_yr 
+            and max(x['Year']) >= end_yr)
+        data_df = data_df.query(
+            f"Year >= {start_yr} & Year <= {end_yr}")
+
+    # # Ensure each trace has only one time series 
+    # # (if not, take the mean of all time series)
+    # data = data.groupby(['trace_ID', 'Year']).mean()
+
+    if 'gamma_shape' in data_df.columns:
+        # Generate descriptive statistics based on 
+        # imported gamma-fitted parameters
+        alpha = data_df['gamma_shape']
+        alpha.loc[alpha<1] = 1
+        beta = 1/data_df['gamma_scale']
+        mode_accum = (alpha-1)/beta
+        var_accum = alpha/beta**2
+        
+        # New df (in long format) with accum data assigned
+        data_long = (
+            data_df.filter(['trace_ID', 'collect_time', 
+            'QC_flag', 'Lat', 'Lon', 
[…]

### 2021-07-16 src/my_mods/rema.py

@@ -1,0 +1,174 @@
+# Module for downloading and manipulating REMA elevation data
+
+# Requisite modules
+from pathlib import Path
+import requests
+import shutil
+import os
+import richdem as rd
+import numpy as np
+import rasterio as rio
+
+def get_REMA(tile_idx, output_dir):
+    """
+    Downloads, unzips, and saves DEM tiles from the Reference Elevation Model of Antarctica (REMA) dataset.
+    Required inputs:
+    tile_idx {pandas.core.frame.DataFrame}: Dataframe with the names (name), tile IDs (tile), and file urls (fileurl) of the requested data for download, taken from the [REMA tile shapefile](http://data.pgc.umn.edu/elev/dem/setsm/REMA/indexes/).
+    output_dir {pathlib.PosixPath}: Path of directory at which to download the requested file.
+    """
+
+    for idx, row in tile_idx.iterrows():
+        f_dir = output_dir.joinpath(row.tile)
+
+        if not f_dir.exists():
+            f_dir.mkdir(parents=True)
+            zip_path = f_dir.joinpath('tmp.tar.gz')
+            r = requests.get(row.fileurl, stream=True)
+            print(f"Downloading tile {f_dir.name}")
+            with open(zip_path, 'wb') as zFile:
+                for chunk in r.iter_content(
+                        chunk_size=1024*1024):
+                    if chunk:
+                        zFile.write(chunk)
+            print(f"Unzipping tile {f_dir.name}")
+            shutil.unpack_archive(zip_path, f_dir)
+            os.remove(zip_path)
+        else:
+            print(f"REMA tile {f_dir.name} already exists locally, moving to next download")
+    print("All requested files downloaded")
+
+
+def calc_topo(dem_path):
+    """
+    Calculates slope and aspect from given DEM and saves output.
+    The function checks to see whether a slope/aspect file has already been created so as to avoid needless processing.
+    
+    Parameters:
+    dem_path (pathlib.PosixPath): The relative or absolute path to an input DEM file.
+
+    Dependencies: 
+    richdem module
+    GDAL binaries
+    pathlib module
+    """
+    slope_path = Path(
+        str(dem_path).replace("dem", "slope"))
+    aspect_path = Path(
+        str(dem_path).replace("dem", "aspect"))
+
+    if ((not slope_path.is_file()) or 
+            (not aspect_path.is_file())):
+        
+        # Load DEM
+        dem = rd.LoadGDAL(str(dem_path))
+        
[…]

### 2021-07-16 docs/analysis.py
Implements walker test for field significance

@@ -329,6 +329,7 @@ import statsmodels.api as sm
 rlm_param = []
 rlm_Tlb = []
 rlm_Tub = []
+pvals = []
 
 for name, series in accum_grid.items():
     X = sm.add_constant(series.index.values)
@@ -340,10 +341,22 @@ for name, series in accum_grid.items():
     # wls_r2.append(mod.rsquared)
     rlm_Tlb.append(mod.conf_int()[1,0])
     rlm_Tub.append(mod.conf_int()[1,1])
+    pvals.append(mod.pvalues[1])
 
 gdf_grid['rlm_T'] = rlm_param
 gdf_grid['rlm_lb'] = rlm_Tlb
 gdf_grid['rlm_ub'] = rlm_Tub
+gdf_grid['rlm_p'] = pvals
+
+# %% Walker field significance tests
+
+a_global = 0.05
+K = gdf_grid.shape[0]
+p_walker = 1 - (1-a_global)**(1/K)
+
+print(f"Walker field significance test (alpha={a_global}) indicates that min p-value must be lower than {p_walker:.2E}")
+print(f"Min p-value in data: {gdf_grid['rlm_p'].min():.2E}")
+print(f"A total of {100*(gdf_grid['rlm_p'] <= p_walker).sum()/ gdf_grid['rlm_p'].count():.0f}% of results meet the Walker criteria")
 
 # %% Data location map
 

### 2021-07-16 src/my_mods/stats.py
Adds option to estimate p-vals in bootstrapping (issues with very small p-values)

@@ -8,7 +8,7 @@ from scipy import signal
 import statsmodels.tsa.stattools as tsa
 from scipy.spatial.distance import pdist
 
-def trend_bs(df, nsim, df_err=pd.DataFrame()):
+def trend_bs(df, nsim, df_err=pd.DataFrame(), pval=False):
     """
     Dpc string goes here.
     """
@@ -18,7 +18,7 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
     trends_bs = pd.DataFrame(columns=df.columns)
     intercepts = pd.DataFrame(columns=df.columns)
 
-    # In no errors exist, create neutral weights
+    # If no errors exist, create neutral weights
     if df_err.empty:
         df_err = pd.DataFrame(
             np.ones(df.shape), index=df.index, 
@@ -68,7 +68,44 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
     trendCI_lb = np.nanpercentile(trends_bs, 2.5, axis=0)
     trendCI_ub = np.nanpercentile(trends_bs, 97.5, axis=0)
 
-    return trend_mu, intercept_mu, trendCI_lb, trendCI_ub
+    if pval:
+
+        tic = time.perf_counter()
+
+        # Generate null data
+        df_null = pd.DataFrame(
+            np.nan_to_num(signal.detrend(df, axis=0)), 
+            index=df.index)
+        
+        null_trends = pd.DataFrame(columns=df_null.columns)
+
+        # Generate null model estimates
+        for _ in range(nsim):
+
+            null_bs = df_null.sample(
+                len(df_null), replace=True).sort_index()
+            coeffs_null = np.polyfit(
+                null_bs.index, null_bs, 1)
+            null_trends = null_trends.append(
+                pd.Series(coeffs_null[0], 
+                index=df_null.columns), 
+                ignore_index=True)
+        
+        # Estimate p-values based on comparison to null models
+        pvals = np.zeros(len(trend_mu))
+        for i, mu in enumerate(trend_mu):
+
+            null_vals = null_trends.iloc[:,i]
+            pvals[i] = (abs(null_vals) >= abs(mu)).sum() \
+                / len(null_vals)
+
+        toc = time.perf_counter()
+        print(f"Execution time of null modeling: {toc-tic} s")
+
+        return trend_mu, intercept_mu, trendCI_lb, trendCI_ub, pvals
+
+    else:    
[…]

### 2021-10-21 docs/grsl-analysis.py
Changes and edits made for revision 2 of grsl manuscript

@@ -1725,6 +1725,144 @@ df_depths['depth_res'] = (
 
 hv.Distribution(df_depths, kdims=['depth_res'], vdims=['Year']).groupby('Year')
 
+#%%[markdown]
+
+# ## Investigations from the Ronne Ice Shelf region
+# 
+# %% Import PAIPR-generated data
+
+# Import 20121018 results
+dir1 = ROOT_DIR.joinpath('data/PAIPR-repeat/20121018/smb/')
+data_raw = paipr.import_PAIPR(dir1)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+data_2012 = paipr.format_PAIPR(
+    data_0, start_yr=2004, end_yr=2011, rm_deep=False).drop(
+    'elev', axis=1)
+a2012_ALL = data_2012.pivot(
+    index='Year', columns='trace_ID', values='accum')
+std2012_ALL = data_2012.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Import 20161024 results
+dir2 = ROOT_DIR.joinpath('data/PAIPR-repeat/20161024/smb/')
+data_raw = paipr.import_PAIPR(dir2)
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+data_2016 = paipr.format_PAIPR(
+    data_0, start_yr=2004, end_yr=2011, rm_deep=False).drop(
+    'elev', axis=1)
+a2016r_ALL = data_2016.pivot(
+    index='Year', columns='trace_ID', values='accum')
+std2016r_ALL = data_2016.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Create gdf of mean results for each trace and 
+# transform to Antarctic Polar Stereographic
+gdf_2012 = paipr.long2gdf(data_2012)
+gdf_2012.to_crs(epsg=3031, inplace=True)
+gdf_2016r = paipr.long2gdf(data_2016)
+gdf_2016r.to_crs(epsg=3031, inplace=True)
+
+# %% Determine overlap between datasets
+
+# Find nearest neighbors between 2011 and 2016 
+# (within 500 m)
+df_dist = so.nearest_neighbor(
[…]

### 2021-10-21 docs/suppl-repeat.py

@@ -1,0 +1,176 @@
+# Script for processing and analyzing additional repeat flightline data
+
+# %% Set environment
+
+# Import requisite modules
+import numpy as np
+import pandas as pd
+import geopandas as gpd
+from pathlib import Path
+import matplotlib.pyplot as plt
+import geoviews as gv
+import holoviews as hv
+from cartopy import crs as ccrs
+from shapely.geometry import Point
+hv.extension('bokeh', 'matplotlib')
+gv.extension('bokeh', 'matplotlib')
+import panel as pn
+import seaborn as sns
+import xarray as xr
+from xrspatial import hillshade
+
+# Define plotting projection to use
+ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
+
+# Set project root directory
+ROOT_DIR = Path(__file__).parents[1]
+
+# Import custom project functions
+import sys
+SRC_DIR = ROOT_DIR.joinpath('src')
+sys.path.append(str(SRC_DIR))
+from my_mods import paipr, rema, viz, stats
+from my_mods import spat_ops as so
+
+# %%
+
+def grsl_clean(
+    data_dir, start_yr=None, end_yr=None, 
+    yr_clip=True, rm_deep=True, cut_QC=2):
+
+    # Import PAIPR results
+    data_raw = paipr.import_PAIPR(data_dir)
+    data_raw.query('QC_flag < @cut_QC', inplace=True)
+    # data_0 = data_raw.query(
+    #     'Year > QC_yr').sort_values(
+    #     ['collect_time', 'Year']).reset_index(drop=True)
+    data_0 = data_raw.sort_values(
+        ['collect_time', 'Year']).reset_index(drop=True)
+    data = paipr.format_PAIPR(
+        data_0, start_yr=start_yr, end_yr=end_yr, 
+        yr_clip=yr_clip, rm_deep=rm_deep).drop(
+        'elev', axis=1)
+    accum_df = data.pivot(
+        index='Year', columns='trace_ID', values='accum')
+    std_df = data.pivot(
+        index='Year', columns='trace_ID', values='std')
+
+    # Create gdf of mean results for each trace and 
[…]

### 2021-10-21 src/my_mods/stats.py

@@ -155,7 +155,7 @@ def acf(df):
 def vario(
     points_gdf, lag_size, 
     d_metric='euclidean', vars='all', 
-    stationarize=False, scale=True):
+    standardize=False, scale=True):
     """A function to calculate the experimental variogram for values associated with a geoDataFrame of points.
 
     Args:
@@ -169,10 +169,22 @@ def vario(
         pandas.core.frame.DataFrame: The calculated semivariance values for each chosen input. Also includes the lag interval (index), the average separation distance (dist), and the number of paired points within each interval (cnt). 
     """
 
+    # Make copy to preserve original input df
+    points_gdf = points_gdf.copy()
+
     # Get column names if 'all' is selected for "vars"
     if vars == "all":
         vars = points_gdf.drop(
             columns='geometry').columns
+
+    if standardize:
+        for var in vars:
+            points_gdf[var] = (
+                (points_gdf[var] - points_gdf[var].mean()) / points_gdf[var].std() )
+
+    # Check if supplied gpd is Points or not
+    if (~points_gdf.geom_type.isin(['Point'])).sum():
+        points_gdf['geometry'] = points_gdf.centroid
 
     # Extact trace coordinates and calculate pairwise distance
     locs_arr = np.array(
@@ -189,9 +201,6 @@ def vario(
                 i_idx[m*i + j - ((i + 2)*(i + 1))//2] = i
                 j_idx[m*i + j - ((i + 2)*(i + 1))//2] = j
 
-
-    if stationarize:
-        pass
     
     # Create dfs for paired-point values
     i_vals = points_gdf[vars].iloc[i_idx].reset_index(

### 2021-10-21 docs/update-valid.py
Modularize (#9)
* Updates the environment.yml file

* Modules (#8)

* Updates custom functions to use proper importing of modules within the new 'my_mods' package

* Adds __init__ to improve import handling

@@ -9,6 +9,8 @@ import geopandas as gpd
 from pathlib import Path
 import matplotlib.pyplot as plt
 from cartopy import crs as ccrs
+import holoviews as hv
+hv.extension('bokeh', 'matplotlib')
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -20,18 +22,19 @@ ROOT_DIR = ROOT_DIR = Path(__file__).parents[1]
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_functions import *
+from my_mods import paipr, viz
+from my_mods import spat_ops as so
 
 # %% Import and format radar data
 
 # Import and format PAIPR results
 dir1 = ROOT_DIR.joinpath('data/PAIPR-outputs/20111109/')
-data_raw = import_PAIPR(dir1)
+data_raw = paipr.import_PAIPR(dir1)
 data_raw.query('QC_flag != 2', inplace=True)
 data_0 = data_raw.query(
     'Year > QC_yr').sort_values(
     ['collect_time', 'Year']).reset_index(drop=True)
-paipr_data = format_PAIPR(
+paipr_data = paipr.format_PAIPR(
     data_0, start_yr=1980, end_yr=2010).drop(
     'elev', axis=1)
 paipr_ALL = paipr_data.pivot(
@@ -41,8 +44,8 @@ std_ALL = paipr_data.pivot(
 
 # Import and format manual results
 dir_0 = ROOT_DIR.joinpath('data/smb_manual/20111109/')
-data_0 = import_PAIPR(dir_0)
-man_data = format_PAIPR(
+data_0 = paipr.import_PAIPR(dir_0)
+man_data = paipr.format_PAIPR(
     data_0, start_yr=1980, end_yr=2010).drop(
     'elev', axis=1)
 man_ALL = man_data.pivot(
@@ -52,9 +55,9 @@ manSTD_ALL = man_data.pivot(
 
 # Create gdf of mean results for each trace and 
 # transform to Antarctic Polar Stereographic
-gdf_paipr = long2gdf(paipr_data)
+gdf_paipr = paipr.long2gdf(paipr_data)
 gdf_paipr.to_crs(epsg=3031, inplace=True)
-gdf_man = long2gdf(man_data)
+gdf_man = paipr.long2gdf(man_data)
[…]

### 2021-10-21 src/my_mods/spat_ops.py

@@ -1,0 +1,300 @@
+# Module for various spatial manipulations and extractions of both raster and vector data
+
+# Import required modules
+import numpy as np
+import geopandas as gpd
+import pandas as pd
+from sklearn.neighbors import BallTree
+import xarray as xr
+
+def pts2grid(geo_df, resolution=5000):
+    import shapely
+    """
+    Description.
+    """
+    xmin, ymin, xmax, ymax = geo_df.total_bounds
+
+    x_pts = np.arange(
+        np.floor(xmin), np.ceil(xmax)+resolution, 
+        step=resolution)
+    y_pts = np.arange(
+        np.floor(ymin), np.ceil(ymax)+resolution, 
+        step=resolution)
+
+    grid_cells = []
+    for x in x_pts:
+        for y in y_pts:
+            x0 = x - resolution
+            y0 = y - resolution
+            grid_cells.append(shapely.geometry.box(x0, y0, x, y))
+    
+    geo_cells = gpd.GeoDataFrame(
+        data=grid_cells, columns=['geometry'],
+        crs=geo_df.crs)
+    # # Diagnostic plot
+    # ax = geo_cells.plot(facecolor='none', edgecolor='grey')
+    # geo_df.plot(ax=ax, column='accum', cmap='viridis', vmax=600)
+
+    geo_sjoin = gpd.sjoin(
+        geo_df, geo_cells, how='inner', op='within')
+    gdf_grid = gpd.GeoDataFrame(
+        data=geo_sjoin[['trace_ID', 'elev', 'accum', 'std', 'index_right']], 
+        geometry=geo_cells.geometry[geo_sjoin['index_right']].values, 
+        crs=geo_df.crs)
+    
+    return gdf_grid
+
+
+def trace_combine(gdf_grid, accum_ALL, std_ALL):
+    """
+    Description.
+    """
+    # Get indices of grids that are populated
+    grid_pop = np.unique(gdf_grid['index_right'])
+
+    # Preallocate return arrays
+    accum_df = pd.DataFrame(
+        columns=np.arange(len(grid_pop)), 
+        index=accum_ALL.index)
+    MoE_df = pd.DataFrame(
+        columns=np.arange(len(grid_pop)), 
+        index=accum_ALL.index)
+    yr_count = pd.DataFrame(
+        columns=np.arange(len(grid_pop)), 
+        index=accum_ALL.index)
+    grid_final = gpd.GeoDataFrame(
[…]

### 2021-10-21 docs/submission-docs/oib-repeat/revision-02/reviewer-response.md
Significance and GRSL-R2 (#12)
* Updates the environment.yml file

* Modules (#8)

* Updates custom functions to use proper importing of modules within the new 'my_mods' package

* Adds __init__ to improve import handling

* Implements walker test for field significance

* Adds option to estimate p-vals in bootstrapping (issues with very small p-values)

* Fixes issues with p-val calculations for small p-walker tests

* Adds function to perform moving block bootstrap

* Adds section to export accum data for mixed effects modeling in R

* Adds word processor docs to ignore file

* Changes and e

# Response to reviewers

>> While the authors have addressed a wide range of comments and concerns by other reviewers, the limited scope for the data used for their analysis remains problematic. While it's fair that a comprehensive analysis of all OIB data is beyond the scope of this short-format paper, the amount of data used remains troublingly limited.  It's not unreasonable to expect the authors to add lines from one or two more regions to the analysis while still fitting within the size limits of a GRSL paper.

We agree with the reviewer that the paper would be improved with additional data from other regions.
We therefore explored the Operation IceBridge dataset again more carefully to see if we could add more analyses to the current study.
This exploration included a preliminary review of 97 additional OIB flights over West Antarctica.

Although the entire Operation IceBridge dataset is considerably larger and more expansive than that used in this study, the requirements for our analysis substantially limit the available data.
The density model used in our study is trained on a specific ice core set with adequate coverage over central WAIS but more limited over other regions.
Our estimated depth-density profiles are therefore less reliable outside the region of the ice core training data.
As this study is specifically interested in the biases between repeat OIB flights separated over time, we could assume a fixed density profile and still investigate how the differences in collection time affect results.
Even with this assumption however, several additional factors severely limited the fraction of OIB data available for use in our study.

First, Snow radar data were collected only on a subset of the OIB flights.
We also require substantial repeat coverage for flights separated in time by a number of years (ideally 5 years to match the separation of the current data used in our study, although we considered candidates within the range of 4--7 years).
Repeat flights themselves are somewhat rare, and many of those that do exist were collected only a year or two apart.
We are also limited to portions of the ice sheet where accumulation rates are sufficiently low to enable (within the upper ~25 m) several years of overlapping accumulation years between the repeat flights (ideally 20 years of overlap to match the current data in our study, but a lower number of years could still provide adequate results).
Finally, the annual horizons in the echograms must be sufficiently well-delineated (e.g. the data must be of sufficient quality) to allow for layer picking with a reasonable level of certainty.
These requirements necessarily limited our analyses to a subset of the full OIB dataset.

In an effort to address the concerns of the reviewer, we performed a survey of OIB Snow radar data in West Antarctica to search for sufficient portions of additional flights meeting these requirements.
Our survey indicated that our requirements remove the vast majority of the OIB dataset from consideration, with other flights failing to meet one or more of these requirements.
Ultimately, we identified 4 additional pairs of Snow radar flights (of the original 97 surveyed candidates) that met our preliminary requirements (data collection separated by at least 4 years, overlapping flight segments of many 10's of kilometers, and in regions of sufficiently low accumulation rates to potentially yield several years of overlapping annual estimates).
These pairs were mostly concentrated in central WAIS in the same general region as our current data, but also included repeat flights on the eastern portion of the Ronne Ice Shelf.
The paired flight collection dates are as follows:

- 2009-10-29 <==> 2016-11-04
- 2010-11-20 <==> 2014-11-06
- 2009-10-18 <==> 2014-11-21
- 2012-10-18 <==> 2016-10-24

Unfortunately, many of these flights only contained annually-resolved or well-delineated data over a small portion of the full flight.
Some of these flights did not have any significant portions of sufficiently-well resolved data that overlap for both pairs of flights (as is the case for 2009-10-18/2014-11-21 and 2010-11-20/2014-11-06).
The two remaining candidates also had substantial issues.
Significant portions of the overlapping sections for 2009-10-29/2016-11-04 have sufficiently high accumulation rates that the generated overlapping time series are too short (<5 years) to include in our analysis while the estimated uncertainties involved in the remaining portions were unacceptably high.
The small section of overlap for 2012-10-18/2016-10-24 not flagged by our automated quality control routine (a segment of ~50 km) also exhibited errors (often 2-3 times the magnitude of the estimated accumulation rates) too high to warrant inclusion in our analysis.

All of these issues and errors combined lead us to not include additional analyses or data within this manuscript.
It may be possible to extract additional useful information from these in a future study/comparison  with substantial changes to our methodology.
We, therefore, elect to not include these data in this submission.
We instead have added additional language to our manuscript to better highlight the spatial limits of the data used in the study and include some justification for this more limited data set (See Section II in the revised manuscript).
We believe these edits better clarify the limitations of our analysis in this regard.

### 2021-10-21 docs/analysis.py

@@ -15,8 +15,8 @@ from cartopy import crs as ccrs
 from cartopy import feature as cf
 # from bokeh.io import output_notebook
 # output_notebook()
-# hv.extension('bokeh')
-# gv.extension('bokeh')
+hv.extension('bokeh')
+gv.extension('bokeh')
 from shapely.geometry import Polygon, Point
 import xarray as xr
 from xrspatial import hillshade
@@ -268,24 +268,64 @@ accum_core = core_ACCUM.loc[yr_start:yr_end,keep_idx]
 gdf_core1979 = gdf_cores.copy().loc[keep_idx,:]
 gdf_core1979['accum'] = accum_core.mean()
 
-# %% Calculate linear trends
+# %% Exporting long format for R mixed effects modeling
+
+# r_accum = pd.melt(
+#     accum_grid.reset_index(), 
+#     id_vars='Year', var_name='ID', value_name='accum')
+
+# loc_df = pd.DataFrame(
+#     {'Easting': gdf_grid.geometry.centroid.x, 
+#     'Northing': gdf_grid.geometry.centroid.y}, 
+#     index=gdf_grid.index)
+
+# r_df = pd.merge(
+#     left=r_accum, right=loc_df, how='left', 
+#     left_on='ID', right_index=True)
+
+# r_df.to_csv(ROOT_DIR.joinpath('data/exports/accum_export.csv'))
+
+# %% Investigate temporal autocorrelation in data
+
+auto_corr = stats.acf(accum_grid)
+auto_corr.mean(axis=1).plot()
+
+# Define moving blocks bootstrap block size
+blck_sz = 3
+
+# %% Calculate linear trends in radar
+
+# Critical p-value for walker field significance
+a_global = 0.05
+K = gdf_grid.shape[0]
+p_walker = 1 - (1-a_global)**(1/K)
+
+# Min power of 10 necessary to ensure sufficient significant digits
+n_sigfig = int(10**(np.ceil(np.log10(1/p_walker))))
 
 # Calculate trends in radar
-trends, _, lb, ub = stats.trend_bs(
-    accum_grid, 1000, df_err=MoE_grid)
+trends, _, lb, ub, pvals = stats.trend_bs(
+    accum_grid, 1000, df_err=MoE_grid, blck_sz=3, 
[…]

### 2021-10-21 src/my_mods/paipr.py

@@ -23,7 +23,9 @@ def import_PAIPR(input_dir):
         data['collect_time'])
     return data
 
-def format_PAIPR(data_df, start_yr=None, end_yr=None):
+def format_PAIPR(
+    data_df, start_yr=None, end_yr=None, 
+    yr_clip=True, rm_deep=True):
     """
 
     """
@@ -72,27 +74,21 @@ def format_PAIPR(data_df, start_yr=None, end_yr=None):
                 accum=data_df['accum_mu'], 
                 std=data_df['accum_std']).reset_index(drop=True))
 
-    # Additional subroutine to remove time series where the deepest 
-    # 3 years have overly large uncertainties (2*std > expected value)
-    # data_tmp = data_long[data_long['Year'] <= (start_yr+2)]
-    # data_log = pd.DataFrame(
-    #     {'trace_ID': data_tmp['trace_ID'], 
-    #     'ERR_log': 2*data_tmp['std'] > data_tmp['accum']})
-    # trace_tmp = data_log.groupby('trace_ID')
-    # IDs_keep = trace_tmp.filter(
-    #     lambda x: not all(x['ERR_log']))['trace_ID'].unique()
-    data_tmp = data_long.groupby(
-        'trace_ID').mean().query('2*std < accum').reset_index()
-    IDs_keep = data_tmp['trace_ID']
-    data_long = data_long[
-        data_long['trace_ID'].isin(IDs_keep)]
+    if rm_deep:
+        # Additional subroutine to remove time series where the deepest 3 years have overly large uncertainties (std > expected value)
+        data_tmp = data_long.groupby(
+            'trace_ID').mean().query('std < accum').reset_index()
+        IDs_keep = data_tmp['trace_ID']
+        data_long = data_long[
+            data_long['trace_ID'].isin(IDs_keep)]
 
-    # Remove time series with fewer than 5 years
-    data_tmp = data_long.join(
-        data_long.groupby('trace_ID')['Year'].count(), 
-        on='trace_ID', rsuffix='_count')
-    data_long = data_tmp.query('Year_count >= 5').drop(
-        'Year_count', axis=1)
+    if yr_clip:
+        # Remove time series with fewer than 5 years
+        data_tmp = data_long.join(
+            data_long.groupby('trace_ID')['Year'].count(), 
+            on='trace_ID', rsuffix='_count')
+        data_long = data_tmp.query('Year_count >= 5').drop(
+            'Year_count', axis=1)
 
[…]

### 2021-10-21 src/my_mods/stats.py

@@ -7,8 +7,24 @@ import numpy as np
 from scipy import signal
 import statsmodels.tsa.stattools as tsa
 from scipy.spatial.distance import pdist
+from scipy.stats import gaussian_kde as kde
 
-def trend_bs(df, nsim, df_err=pd.DataFrame()):
+def blck_resample(df, blck_sz):
+
+    k = df.shape[0] // blck_sz
+    start_idx = np.random.randint(0,df.shape[0]-blck_sz, k)
+    idx = []
+    for num in start_idx:
+        for el in range(blck_sz):
+            idx.append(num + el)
+    
+    df_sample = df.iloc[idx,:].sort_index()
+
+    return df_sample
+
+def trend_bs(
+    df, nsim, df_err=pd.DataFrame(), blck_sz=1,
+    pvals=False, n_pvals=None):
     """
     Dpc string goes here.
     """
@@ -18,7 +34,8 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
     trends_bs = pd.DataFrame(columns=df.columns)
     intercepts = pd.DataFrame(columns=df.columns)
 
-    # In no errors exist, create neutral weights
+    # If no errors exist, create neutral weights
+
     if df_err.empty:
         df_err = pd.DataFrame(
             np.ones(df.shape), index=df.index, 
@@ -27,8 +44,9 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
     # Peform bootstrapping
     for _ in range(nsim):
         # Randomly resample data
-        data_bs = df.sample(
-            len(df), replace=True).sort_index()
+        # data_bs = df.sample(
+        #     len(df), replace=True).sort_index()
+        data_bs = blck_resample(df, blck_sz=blck_sz)
         
         # Generate mean weights
         weights_bs = (
@@ -68,7 +86,53 @@ def trend_bs(df, nsim, df_err=pd.DataFrame()):
     trendCI_lb = np.nanpercentile(trends_bs, 2.5, axis=0)
     trendCI_ub = np.nanpercentile(trends_bs, 97.5, axis=0)
 
-    return trend_mu, intercept_mu, trendCI_lb, trendCI_ub
+    if pvals:
+
+        tic = time.perf_counter()
+
+        # Generate null data
+        df_null = pd.DataFrame(
+            signal.detrend(df.fillna(df.mean()), axis=0), 
+            index=df.index)
+        
+        null_trends = pd.DataFrame(columns=df_null.columns)
+
[…]

### 2022-04-07 docs/grsl-analysis.py
Grsl submit (#13)
* Add old Word docs to ignore

* Updates scripts for final submission with GRSL manuscript

* Updates docs for final press version of GRSL article

@@ -14,7 +14,6 @@ from cartopy import crs as ccrs
 from shapely.geometry import Point
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
-import panel as pn
 import seaborn as sns
 import xarray as xr
 from xrspatial import hillshade
@@ -126,17 +125,6 @@ chunk_centers = gpd.GeoDataFrame({
             -4.639E5, -4.639E5]), 
     crs="EPSG:3031")
 
-# # Assign flight chunk label based on location
-# chunk_centers = gpd.GeoDataFrame({
-#     'Site': ['PIG', 'A', 'B', 'C', 'D', 'E', 'F'],
-#     'Name': ['PIG', 'A', 'B', 'C', 'D', 'E', 'F']},
-#     geometry=gpd.points_from_xy(
-#         [-1.297E6, -1.177E6, -1.115E6, -1.024E6, 
-#             -1.093E6, -1.159E6, -1.263E6], 
-#         [-1.409E5, -2.898E5, -3.681E5, -4.639E5, 
-#             -4.639E5, -4.640E5, -4.668E5]), 
-#     crs="EPSG:3031")
-
 #%% Data map components
 
 # Accumulation plots for both flights
@@ -243,8 +231,10 @@ hill_plt = hv.Image(xr_HS.values, bounds=tpl_bnds).opts(
 
 # %% Get ice velocities of trace locations
 
-xr_vice = xr.open_dataset(ROOT_DIR.joinpath(
-    'data/ice-velocities/antarctica_ice_velocity_450m_v2.nc'))
+xr_vice = xr.open_dataset(
+    ROOT_DIR.joinpath(
+        'data/ice-velocities/', 
+        'antarctica_ice_velocity_450m_v2.nc'))
 xr_clip = xr_vice.sel(
     x=slice(
         gdf_PAIPR.total_bounds[0], 
@@ -260,11 +250,13 @@ vice_pts = gpd.GeoDataFrame(
 vice_pts['Vxy'] = np.sqrt(
     vice_pts['VX']**2 + vice_pts['VY']**2)
 
-gv.Points(vice_pts, crs=ANT_proj, 
+vice_plt = gv.Points(vice_pts, crs=ANT_proj, 
 vdims=['Vxy', 'VX', 'VY']).opts(
     projection=ANT_proj, color='Vxy', cmap='viridis', 
[…]

### 2022-04-07 docs/suppl-repeat.py

@@ -7,17 +7,11 @@ import numpy as np
 import pandas as pd
 import geopandas as gpd
 from pathlib import Path
-import matplotlib.pyplot as plt
 import geoviews as gv
 import holoviews as hv
 from cartopy import crs as ccrs
-from shapely.geometry import Point
 hv.extension('bokeh', 'matplotlib')
 gv.extension('bokeh', 'matplotlib')
-import panel as pn
-import seaborn as sns
-import xarray as xr
-from xrspatial import hillshade
 
 # Define plotting projection to use
 ANT_proj = ccrs.SouthPolarStereo(true_scale_latitude=-71)
@@ -29,7 +23,7 @@ ROOT_DIR = Path(__file__).parents[1]
 import sys
 SRC_DIR = ROOT_DIR.joinpath('src')
 sys.path.append(str(SRC_DIR))
-from my_mods import paipr, rema, viz, stats
+from my_mods import paipr
 from my_mods import spat_ops as so
 
 # %%
@@ -41,9 +35,7 @@ def grsl_clean(
     # Import PAIPR results
     data_raw = paipr.import_PAIPR(data_dir)
     data_raw.query('QC_flag < @cut_QC', inplace=True)
-    # data_0 = data_raw.query(
-    #     'Year > QC_yr').sort_values(
-    #     ['collect_time', 'Year']).reset_index(drop=True)
+
     data_0 = data_raw.sort_values(
         ['collect_time', 'Year']).reset_index(drop=True)
     data = paipr.format_PAIPR(
@@ -73,7 +65,7 @@ aPIG2016_ALL, stdPIG2016_ALL, gdf_PIG2016 = grsl_clean(
         'data/PAIPR-repeat/20161104/smb'), 
     start_yr=2004, end_yr=2008, yr_clip=False, rm_deep=True)
 
-# Find nearest neighbors between 2011 and 2016 
+# Find nearest neighbors between 2009 and 2016 
 # (within 500 m)
 df_dist = so.nearest_neighbor(
     gdf_PIG2009, gdf_PIG2016, return_dist=True)
@@ -122,7 +114,6 @@ plt_res = gv.Points(
         projection=ANT_proj, color='accum_res', size=12,
         bgcolor='silver', 
         colorbar=True, 
-        # cmap='seismic_r', 
         cmap='BrBG',
         symmetric=True, tools=['hover'], 
         width=750, height=750)

### 2022-04-07 src/accum_analysis.R
Clears old R files in preparation for new imports

@@ -1,367 +1,0 @@
-# Script to generate data for use in notebooks
-
-# Attach required packages
-library(tidyverse)
-library(here)
-library(sf)
-library(tcltk)
-library(readxl)
-library(broom)
-library(rgdal)
-library(raster)
-library(tictoc)
-
-tic("Total runtime")
-
-##----
-## Import and clean data
-
-# Function to import all .csv files of long format PAIPR results from a given directory, and
-# concatenate the results into a single tibble
-import_PAIPR = function(input_dir) {
-  files = list.files(input_dir, pattern = "*.csv", full.names = TRUE)
-  data = tibble()
-  for (f in files) {
-    data_f = read_csv(f)
-    data = rbind(data, data_f)
-  }
-  return(data)
-}
-
-# Import and format PAIPR-generated SMB results
-data = import_PAIPR( here("data/gamma_20111109"))
-data$Year = as.integer(data$Year)
-
-# Generate descriptive statistics based on imported gamma-fitted parameters
-alpha = data$gamma_shape
-alpha[which(alpha<1)] = 1
-beta = 1/data$gamma_scale
-mode.accum = (alpha-1)/beta
-var.accum = alpha/beta^2
-
-# Make df of relevant data in long format
-accum.long = data %>% dplyr::select(Lat, Lon, elev, Year) %>% 
-  mutate(accum = mode.accum, sd = sqrt(var.accum))
-
-# Find duplicate annual accum records and average into single record
-accum.long = accum.long %>% group_by(Lat, Lon, elev, Year) %>%
-  summarise(accum = mean(accum, na.rm = TRUE), sd = mean(sd, na.rm = TRUE)) %>% ungroup()
-
-# This removes years prior to a cutoff, and clips the results 
-# to only time series that extend to cutoff
-yr.first = 1985
-yr.last = 2007
-accum.long = accum.long %>% filter(Year>=yr.first) %>% filter(Year <= yr.last)
-data.cutoff = accum.long %>% group_by(Lat, Lon) %>% summarise(yr.start = min(Year))
-
-# Nest year/accum data into list-columns for each location  
-# (insert # prior to "filter" pipe to avoid cutoff)
-accum.loc = nest(accum.long, data = c(Year, accum, sd)) %>% filter(data.cutoff$yr.start <= yr.first)
-rm(data, data.cutoff, alpha, beta, mode.accum, var.accum)
-
-
-
[…]

### 2022-04-07 notebooks/draft-modeling.Rmd
Merges INLA work from r-WAIS into this repo

---
title: "WAIS Summary"
output: html_notebook
---

## Model setup

Here's a review of the statistical underpinnings of the model as I see them.
Feel free to critique/correct any misconceptions or errors you see here.
Anywhere throughout the document that I have specific questions, I include them as **bold text**.

The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension.
I model accumulation as a gamma distribution to ensure a zero-bounded function with increased variance at higher expected values.
$$\mathbf{Y} \sim Gamma(a,b)$$
Rather than using the shape and rate parameters $a$ and $b$, we can express this gamma distribution in terms of the expected y value $E(y_i) = a_i/b_i = \mu_i$ and its associated variance (as expressed using the precision parameter $\phi$) $Var(y_i) = a_i/b_i^2 = \mu_i^2/\phi$.
This $\phi$ is the same precision parameter as output from `inla` (there is also a scale parameter $s$ that has a default of 1 where $Var(y_i) = 1/\tau$ and $\tau = (s\phi)/\mu_i^2$).
With rearranging and substituting, we have $a_i = \phi$ and $b_i = \phi/\mu_i$ and we can therefore express the distribution in accumulation observations in terms of $\phi$ and $\mu_i$.
$$y_i \sim Gamma(\mu_i, \phi)$$

For each observation $y_i \in \mathbf{Y}$, we can construct a linear predictor $\eta_i$ for the expectation value $\mu_i$.
As we assume a gamma-distribution for accumulation, we associate $\eta_i$ and $\mu_i$ via the log link function.
$$\log(\mu_i) = \eta_i$$ 
We can further model $\eta_i$ as a generalized linear model consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case we are modeling a spatial random effect $u_i$ and a temporal random effect $\omega_i$).
$$
\eta_{i} = \alpha +  \sum^M_{m=1} \beta_m x_{mi} + u_i + \omega_{i}
$$

These latent random effects $u_i$ and $\omega_i$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
We can model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of some hyperparameters $\psi$ and an associated prior distribution $p(\psi)$.
This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{ij} = Cov(\theta_i, \theta_j)$.

$$
\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
$$

If we assume adherence to Markovian properties, our precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of our process, i.e. $Q_{ij} \neq 0 \iff j \in \{i,N(i)\}$.
In this assumption, we specify a Gaussian-Markov Random Field for both our spatial and temporal structures, permitting vast improvements in computational efficiency.
We also model the spatial and temporal random effects independently, which again greatly improves computation time, but will fail to capture any co-dependencies between these effects.

The spatial GMRF $u_i$ is derived from a stochastic partial differential equation...*still working on the wording/thought process here...*

We model the temporal random effect $\omega_i$ as a first-order autoregressive model^[I will eventually want to update this to an AR(2) process due to the known temporal autocorrelation in our data] (representing how the spatial field evolves in time) where
$$\omega_t = \rho \omega_{t-1}$$

Our objective, therefore, is to find the joint posterior distribution of the various parameters of interest, as this will allow me to perform inference and make predictions at unobserved locations.
$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s\mid \mathbf{Y})$$

## Imports

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(cetcolor)
library(sf)
library(terra)
library(INLA)
library(INLAutils)
library(spdep)
```

## Data loading and pre-processing

```{r data-load, message=FALSE}
data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1) %>% 
  mutate(Year = as.integer(Year))
gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()

# Remove points in far northern extent
accum_bnds = st_coordinates(gdf_traces)
gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))

# Small-scale testing bounds
gdf_traces = gdf_traces %>% st_crop(xmin=-1.4e6, xmax=-1.25e6,
                                    ymin=min(accum_bnds[,2]), ymax=-3e5)

# Remove unrealistically high elevations and overly large accum sites
trace_drop = gdf_traces %>% filter(elev > 3000 | accum > 1000)

# Add East/North coordinates and drop filtered sites
pts_tmp = st_coordinates(gdf_traces)
coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
gdf_traces = gdf_traces %>% filter(!trace_ID %in% trace_drop$trace_ID)
```

We need to import the other covariate data for the region.
This consists of topographic features from REMA and climatic variables from RACMO.
```{r}
REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
crs(dem) = st_crs(gdf_traces)$wkt

slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
crs(slp) = crs(dem)

aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
crs(aspct) = crs(dem)

topo.stk = c(dem, slp, aspct)

topo.pts = extract(topo.stk, vect(gdf_traces$geometry)) %>% as_tibble() %>% 
  rename(dem=REMA_200m_dem) %>% 
  mutate(trace.ID=gdf_traces$trace_ID, aspect=sin(pi/180*aspect))
topo.pts = topo.pts %>% select(-ID)

RACMO.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data", 
                      "/RACMO/2.3p2_yearly_ANT27_1979_2016/interim")
files = list.files(path = RACMO.dir, full.names = TRUE, 
                   pattern = "\\.tif$")

r.smb = rast(files[1])
names(r.smb) = 1979:2016
smb.pts = extract(r.smb, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% mutate(trace.ID=gdf_traces$trace_ID) %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'SMB') %>% 
  mutate(Year = as.integer(Year))

r.u10m = rast(files[2])
names(r.u10m) = 1979:2016
u10m.pts = extract(r.u10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'u10m') %>% 
  mutate(Year = as.integer(Year))

r.v10m = rast(files[3])
names(r.v10m) = 1979:2016
v10m.pts = extract(r.v10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% pivot_longer(cols = num_range(prefix='', range=1979:2016), 
                               names_to = 'Year', values_to = 'v10m') %>% 
  mutate(Year = as.integer(Year))

racmo.pts = smb.pts %>% 
  mutate(u10m=u10m.pts$u10m, v10m=v10m.pts$v10m) %>% 
  select(-ID)

data = data %>% 
  left_join(racmo.pts, by = c("trace_ID" = "trace.ID", "Year")) %>% 
  left_join(topo.pts, by = c("trace_ID" = "trace.ID"))

# data = data %>% mutate(S_x = abs(slope)*cos(asin(aspect)), 
#                        S_y = abs(slope)*aspect) %>% 
#   mutate(MSWD = S_x*u10m + S_y*v10m)
```

While I am testing/developing things, I'm only using a small subset of the data (around 1/2 a percent).
This still gives a few thousand observations to work with though.
Also for initial testing of everything, I am only including a single covariate (elevation).
Once I have a fully fleshed-out workflow though, I intend to add a few more (slope, wind vectors, etc.).
I center this covariate about its mean and also normalize the discrete time dimension, setting the min year as Year 1.

```{r}
set.seed(777)
gdf.tmp = gdf_traces %>% sample_frac(0.01)
data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year < 1999) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m)
# dat = data %>% select(trace_ID, East, North, Year, accum, 
#                       std, elev, dem, slope, aspect, u10m, v10m)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

## INLA  modeling

I create the mesh used for defining the neighborhoods (this is important for generating the sparse precision matrix $Q$, enabling the GMRF approximation).
```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 75000), cutoff = 5000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")
```

I then define the SPDE object and create the identifier matrix $A$.
**Question: Am I defining these priors correctly? I'm using the PC priors, which I think are defined in the untransformed spatial scale if I followed things right.**
```{r}
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
  prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05
# iset = inla.spde.make.index('i', n.spde = spde$n.spde,
#   n.group = max(dat$Year))

# Create projector matrix for SPDE mesh
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))

# Assign index vectors for SPDE model
spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)

# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m), 
                 list(time = dat.train$Year), 
                 spat.idx),
  tag = 'dat') 
```

I also define a PC-prior for the autoregressive temporal autocorrelation.
In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data.
**Questions: I want to make sure I am defining the priors properly, so let me know what could be improved here. Also, I am currently modeling this as a separable spacetime model with independent random effects for space and time. I largely followed the first example in Section 8.6.1 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models). It also shows an example of doing this using a grouping method which I think might incorporate some interactions between space and time autocorrelations, but the fully indepedent separable model is MUCH faster.**
```{r}
## Define other priors (spatial field priors defined earlier)

# Prior on autoregressive rho parameter
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Prior on fixed effects intercept (not currently implemented)
prior.fixed = list()

# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Model formula
formula = y ~ -1 + 
  Intercept + dem + slope + aspect + #fixed effects
  slope*aspect + #fixed effects interactions
  f(time, model = 'ar1', hyper = time.spec) + #temporal random effect
  f(spat.field, model = spde) #spatial random effect
```

```{r}
# Model fitting
system.time(
  mod.sep_ind <- inla(formula, 
                  data = inla.stack.data(dat.stack), 
                  family = 'Gamma',
                  quantiles = NULL, 
                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
                  # control.fixed = prior.fixed, 
                  control.family = list(hyper = list(prec = prec.prior)), 
                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE))
)
```

### Form 2

Rather than treating space and time wholly independently, we can also model the spatial component at each time step separately.
This still does not include interactions between space and time effects, but does allow the spatial dependence to vary over time.
The following implementation follows examples shown in  Section 8.6.2 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models) amd Section 7.1 of [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/ch-spacetime.html#discrete-time-domain).

```{r}
set.seed(777)
gdf.tmp = gdf_traces %>% sample_frac(0.005)
data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
  filter(Year >= 1989) %>% filter(Year < 1999) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m)
# dat = data %>% select(trace_ID, East, North, Year, accum, 
#                       std, elev, dem, slope, aspect, u10m, v10m)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 100000), cutoff = 7500)

spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
  prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05

# Create projector matrix for SPDE mesh
A.spat = inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat.train$East, dat.train$North),
                          group = dat.train$Year)
A.year = inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat.train$East, dat.train$North))

# Assign index vectors for SPDE model
spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde,
                                n.group = length(unique(dat.train$Year)))

# Assign index vectors for SPDE model for year covariate
year.idx = inla.spde.make.index(name = "year.diff", n.spde = spde$n.spde)

# Make data stack
# dat.stack <- inla.stack(
#   data = list(y = dat.train$accum),
#   A = list(1, 1, 1, A.spat),
#   effects = list(list(Intercept = rep(1, nrow(dat.train))),
#                  tibble(elev = dat.train$elev, dem=dat.train$dem,
#                         slope=dat.train$slope, aspect=dat.train$aspect,
#                         u10m=dat.train$u10m, v10m=dat.train$v10m),
#                  list(time = dat.train$Year),
#                  spat.idx),
#   tag = 'dat')
dat.stack = inla.stack(
  data = list(y = dat.train$accum),
  A = list(1, 1, A.spat, A.year),
  effects = list(list(Intercept = rep(1, nrow(dat.train))),
                 tibble(elev = dat.train$elev, dem=dat.train$dem,
                        slope=dat.train$slope, aspect=dat.train$aspect,
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        year=dat.train$Year),
                 spat.idx, 
                 year.idx), 
  tag = 'dat')

# Prior on autoregressive rho parameter
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Prior on fixed effects intercept (not currently implemented)
prior.fixed = list()

# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Model formula
formula = y ~ -1 +
  Intercept + dem + slope + aspect + #fixed effects
  # slope*aspect + #fixed effects interactions
  f(spat.field, model = spde, group = spat.field.group,
    control.group = list(model = 'ar1', hyper = time.spec)) + 
  f(year.diff, year, model = spde, constr = FALSE)

# Model fitting
system.time(
  mod.sep_group <- inla(formula,
                  data = inla.stack.data(dat.stack),
                  family = 'Gamma',
                  quantiles = NULL,
                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1),
                  # control.fixed = prior.fixed,
                  control.family = list(hyper = list(prec = prec.prior)),
                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE))
)
```

On a (very) small testing set (use commented section in `data-load` chunk) of ~100 locations, the grouped separable model does perform substantially better based on WAIC score, but at the cost of around 50x the computation time (this difference may be even larger on a larger dataset).
Fitted fixed effects are of a similar magnitude between the two models.
This model produces greater precision in the gamma observations and roughly similar hyperparameters for the spatial field (this model has a higher range but lower range variance compared to the independent separable model).
I'll need think more about what the GroupRho parameter for the spatial field means, and how it compares to the independent separable model.

### Direct temporal trend analysis

This rendition of the model follows the areal example for disease mapping found in Section 3.3 of [A tutorial in spatial and spatio-temporal models with R-INLA](https://discovery.ucl.ac.uk/id/eprint/1415919/1/Baio_BlaCamBaiRue.pdf).
The idea here is to directly model the temporal trend as a global fixed effect, but then model the deviations from the global value as a spatially-correlated random effect.
The linear predictor therefore looks like:
$$
\eta_{it} = \alpha + \sum^M_{m=1} \beta_m x_{mi} + (\beta+\delta_i) \times time \ + u_i + \omega_i
$$

This model therefore accounts for fixed effects, a global fixed effect from time, a spatially-varying deviance from the global time effect, a spatial random effect, and a temporal random effect.

Additionally thoughts on doing this are found in this [R-inla discussion post](https://groups.google.com/g/r-inla-discussion-group/c/X_g4NuqqHfY).
The [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/index.html) also discusses how to implement spatially-varying model coefficients in INLA.

```{r}

mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 100000), cutoff = 1000)

spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
  prior.sigma = c(400, 0.05))

# year.spde = inla.spde2.matern(mesh = mesh)
year.spde = spde

# Create projector matrix for SPDE mesh
A.spat = inla.spde.make.A(mesh = mesh,
                          loc = cbind(dat.train$East, dat.train$North))

# Assign index vectors for SPDE model
spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)

# Assign index vectors for SPDE model for year covariate
year.idx = inla.spde.make.index(name = "year.diff", n.spde = year.spde$n.spde)

# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum),
  A = list(1, 1, A.spat, A.spat, 1),
  effects = list(list(Intercept = rep(1, nrow(dat.train))),
                 tibble(elev = dat.train$elev, dem=dat.train$dem,
                        slope=dat.train$slope, aspect=dat.train$aspect,
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        year=dat.train$Year),
                 year.idx, 
                 spat.idx, 
                 list(time = dat.train$Year)),
  tag = 'dat')

# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Prior on autoregressive rho parameter
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Model formula
# formula = y ~ -1 + 
#   Intercept + dem + slope + aspect + #fixed effects
#   slope*aspect + #fixed effects interactions
#   f(year.diff, year, model = "iid") + year + 
#   f(spat.field, model = spde) + #spatial random effect
#   f(time, model = 'ar1', hyper = time.spec)#temporal random effect

formula = y ~ -1 + Intercept + 
  dem + slope + aspect + #fixed effects
  # f(time, model = 'ar1', hyper = time.spec) + #AR temporal random effect
  # f(year.diff, year, model = spde, constr = FALSE) + #spatially varying temporal trend
  year + f(year.diff, year, model = spde) + #spatially varying temporal trend
  f(spat.field, model = spde) #spatial random effect

formula = y ~ -1 + 
  f(spat.field, model = spde) + 
  year + f(year.diff, year, model = spde)

# Model fitting
system.time(
  mod.trends <- inla(formula,
                  data = inla.stack.data(dat.stack),
                  family = 'Gamma',
                  quantiles = NULL,
                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), 
                                           link = 1),
                  # control.fixed = prior.fixed,
                  control.family = list(hyper = list(prec = prec.prior)),
                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE,
                                         config=TRUE))
)
```

```{r}

fitted.idx = inla.stack.index(dat.stack, 'dat')$data

test.fitted = tibble(mu = mod.trends$summary.fitted.values$mean[fitted.idx]) %>% 
  bind_cols(dat.train %>% select(Year, East, North)) %>% 
  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
test.rand = tibble(t.trend = exp(mod.trends$summary.random$year.diff$mean)-1, 
                   spat.mu = exp(mod.trends$summary.random$spat.field$mean), 
                   East = mesh$loc[,1], North = mesh$loc[,2]) %>% 
  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
ggplot(test.rand) + geom_sf(aes(color=t.trend)) + scale_color_viridis_c()
```

## Model interpretation

```{r}
mod.train = mod.sep_ind
```

Below is a summary of the model results.
```{r}
summary(mod.train)
```

Here's a rundown of how I understand these results.
Let me know if I'm missing something or there are otherwise gaps in my thought process.

The fixed effects are in a log-transformed scale.
This means that the intercept (which should represent the average accumulation when other fixed and random effects are at their average values) is ~396 mm/a when in the original untransformed scale.
Elevation has no predictive power beyond what is already explained spatially, but other covariates would need to be similarly transformed if I wanted them on the original scale.

The DIC is pretty meaningless without other models to compare to, but lower is better when I do start comparing to other models.
**Question: What utility is the marginal log-likelihood value?**

The precision parameter for the Gamma observations is the $\phi$ parameter from earlier. For a given expected accumulation $\mu_i=e^{\eta_i}$ therefore, the variance should be $Var_i = \frac{\mu_i^2}{\phi}$.
**I don't think this parameter requires other transformations, but I am not sure.**

**I'm not sure what to do with the precision for time parameter. I mean, this obviously represents some uncertainty associated with the temporal dimension, but I don't know what it is or how to use it. Is this error not already incorporated in the precision parameter for the Gamma observations (which if I understand correctly should represent the total variance in the raw untransformed accumulation scale)? If not, how do I incorporate it?**

Rho for time is the estimated temporal autocorrelation in the data, showing I high degree of correlation one year to the next.

The range for the spatial field shows that data are correlated to a pretty far distance of >100 km.
**I'm not sure what the st dev for the spatial field is though. I would think that this would represent how strongly the data vary spatially when accounting for other covariates, but this seems really small in that case. Even the measurement uncertainty on accumulation is around 4 orders of magnitude greater than this, so even if this is on the log scale it doesn't explain it. Clearly I'm missing/misunderstanding something here.**

## Model validation

I compare model results to observations not included in model training.
I therefore update the model components to generate predictions for these data points as well.
**Question: Does this set up properly account for the time dimension? I've updated the projector matrix to use the locations of the test data. Do I need to do something similar to include the time? I'm guessing no, as I include it directly as an effect in the `inla.stack` call.**
```{r}
# Create projector matrix for test data
A.test = inla.spde.make.A(mesh = mesh, loc = cbind(dat.test$East, dat.test$North))

# Create test data stack and join to data stack
test.stack <- inla.stack(
  data = list(y = NA), 
  A = list(1, 1, 1, A.test),
  effects = list(list(Intercept = rep(1, nrow(dat.test))), 
                 tibble(elev = dat.test$elev, dem=dat.test$dem, 
                        slope=dat.test$slope, aspect=dat.test$aspect, 
                        u10m=dat.test$u10m, v10m=dat.test$v10m),
                 list(time = dat.test$Year), 
                 spat.idx),
  tag = 'test')
val.stack <- inla.stack(dat.stack, test.stack)

# Model fitting to test data
system.time(
  mod.test <- inla(formula, 
                  data = inla.stack.data(val.stack), 
                  family = 'Gamma',
                  quantiles = NULL, 
                  control.predictor = list(compute = TRUE, A = inla.stack.A(val.stack), link = 1), 
                  control.family = list(hyper = list(prec = prec.prior)), 
                  control.inla = list(strategy = 'adaptive'), 
                  control.mode = list(theta = results$mode$theta, restart = FALSE))
)
```

Below is a function that, given an input of mean estimates, sd on mean estimates, precision parameter for Gamma observations, and sd on precision parameter, returns $n$ samples from the posterior predictive distribution for the given inputs.
```{r}
pp.dist = function(data, prec, prec.sd, n=5000, return_sample=FALSE) {
  if (return_sample==TRUE) {
    ppd = matrix(data=NA, nrow=nrow(data), ncol=n)
  } else {
    ppd = tibble(mu=rep(0,nrow(data)), map=rep(0,nrow(data)), 
                 CI.low=rep(0,nrow(data)), CI.high=rep(0,nrow(data)))
  }

  for (i in 1:nrow(data)) {
    dat.i = data[i,]
    mu = rnorm(n, mean=dat.i$mu, sd=dat.i$mu.sd)
    a = rnorm(n, mean=prec, sd=prec.sd)
    b = a/mu
    dist = rgamma(n=n, shape = a, rate = b)

    if (return_sample==TRUE) {
      ppd[i,] = dist
    } else {
      dist.mu = mean(dist)
      d = density(dist)
      dist.map = d$x[which.max(d$y)]
      dist.hdi = HDInterval::hdi(dist, credMass=0.89)
      ppd[i,] = tibble_row(mu=dist.mu, map=dist.map, 
                           CI.low=as.numeric(dist.hdi[1]), 
                           CI.high=as.numeric(dist.hdi[2]))
    }
  }
  return(ppd)
}
```

This is an example of generating posterior predictive distributions for a couple of data points.
**Am I doing this right?**
```{r message=FALSE}
# Get inla grid indices where test data resides
test.idx = inla.stack.index(val.stack, 'test')$data

# Create df of model predictions for test data
test.fitted = tibble(mu = mod.test$summary.fitted.values$mean[test.idx], 
                     mu.sd = mod.test$summary.fitted.values$sd[test.idx], 
                     accum = dat.test$accum, accum.sd = dat.test$std)

# Create matrix of model posterior  samples for a few test data
system.time(
  post.pred_samp <- pp.dist(data=head(test.fitted), prec=mod.test$summary.hyperpar$mean[1], 
                       prec.sd=mod.test$summary.hyperpar$sd[1], return_sample = TRUE)
)

pp.samples = as_tibble(t(post.pred_samp), .name_repair = 'universal') %>%
  rename(R1=...1, R2=...2, R3=...3, R4=...4, R5=...5, R6=...6) %>% 
  mutate(Sample.ID = 1:ncol(post.pred_samp)) %>% 
  pivot_longer(cols = starts_with("R"), names_to = "R.name", values_to = "accum.rvs")

accum.obs = tibble()
for (i in 1:length(unique(pp.samples$R.name))) {
  r.name = paste("R", i, sep = '')
  accum.i = rnorm(max(pp.samples$Sample.ID), 
                  mean=test.fitted$accum[i], 
                  sd=test.fitted$accum.sd[i])
  accum.obs = accum.obs %>% 
    bind_rows(tibble(Sample.ID = 1:length(accum.i), R.name = r.name, 
                     accum.obs = accum.i))
}

pp.samples = pp.samples %>% left_join(accum.obs, by = c("R.name", "Sample.ID")) %>% 
  mutate(R.name = as.factor(R.name))

# ggplot(pp.samples) + 
#   geom_density(aes(x=accum.rvs, color=ROW, fill=ROW), alpha=0.1) + 
#   xlab('Posterior predictive accumulation')

vars = c("INLA"="red", "PAIPR"="blue")
ggplot(pp.samples) + 
  geom_density(aes(x=accum.rvs, color='INLA', fill='INLA'), alpha=0.1) + 
  geom_density(aes(x=accum.obs, color='PAIPR', fill='PAIPR'), alpha=0.1) + 
  scale_colour_manual(name="Data sources:", values=vars) + 
  scale_fill_manual(name="Data sources:", values=vars) + 
  xlab("Accumulation") + 
  facet_wrap(~R.name)
```

And these return summary statistics for the ppc distributions for all the test data.
```{r}
# Generate summaries of posterior predictive checks for test data
system.time(
  post.pred_test <- pp.dist(data=test.fitted, prec=mod.test$summary.hyperpar$mean[1], 
                       prec.sd=mod.test$summary.hyperpar$sd[1])
)
```

Below are some plots visually summarizing the performance of the test data
```{r}
res.test = dat.test %>% 
  mutate(mu = test.fitted$mu, 
         mu.pp=post.pred_test$mu, 
         map.pp=post.pred_test$map, 
         CI.low=post.pred_test$CI.low, 
         CI.hi=post.pred_test$CI.high) %>% 
  mutate(y.res = mu-accum, pp.res=map.pp-accum)

ggplot(data=res.test, aes(accum, map.pp)) + 
  geom_point() + 
  geom_abline(slope=1, intercept=0, color='red') + 
  # geom_smooth(method = 'lm') + 
  # lims(x=c(200,600), y=c(200,600)) + 
  xlab("Accum (obs)") + ylab("Accum (PP MAP)")

ggplot(res.test, aes(mu, map.pp)) + geom_point() + 
  geom_abline(slope=1, intercept=0, color='red') + 
  # lims(x=c(200,600), y=c(200,600)) + 
  xlab("Accum (expected)") + ylab("Accum (post. predictive MAP)")

d.map = density(res.test$pp.res)
map.mode = d.map$x[which.max(d.map$y)]
d.mu = density(res.test$y.res)
mu.mode = d.mu$x[which.max(d.mu$y)]
ggplot(res.test) + geom_density(aes(pp.res)) + 
  geom_vline(xintercept = map.mode, linetype=2, color='red') + 
  annotate(geom="text", x=map.mode-67, y=max(d.map$y), 
           label=paste("MAP =", format(map.mode, digits=3)), 
           color="red") + 
  geom_vline(xintercept = mu.mode, linetype=2, color='blue') + 
  annotate(geom="text", x=mu.mode+33, y=max(d.map$y), 
           label=paste("Mu =", format(mu.mode, digits=3)), 
           color="blue") + 
  xlab("Mean post. pred. accum residuals") + ylab('density')
```

## Random effects

Here are plots showing the spatial random field effect and the error on the estimates.
```{r}
spat.tbl = tibble(E = mesh$loc[,1], N = mesh$loc[,2], 
                  spat.mu = mod.test$summary.random$spat.field$mean, 
                   spat.sd = mod.test$summary.random$spat.field$sd) %>% 
  st_as_sf(coords=c("E", "N"), crs=st_crs(dat.train))

ggplot(spat.tbl) + geom_sf(aes(color=spat.mu)) + 
  scale_color_gradientn(colours = cet_pal(5, name = "d8"))
ggplot(spat.tbl) + geom_sf(aes(color=spat.sd)) + 
  scale_color_viridis_c(option='plasma')
```
**Again, the magnitude of the spatial effect seems pretty small. Am I missing something here?**

Below is a plot of the temporal random effect and its estimated uncertainty.
```{r}
time.tbl = tibble(Year=yr.min+(1:nrow(mod.test$summary.random$time)), 
                  time.mu = mod.test$summary.random$time$mean, 
                   time.sd = mod.test$summary.random$time$sd)

ggplot(time.tbl, aes(x=Year, y=time.mu)) + 
  geom_line() + 
  geom_ribbon(aes(ymin=time.mu-time.sd,ymax=time.mu+time.sd), alpha=0.3) + 
  ylab("Magnitude of temporal random effect")
```
**The magnitude of this temporal random effect also seems quite small. Do these need to be transformed somehow to some other scale? But even when using a log link function here, the end result is still very small compared to both the magnitude of the data and the estimated precision of the observations).**

**Also, how do I interpret/work with investigating the spatial and temporal random variables at the same time?**

## New predictions

For new predictions, we first must create a grid of prediction points and project that grid onto a new projector matrix.
```{r}
pred.bbox = st_bbox(gdf_traces)

stepsize = 5000
buffer = 10*stepsize
E.range = round(c(pred.bbox[1]-buffer, pred.bbox[3]+buffer))
N.range = round(c(pred.bbox[2]-buffer, pred.bbox[4]+buffer))

nxy <- round(c(diff(E.range), diff(N.range)) / stepsize) # Calculate the number of cells in the x and y ranges

# Project the spatial field on the mesh vertices using the inla.mesh.projector() function
projgrid <- inla.mesh.projector(mesh,
                                xlim = E.range,
                                ylim = N.range,
                                dims = nxy)

```

```{r}
# Find index of grid points within the country border
grid.pts = tibble(x = projgrid$lattice$loc[,1],
                  y = projgrid$lattice$loc[,2],
                  grid.ID = 1:nrow(projgrid$lattice$loc))
grid.sf = st_as_sf(grid.pts, coords=c('x','y'))
st_crs(grid.sf) = st_crs(gdf_traces)

# # Extract topo covariates from rasters for each grid point
# topo.pts = extract(topo.stk, vect(grid.sf$geometry))
# grid.sf = grid.sf %>% mutate(dem=topo.pts$REMA_200m_dem, 
#                                slope=topo.pts$slope, 
#                                aspect=topo.pts$aspect)
# 
# # Extract clim covariates from rasters for each grid point
# racmo.pts = extract(racmo.stk, vect(grid.sf$geometry))
# grid.sf = grid.sf %>% mutate(smb.racmo=racmo.pts$smb, 
#                                u10m=racmo.pts$u10m, 
#                                v10m=racmo.pts$v10m)
# grid.sf = grid.sf %>% mutate(S_x = abs(slope)*cos(aspect*pi/180), 
#                                    S_y = abs(slope)*sin(aspect*pi/180)) %>% 
#   mutate(MSWD = S_x*u10m + S_y*v10m)

```

### 2022-04-07 src/preprocess.py

@@ -1,0 +1,52 @@
+# Script for preprocessing WAIS data in preparation to analysis in R using INLA.
+
+## Imports
+from pathlib import Path
+from importlib import util
+import pandas as pd
+
+# Define environment directories
+ROOT_DIR = Path().absolute().parents[0]
+EXT_DIR = Path('/home/durbank/Documents/Research/Antarctica/WAIS-central/')
+
+# Import custom paipr processing module
+spec = util.spec_from_file_location(
+    "paipr", EXT_DIR.joinpath('src/my_mods/paipr.py'))
+paipr = util.module_from_spec(spec)
+spec.loader.exec_module(paipr)
+
+# Import raw data
+data_list = [folder for folder in EXT_DIR.joinpath(
+    'data/PAIPR-outputs').glob('*')]
+data_raw = pd.DataFrame()
+for folder in data_list:
+    data = paipr.import_PAIPR(folder)
+    data_raw = data_raw.append(data)
+
+# Remove results for below QC data reliability
+data_raw.query('QC_flag != 2', inplace=True)
+data_0 = data_raw.query(
+    'Year > QC_yr').sort_values(
+    ['collect_time', 'Year']).reset_index(drop=True)
+
+# Format and sort results for further processing
+data_form = paipr.format_PAIPR(data_0)
+
+# Create time series arrays for annual accumulation 
+# and error
+data_all = data_form.drop(['QC_flag', 'QC_med'], axis=1)
+accum_ALL = data_form.pivot(
+    index='Year', columns='trace_ID', values='accum')
+std_ALL = data_form.pivot(
+    index='Year', columns='trace_ID', values='std')
+
+# Create gdf of mean results for each trace and 
+# transform to Antarctic Polar Stereographic
+gdf_traces = paipr.long2gdf(data_form)
+gdf_traces.to_crs(epsg=3031, inplace=True)
+
+# Save data to disk
+data_all.to_csv(ROOT_DIR.joinpath('data/paipr-long.csv'))
+accum_ALL.to_csv(ROOT_DIR.joinpath('data/paipr-accum.csv'))
+std_ALL.to_csv(ROOT_DIR.joinpath('data/paipr-std.csv'))
+gdf_traces.to_file(ROOT_DIR.joinpath('data/traces.geojson'), driver='GeoJSON')

### 2022-04-11 notebooks/draft-modeling.Rmd
Breaks out pre-processing into separate scripts

@@
 
-## Data loading and pre-processing
-
-```{r data-load, message=FALSE}
-data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1) %>% 
-  mutate(Year = as.integer(Year))
-gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()
-
-# Remove points in far northern extent
-accum_bnds = st_coordinates(gdf_traces)
-gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
-                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))
-
-# Small-scale testing bounds
-gdf_traces = gdf_traces %>% st_crop(xmin=-1.4e6, xmax=-1.25e6,
-                                    ymin=min(accum_bnds[,2]), ymax=-3e5)
-
-# Remove unrealistically high elevations and overly large accum sites
-trace_drop = gdf_traces %>% filter(elev > 3000 | accum > 1000)
-
-# Add East/North coordinates and drop filtered sites
-pts_tmp = st_coordinates(gdf_traces)
-coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
-data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
-gdf_traces = gdf_traces %>% filter(!trace_ID %in% trace_drop$trace_ID)
-```
-
-We need to import the other covariate data for the region.
-This consists of topographic features from REMA and climatic variables from RACMO.
+## Data loading
 ```{r}
-REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
-dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
-crs(dem) = st_crs(gdf_traces)$wkt
-
-slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
-crs(slp) = crs(dem)
-
-aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
-crs(aspct) = crs(dem)
-
-topo.stk = c(dem, slp, aspct)
-
-topo.pts = extract(topo.stk, vect(gdf_traces$geometry)) %>% as_tibble() %>% 
-  rename(dem=REMA_200m_dem) %>% 
-  mutate(trace.ID=gdf_traces$trace_ID, aspect=sin(pi/180*aspect))
-topo.pts = topo.pts %>% select(-ID)
-
-RACMO.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data", 
-                      "/RACMO/2.3p2_yearly_ANT27_1979_2016/interim")
-files = list.files(path = RACMO.dir, full.names = TRUE, 
-                   pattern = "\\.tif$")
-
-r.smb = rast(files[1])
-names(r.smb) = 1979:2016
-smb.pts = extract(r.smb, vect(gdf_traces$geometry)) %>% 
-  as_tibble() %>% mutate(trace.ID=gdf_traces$trace_ID) %>% 
-  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
-               names_to = 'Year', values_to = 'SMB') %>% 
-  mutate(Year = as.integer(Year))
-
-r.u10m = rast(files[2])
-names(r.u10m) = 1979:2016
-u10m.pts = extract(r.u10m, vect(gdf_traces$geometry)) %>% 
-  as_tibble() %>% 
-  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
-               names_to = 'Year', values_to = 'u10m') %>% 
-  mutate(Year = as.integer(Year))
-
-r.v10m = rast(files[3])
-names(r.v10m) = 1979:2016
-v10m.pts = extract(r.v10m, vect(gdf_traces$geometry)) %>% 
-  as_tibble() %>% pivot_longer(cols = num_range(prefix='', range=1979:2016), 
-                               names_to = 'Year', values_to = 'v10m') %>% 
-  mutate(Year = as.integer(Year))
-
-racmo.pts = smb.pts %>% 
-  mutate(u10m=u10m.pts$u10m, v10m=v10m.pts$v10m) %>% 
-  select(-ID)
-
-data = data %>% 
-  left_join(racmo.pts, by = c("trace_ID" = "trace.ID", "Year")) %>% 
-  left_join(topo.pts, by = c("trace_ID" = "trace.ID"))
-
-# data = data %>% mutate(S_x = abs(slope)*cos(asin(aspect)), 
-#                        S_y = abs(slope)*aspect) %>% 
-#   mutate(MSWD = S_x*u10m + S_y*v10m)
+data = readRDS(here('data/Rdata-clean.rds'))
+gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
 ```
@@
 set.seed(777)
-gdf.tmp = gdf_traces %>% sample_frac(0.01)
-data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
-  filter(Year >= 1979) %>% filter(Year < 1999) %>% arrange(trace_ID, Year)
+gdf_traces = gdf_traces %>% sample_frac(0.01)
+data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
+  filter(Year >= 1979) %>% filter(Year < 2014) %>% arrange(trace_ID, Year)
+```
 
+We next subset to the variables of interest, center, standardize/normalize the covariates, and perform a train/test split.
+```{r}
 # Select variables of interest
-dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
-                      std, elev, dem, slope, aspect, u10m, v10m)
-# dat = data %>% select(trace_ID, East, North, Year, accum, 
-#                       std, elev, dem, slope, aspect, u10m, v10m)
+dat = data %>% select(trace_ID, East, North, Year, accum, 
+                      std, elev, dem, slope, aspect, u10m, v10m, 
+                      mu.u10, mu.v10)
 
@@
          dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
-         u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
+         u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
+         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE))
+         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
+
+# Scale covariates by sd
+dat = dat %>% 
+  mutate(Year = Year, elev = elev/sd(elev), dem=dem/sd(dem), 
+         slope=slope/sd(slope), u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), 
+         mu.u10=sd(mu.u10), mu.v10=sd(mu.v10))
 
@@
 
-I then define the SPDE object and create the identifier matrix $A$.
-**Question: Am I defining these priors correctly? I'm using the PC priors, which I think are defined in the untransformed spatial scale if I followed things right.**
-```{r}
-spde = inla.spde2.pcmatern(mesh = mesh, 
-  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
-  prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05
-# iset = inla.spde.make.index('i', n.spde = spde$n.spde,
-#   n.group = max(dat$Year))
+<!-- I then define the SPDE object and create the identifier matrix $A$. -->
+<!-- **Question: Am I defining these priors correctly? I'm using the PC priors, which I think are defined in the untransformed spatial scale if I followed things right.** -->
+<!-- ```{r} -->
+<!-- spde = inla.spde2.pcmatern(mesh = mesh,  -->
+<!--   prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05 -->
+<!--   prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05 -->
+<!-- # iset = inla.spde.make.index('i', n.spde = spde$n.spde, -->
+<!-- #   n.group = max(dat$Year)) -->
 
-# Create projector matrix for SPDE mesh
-A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
+<!-- # Create projector matrix for SPDE mesh -->
+<!-- A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North)) -->
 
-# Assign index vectors for SPDE model
-spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)
+<!-- # Assign index vectors for SPDE model -->
+<!-- spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde) -->
 
-# Make data stack
-dat.stack <- inla.stack(
-  data = list(y = dat.train$accum), 
-  A = list(1, 1, 1, A.spat),
-  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
-                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
-                        slope=dat.train$slope, aspect=dat.train$aspect, 
-                        u10m=dat.train$u10m, v10m=dat.train$v10m), 
-                 list(time = dat.train$Year), 
-                 spat.idx),
-  tag = 'dat') 
-```
+<!-- # Make data stack -->
+<!-- dat.stack <- inla.stack( -->
+<!--   data = list(y = dat.train$accum),  -->
+<!--   A = list(1, 1, 1, A.spat), -->
+<!--   effects = list(list(Intercept = rep(1, nrow(dat.train))),  -->
+<!--                  tibble(elev = dat.train$elev, dem=dat.train$dem,  -->
+<!--                         slope=dat.train$slope, aspect=dat.train$aspect,  -->
+<!--                         u10m=dat.train$u10m, v10m=dat.train$v10m),  -->
+<!--                  list(time = dat.train$Year),  -->
+<!--                  spat.idx), -->
+<!--   tag = 'dat')  -->
+<!-- ``` -->
 
-I also define a PC-prior for the autoregressive temporal autocorrelation.
-In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data.
-**Questions: I want to make sure I am defining the priors properly, so let me know what could be improved here.
+<!-- I also define a PC-prior for the autoregressive temporal autocorrelation. -->
+<!-- In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data. -->
+<!-- **Questions: I want to make sure I am defining the priors properly, so let me know what could be improved here.
 Also, I am currently modeling this as a separable spacetime model with independent random effects for space and time.
 I largely followed the first example in Section 8.6.1 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models).
-It also shows an example of doing this using a grouping method which I think might incorporate some interactions between space and time autocorrelations, but the fully indepedent separable model is MUCH faster.**
-```{r}
-## Define other priors (spatial field priors defined earlier)
+It also shows an example of doing this using a grouping method which I think might incorporate some interactions between space and time autocorrelations, but the fully indepedent separable model is MUCH faster.** -->
+<!-- ```{r} -->
+<!-- ## Define other priors (spatial field priors defined earlier) -->
 
-# Prior on autoregressive rho parameter
-time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
+<!-- # Prior on autoregressive rho parameter -->
+<!-- time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9))) -->
 
-# Prior on fixed effects intercept (not currently implemented)
-prior.fixed = list()
+<!-- # Prior on fixed effects intercept (not currently implemented) -->
+<!-- prior.fixed = list() -->
 
-# Prior on precision parameter for Gamma observations
-prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%
+<!-- # Prior on precision parameter for Gamma observations -->
+<!-- prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10% -->
 
-# Model formula
-formula = y ~ -1 + 
-  Intercept + dem + slope + aspect + #fixed effects
-  slope*aspect + #fixed effects interactions
-  f(time, model = 'ar1', hyper = time.spec) + #temporal random effect
-  f(spat.field, model = spde) #spatial random effect
-```
+<!-- # Model formula -->
+<!-- formula = y ~ -1 +  -->
+<!--   Intercept + dem + slope + aspect + #fixed effects -->
+<!--   slope*aspect + #fixed effects interactions -->
+<!--   f(time, model = 'ar1', hyper = time.spec) + #temporal random effect -->
+<!--   f(spat.field, model = spde) #spatial random effect -->
+<!-- ``` -->
 
-```{r}
-# Model fitting
-system.time(
-  mod.sep_ind <- inla(formula, 
-                  data = inla.stack.data(dat.stack), 
-                  family = 'Gamma',
-                  quantiles = NULL, 
-                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
-                  # control.fixed = prior.fixed, 
-                  control.family = list(hyper = list(prec = prec.prior)), 
-                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE))
-)
-```
+<!-- ```{r} -->
+<!-- # Model fitting -->
+<!-- system.time( -->
+<!--   mod.sep_ind <- inla(formula,  -->
+<!--                   data = inla.stack.data(dat.stack),  -->
+<!--                   family = 'Gamma', -->
+<!--                   quantiles = NULL,  -->
+<!--                   control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1),  -->
+<!--                   # control.fixed = prior.fixed,  -->
+<!--                   control.family = list(hyper = list(prec = prec.prior)),  -->
+<!--                   control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE)) -->
+<!-- ) -->
+<!-- ``` -->
 
-### Form 2
+<!-- ### Form 2 -->
 
-Rather than treating space and time wholly independently, we can also model the spatial component at each time step separately.
-This still does not include interactions between space and time effects, but does allow the spatial dependence to vary over time.
-The following implementation follows examples shown in  Section 8.6.2 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models) amd Section 7.1 of [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/ch-spacetime.html#discrete-time-domain).
+<!-- Rather than treating space and time wholly independently, we can also model the spatial component at each time step separately. -->
+<!-- This still does not include interactions between space and time effects, but does allow the spatial dependence to vary over time. -->
+<!-- The following implementation follows examples shown in  Section 8.6.2 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models) amd Section 7.1 of [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/ch-spacetime.html#discrete-time-domain).
+-->
 
-```{r}
-set.seed(777)
-gdf.tmp = gdf_traces %>% sample_frac(0.005)
-data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
-  filter(Year >= 1989) %>% filter(Year < 1999) %>% arrange(trace_ID, Year)
+<!-- ```{r} -->
+<!-- set.seed(777) -->
+<!-- gdf.tmp = gdf_traces %>% sample_frac(0.005) -->
+<!-- data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>%  -->
+<!--   filter(Year >= 1989) %>% filter(Year < 1999) %>% arrange(trace_ID, Year) -->
 
-# Select variables of interest
-dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
-                      std, elev, dem, slope, aspect, u10m, v10m)
-# dat = data %>% select(trace_ID, East, North, Year, accum, 
-#                       std, elev, dem, slope, aspect, u10m, v10m)
+<!-- # Select variables of interest -->
+<!-- dat = data.tmp %>% select(trace_ID, East, North, Year, accum,  -->
+<!--                       std, elev, dem, slope, aspect, u10m, v10m) -->
+<!-- # dat = data %>% select(trace_ID, East, North, Year, accum,  -->
+<!-- #                       std, elev, dem, slope, aspect, u10m, v10m) -->
 
-# Center covariates
-yr.min = min(dat$Year)-1
-dat = dat %>% 
-  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
-         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
-         u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
+<!-- # Center covariates -->
+<!-- yr.min = min(dat$Year)-1 -->
+<!-- dat = dat %>%  -->
+<!--   mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE),  -->
+<!--          dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE),  -->
+<!--          u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE)) -->
 
-# Split into training and testing sets
-dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
-dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
-dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
-```
+<!-- # Split into training and testing sets -->
+<!-- dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID) -->
+<!-- dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID) -->
+<!-- dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID) -->
+<!-- ``` -->
 
-```{r}
-mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
-                    max.edge = c(25000, 100000), cutoff = 7500)
+<!-- ```{r} -->
+<!-- mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),  -->
+<!--                     max.edge = c(25000, 100000), cutoff = 7500) -->
 
-spde = inla.spde2.pcmatern(mesh = mesh, 
-  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
-  prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05
+<!-- spde = inla.spde2.pcmatern(mesh = mesh,  -->
+<!--   prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05 -->
+<!--   prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05 -->
 
-# Create projector matrix for SPDE mesh
-A.spat = inla.spde.make.A(mesh = mesh,
-                          loc = cbind(dat.train$East, dat.train$North),
-                          group = dat.train$Year)
-A.year = inla.spde.make.A(mesh = mesh,
-                          loc = cbind(dat.train$East, dat.train$North))
+<!-- # Create projector matrix for SPDE mesh -->
+<!-- A.spat = inla.spde.make.A(mesh = mesh, -->
+<!--                           loc = cbind(dat.train$East, dat.train$North), -->
+<!--                           group = dat.train$Year) -->
+<!-- A.year = inla.spde.make.A(mesh = mesh, -->
+<!--                           loc = cbind(dat.train$East, dat.train$North)) -->
 
-# Assign index vectors for SPDE model
-spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde,
-                                n.group = length(unique(dat.train$Year)))
+<!-- # Assign index vectors for SPDE model -->
+<!-- spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde, -->
+<!--                                 n.group = length(unique(dat.train$Year))) -->
 
-# Assign index vectors for SPDE model for year covariate
-year.idx = inla.spde.make.index(name = "year.diff", n.spde = spde$n.spde)
+<!-- # Assign index vectors for SPDE model for year covariate -->
+<!-- year.idx = inla.spde.make.index(name = "year.diff", n.spde = spde$n.spde) -->
 
-# Make data stack
-# dat.stack <- inla.stack(
-#   data = list(y = dat.train$accum),
-#   A = list(1, 1, 1, A.spat),
-#   effects = list(list(Intercept = rep(1, nrow(dat.train))),
-#                  tibble(elev = dat.train$elev, dem=dat.train$dem,
-#                         slope=dat.train$slope, aspect=dat.train$aspect,
-#                         u10m=dat.train$u10m, v10m=dat.train$v10m),
-#                  list(time = dat.train$Year),
-#                  spat.idx),
-#   tag = 'dat')
-dat.stack = inla.stack(
-  data = list(y = dat.train$accum),
-  A = list(1, 1, A.spat, A.year),
-  effects = list(list(Intercept = rep(1, nrow(dat.train))),
-                 tibble(elev = dat.train$elev, dem=dat.train$dem,
-                        slope=dat.train$slope, aspect=dat.train$aspect,
-                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
-                        year=dat.train$Year),
-                 spat.idx, 
-                 year.idx), 
-  tag = 'dat')
+<!-- # Make data stack -->
+<!-- # dat.stack <- inla.stack( -->
+<!-- #   data = list(y = dat.train$accum), -->
+<!-- #   A = list(1, 1, 1, A.spat), -->
+<!-- #   effects = list(list(Intercept = rep(1, nrow(dat.train))), -->
+<!-- #                  tibble(elev = dat.train$elev, dem=dat.train$dem, -->
+<!-- #                         slope=dat.train$slope, aspect=dat.train$aspect, -->
+<!-- #                         u10m=dat.train$u10m, v10m=dat.train$v10m), -->
+<!-- #                  list(time = dat.train$Year), -->
+<!-- #                  spat.idx), -->
+<!-- #   tag = 'dat') -->
+<!-- dat.stack = inla.stack( -->
+<!--   data = list(y = dat.train$accum), -->
+<!--   A = list(1, 1, A.spat, A.year), -->
+<!--   effects = list(list(Intercept = rep(1, nrow(dat.train))), -->
+<!--                  tibble(elev = dat.train$elev, dem=dat.train$dem, -->
+<!--                         slope=dat.train$slope, aspect=dat.train$aspect, -->
+<!--                         u10m=dat.train$u10m, v10m=dat.train$v10m,  -->
+<!--                         year=dat.train$Year), -->
+<!--                  spat.idx,  -->
+<!--                  year.idx),  -->
+<!--   tag = 'dat') -->
 
-# Prior on autoregressive rho parameter
-time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
+<!-- # Prior on autoregressive rho parameter -->
+<!-- time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9))) -->
 
-# Prior on fixed effects intercept (not currently implemented)
-prior.fixed = list()
+<!-- # Prior on fixed effects intercept (not currently implemented) -->
+<!-- prior.fixed = list() -->
 
-# Prior on precision parameter for Gamma observations
-prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%
+<!-- # Prior on precision parameter for Gamma observations -->
+<!-- prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10% -->
 
-# Model formula
-formula = y ~ -1 +
-  Intercept + dem + slope + aspect + #fixed effects
-  # slope*aspect + #fixed effects interactions
-  f(spat.field, model = spde, group = spat.field.group,
-    control.group = list(model = 'ar1', hyper = time.spec)) + 
-  f(year.diff, year, model = spde, constr = FALSE)
+<!-- # Model formula -->
+<!-- formula = y ~ -1 + -->
+<!--   Intercept + dem + slope + aspect + #fixed effects -->
+<!--   # slope*aspect + #fixed effects interactions -->
+<!--   f(spat.field, model = spde, group = spat.field.group, -->
+<!--     control.group = list(model = 'ar1', hyper = time.spec)) +  -->
+<!--   f(year.diff, year, model = spde, constr = FALSE) -->
 
-# Model fitting
-system.time(
-  mod.sep_group <- inla(formula,
-                  data = inla.stack.data(dat.stack),
-                  family = 'Gamma',
-                  quantiles = NULL,
-                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1),
-                  # control.fixed = prior.fixed,
-                  control.family = list(hyper = list(prec = prec.prior)),
-                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE))
-)
-```
+<!-- # Model fitting -->
+<!-- system.time( -->
+<!--   mod.sep_group <- inla(formula, -->
+<!--                   data = inla.stack.data(dat.stack), -->
+<!--                   family = 'Gamma', -->
+<!--                   quantiles = NULL, -->
+<!--                   control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), -->
+<!--                   # control.fixed = prior.fixed, -->
+<!--                   control.family = list(hyper = list(prec = prec.prior)), -->
+<!--                   control.compute = list(waic = TRUE, return.marginals.predictor=FALSE, config=TRUE)) -->
+<!-- ) -->
+<!-- ``` -->
 
-On a (very) small testing set (use commented section in `data-load` chunk) of ~100 locations, the grouped separable model does perform substantially better based on WAIC score, but at the cost of around 50x the computation time (this difference may be even larger on a larger dataset).
-Fitted fixed effects are of a similar magnitude between the two models.
-This model produces greater precision in the gamma observations and roughly similar hyperparameters for the spatial field (this model has a higher range but lower range variance compared to the independent separable model).
-I'll need think more about what the GroupRho parameter for the spatial field means, and how it compares to the independent separable model.
+<!-- On a (very) small testing set (use commented section in `data-load` chunk) of ~100 locations, the grouped separable model does perform substantially better based on WAIC score, but at the cost of around 50x the computation time (this difference may be even larger on a larger dataset).
+-->
+<!-- Fitted fixed effects are of a similar magnitude between the two models. -->
+<!-- This model produces greater precision in the gamma observations and roughly similar hyperparameters for the spatial field (this model has a higher range but lower range variance compared to the independent separable model).
+-->
+<!-- I'll need think more about what the GroupRho parameter for the spatial field means, and how it compares to the independent separable model. -->
 
@@
 ```{r}
-
 mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
-                    max.edge = c(25000, 100000), cutoff = 1000)
+                    max.edge = c(25000, 75000), cutoff = 5000)
+plot(mesh)
+points(dat.train %>% select(East, North), col = "red")
 
 spde = inla.spde2.pcmatern(mesh = mesh, 
-  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
-  prior.sigma = c(400, 0.05))
+  prior.range = c(15000, 0.01), # P(range < 10 km) = 0.01
+  prior.sigma = c(10, 0.05))
 
-# year.spde = inla.spde2.matern(mesh = mesh)
-year.spde = spde
+spde.yr = inla.spde2.pcmatern(mesh = mesh, 
+  prior.range = c(5000, 0.01), # P(range < 10 km) = 0.01
+  prior.sigma = c(1, 0.01), 
+  constr = TRUE)
 
-# Create projector matrix for SPDE mesh
-A.spat = inla.spde.make.A(mesh = mesh,
-                          loc = cbind(dat.train$East, dat.train$North))
+# Index for SPDE model and projector matrix
+spat.idx = inla.spde.make.index(name = "spat.idx", n.spde = spde$n.spde)
+A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
 
-# Assign index vectors for SPDE model
-spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)
-
-# Assign index vectors for SPDE model for year covariate
-year.idx = inla.spde.make.index(name = "year.diff", n.spde = year.spde$n.spde)
+# Index for SPDE model and projector matrix
+spYR.idx = inla.spde.make.index(name = "spYR.idx", n.spde = spde.yr$n.spde)
+A.spYR = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
 
@@
   data = list(y = dat.train$accum), 
-  A = list(1, 1, A.spat, A.spat, 1),
+  A = list(1, 1, 1, A.spat, A.spYR),
   effects = list(list(Intercept = rep(1, nrow(dat.train))), 
@@
                         u10m=dat.train$u10m, v10m=dat.train$v10m, 
-                        year=dat.train$Year),
-                 year.idx, 
+                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10, 
+                        Year=dat.train$Year), 
+                 list(time = dat.train$Year), 
                  spat.idx, 
-                 list(time = dat.train$Year)),
+                 spYR.idx),
   tag = 'dat')
 
-# Prior on precision parameter for Gamma observations
+time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
+prior.fixed = list()
 prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%
 
-# Prior on autoregressive rho parameter
-time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
+# form.iid = y ~ -1 + Intercept + dem + slope + 
+#   # f(spat.idx, model = spde) + 
+#   Year + f(spat.idx, Year, model = spde.yr) +
+#   f(time, model = 'ar1', hyper = time.spec)
 
-# Model formula
-# formula = y ~ -1 + 
-#   Intercept + dem + slope + aspect + #fixed effects
-#   slope*aspect + #fixed effects interactions
-#   f(year.diff, year, model = "iid") + year + 
-#   f(spat.field, model = spde) + #spatial random effect
-#   f(time, model = 'ar1', hyper = time.spec)#temporal random effect
+form.iid = y ~ -1 + Intercept + dem + 
+  aspect + 
+  f(spat.idx, model = spde) +
+  Year + f(spYR.idx, Year, model = spde.yr) +
+  # f(spat.idx, Year, model = spde.yr) +
+  f(time, model = 'ar1', hyper = time.spec)
+```
 
-formula = y ~ -1 + Intercept + 
-  dem + slope + aspect + #fixed effects
-  # f(time, model = 'ar1', hyper = time.spec) + #AR temporal random effect
-  # f(year.diff, year, model = spde, constr = FALSE) + #spatially varying temporal trend
-  year + f(year.diff, year, model = spde) + #spatially varying temporal trend
-  f(spat.field, model = spde) #spatial random effect
-
-formula = y ~ -1 + 
-  f(spat.field, model = spde) + 
-  year + f(year.diff, year, model = spde)
-
-# Model fitting
-system.time(
-  mod.trends <- inla(formula,
+```{r}
+mod.iid = inla(form.iid, 
                data = inla.stack.data(dat.stack), 
@@
                quantiles = NULL, 
-                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), 
-                                           link = 1),
+               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
                # control.fixed = prior.fixed, 
                control.family = list(hyper = list(prec = prec.prior)), 
-                  control.compute = list(waic = TRUE, return.marginals.predictor=FALSE,
-                                         config=TRUE))
-)
+               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))
+
+```
+
+## Evalulate model performance
+
+```{r}
+summary(mod.iid)
+```
+
+```{r}
+
+t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {
+
+  if (any(class(mod.marg) == "matrix")) {
+    mod.marg = list(mod.marg)
+    if (!is.null(param.names)) {
+      names(mod.marg = param.names)
+    } else {
+      names(mod.marg) = "UNNAMED"
+    }
+  }
+
+  if (is.null(param.names)) {
+    param.names = names(mod.marg)
+  }
+
+    if (is.list(t.func) != TRUE) {
+    t.func = list(t.func)
+  }
+  if (length(t.func) == 1) {
+    t.func = rep(t.func, length(param.names))
+  }
+
+  # Preallocate output tbl
+  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))
+
+  marg.names = names(mod.marg)
+
+  for (i in 1:length(marg.names)) {
+    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
+    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
+  }
+  return(out)
+}
+```
+
+```{r}
+marg.fixed = t.marg(mod.iid$marginals.fixed, function(x) exp(x))
+ggplot(marg.fixed, aes(x=x,y=y)) + geom_line() + facet_wrap(vars(Param), scales = "free")
+
+HP.names = c("Gamma precision", "Range for spat.idx", "Stdev for spat.idx", 
+             "Range for spYR.idx", "Stdev for spYR.idx", "Time precision", "Rho for time")
+marg.hyper = t.marg(mod.iid$marginals.hyperpar, function(x) (x), param.names = HP.names)
+ggplot(marg.hyper, aes(x=x,y=y)) + geom_line() + facet_wrap(vars(Param), scales = "free")
+```
+
+```{r}
+p.selection = c("Intercept", "dem", "aspect", "Gamma precision", "Range for spat.idx", 
+                "Stdev for spat.idx", "Time precision", "Rho for time")
+tmp = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
+```
+```{r fig, fig.width=16}
+ggplot(tmp, aes(x=x,y=y)) + geom_line() + 
+  facet_wrap(vars(Param), scales = "free") + 
+  theme(text = element_text(size = 20))  
+```
+
+```{r}
+Bi.sp = tibble(East = mesh$loc[,1], North = mesh$loc[,2], 
+                 B.i = mod.iid$summary.random$spYR.idx$mean) %>% 
+  mutate(B.tot = B.i+mod.iid$summary.fixed["Year",]$mean)
+ggplot(Bi.sp) + geom_density(aes(x=B.tot))
+# ggplot(Bi.sp) + geom_density(aes(x=B.i))
+
+```
+
+Compare results to a standard log-link gamma GLM to ensure correct interpretation
+```{r}
+# Formula without random variables
+f.base = y ~ -1 + Intercept + Year + dem + aspect
+
+# Fitting a frequentist GLM to data
+mod.glm = glm(f.base, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1), 
+               family = Gamma(link = "log"))
+
+summary(mod.glm)
+```
+
+All of this looks good except for when looking at the individual summaries spatial random effects on the temporal trend.
+In this case, the sd for individual estimates is some 4-5 orders of magnitude larger than the mean, making them virtually worthless.
+This is visually shown in the below plot (and recall that this is the untransformed values, so after exponentiation this just becomes ridiculous).
+```{r}
+plot(mod.iid$marginals.random$spYR.idx$index.7, type='l')
 ```
@@
 
-test.fitted = tibble(mu = mod.trends$summary.fitted.values$mean[fitted.idx]) %>% 
+test.fitted = tibble(mu = mod.iid$summary.fitted.values$mean[fitted.idx]) %>% 
   bind_cols(dat.train %>% select(Year, East, North)) %>% 
   st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
-test.rand = tibble(t.trend = exp(mod.trends$summary.random$year.diff$mean)-1, 
-                   spat.mu = exp(mod.trends$summary.random$spat.field$mean), 
+test.rand = tibble(t.trend = exp(mod.iid$summary.random$spYR.idx$mean + mod.iid$summary.fixed['Year',]$mean), 
+                   spat.mu = exp(mod.iid$summary.random$spat.idx$mode), 
                    East = mesh$loc[,1], North = mesh$loc[,2]) %>% 
   st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
-ggplot(test.rand) + geom_sf(aes(color=t.trend)) + scale_color_viridis_c()
+ggplot(test.rand) + geom_sf(aes(color=spat.mu)) + scale_color_viridis_c()
 ```
@@
 ## Model interpretation
-
-```{r}
-mod.train = mod.sep_ind
-```
-
-Below is a summary of the model results.
-```{r}
-summary(mod.train)
-```
 
@@
   data = list(y = NA), 
-  A = list(1, 1, 1, A.test),
+  A = list(1, 1, 1, A.test, A.test),
   effects = list(list(Intercept = rep(1, nrow(dat.test))), 
@@
                         slope=dat.test$slope, aspect=dat.test$aspect, 
-                        u10m=dat.test$u10m, v10m=dat.test$v10m),
+                        u10m=dat.test$u10m, v10m=dat.test$v10m,
+                        mu.u10=dat.test$mu.u10, mu.v10=dat.test$mu.v10, 
+                        Year=dat.test$Year), 
                  list(time = dat.test$Year), 
-                 spat.idx),
+                 spat.idx, 
+                 spYR.idx),
   tag = 'test')
@@
 # Model fitting to test data
-system.time(
-  mod.test <- inla(formula, 
+
+mod.test <- inla(form.iid, 
                  data = inla.stack.data(val.stack), 
@@
                  control.inla = list(strategy = 'adaptive'), 
-                  control.mode = list(theta = results$mode$theta, restart = FALSE))
-)
+                 control.mode = list(theta = mod.iid$mode$theta, restart = FALSE))
+
 ```
 
-Below is a function that, given an input of mean estimates, sd on mean estimates, precision parameter for Gamma observations, and sd on precision parameter, returns $n$ samples from the posterior predictive distribution for the given inputs.
-```{r}
-pp.dist = function(data, prec, prec.sd, n=5000, return_sample=FALSE) {
-  if (return_sample==TRUE) {
-    ppd = matrix(data=NA, nrow=nrow(data), ncol=n)
-  } else {
-    ppd = tibble(mu=rep(0,nrow(data)), map=rep(0,nrow(data)), 
-                 CI.low=rep(0,nrow(data)), CI.high=rep(0,nrow(data)))
-  }
+<!-- Below is a function that, given an input of mean estimates, sd on mean estimates, precision parameter for Gamma observations, and sd on precision parameter, returns $n$ samples from the posterior predictive distribution for the given inputs.
+-->
+<!-- ```{r} -->
+<!-- pp.dist = function(data, prec, prec.sd, n=5000, return_sample=FALSE) { -->
+<!--   if (return_sample==TRUE) { -->
+<!--     ppd = matrix(data=NA, nrow=nrow(data), ncol=n) -->
+<!--   } else { -->
+<!--     ppd = tibble(mu=rep(0,nrow(data)), map=rep(0,nrow(data)),  -->
+<!--                  CI.low=rep(0,nrow(data)), CI.high=rep(0,nrow(data))) -->
+<!--   } -->
 
-  for (i in 1:nrow(data)) {
-    dat.i = data[i,]
-    mu = rnorm(n, mean=dat.i$mu, sd=dat.i$mu.sd)
-    a = rnorm(n, mean=prec, sd=prec.sd)
-    b = a/mu
-    dist = rgamma(n=n, shape = a, rate = b)
+<!--   for (i in 1:nrow(data)) { -->
+<!--     dat.i = data[i,] -->
+<!--     mu = rnorm(n, mean=dat.i$mu, sd=dat.i$mu.sd) -->
+<!--     a = rnorm(n, mean=prec, sd=prec.sd) -->
+<!--     b = a/mu -->
+<!--     dist = rgamma(n=n, shape = a, rate = b) -->
 
-    if (return_sample==TRUE) {
-      ppd[i,] = dist
-    } else {
-      dist.mu = mean(dist)
-      d = density(dist)
-      dist.map = d$x[which.max(d$y)]
-      dist.hdi = HDInterval::hdi(dist, credMass=0.89)
-      ppd[i,] = tibble_row(mu=dist.mu, map=dist.map, 
-                           CI.low=as.numeric(dist.hdi[1]), 
-                           CI.high=as.numeric(dist.hdi[2]))
-    }
-  }
-  return(ppd)
-}
-```
+<!--     if (return_sample==TRUE) { -->
+<!--       ppd[i,] = dist -->
+<!--     } else { -->
+<!--       dist.mu = mean(dist) -->
+<!--       d = density(dist) -->
+<!--       dist.map = d$x[which.max(d$y)] -->
+<!--       dist.hdi = HDInterval::hdi(dist, credMass=0.89) -->
+<!--       ppd[i,] = tibble_row(mu=dist.mu, map=dist.map,  -->
+<!--                            CI.low=as.numeric(dist.hdi[1]),  -->
+<!--                            CI.high=as.numeric(dist.hdi[2])) -->
+<!--     } -->
+<!--   } -->
+<!--   return(ppd) -->
+<!-- } -->
+<!-- ``` -->
 
-This is an example of generating posterior predictive distributions for a couple of data points.
-**Am I doing this right?**
-```{r message=FALSE}
-# Get inla grid indices where test data resides
-test.idx = inla.stack.index(val.stack, 'test')$data
+<!-- This is an example of generating posterior predictive distributions for a couple of data points. -->
+<!-- **Am I doing this right?** -->
+<!-- ```{r message=FALSE} -->
+<!-- # Get inla grid indices where test data resides -->
+<!-- test.idx = inla.stack.index(val.stack, 'test')$data -->
 
-# Create df of model predictions for test data
-test.fitted = tibble(mu = mod.test$summary.fitted.values$mean[test.idx], 
-                     mu.sd = mod.test$summary.fitted.values$sd[test.idx], 
-                     accum = dat.test$accum, accum.sd = dat.test$std)
+<!-- # Create df of model predictions for test data -->
+<!-- test.fitted = tibble(mu = mod.test$summary.fitted.values$mean[test.idx],  -->
+<!--                      mu.sd = mod.test$summary.fitted.values$sd[test.idx],  -->
+<!--                      accum = dat.test$accum, accum.sd = dat.test$std) -->
 
-# Create matrix of model posterior  samples for a few test data
-system.time(
-  post.pred_samp <- pp.dist(data=head(test.fitted), prec=mod.test$summary.hyperpar$mean[1], 
-                       prec.sd=mod.test$summary.hyperpar$sd[1], return_sample = TRUE)
-)
+<!-- # Create matrix of model posterior  samples for a few test data -->
+<!-- system.time( -->
+<!--   post.pred_samp <- pp.dist(data=head(test.fitted), prec=mod.test$summary.hyperpar$mean[1],  -->
+<!--                        prec.sd=mod.test$summary.hyperpar$sd[1], return_sample = TRUE) -->
+<!-- ) -->
 
-pp.samples = as_tibble(t(post.pred_samp), .name_repair = 'universal') %>%
-  rename(R1=...1, R2=...2, R3=...3, R4=...4, R5=...5, R6=...6) %>% 
-  mutate(Sample.ID = 1:ncol(post.pred_samp)) %>% 
-  pivot_longer(cols = starts_with("R"), names_to = "R.name", values_to = "accum.rvs")
+<!-- pp.samples = as_tibble(t(post.pred_samp), .name_repair = 'universal') %>% -->
+<!--   rename(R1=...1, R2=...2, R3=...3, R4=...4, R5=...5, R6=...6) %>%  -->
+<!--   mutate(Sample.ID = 1:ncol(post.pred_samp)) %>%  -->
+<!--   pivot_longer(cols = starts_with("R"), names_to = "R.name", values_to = "accum.rvs") -->
 
-accum.obs = tibble()
-for (i in 1:length(unique(pp.samples$R.name))) {
-  r.name = paste("R", i, sep = '')
-  accum.i = rnorm(max(pp.samples$Sample.ID), 
-                  mean=test.fitted$accum[i], 
-                  sd=test.fitted$accum.sd[i])
-  accum.obs = accum.obs %>% 
-    bind_rows(tibble(Sample.ID = 1:length(accum.i), R.name = r.name, 
-                     accum.obs = accum.i))
-}
+<!-- accum.obs = tibble() -->
+<!-- for (i in 1:length(unique(pp.samples$R.name))) { -->
+<!--   r.name = paste("R", i, sep = '') -->
+<!--   accum.i = rnorm(max(pp.samples$Sample.ID),  -->
+<!--                   mean=test.fitted$accum[i],  -->
+<!--                   sd=test.fitted$accum.sd[i]) -->
+<!--   accum.obs = accum.obs %>%  -->
+<!--     bind_rows(tibble(Sample.ID = 1:length(accum.i), R.name = r.name,  -->
+<!--                      accum.obs = accum.i)) -->
+<!-- } -->
 
-pp.samples = pp.samples %>% left_join(accum.obs, by = c("R.name", "Sample.ID")) %>% 
-  mutate(R.name = as.factor(R.name))
+<!-- pp.samples = pp.samples %>% left_join(accum.obs, by = c("R.name", "Sample.ID")) %>%  -->
+<!--   mutate(R.name = as.factor(R.name)) -->
 
-# ggplot(pp.samples) + 
-#   geom_density(aes(x=accum.rvs, color=ROW, fill=ROW), alpha=0.1) + 
-#   xlab('Posterior predictive accumulation')
+<!-- # ggplot(pp.samples) +  -->
+<!-- #   geom_density(aes(x=accum.rvs, color=ROW, fill=ROW), alpha=0.1) +  -->
+<!-- #   xlab('Posterior predictive accumulation') -->
 
-vars = c("INLA"="red", "PAIPR"="blue")
-ggplot(pp.samples) + 
-  geom_density(aes(x=accum.rvs, color='INLA', fill='INLA'), alpha=0.1) + 
-  geom_density(aes(x=accum.obs, color='PAIPR', fill='PAIPR'), alpha=0.1) + 
-  scale_colour_manual(name="Data sources:", values=vars) + 
-  scale_fill_manual(name="Data sources:", values=vars) + 
-  xlab("Accumulation") + 
-  facet_wrap(~R.name)
-```
+<!-- vars = c("INLA"="red", "PAIPR"="blue") -->
+<!-- ggplot(pp.samples) +  -->
+<!--   geom_density(aes(x=accum.rvs, color='INLA', fill='INLA'), alpha=0.1) +  -->
+<!--   geom_density(aes(x=accum.obs, color='PAIPR', fill='PAIPR'), alpha=0.1) +  -->
+<!--   scale_colour_manual(name="Data sources:", values=vars) +  -->
+<!--   scale_fill_manual(name="Data sources:", values=vars) +  -->
+<!--   xlab("Accumulation") +  -->
+<!--   facet_wrap(~R.name) -->
+<!-- ``` -->
 
-And these return summary statistics for the ppc distributions for all the test data.
-```{r}
-# Generate summaries of posterior predictive checks for test data
-system.time(
-  post.pred_test <- pp.dist(data=test.fitted, prec=mod.test$summary.hyperpar$mean[1], 
-                       prec.sd=mod.test$summary.hyperpar$sd[1])
-)
-```
+<!-- And these return summary statistics for the ppc distributions for all the test data. -->
+<!-- ```{r} -->
+<!-- # Generate summaries of posterior predictive checks for test data -->
+<!-- system.time( -->
+<!--   post.pred_test <- pp.dist(data=test.fitted, prec=mod.test$summary.hyperpar$mean[1],  -->
+<!--                        prec.sd=mod.test$summary.hyperpar$sd[1]) -->
+<!-- ) -->
+<!-- ``` -->
 
-Below are some plots visually summarizing the performance of the test data
-```{r}
-res.test = dat.test %>% 
-  mutate(mu = test.fitted$mu, 
-         mu.pp=post.pred_test$mu, 
-         map.pp=post.pred_test$map, 
-         CI.low=post.pred_test$CI.low, 
-         CI.hi=post.pred_test$CI.high) %>% 
-  mutate(y.res = mu-accum, pp.res=map.pp-accum)
+<!-- Below are some plots visually summarizing the performance of the test data -->
+<!-- ```{r} -->
+<!-- res.test = dat.test %>%  -->
+<!--   mutate(mu = test.fitted$mu,  -->
+<!--          mu.pp=post.pred_test$mu,  -->
+<!--          map.pp=post.pred_test$map,  -->
+<!--          CI.low=post.pred_test$CI.low,  -->
+<!--          CI.hi=post.pred_test$CI.high) %>%  -->
+<!--   mutate(y.res = mu-accum, pp.res=map.pp-accum) -->
 
-ggplot(data=res.test, aes(accum, map.pp)) + 
-  geom_point() + 
-  geom_abline(slope=1, intercept=0, color='red') + 
-  # geom_smooth(method = 'lm') + 
-  # lims(x=c(200,600), y=c(200,600)) + 
-  xlab("Accum (obs)") + ylab("Accum (PP MAP)")
+<!-- ggplot(data=res.test, aes(accum, map.pp)) +  -->
+<!--   geom_point() +  -->
+<!--   geom_abline(slope=1, intercept=0, color='red') +  -->
+<!--   # geom_smooth(method = 'lm') +  -->
+<!--   # lims(x=c(200,600), y=c(200,600)) +  -->
+<!--   xlab("Accum (obs)") + ylab("Accum (PP MAP)") -->
 
-ggplot(res.test, aes(mu, map.pp)) + geom_point() + 
-  geom_abline(slope=1, intercept=0, color='red') + 
-  # lims(x=c(200,600), y=c(200,600)) + 
-  xlab("Accum (expected)") + ylab("Accum (post. predictive MAP)")
+<!-- ggplot(res.test, aes(mu, map.pp)) + geom_point() +  -->
+<!--   geom_abline(slope=1, intercept=0, color='red') +  -->
+<!--   # lims(x=c(200,600), y=c(200,600)) +  -->
+<!--   xlab("Accum (expected)") + ylab("Accum (post. predictive MAP)") -->
 
-d.map = density(res.test$pp.res)
-map.mode = d.map$x[which.max(d.map$y)]
-d.mu = density(res.test$y.res)
-mu.mode = d.mu$x[which.max(d.mu$y)]
-ggplot(res.test) + geom_density(aes(pp.res)) + 
-  geom_vline(xintercept = map.mode, linetype=2, color='red') + 
-  annotate(geom="text", x=map.mode-67, y=max(d.map$y), 
-           label=paste("MAP =", format(map.mode, digits=3)), 
-           color="red") + 
-  geom_vline(xintercept = mu.mode, linetype=2, color='blue') + 
-  annotate(geom="text", x=mu.mode+33, y=max(d.map$y), 
-           label=paste("Mu =", format(mu.mode, digits=3)), 
-           color="blue") + 
-  xlab("Mean post. pred. accum residuals") + ylab('density')
-```
+<!-- d.map = density(res.test$pp.res) -->
+<!-- map.mode = d.map$x[which.max(d.map$y)] -->
+<!-- d.mu = density(res.test$y.res) -->
+<!-- mu.mode = d.mu$x[which.max(d.mu$y)] -->
+<!-- ggplot(res.test) + geom_density(aes(pp.res)) +  -->
+<!--   geom_vline(xintercept = map.mode, linetype=2, color='red') +  -->
+<!--   annotate(geom="text", x=map.mode-67, y=max(d.map$y),  -->
+<!--            label=paste("MAP =", format(map.mode, digits=3)),  -->
+<!--            color="red") +  -->
+<!--   geom_vline(xintercept = mu.mode, linetype=2, color='blue') +  -->
+<!--   annotate(geom="text", x=mu.mode+33, y=max(d.map$y),  -->
+<!--            label=paste("Mu =", format(mu.mode, digits=3)),  -->
+<!--            color="blue") +  -->
+<!--   xlab("Mean post. pred. accum residuals") + ylab('density') -->
+<!-- ``` -->
 
@@
 spat.tbl = tibble(E = mesh$loc[,1], N = mesh$loc[,2], 
-                  spat.mu = mod.test$summary.random$spat.field$mean, 
-                   spat.sd = mod.test$summary.random$spat.field$sd) %>% 
+                  spat.mu = mod.test$summary.random$spat.idx$mode, 
+                   spat.sd = mod.test$summary.random$spat.idx$sd) %>% 
   st_as_sf(coords=c("E", "N"), crs=st_crs(dat.train))
@@
 time.tbl = tibble(Year=yr.min+(1:nrow(mod.test$summary.random$time)), 
-                  time.mu = mod.test$summary.random$time$mean, 
+                  time.mu = mod.test$summary.random$time$mode, 
                    time.sd = mod.test$summary.random$time$sd)
@@
 
+# Add time coordinates to prediction grid
+
 # # Extract topo covariates from rasters for each grid point

### 2022-05-12 notebooks/reports/phil_20220506.Rmd
Adds report detailing INLA model and questions for Phil

---
title: "WAIS Accum Modeling in INLA"
output: 
  html_notebook:
    code_folding: hide
---

Here is the breakdown so far of what I am trying to do in INLA with modeling annual accumulation across central West Antarctica.
The main parameter of interest for me is the spatial distribution in the change in accumulation over time (with the uncertainty in this trend across the region) while accounting for the temporal and spatial autocorrelation in the data.
My main questions are (1) do I understand the model properly and is it doing/estimating what I think it is? (2) are there better ways that you know of to implement the model for my targets of interest? and (3) I'm seeing some weird behavior in some of the marginal posterior distributions that I would like your thoughts on.

## Model setup

*Here's a overview of the statistical underpinnings of the model as I see them.*
*Feel free to critique/correct any misconceptions or errors you see here (and any issues with equations/notations, as I am new to this and have also been revising/updating it a fair bit).*
*Many thanks!*

The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension.
I model accumulation as a gamma distribution to ensure a zero-bounded function with increased variance at higher expected values.
$$\mathbf{Y} \sim Gamma(a,b)$$
Rather than using the shape and rate parameters $a$ and $b$, I express this gamma distribution in terms of the expected y value $E(y_i) = a_i/b_i = \mu_i$ and its associated variance (as expressed in INLA using the precision parameter $\phi$) $Var(y_i) = a_i/b_i^2 = \mu_i^2/\phi$.
This $\phi$ is the same precision parameter as output from `inla` (there is also a scale parameter $s$ that has a default of 1 where $Var(y_i) = 1/\tau$ and $\tau = (s\phi)/\mu_i^2$).
With rearranging and substituting, I have $a_i = \phi$ and $b_i = \phi/\mu_i$ and I can therefore express the distribution in accumulation observations in terms of $\phi$ and $\mu_i$.
$$y_i \sim Gamma(\mu_i, \phi)$$

For each observation $y_{i,t} \in \mathbf{Y}$, I construct a linear predictor $\eta_{i,t}$ for the expectation value $\mu_{i,t}$.
As I assume a gamma-distribution for accumulation, I associate $\eta_{i,t}$ and $\mu_{i,t}$ via the log link function.
$$\log(\mu_{i,t}) = \eta_{i,t}$$ 
I can further model $\eta_{i,t}$ as a generalized linear model consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case modeling a spatial random effect $u(\cdot)$ and a temporal random effect $\omega(\cdot)$).
$$
\eta_{i,t} = \alpha + \sum^M_{m=1} \beta_m x_{mi} + u(s_i) + \omega(t)
$$

These latent random effects $u_i$ and $\omega_i$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
I can model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of some hyperparameters $\pmb{\psi}$ and an associated prior distribution $p(\pmb{\psi})$.
This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{jk} = Cov(\theta_j, \theta_k)$.

$$
\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
$$

If I assume adherence to Markovian properties, the precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of the process, i.e. $Q_{jk} \neq 0 \iff k \in \{j,N(j)\}$.
In this assumption, I specify a Gaussian-Markov Random Field for both the spatial and temporal structures, permitting vast improvements in computational efficiency.
I also model the spatial and temporal random effects independently, which again greatly improves computation time, but will fail to capture any co-dependencies between these effects.

I model the temporal random effect $\omega$ as a first-order autoregressive model^[I will eventually want to update this to an AR(2) process due to the known temporal autocorrelation in the data] (representing how the spatial field evolves in time) where
$$\omega_t = \rho \omega_{t-1}$$

The objective, therefore, is to find the joint posterior distribution of the various parameters of interest, as this will allow me to perform inference and make predictions at unobserved locations.
$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s \mid \mathbf{Y})$$

### Direct temporal trend analysis

*This is where I've (maybe?) diverged a bit from what I discussed above.*

As I am most interested in estimating the spatially-varying trend in accumulation over time, I've attempted to adapt/implement a model to explicitly do this.
This rendition of the model follows the areal example for disease mapping found in Section 3.3 of [A tutorial in spatial and spatio-temporal models with R-INLA](https://discovery.ucl.ac.uk/id/eprint/1415919/1/Baio_BlaCamBaiRue.pdf), along with some inspiration from an [INLA discussion group thread](https://groups.google.com/g/r-inla-discussion-group/c/X_g4NuqqHfY).
The [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in "Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA" also discusses how to implement spatially-varying model coefficients in INLA, but near as I can tell it doesn't explicitly discuss extracting the time coefficient (but maybe I missed it, or it is trivial to add this?).

The idea here (i.e. in following the disease mapping example) is to directly model the temporal trend as a global fixed effect, but then model the deviations from the global value as a spatially-correlated random effect.
I want to do this while also accounting for both spatial and temporal correlations in the expected values of accumulation.
The linear predictor therefore now looks like:
$$
\eta_{i,t} = \alpha + \sum^M_{m=1} \beta_m x_{mi} + (\beta_G + \delta_i) \times time \ + u(s_i) + \omega(t)
$$

This model therefore accounts for fixed effects, a global fixed effect from time $\beta_G$, a spatially-varying deviation (modeled as a spatial random effect indexed on time) from the global time effect $\delta_i$, a spatial random effect, and a temporal random effect.

*When looking at the model implementation below, is the above model what I am actually doing?*

## Model encoding

### Imports

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(sf)
library(cetcolor)
# library(terra)
library(INLA)
library(spdep)
```

### Data loading
```{r}
data = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
```

While I am testing/developing things, I'm only using a small subset of the data (around 1/2 a percent).
These are spatially distributed across the full region but further limited to the time period 1979--2010.
This still gives a few thousand observations to work with though.
Also for initial testing of everything, I am only including select time-invariant covariates.
Once I have a fully fleshed-out workflow though, I intend to add a few more, including some time-varying climate estimates (e.g. wind vectors).
I center the covariates about the mean and scale them with the standard deviation, and also normalize the discrete time dimension, setting the min year as Year 1.
I also perform a training/validation split for later use.

```{r}
set.seed(777)
gdf_traces = gdf_traces %>% sample_frac(0.01)
data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year < 2010) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m, 
                      mu.u10, mu.v10)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(Year = Year, elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), 
         mu.u10=sd(mu.u10), mu.v10=sd(mu.v10))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

### INLA  modeling

I create the mesh used for defining the neighborhoods (this is important for generating the sparse precision matrix $Q$, enabling the GMRF approximation).
I also create spatial projector matrices for the spatial random effect `A.spat` and the spatial effect on the deviation from the global time trend `A.spYR`.
I assign priors here using PC matern covariance.
*I may also need to revisit the priors specification, as I'm not sure I have the right assumption of units/transformations for all of them.*

*These next sections are where I'm not sure I am telling INLA to do what I think I am.*
```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 75000), cutoff = 5000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(15000, 0.01), # P(range < 15 km) = 0.01
  prior.sigma = c(10, 0.05))

spde.yr = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(5000, 0.01), # P(range < 5 km) = 0.01
  prior.sigma = c(1, 0.01), 
  constr = TRUE) #Adds an integrate-to-zero constraint (this seemed appropriate as this will model deviations from the global mean effect for time)

# Index for SPDE model and projector matrix
spat.idx = inla.spde.make.index(name = "spat.idx", n.spde = spde$n.spde)
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))

# Index for SPDE model and projector matrix
spYR.idx = inla.spde.make.index(name = "spYR.idx", n.spde = spde.yr$n.spde)
A.spYR = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
```

```{r}
# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat, A.spYR),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10, 
                        Year=dat.train$Year), 
                 list(time = dat.train$Year), 
                 spat.idx, 
                 spYR.idx),
  tag = 'dat')

# Priors for temporal autocorrelation, fixed effects (currently empty), and gamma precision
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
prior.fixed = list()
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Model formula
form = y ~ -1 + Intercept + #modeling the intercept
  dem + aspect + #fixed effects
  f(spat.idx, model = spde) + #spatial random effect
  Year + f(spYR.idx, Year, model = spde.yr) + #Global effect from year and deviation from global (modeled as a spatial random effect with with weights given by Year)
  f(time, model = 'ar1', hyper = time.spec) #Temporal random effect modeled as AR1
```
*The above model formula is a main place I'm not sure I am specifying what I think I am for INLA. Is this doing what I think I'm doing i.e. the formula shown in "Direct temporal trend analysis?"*

## Run the model

```{r}
mod = inla(form, 
               data = inla.stack.data(dat.stack), 
               family = 'Gamma',
               quantiles = NULL, 
               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
               # control.fixed = prior.fixed, 
               control.family = list(hyper = list(prec = prec.prior)), 
               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))

```

## Evalulate model performance

```{r}
summary(mod)
```

Compare results to a standard log-link gamma GLM to ensure correct interpretation of the fixed effects.
```{r}
# Formula without random variables
f.glm = y ~ -1 + Intercept + Year + dem + aspect

# Fitting a frequentist GLM to data
mod.glm = glm(f.glm, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1), 
               family = Gamma(link = "log"))

summary(mod.glm)
```

### Posterior plots

Below is just a helper function for extracting and transforming marginal distributions to more intuitive scales.
I've then included some plots of the transformed marginal posteriors.
The hyperparameters for precisions and the sd's of spatial indices are left untransformed in these plots.
```{r}

t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {

  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }

  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }

    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }

  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))

  marg.names = names(mod.marg)

  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", "Range for spat.idx", "Stdev for spat.idx", 
             "Range for spYR.idx", "Stdev for spYR.idx", "Time precision", "Rho for time")
marg.hyper = t.marg(mod$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "Year", "dem", "aspect", "Gamma precision", "Range for spat.idx", 
                "Stdev for spat.idx", "Time precision", "Rho for time", "Stdev for spYR.idx", 
                "Range for spYR.idx")
tmp = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r fig, fig.width=16}
ggplot(tmp, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

If I understand correctly (and used the right transformations for the links), the following posteriors should show:

- Intercept: the mean accumulation when all other effects are at their mean values (this will be in the original units of mm w.e. per year) *this value of ~430 mm/a seems reasonable*
- Gamma precision: the expected residual variability in accumulation from random noise not accounted for in the expected value $\mu_{i,t}$ *(for the average accumulation, this would be a sd of ~86 mm/a, which seems reasonable)*
- fixed effect coefficients: should be multiplicative e.g. a one $\sigma$ increase in elevation (`dem` in the model) results in ~14% reduction in accumulation ($0.86-1= -0.14$) *these values also seem fairly reasonable*
- range: the area-of-effect for spatial correlation (in meters) *again seems reasonable (~80 km) for spatial correlation in accumulation*
- rho: the temporal autocorrelation *we assume this to be pretty high, so ~0.9 again makes sense*
- `stdev for spat.idx`: the standard deviation in accumulation values (in original mm/a?) due to spatial autocorrelation i.e. how much variability we expect due to spatial correlation? *if that's true, this seems rather small at ~0.15 mm/a, especially given the value for precision.*
  - *After additional consideration, I think this might be on an untransformed log scale, in which case this represents a $\pm16$% impact on the expected accumulation, which seems reasonable to me. Any clarification/confirmation you can offer would be wonderful though.*
- `stdev for spYR.idx`: the standard deviation in the magnitude of deviation from the global accumulation trend with time i.e. how much variability we expect in the `Year` fixed coefficient due to spatial correlation. *I don't really have any idea about the units here though, is this perhaps a % change as well? If that's true, then the temporal trends STRONGLY vary based on location, given the narrow range in the global coefficient for Year of ~0.5% loss per year and that a decent amount of probability in the deviations shows 25+%. Could use some insight on this for sure.*
  - *Or might this be a % change to the trend value itself? In which case it would represent say a 25% increase/decrease relative to the -0.5% baseline?*

For the most part, these posteriors seem reasonable and fit with what I would broadly expect.
I'm having some trouble interpreting/assessing a couple of them though (as noted above), and would greatly appreciate your thoughts/feedback on it.

I'm also noticing some other weird things that I would like your thoughts on.
Regarding the deviations from the global time trend, I did a bit more digging and found that the mean values across the region for these deviations are minuscule (shown in the plot below), on the order of $10^{-17}$, while the sd's in these values are much much higher ($10^{-1}$).
The deviations also do not sum to zero (as I think they should as I give this constraint when defining the SPDE model in Line 154), but maybe this is just an artifact of the values being so close to zero anyways?
```{r}
Bi.sp = tibble(East = mesh$loc[,1], North = mesh$loc[,2], 
               delta.i = mod$summary.random$spYR.idx$mean, 
               delta.sd = mod$summary.random$spYR.idx$sd) %>% 
  mutate(B.G = delta.i+mod$summary.fixed["Year",]$mean)
# ggplot(Bi.sp) + geom_density(aes(x=B.G))
ggplot(Bi.sp) + geom_density(aes(x=delta.i))

ggplot(Bi.sp) + geom_density(aes(x=delta.sd))
```

This is further highlighted by the below plot of a representative marginal distribution in $\beta_G + d_i$ for a single location (after applying an inverse link transformation, so this now shows the % change in accumulation per year due to time for this location).
```{r}
B.ex = t.marg(mod$summary.fixed$mean[4]+mod$marginals.random$spYR.idx$index.7, 
              function(x) exp(x)) %>% 
  mutate(x = x-1)
ggplot(B.ex) + geom_line(aes(x=x,y=y))
```

So the modal deviation is negligible relative to the global time trend (-0.5%), but the error on this effect is massive with very fat tails, resulting in a plausible effect anywhere from a ~50% loss to a 100% gain in accumulation per year for this particular location.
All the other locations have similarly massive uncertainties.
This seems like far too wide a range to be useful.
I'm now trying to figure out if this results from a mis-specfication of the model, or if INLA itself is just struggling with these data and this particular parameter for whatever reason.

## Additional plots and comparisons (in case you find them helpful)

Plots of the fitted mean values and the gamma precision uncertainites (both in original units) for Years 1, 11, 21, and 31 (1979, 1989, 1999, 2009).
```{r}
fitted.idx = inla.stack.index(dat.stack, 'dat')$data

val.fitted = tibble(mu = mod$summary.fitted.values$mean[fitted.idx]) %>% 
  mutate(sd = sqrt(mu^2/mod$summary.hyperpar$mean[1])) %>% 
  bind_cols(dat.train %>% select(Year, East, North)) %>% 
  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))

ggplot(val.fitted %>% filter(Year %in% c(1,11,21,31))) + 
  geom_sf(aes(color=mu)) + scale_color_viridis_c() + facet_wrap(~ Year)

ggplot(val.fitted %>% filter(Year %in% c(1,11,21,31))) + 
  geom_sf(aes(color=sd)) + scale_color_viridis_c(option="plasma") + facet_wrap(~ Year)
```

Plots of the mean spatial random effect for accumulation and for the temporal trend in accumulation.
I am using the assumption that these values are on the log scale, and am therefore representing them transformed to a % increase/decrease in the expected accumulation.
```{r}
test.rand = tibble(trend.delta = exp(mod$summary.random$spYR.idx$mean) - 1, 
                   spat.mu = exp(mod$summary.random$spat.idx$mean) - 1, 
                   East = mesh$loc[,1], North = mesh$loc[,2]) %>% 
  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
ggplot(test.rand) + geom_sf(aes(color=spat.mu)) + 
  scale_color_gradientn(colours = cet_pal(5, name = "d1"))
ggplot(test.rand) + geom_sf(aes(color=trend.delta)) + 
  scale_color_continuous()
```

Plots of representative marginal distributions for the spatial random effect of individual select locations.
The selected locations represent the min/mid/max effects across the region.
These are again transformed to represent the % change in expected accumulation value due to the spatial random effect.
```{r}
marg.spat = t.marg(mod$marginals.random$spat.idx$index.205, function(x) exp(x)) %>% 
  mutate(Param = "idx.205")
marg.spat = marg.spat %>% bind_rows(t.marg(mod$marginals.random$spat.idx$index.1147, function(x) exp(x)) 
                                    %>% mutate(Param = "idx.1147")) %>% 
  bind_rows(t.marg(mod$marginals.random$spat.idx$index.214, function(x) exp(x)) 
            %>% mutate(Param = "idx.214"))
marg.spat = marg.spat %>% mutate(x = x-1)

ggplot(marg.spat, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param))  
```

And similar plots for selected years of the temporal random effect (1979, 1989, 1999, 2009).
```{r}
marg.time = t.marg(mod$marginals.random$time$index.1, function(x) exp(x)) %>% mutate(Param = "1979")
marg.time = marg.time %>% bind_rows(t.marg(mod$marginals.random$time$index.11, function(x) exp(x)) 
                                    %>% mutate(Param = "1989")) %>% 
  bind_rows(t.marg(mod$marginals.random$time$index.21, function(x) exp(x)) 
            %>% mutate(Param = "1999")) %>% 
  bind_rows(t.marg(mod$marginals.random$time$index.31, function(x) exp(x)) 
            %>% mutate(Param = "2009"))
marg.time = marg.time %>% mutate(x=x-1)

ggplot(marg.time, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param))  
```

Below is a plot of the temporal random effect and its estimated uncertainty ($\pm \sigma$) across the full time series.
```{r}
time.tbl = tibble(Year=yr.min+(1:nrow(mod$summary.random$time)), 
                  time.mu = mod$summary.random$time$mean, 
                   time.sd = mod$summary.random$time$sd)

ggplot(time.tbl, aes(x=Year, y=time.mu)) + 
  geom_line() + 
  geom_ribbon(aes(ymin=time.mu-time.sd,ymax=time.mu+time.sd), alpha=0.3) + 
  ylab("Magnitude of temporal random effect")
```

## Summary of questions

The following is just a list of the main questions I still have and would appreciate any info/feedback you might have on them.

1. Does the model I have devised seem reasonable, or are there (easy/quick) ways to better get out the information of interest?
2. Does the model as specified (particularly in regards to the spatial variation in time trends) actually calculate/return what I think it does?
3. Are my interpretations of model output/transformations correct?
4. Thoughts/suggestions regarding the very small mean values and comparatively very large uncertainties of $\delta_i$ (the spatial deviations in time trend coefficients) seen in the data?

### 2022-05-12 notebooks/compare-racmo.Rmd
Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

---
title: "Comparison of WAIS accumulation trends in PAIPR to RACMO"
output: 
  html_notebook:
    code_folding: hide
---

```{r}
library(here)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(sf)
library(cetcolor)
```

```{r}
dat = readRDS(here('data/Rdata-clean.rds'))
dat.gdf = readRDS(here('data/Rdata-gdf_trace.rds'))

# Remove rows with no RACMO SMB or PAIPR accum estimates
dat = dat %>% drop_na(SMB, accum)

# Remove high accumulation outliers
dat = dat %>% filter(accum <= 1000)

# Only select traces with full coverage 1979-2009
yr.cutoff = 2008
dat = dat %>% group_by(trace_ID) %>% filter(Year >= 1979) %>% filter(Year <= yr.cutoff) %>% 
  filter(all(1979:yr.cutoff %in% Year))

# Remove (now) missing traces from gdf
dat.gdf = dat.gdf %>% filter(trace_ID %in% dat$trace_ID)

# Add mean SMB results to gdf
dat.gdf = left_join(dat.gdf %>% select(-accum), 
                    dat %>% group_by(trace_ID) %>% summarise(accum=mean(accum), SMB=mean(SMB)))
dat.gdf = dat.gdf %>% mutate(accum.res = SMB-accum)

# Calculate residuals
dat = dat %>% mutate(res = SMB-accum)
```

Direct 1:1 plot of PAIPR to RACMO (averaged over all years).
```{r}
ggplot(dat.gdf) + geom_point(aes(x=accum, y=SMB), color='blue', alpha=0.25) + 
  geom_smooth(aes(x=accum, y=SMB), method = 'lm') + geom_abline(slope=1, color='red') + 
  lims(x=c(100,600), y=c(100,600))

ggplot(dat.gdf) + geom_sf(aes(color=accum.res)) + scale_color_gradient2()
```

```{r PAIPR-RACMO_plots}
# ggplot(dat) + geom_density(aes(x=res, group=Year, fill=Year), color=NA, alpha=0.2) + 
#   scale_fill_viridis_c() + 
#   geom_density(aes(x=res), color='black')

ggplot(dat, aes(x=factor(Year), y=res)) + 
  geom_hline(yintercept = 0) + 
  geom_violin(aes(fill=Year), position="dodge", alpha=0.5) + 
  stat_summary(fun=median, geom="point", size=1, color="red") + 
  scale_fill_viridis_c()
```

```{r}
accum.glm = dat %>% 
  mutate(Year = Year-mean(Year)) %>% 
  group_by(trace_ID) %>% 
  do(tidy(glm(accum ~ Year, family=Gamma(link='log'), data = .))) %>% 
  filter(term=='Year') %>% select(-term) %>% mutate(est.trans = exp(estimate)-1)
SMB.glm = dat %>% 
  mutate(Year = Year-mean(Year)) %>% 
  group_by(trace_ID) %>% 
  do(tidy(glm(SMB ~ Year, family=Gamma(link='log'), data = .))) %>% 
  filter(term=='Year') %>% select(-term) %>% mutate(est.trans = exp(estimate)-1)

dat.gdf = inner_join(dat.gdf, 
                     accum.glm %>% select(trace_ID, est.trans) %>% rename(estimate=est.trans), 
                     by="trace_ID", suffix=c(".raw", ".accum")) %>% 
  left_join(SMB.glm %>% select(trace_ID, est.trans) %>% rename(estimate=est.trans), 
            by="trace_ID", suffix=c(".accum", ".SMB"))
```

```{r}
ggplot(dat.gdf) + geom_density(aes(x=estimate.accum), fill='blue', alpha=0.5) + 
  geom_density(aes(x=estimate.SMB), fill='red', alpha=0.5)
```

```{r}
dat.gdf = dat.gdf %>% mutate(trend.res = estimate.accum-estimate.SMB)
ggplot(dat.gdf) + geom_density(aes(x=trend.res), fill='blue', alpha=0.5) + 
  geom_vline(xintercept = 0)

ggplot(dat.gdf) + 
  geom_sf(aes(color=trend.res))  + 
  # scale_color_gradientn(colours = cet_pal(7, name = "d1"))
  scale_color_gradient2()
```

```{r}
ggplot(dat %>% group_by(Year) %>% summarise(SMB=median(SMB), accum=median(accum))) + 
  geom_line(aes(x=Year, y=accum), color='blue') + 
  geom_line(aes(x=Year, y=SMB), color='red')

ggplot(dat %>% group_by(trace_ID)) + 
  geom_line(aes(x=Year, y=accum, group=trace_ID), color='blue', alpha=0.01) + 
  geom_line(aes(x=Year, y=SMB, group=trace_ID), color='red', alpha=0.01)
```

```{r}
ggplot(dat.gdf) + geom_sf(aes(color=trace_ID)) + scale_color_viridis_c()
```

```{r}
ri = dat$trace_ID[sample.int(nrow(dat),1)]
ggplot(dat %>% filter(trace_ID==ri)) + 
  geom_line(aes(x=Year, y=accum), color='blue') + 
  geom_line(aes(x=Year, y=SMB), color='red') + 
  ggtitle(label = ri)
```

### 2022-05-27 notebooks/compare-racmo.Rmd
Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

@@
 ggplot(dat.gdf) + 
+  geom_sf(aes(color=estimate.accum))  + 
+  # scale_color_gradientn(colours = cet_pal(7, name = "d1"))
+  scale_color_gradient2()
+
+ggplot(dat.gdf) + 
   geom_sf(aes(color=trend.res))  + 
   # scale_color_gradientn(colours = cet_pal(7, name = "d1"))
-  scale_color_gradient2()
+  scale_color_gradient2(limits=c(-0.025, 0.005))
 ```

### 2022-05-27 tutorials/sp-regress.Rmd
Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

---
title: "R Notebook"
output: html_notebook
---

This is a reproduction of the example in the [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in "Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA".
I do this as an understanding check prior to adapting and applying it to my data.

```{r}
library(INLA)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(spdep)
```

Define custom function (found in `spde-book-files/R/spde-book-functions.R`) for generated spatially correlated data.
```{r}
book.rMatern <- function(n, coords, sigma=1, range, kappa = sqrt(8*nu)/range, variance = sigma^2, nu=1) {
    m <- as.matrix(dist(coords))
    m <- exp((1-nu)*log(2) + nu*log(kappa*m)-
             lgamma(nu))*besselK(m*kappa, nu)
    diag(m) <- 1
    return(drop(crossprod(chol(variance*m),
                          matrix(rnorm(nrow(coords)*n), ncol=n))))
}
```

Generate random spatial locations
```{r}
n <- 150
set.seed(1)
coo <- matrix(runif(2 * n), n)
```

Generate spatially and temporally correlated data
```{r}
kappa <- c(10, 12)
sigma2 <- c(1 / 2, 1 / 4)
k <- 15
rho <- c(0.7, 0.5)

set.seed(2)
beta0 <- book.rMatern(k, coo, range = sqrt(8) / kappa[1],
  sigma = sqrt(sigma2[1]))

set.seed(3)
beta1 <- book.rMatern(k, coo, range = sqrt(8) / kappa[2],
  sigma = sqrt(sigma2[2]))
beta0[, 1] <- beta0[, 1] / (1 - rho[1]^2)
beta1[, 1] <- beta1[, 1] / (1 - rho[2]^2)

for (j in 2:k) {
  beta0[, j] <- beta0[, j - 1] * rho[1] + beta0[, j] *
    (1 - rho[1]^2)
  beta1[, j] <- beta1[, j - 1] * rho[2] + beta1[, j] *
    (1 - rho[2]^2)
}

set.seed(4)
# Simulate the covariate values
hh <- runif(n * k) 
mu.beta <- c(-5, 1)
taue <- 20

set.seed(5)
# Error in the observation
error <- rnorm(n * k, 0, sqrt(1 / taue))

# Dynamic regression part
y <- (mu.beta[1] + beta0) + (mu.beta[2] + beta1) * hh + 
  error

# DGK Model year as covariate
year.val = matrix(data=(1:k)/k, nrow=n, ncol=k, byrow=TRUE)
# y = (mu.beta[1] + beta0) + (mu.beta[2] + beta1) * year.val + 
#   error

# Arrange in tbl
dat = tibble(loc.idx=rep(1:n,k), E=rep(coo[,1],k), N=rep(coo[,2],k), Year=rep(1:k, each=n), y=c(y), hh=hh, yr.val = c(year.val))
```

Exploration of the data.
```{r}
h.trends = dat %>% 
  group_by(loc.idx) %>% 
  do(tidy(lm(y ~ hh, data = .))) %>% 
  filter(term=='hh') %>% select(-term)
dat.mu = dat %>% group_by(loc.idx) %>% 
  summarize(E=mean(E), N=mean(N), y=mean(y)) %>% 
  left_join(h.trends %>% select(loc.idx, estimate))

ggplot(dat.mu) + geom_point(aes(x=E, y=N, color=y)) + scale_color_viridis_c()
ggplot(dat.mu) + geom_point(aes(x=E, y=N, color=estimate)) + scale_color_gradient2()
ggplot(dat %>% group_by(loc.idx)) + geom_line(aes(x=Year, y=y, group=loc.idx), alpha=0.1)
```

```{r}
mesh <- inla.mesh.2d(dat %>% select(E,N), max.edge = c(0.15, 0.3),
  offset = c(0.05, 0.3), cutoff = 0.07)

plot(mesh)
points(dat %>% select(E, N), col = "red")

spde <- inla.spde2.pcmatern(mesh = mesh,
  prior.range = c(0.05, 0.01), # P(practic.range < 0.05) = 0.01
  prior.sigma = c(1, 0.01)) # P(sigma > 1) = 0.01

i0 <- inla.spde.make.index('i0', spde$n.spde, n.group = length(unique(dat$Year)))
i1 <- inla.spde.make.index('i1', spde$n.spde, n.group = length(unique(dat$Year)))
```

```{r}
A0 <- inla.spde.make.A(mesh, loc=cbind(dat$E, dat$N), group = dat$Year)
A1 <- inla.spde.make.A(mesh, loc=cbind(dat$E, dat$N), 
                       group = dat$Year, weights = dat$yr.val)

dat.stack <- inla.stack(
  data = list(y = dat$y), 
  A = list(A0, A1, 1), 
  effects = list(i0, 
                 i1, 
                 tibble(Intercept = 1, mu1 = 1, 
                        yr.val=dat$yr.val, h = dat$hh)),
  tag = 'dat')

form <- y ~ 0 + mu1 + yr.val + # to fit mu_beta
  f(i0, model = spde, group = i0.group, 
    control.group = list(model = 'ar1')) + 
  f(i1, model = spde, group = i1.group, 
    control.group = list(model = 'ar1'))

# Assigned initial starting values for parameter estimates
theta.ini <- c(
  log(taue), # likelihood log precision
  log(sqrt(8) / kappa[1]), # log range 1
  log(sqrt(sigma2[1])), # log stdev 1
  log((1 + rho[1])/(1 - rho[1])), # log trans. rho 1
  log(sqrt(8) / kappa[2]), # log range 1
  log(sqrt(sigma2[2])), # log stdev 1
  log((1 + rho[2]) / (1 - rho[2])))# log trans. rho 2
```

```{r model, cache=TRUE}
mod = inla(form, family = 'gaussian',
           data = inla.stack.data(dat.stack), 
           control.predictor = list(A = inla.stack.A(dat.stack)), 
           control.mode = list(theta = theta.ini, restart=TRUE), # initial theta value
           control.inla = list(int.strategy = 'eb'))# no integr. wrt theta)
```

Extract model results of interest
```{r}
# Get indices of observed data
fitted.idx = inla.stack.index(dat.stack, 'dat')$data

dat = dat %>% mutate(y.mod=mod$summary.fitted.values$mean[fitted.idx], 
                     i0.mu = mod$summary.random$i0$mean[fitted.idx], 
                     i1.mu = mod$summary.random$i1$mean[fitted.idx])
```

## Adaptations

I'm adapting the example to better align it with my particular data to see if I can still model and understand it.

Generate random spatial locations
```{r}
n <- 150
set.seed(1)
coo <- matrix(runif(2 * n), n)
```

Generate spatially and temporally correlated data.
The differences here with the original are (1) we are focused on the coefficient with respect to time (rather than `hh`) and (2) the time coefficient only varies spatially (with no need therefore of a temporal autocorrelation component for it). 
```{r}
kappa <- c(10, 12)
sigma2 <- c(0.75, 1.1)
k <- 15
rho <- c(0.7, 0.5)

set.seed(2)
beta0 <- book.rMatern(k, coo, range = sqrt(8) / kappa[1],
  sigma = sigma2[1])

set.seed(3)
beta1 <- book.rMatern(1, coo, range = sqrt(8) / kappa[2],
  sigma = sigma2[2])
beta0[, 1] <- beta0[, 1] / (1 - rho[1]^2)
# beta1[, 1] <- beta1[, 1] / (1 - rho[2]^2)

for (j in 2:k) {
  beta0[, j] <- beta0[, j - 1] * rho[1] + beta0[, j] *
    (1 - rho[1]^2)
  # beta1[, j] <- beta1[, j - 1] * rho[2] + beta1[, j] *
  #   (1 - rho[2]^2)
}

# Remove temporal dependence of beta1 through averaging (each year has same value for beta1)
# beta1 = as.vector(rowMeans(beta1))

set.seed(4)
# Simulate the covariate values
hh <- runif(n * k) 
mu.beta <- c(-5, 1)
taue <- 20

set.seed(5)
# Error in the observation
error <- rnorm(n * k, 0, sqrt(1 / taue))

# DGK Model year as covariate
Year = matrix(data=(1:k), nrow=n, ncol=k, byrow=TRUE)
y = (mu.beta[1] + beta0) + (mu.beta[2] + beta1) * Year +
  error
# year.val = matrix(data=(1:k)/k, nrow=n, ncol=k, byrow=TRUE)
# y = (mu.beta[1] + beta0) + (mu.beta[2] + beta1) * year.val +
#   error

# Arrange in tbl
dat = tibble(loc.idx=rep(1:n,k), E=rep(coo[,1],k), N=rep(coo[,2],k), 
             Year=rep(1:k, each=n), y=c(y), yr.val = c(year.val))
```

Exploration of the data.
```{r}
yr.trends = dat %>% 
  group_by(loc.idx) %>% 
  do(tidy(lm(y ~ Year, data = .))) %>% 
  filter(term=='Year') %>% select(-term)
dat.mu = dat %>% group_by(loc.idx) %>% 
  summarize(E=mean(E), N=mean(N), y=mean(y)) %>% 
  left_join(yr.trends %>% select(loc.idx, estimate))

ggplot(dat.mu) + geom_point(aes(x=E, y=N, color=y)) + scale_color_viridis_c()
ggplot(dat.mu) + geom_point(aes(x=E, y=N, color=estimate))
ggplot(dat %>% group_by(loc.idx)) + geom_line(aes(x=Year, y=y, group=loc.idx), alpha=0.1)
```

```{r}
mesh <- inla.mesh.2d(dat %>% select(E,N), max.edge = c(0.15, 0.3),
  offset = c(0.05, 0.3), cutoff = 0.07)

plot(mesh)
points(dat %>% select(E, N), col = "red")

spde <- inla.spde2.pcmatern(mesh = mesh,
  prior.range = c(0.05, 0.01), # P(practic.range < 0.05) = 0.01
  prior.sigma = c(1, 0.01)) # P(sigma > 1) = 0.01

i0 <- inla.spde.make.index('i0', spde$n.spde, n.group = length(unique(dat$Year)))
i1 <- inla.spde.make.index('i1', spde$n.spde)

A0 <- inla.spde.make.A(mesh, loc=cbind(dat$E, dat$N), group = dat$Year)
A1 <- inla.spde.make.A(mesh, loc=cbind(dat$E, dat$N), weights = dat$Year)

dat.stack <- inla.stack(
  data = list(y = dat$y), 
  A = list(A0, A1, 1), 
  effects = list(i0, 
                 i1, 
                 tibble(Intercept = 1, mu1 = 1, 
                        Year=dat$Year)),
  tag = 'dat')

form <- y ~ 0 + mu1 + Year + # to fit mu_beta
  f(i0, model = spde, group = i0.group, 
    control.group = list(model = 'ar1')) + 
  f(i1, model = spde)
```

```{r}
mod.2 = inla(form, family = 'gaussian',
           data = inla.stack.data(dat.stack), 
           control.predictor = list(A = inla.stack.A(dat.stack)), 
           control.inla = list(int.strategy = 'eb'), # no integr. wrt theta)
           control.compute = list(config=TRUE))
```

```{r}
# i=sample.int(max(dat$loc.idx), size=1)
# ggplot(dat %>% filter(loc.idx==i)) + geom_line(aes(x=Year, y=y))
```

Posterior checks on the linear coefficient (combined global and spatial components) for time.
```{r}
# Posterior draws for year coefficients
draws=500
n.locs = spde$n.spde
tmp = inla.posterior.sample(draws, result = mod.2)
tmp2 = inla.posterior.sample.eval(c("Year", "i1"), tmp)
tmp.glob = unname(tmp2[1,])
tmp.i = unname(tmp2[2:(n.locs+1),])
tmp.tot = tmp.glob + tmp.i

coeff.ppc = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp.tot))

ggplot(coeff.ppc %>% group_by(Sample)) + 
  geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
  geom_density(data=dat.mu, aes(x=estimate), color='red') 
```

### 2022-06-01 notebooks/kronecker-tests.Rmd
Adds notebook for experimenting with kronecker products for st modeling

---
title: "Testing spacetime model with Kronecker sums on WAIS data"
output: html_notebook
---

This attempts to model the time coefficients of accumulation data following the example found in the [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in "Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA".
I adapt this to my particular use case and will eventually further modify it to use a Gamma distribution with log link.

```{r}
library(here)
library(INLA)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(spdep)
```

```{r}
data = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))

set.seed(777)
gdf_traces = gdf_traces %>% sample_frac(0.01)

# # Subset to smaller region
# tmp = st_bbox(gdf_traces)
# East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[1] + (tmp[3]-tmp[1])/2)
# North = c(tmp[2] + (tmp[4]-tmp[2])/2, tmp[4] - (tmp[4]-tmp[2])/8)
# tmp.df = tibble(East=East, North=North)
# poly = tmp.df %>%
#   st_as_sf(coords = c("East", "North"),
#            crs = 3031) %>%
#   st_bbox() %>%
#   st_as_sfc()
# gdf.idx = sapply(st_intersects(gdf_traces, poly),function(x){length(x)==0})
# gdf_traces = gdf_traces[!gdf.idx,]

data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year < 2010) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, 
                      # u10m, v10m, mu.u10, mu.v10
                      )

# Center covariates
dat = dat %>% 
  mutate(elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
         ) %>% 
  mutate(Year.mod = Year-mean(Year), Year.idx = (Year-min(Year)+1))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), 
         # u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), mu.u10=sd(mu.u10), mu.v10=sd(mu.v10)
         )

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

```{r}
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(60000, 120000), cutoff = 12000)
# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
#                     max.edge = c(30000, 90000), cutoff = 1000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

# These values will also need to be adjusted when going back to Gamma?
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(20000, 0.01), # P(range < 15 km) = 0.01
  prior.sigma = c(100, 0.05))

st.idx <- inla.spde.make.index('st.idx', spde$n.spde, n.group = length(unique(dat.train$Year.idx)))
yr.idx <- inla.spde.make.index('yr.idx', spde$n.spde)

A.st <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         group = dat.train$Year.idx)
A.yr <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         weights = dat.train$Year.mod)

dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, A.st, A.yr),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        # slope=dat.train$slope, aspect=dat.train$aspect, 
                        Year=dat.train$Year.mod),
                 st.idx, 
                 yr.idx),
  tag = 'dat')

# Model formula
form = y ~ -1 + Intercept + #modeling the intercept
  # dem + aspect + #fixed effects
  f(st.idx, model = spde, group=st.idx.group, 
    control.group=list(model='ar1')) + #spatial random effect with AR1 temporal autocorrelation
  f(yr.idx, model = spde) #Global effect from year and deviation from global (modeled as a spatial random effect with weights given by Year)
```

## Gamma modeling

```{r mod-gamma, cache=TRUE}
# For testing, initially use gaussian

# mod.gamma = inla(form,
#                data = inla.stack.data(dat.stack),
#                family = 'Gamma',
#                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
#                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
#                control.compute = list(waic=TRUE, config=TRUE))
mod.gamma = readRDS(here('data/interim_results/mod-gamma.rds'))

summary(mod.gamma)
```

### Comparisons to GLM models

```{r}
f.glm = y ~ -1 + Intercept + Year

glm.gauss = glm(f.glm, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1, Year=Year.mod))
glm.logGaus = mod.glm = glm(f.glm, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1, Year=Year.mod), 
               family = gaussian(link = "log"))
glm.gamma = glm(f.glm, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1, Year=Year.mod), 
               family = Gamma(link = "log"))
summary(glm.gamma)
```

### Posterior checks

Directly calculate linear coefficients of time from raw data.
```{r}
yr.trends = dat.train %>% 
  group_by(trace_ID) %>% 
  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
  filter(term=='Year.mod') %>% select(-term)
dat.mu = dat.train %>% group_by(trace_ID) %>% 
  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
  mutate(log.est = log(1+(estimate/accum)))
```

```{r}
# Posterior draws for year coefficients
draws=100
n.locs = spde$n.spde
mod.samples = inla.posterior.sample(draws, result = mod.gamma)
tmp = inla.posterior.sample.eval(c("yr.idx"), mod.samples)
coeff.ppc = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))

ggplot(coeff.ppc %>% group_by(Sample)) + 
  geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
  geom_density(data=dat.mu, aes(x=log.est), color='red') + 
  xlim(c(-0.03, .03))
```

### Spatial comparisons

```{r}
tmp = tibble(Source="model", 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.gamma$summary.random$yr.idx$`0.5quant`, 
             LB=mod.gamma$summary.random$yr.idx$`0.025quant`, 
             UB=mod.gamma$summary.random$yr.idx$`0.975quant`) %>% 
  # mutate(weak=(LB<0 && UB>0)) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

trends.sf = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% mutate(LB=NA, UB=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)

ggplot(trends.sf %>% filter(Source=="data")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.sf %>% filter(Source=="model")) + 
  geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.sf %>% filter(Source=="model")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.sf %>% filter(Source=="model")) + 
  geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
```
## Spacetime random effect removed

It's possible that I am removing temporal signal from my estimates when I model the spacetime auto-correlations.
Here I run the model without this random effect for comparison (so only using a spatially varying temporal coefficient).

```{r}
# Model formula
f.noST = y ~ -1 + Intercept + 
  # f(st.idx, model = spde, group=st.idx.group, 
  #   control.group=list(model='ar1')) + 
  f(yr.idx, model = spde)

mod.noST = inla(f.noST,
                data = inla.stack.data(dat.stack),
                family = 'Gamma',
                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
                control.compute = list(waic=TRUE, config=TRUE))

summary(mod.noST)
```

### Posterior checks

```{r}
# Posterior draws for year coefficients
draws=100
n.locs = spde$n.spde
samples.noST = inla.posterior.sample(draws, result = mod.noST)
tmp = inla.posterior.sample.eval(c("yr.idx"), samples.noST)
ppc.noST = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))

ggplot(ppc.noST %>% group_by(Sample)) + 
  geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
  geom_density(data=dat.mu, aes(x=log.est), color='red') + 
  xlim(c(-0.03, .03))
```

### Spatial comparisons

```{r}
tmp = tibble(Source="model", 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.noST$summary.random$yr.idx$`0.5quant`, 
             LB=mod.noST$summary.random$yr.idx$`0.025quant`, 
             UB=mod.noST$summary.random$yr.idx$`0.975quant`) %>% 
  # mutate(weak=(LB<0 && UB>0)) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

trends.noST = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% mutate(LB=NA, UB=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)

ggplot(trends.noST %>% filter(Source=="data")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
ggplot(trends.noST %>% filter(Source=="model")) + 
  geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
```

## Lognormal modeling

```{r mod-gauss, cache=TRUE}

# stack.LN <- inla.stack(
#   data = list(y = log(dat.train$accum)), 
#   A = list(1, 1, A.st, A.yr),
#   effects = list(list(Intercept = rep(1, nrow(dat.train))), 
#                  tibble(elev = dat.train$elev, dem=dat.train$dem, 
#                         # slope=dat.train$slope, aspect=dat.train$aspect, 
#                         Year=dat.train$Year.mod),
#                  st.idx, 
#                  yr.idx),
#   tag = 'dat')
# 
# # For testing, initially use gaussian
# mod.gauss = inla(form, 
#                data = inla.stack.data(stack.LN), 
#                family = 'Gaussian',
#                control.predictor = list(compute = TRUE, A = inla.stack.A(stack.LN)), 
#                control.compute = list(waic=TRUE, config=TRUE))
# 
# # mod.gauss = inla(form, 
# #                  data = inla.stack.data(dat.stack), 
# #                  )
# 
# summary(mod.gauss)
```

### 2022-06-01 tutorials/sp-regress.Rmd
Adds notebook for experimenting with kronecker products for st modeling

@@
 hh <- runif(n * k) 
-mu.beta <- c(-5, 1)
+mu.beta <- c(400, -1)
 taue <- 20
@@
 dat = tibble(loc.idx=rep(1:n,k), E=rep(coo[,1],k), N=rep(coo[,2],k), 
-             Year=rep(1:k, each=n), y=c(y), yr.val = c(year.val))
+             Year=rep(1:k, each=n), y=c(y))
 ```
@@
   offset = c(0.05, 0.3), cutoff = 0.07)
+# mesh <- inla.mesh.2d(dat %>% select(E,N), max.edge = c(0.25, 0.5),
+#   offset = c(0.05, 0.3), cutoff = 0.15)
 
@@
 dat.stack <- inla.stack(
-  data = list(y = dat$y), 
+  # data = list(y = dat$y), 
+  data = list(y = log(dat$y)),
   A = list(A0, A1, 1), 
@@
            data = inla.stack.data(dat.stack), 
+           # control.family = list(control.link=list(model='log')), 
            control.predictor = list(A = inla.stack.A(dat.stack)), 
@@
 ```{r}
+summary(mod.2)
 # i=sample.int(max(dat$loc.idx), size=1)

### 2022-07-11 notebooks/reports/summer_20220608.Rmd
Adds notebooks for comparisons to kronecker product and explanation to Summer

---
title: "08 June Update"
output: html_notebook
---

## Modeling

```{r}
library(here)
library(INLA)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(spdep)
```
### Data imports and preprocessing

```{r}
data = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))

# Keep only records that extend to at least a given year
min.yr = 1979
keep.trace = data %>% group_by(trace_ID) %>% summarise(yr.start=min(Year)) %>% 
  filter(yr.start <= min.yr)
gdf_traces = gdf_traces %>% filter(trace_ID %in% keep.trace$trace_ID)

# Subset to 1-km spacing (improved computation and convergence)
skip.int = 25
gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
gdf_traces = gdf_traces[gdf.idx,]

# Match filtering performed on traces
data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
  arrange(trace_ID, Year)

# TMP: Filter out data prior to 1979 (improves computational speed)
data = data %>% filter(Year >= min.yr)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, SMB, elev, dem, slope, aspect, 
                      # u10m, v10m, mu.u10, mu.v10
                      )

# Center covariates
dat = dat %>% 
  mutate(elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
         ) %>% 
  mutate(Year.mod = Year-mean(Year), Year.idx = (Year-min(Year)+1))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), 
         # u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), mu.u10=sd(mu.u10), mu.v10=sd(mu.v10)
         )

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

Directly calculate linear coefficients of time from raw data.
```{r}
yr.trends = dat.train %>% 
  group_by(trace_ID) %>% 
  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
  filter(term=='Year.mod') %>% select(-term)
dat.mu = dat.train %>% group_by(trace_ID) %>% 
  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
  mutate(log.est = log(1+(estimate/accum)))
```

### Model setup

```{r}
# Define mesh
# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
#                     max.edge = c(18000, 60000), cutoff = 2500)
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
                    max.edge = c(50000, 125000), cutoff = 10000)

plot(mesh)
points(dat.train %>% select(East, North), col = "red")

# These values will also need to be adjusted when going back to Gamma?
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(50000, 0.01), # P(range < 50 km) = 0.01
  prior.sigma = c(2, 0.05))

# Index and projector matrix for spatial random effect
spat.idx = inla.spde.make.index('spat.idx', spde$n.spde)
A.spat = inla.spde.make.A(mesh, 
                          loc=cbind(dat.train$East, dat.train$North))

# Index and projector matrix for spatially varying temporal trend
yr.idx <- inla.spde.make.index('yr.idx', spde$n.spde)
A.yr <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         weights = dat.train$Year.mod)

# Make data stack
stack.sep <- inla.stack(
  data = list(y = dat.train$accum), 
  # A = list(1, 1, 1, A.yr),
  A = list(1, 1, 1, A.spat, A.yr),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        Year=dat.train$Year.mod),
                 list(time = dat.train$Year.mod),
                 spat.idx, 
                 yr.idx),
  tag = 'dat')

# Priors for temporal autocorrelation
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))

# Model formula
f.sep = y ~ -1 + Intercept + 
  dem + 
  f(spat.idx, model = spde) +
  f(time, model = 'ar1', hyper = time.spec) +  #Temporal random effect modeled as AR1
  f(yr.idx, model = spde)
```

```{r}
mod.sep = inla(f.sep,
                data = inla.stack.data(stack.sep),
                family = 'Gamma',
                control.predictor = list(compute = TRUE, A = inla.stack.A(stack.sep), link=1),
                control.inla = list(int.strategy='ccd', strategy='simplified.laplace'), 
                control.compute = list(waic=TRUE, config=TRUE))

summary(mod.sep)
```

```{r}
f.glm = y ~ -1 + Intercept + Year + dem
mod.glm = glm(f.glm, 
                data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1, Year=Year.mod), 
               family = Gamma(link = "log"))
summary(mod.glm)
```

## Model results

### Posterior summaries

```{r}
t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {

  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }

  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }

    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }

  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))

  marg.names = names(mod.marg)

  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod.sep$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", 
             "Range for spat.idx", "Stdev for spat.idx", 
             "Time precision", "Rho for time", 
             "Range for yr.idx", "Stdev for yr.idx")
marg.hyper = t.marg(mod.sep$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "dem", "Gamma precision", 
                "Range for spat.idx", "Stdev for spat.idx", 
                "Time precision", "Rho for time", 
                "Range for yr.idx", "Stdev for yr.idx")
tmp = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r fig, fig.width=16}
ggplot(tmp, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

## Investigating the components of prediction

```{r}

pred.trace = gdf_traces[seq(1,nrow(gdf_traces),3),] %>% mutate(color=as.factor(trace_ID))
dat.pred = dat %>% filter(trace_ID %in% pred.trace$trace_ID) %>% mutate(color=as.factor(trace_ID))

Years = sort(unique(dat.train$Year))
loc.pred = st_coordinates(pred.trace) %x% rep(1, length(Years))
covar.pred = dat.pred %>% group_by(trace_ID) %>% summarize_all(mean, na.rm=TRUE)
yr.mod = rep(Years - mean(Years), nrow(pred.trace))
st.pred = tibble(X=loc.pred[,1], Y=loc.pred[,2], Year = rep(Years, nrow(pred.trace)), dem=rep(covar.pred$dem, each=length(Years)))

proj.pred = inla.mesh.projector(mesh=mesh, loc = loc.pred)
spat.RE = inla.mesh.project(proj.pred, mod.sep$summary.random$spat.idx$mean)
time.RE = rep(mod.sep$summary.random$time$mean, nrow(pred.trace))
trnd.RE = inla.mesh.project(proj.pred, mod.sep$summary.random$yr.idx$mean)

# Construct linear predictor for each spacetime point
lin.est = mod.sep$summary.fixed$mean[1] + 
  mod.sep$summary.fixed$mean[2]*st.pred$dem + 
  spat.RE + time.RE + 
  trnd.RE*yr.mod

st.pred = st.pred %>% 
  mutate(fx.RE=mod.sep$summary.fixed$mean[2]*st.pred$dem, spat.RE=spat.RE, 
         time.RE=time.RE, trnd.RE=trnd.RE*yr.mod, pred=exp(lin.est)) %>% 
  mutate(pred.sd=sqrt(pred^2/mod.sep$summary.hyperpar$mean[1]), 
         color=rep(pred.trace$color, each=length(Years)))

spat.pred = st.pred %>% group_by(color) %>% summarize_all(mean, rm.na=TRUE)
spat.pred[,'trend'] = trnd.RE[seq(1,length(trnd.RE), by=37)]
ggplot(spat.pred) + geom_point(aes(x=X, y=Y, color=pred)) + scale_color_viridis_c()
ggplot(spat.pred) + geom_point(aes(x=X, y=Y, color=fx.RE)) + scale_color_gradient2()
ggplot(spat.pred) + geom_point(aes(x=X, y=Y, color=spat.RE)) + scale_color_gradient2()
ggplot(spat.pred) + geom_point(aes(x=X, y=Y, color=trend)) + scale_color_gradient2()
```

```{r}
mod.time = tibble(Year=Years, mu=mod.sep$summary.random$time$mean, sd=mod.sep$summary.random$time$sd)
# ggplot(st.pred %>% group_by(Year) %>% summarise(time.RE=mean(time.RE))) + 
#   geom_line(aes(x=Year, y=time.RE))
ggplot(mod.time) + geom_hline(yintercept = 0, color='black') + geom_ribbon(aes(x=Year, ymin=mu-sd, ymax=mu+sd), fill='red', alpha=0.2) + 
  geom_line(aes(x=Year, y=mu), color='red')
```

```{r}
# pts.IDs = c(2500, 7798, 10298, 12814, 16053, 18553)
pts.IDs = sample(unique(st.pred$color), 3)
pts.trace = pred.trace %>% filter(color %in% pts.IDs)
dat.pts = dat.pred %>% filter(color %in% pts.IDs)
st.pts = st.pred %>% filter(color %in% pts.IDs)

ggplot(pts.trace) + geom_sf(data=gdf_traces, color='black') + geom_sf(aes(color=color)) + scale_color_discrete()
ggplot(st.pts) + geom_line(data=dat.pts, aes(x=Year, y=accum, group=color, color=color), size=0.5, linetype="dashed") + 
  geom_ribbon(aes(x=Year, ymin=pred-pred.sd, ymax=pred+pred.sd, group=color, fill=color), alpha=0.25)
  # geom_line(aes(x=Year, y=pred, group=color, color=color), linetype="dashed")
```

### Modeled time trends

Calculate probability that trend is less than zero.
```{r}
P.neg = vector(mode = "numeric", length = length(mod.sep$marginals.random$yr.idx))
for (i in 1:length(P.neg)) {
  marg.i = inla.tmarginal(function(x) x, 
                          mod.sep$marginals.random$yr.idx[[i]], 
                          n=500)
  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
}
```

Map of estimated trends with probability of less than zero
```{r}
tmp = tibble(Source="model", 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             spat.med=mod.sep$summary.random$spat.idx$`0.5quant`, 
             med=mod.sep$summary.random$yr.idx$`0.5quant`, 
             P.neg=P.neg, 
             LB=mod.sep$summary.random$yr.idx$`0.025quant`, 
             UB=mod.sep$summary.random$yr.idx$`0.975quant`) %>% 
  # mutate(weak=(LB<0 && UB>0)) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

trends.sep = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% 
  mutate(spat.med=NA, LB=NA, UB=NA, P.neg=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)
```

```{r}
ggplot(trends.sep) + geom_density(aes(x=med, group=Source, color=Source))
ggplot(trends.sep %>% filter(Source=="data")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
ggplot(trends.sep %>% filter(Source=="model")) +
  geom_sf(aes(color=spat.med)) + scale_color_gradient2()
ggplot(trends.sep %>% filter(Source=="model")) + 
  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
ggplot(trends.sep %>% filter(Source=="model")) +
  geom_sf(aes(color=(UB-LB)/2)) + 
  scale_color_viridis_c(option = "plasma")
# Change to log scale

ggplot(trends.sep %>% filter(Source=='model')) + 
  geom_sf(aes(color=P.neg)) + scale_color_viridis_c()
ggplot(trends.sep %>% filter(Source=="model")) + geom_density(aes(x=P.neg))
```
Probability of getting an erroneous trend when drilling an ice core.

TODO LIST:

- Keep exploring/checking the data
- Think through what is the story that I want to have in this paper (what's the through narrative?)
- Is there a clean way to state probability of being wrong with an ice core (analogous to Bayesian AB testing)
- What do I sacrifice if I don't include reanalysis comparisons?

Extract marginal distributions for subset of coefficients (for plotting purposes)
```{r}
n=500
idx.set = as.integer(seq(1,spde$n.spde, length.out=n))
marg.set = mod.sep$marginals.random$yr.idx[idx.set]

# Create tbl of marginals
marg.tbl = tibble()
for (i in 1:length(idx.set)) {
  set.i = marg.set[[i]]
  tbl.i = tibble(Index = idx.set[i], X=set.i[,1], Y=set.i[,2])
  marg.tbl = marg.tbl %>% bind_rows(tbl.i)
}

ggplot(marg.tbl) + geom_line(aes(x=X, y=Y, group=Index), color='black', alpha=0.01)
```

Compare raw coefficients with fitted ones
```{r}
dat.obs = trends.sep %>% filter(Source=="data")
dat.mod = trends.sep %>% filter(Source=="model")

NN.idx = st_nearest_feature(dat.obs, dat.mod)
trend.comp = dat.obs %>% select(geometry) %>% 
  mutate(T.obs=dat.obs$med, T.mod=dat.mod$med[NN.idx], P.neg=dat.mod$P.neg[NN.idx],  
         LB=dat.mod$LB[NN.idx], UB=dat.mod$UB[NN.idx])

ggplot(trend.comp) + geom_density(aes(x=T.mod-T.obs)) + xlim(c(-0.05,0.05))
ggplot(trend.comp) + geom_point(aes(x=T.obs, y=T.mod)) + geom_abline(slope=1, color='red') + 
    lims(x=c(-0.1,0.1), y=c(-0.07,0.07))
ggplot(trend.comp) + geom_sf(aes(color=T.mod-T.obs)) + scale_color_gradient2(limits=c(-0.1,0.1))
```

```{r}
# yr.start = min(dat.train$Year)-1
time.tbl = tibble(Year=1979:2009, med=mod.sep$summary.random$time$`0.5quant`, 
                  LB=mod.sep$summary.random$time$`0.025quant`, 
                  UB=mod.sep$summary.random$time$`0.975quant`)

ggplot(time.tbl, aes(x=Year, y=med)) + 
  geom_line(linetype="dashed") + 
  geom_ribbon(aes(ymin=LB, ymax=UB), alpha=0.3) + 
  ylab("Magnitude of temporal random effect")
```

## Comparisons with RACMO

Calculate (log) trends in RACMO
```{r}
dat.racmo = dat %>% 
  select(trace_ID, East, North, Year, accum, SMB) %>% drop_na(SMB, accum)
yr.cutoff = 2008
dat.racmo = dat.racmo %>% group_by(trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year <= yr.cutoff) %>% 
  filter(all(1979:yr.cutoff %in% Year)) %>% 
  mutate(Year.mod = Year-mean(Year))

tmp = dat.racmo %>% 
  group_by(trace_ID) %>% 
  do(tidy(lm(SMB ~ Year.mod, data = .))) %>% 
  filter(term=='Year.mod') %>% select(-term)
racmo.gdf = dat.racmo %>% group_by(trace_ID) %>% 
  summarize(East=mean(East), North=mean(North), accum=mean(accum), SMB=mean(SMB)) %>% 
  left_join(tmp %>% select(trace_ID, estimate)) %>% 
  mutate(log.est = log(1+(estimate/SMB))) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

NN.idx = st_nearest_feature(racmo.gdf, dat.mod)
```

### 2022-07-11 scratch/mod-tests.Rmd
Adds notebook for comparing separable models with and without extra spatial random variable

---
title: "Model comparisons between separable and kroenecker products"
output: html_notebook
---

```{r}
library(here)
library(INLA)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(spdep)
```
### Data imports and preprocessing

```{r}
data = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))

# # Clip bounds
# bbox = st_bbox(gdf_traces)
# clipper = c((bbox[1]+0.3*(bbox[3]-bbox[1])), 
#             (bbox[2]+0.5*(bbox[4]-bbox[2])), 
#             (bbox[3]-0.2*(bbox[3]-bbox[1])), 
#             bbox[4])
# gdf_traces = st_crop(gdf_traces, clipper)

skip.int = 25
gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
gdf_traces = gdf_traces[gdf.idx,]

data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
  filter(Year >= 1975) %>% filter(Year < 2015) %>% arrange(trace_ID, Year)

# Remove gdf rows where all data have been filtered out
gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, SMB, elev, dem, slope, aspect, 
                      # u10m, v10m, mu.u10, mu.v10
                      )

# Center covariates
dat = dat %>% 
  mutate(elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
         ) %>% 
  mutate(Year.mod = Year-mean(Year), Year.idx = (Year-min(Year)+1))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), 
         # u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), mu.u10=sd(mu.u10), mu.v10=sd(mu.v10)
         )

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

Directly calculate linear coefficients of time from raw data.
```{r}
yr.trends = dat.train %>% 
  group_by(trace_ID) %>% 
  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
  filter(term=='Year.mod') %>% select(-term)
dat.mu = dat.train %>% group_by(trace_ID) %>% 
  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
  mutate(log.est = log(1+(estimate/accum)))
```

### Model setup

```{r}
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
                    max.edge = c(60000, 120000), cutoff = 10000)
# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
#                     max.edge = c(30000, 90000), cutoff = 2000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

# These values will also need to be adjusted when going back to Gamma?
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(20000, 0.01), # P(range < 20 km) = 0.01
  prior.sigma = c(10, 0.05))
```

## Separable formulation

```{r}

spat.idx = inla.spde.make.index('spat.idx', spde$n.spde)
A.spat = inla.spde.make.A(mesh, 
                          loc=cbind(dat.train$East, dat.train$North))

yr.idx <- inla.spde.make.index('yr.idx', spde$n.spde)
A.yr <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         weights = dat.train$Year.mod)

# Make data stack
stack.sep <- inla.stack(
  data = list(y = dat.train$accum), 
  # A = list(1, 1, 1, A.yr),
  A = list(1, 1, 1, A.spat, A.yr),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        Year=dat.train$Year.mod),
                 list(time = dat.train$Year.mod),
                 spat.idx, 
                 yr.idx),
  tag = 'dat')

# Priors for temporal autocorrelation
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))

# Model formula
f.sep = y ~ -1 + Intercept + 
  dem + 
  f(spat.idx, model = spde) + 
  f(time, model = 'ar1', hyper = time.spec) +  #Temporal random effect modeled as AR1
  f(yr.idx, model = spde)
```

## Run models

```{r}
mod.sep = inla(f.sep,
                data = inla.stack.data(stack.sep),
                family = 'Gamma',
                control.predictor = list(compute = TRUE, A = inla.stack.A(stack.sep), link=1),
                control.inla = list(int.strategy='eb'), # Option improve computation time at cost of accuracy/precision
                control.compute = list(waic=TRUE, config=TRUE))
```

```{r}
# Model formula
f.yridx = y ~ -1 + Intercept + 
  dem + 
  f(time, model = 'ar1', hyper = time.spec) +  #Temporal random effect modeled as AR1
  f(yr.idx, model = spde)

mod.yridx = inla(f.yridx,
                 data = inla.stack.data(stack.sep),
                 family = 'Gamma',
                 control.predictor = list(compute = TRUE, A = inla.stack.A(stack.sep), link=1),
                 control.inla = list(int.strategy='eb'), # Option improve computation time at cost of accuracy/precision
                 control.compute = list(waic=TRUE, config=TRUE))
```

```{r}
summary(mod.sep)
```

```{r}
summary(mod.yridx)
```
## Kroenecker formulation

```{r}
# st.idx <- inla.spde.make.index('st.idx', spde$n.spde, 
#                                n.group = length(unique(dat.train$Year.idx)))
# A.st <- inla.spde.make.A(mesh,
#                          loc=cbind(dat.train$East, dat.train$North),
#                          group = dat.train$Year.idx)
# 
# yr.idx <- inla.spde.make.index('yr.idx', spde$n.spde)
# A.yr <- inla.spde.make.A(mesh, 
#                          loc=cbind(dat.train$East, dat.train$North), 
#                          weights = dat.train$Year.mod)
# 
# # Make data stack
# stack.kroen = inla.stack(
#   data = list(y = dat.train$accum), 
#   # A = list(1, 1, 1, A.yr),
#   A = list(1, 1, A.st, A.yr),
#   effects = list(list(Intercept = rep(1, nrow(dat.train))), 
#                  tibble(elev = dat.train$elev, dem=dat.train$dem, 
#                         slope=dat.train$slope, aspect=dat.train$aspect, 
#                         Year=dat.train$Year.mod),
#                  st.idx, 
#                  yr.idx),
#   tag = 'dat')
# 
# # Priors for temporal autocorrelation
# time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
# 
# # Model formula
# f.kroen = y ~ -1 + Intercept + 
#   dem + 
#   f(st.idx, model = spde, group=st.idx.group, 
#     control.group=list(model='ar1', hyper=time.spec)) +
#   f(yr.idx, model = spde)
```

```{r}
# mod.kroen = inla(f.kroen,
#                 data = inla.stack.data(stack.kroen),
#                 family = 'Gamma',
#                 control.predictor = list(compute = TRUE, A = inla.stack.A(stack.kroen), link=1),
#                 control.inla = list(int.strategy='eb'), # Option improve computation time at cost of accuracy/precision
#                 control.compute = list(waic=TRUE, config=TRUE))
# 
# summary(mod.kroen)
```

## Model results

```{r}
t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {

  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }

  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }

    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }

  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))

  marg.names = names(mod.marg)

  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod.sep$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", 
             "Range for spat.idx", "Stdev for spat.idx", 
             "Time precision", "Rho for time", 
             "Range for yr.idx", "Stdev for yr.idx")
marg.hyper = t.marg(mod.sep$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "dem", "Gamma precision", 
                "Range for spat.idx", "Stdev for spat.idx", 
                "Time precision", "Rho for time", 
                "Range for yr.idx", "Stdev for yr.idx")

post.sep = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r post-sep, fig.width=16}
ggplot(post.sep, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

Posteriors for model with no extra spatial random effect.
```{r}
marg.fixed = t.marg(mod.yridx$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", 
             "Time precision", "Rho for time", 
             "Range for yr.idx", "Stdev for yr.idx")
marg.hyper = t.marg(mod.yridx$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "dem", "Gamma precision", 
                "Time precision", "Rho for time", 
                "Range for yr.idx", "Stdev for yr.idx")

post.yridx = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r post-yridx, fig.width=16}
ggplot(post.yridx, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

## Investigating the components of prediction

```{r}
pred.trace = gdf_traces[seq(1,nrow(gdf_traces),3),] %>% mutate(color=as.factor(trace_ID))
dat.pred = dat %>% filter(trace_ID %in% pred.trace$trace_ID) %>% mutate(color=as.factor(trace_ID))

Years = sort(unique(dat.train$Year))
loc.pred = st_coordinates(pred.trace) %x% rep(1, length(Years))
covar.pred = dat.pred %>% select(trace_ID, dem) %>%
  group_by(trace_ID) %>% filter(row_number()==1)
yr.mod = rep(Years - mean(Years), nrow(pred.trace))
st.pred = tibble(X=loc.pred[,1], Y=loc.pred[,2], Year = rep(Years, nrow(pred.trace)), 
                 dem=rep(covar.pred$dem, each=length(Years)))

proj.pred = inla.mesh.projector(mesh=mesh, loc = loc.pred)
```

```{r}
spat.RE_sep = inla.mesh.project(proj.pred, mod.sep$summary.random$spat.idx$mean)
time.RE_sep = rep(mod.sep$summary.random$time$mean, nrow(pred.trace))
trnd.RE_sep = inla.mesh.project(proj.pred, mod.sep$summary.random$yr.idx$mean)

# Construct linear predictor for each spacetime point
lin.est_sep = mod.sep$summary.fixed$mean[1] + 
  mod.sep$summary.fixed$mean[2]*st.pred$dem + 
  spat.RE_sep + time.RE_sep + 
  trnd.RE_sep*yr.mod

pred.sep = st.pred %>% 
  mutate(fx.RE=mod.sep$summary.fixed$mean[2]*st.pred$dem, spat.RE=spat.RE_sep, 
         time.RE=time.RE_sep, trnd.RE=trnd.RE_sep*yr.mod, pred=exp(lin.est_sep)) %>% 
  mutate(pred.sd=sqrt(pred^2/mod.sep$summary.hyperpar$mean[1]), 
         Model='sep', 
         color=rep(pred.trace$color, each=length(Years)))
```

```{r}
time.RE_yr = rep(mod.yridx$summary.random$time$mean, nrow(pred.trace))
trnd.RE_yr = inla.mesh.project(proj.pred, mod.yridx$summary.random$yr.idx$mean)

# Construct linear predictor for each spacetime point
lin.est_yr = mod.yridx$summary.fixed$mean[1] + 
  mod.yridx$summary.fixed$mean[2]*st.pred$dem + 
  time.RE_yr + 
  trnd.RE_yr*yr.mod

pred.yr = st.pred %>% 
  mutate(fx.RE=mod.yridx$summary.fixed$mean[2]*st.pred$dem, spat.RE=NA, 
         time.RE=time.RE_yr, trnd.RE=trnd.RE_yr*yr.mod, pred=exp(lin.est_yr)) %>% 
  mutate(pred.sd=sqrt(pred^2/mod.yridx$summary.hyperpar$mean[1]), 
         Model='yr.idx', 
         color=rep(pred.trace$color, each=length(Years)))

pred.comb = pred.sep %>% bind_rows(pred.yr)

spat.pred = pred.comb %>% group_by(Model, color) %>% summarize_all(mean, rm.na=TRUE)
spat.pred[1:(nrow(spat.pred)/2),'trend'] = trnd.RE_sep[seq(1,length(trnd.RE_sep), 
                                      by=(length(trnd.RE_sep)/(nrow(spat.pred)/2)))]
spat.pred[(1+nrow(spat.pred)/2):nrow(spat.pred),'trend'] = 
  trnd.RE_yr[seq(1,length(trnd.RE_yr), 
                  by=(length(trnd.RE_yr)/(nrow(spat.pred)/2)))]
```

```{r}
ggplot(spat.pred, aes(x=X,y=Y, color=pred)) + 
  geom_point() + scale_color_viridis_c() + 
  facet_wrap(vars(Model))
ggplot(spat.pred, aes(x=X,y=Y, color=fx.RE)) + 
  geom_point() + scale_color_gradient2() + 
  facet_wrap(vars(Model))
ggplot(spat.pred %>% filter(Model=="sep")) + 
  geom_point(aes(x=X, y=Y, color=spat.RE)) + scale_color_gradient2()
ggplot(spat.pred, aes(x=X,y=Y, color=trend)) + 
  geom_point() + scale_color_gradient2() + 
  facet_wrap(vars(Model))
```

```{r}
mod.time = tibble(Year=Years, mu=mod.sep$summary.random$time$mean, 
                  sd=mod.sep$summary.random$time$sd, Model='sep')
tmp.time = tibble(Year=Years, mu=mod.yridx$summary.random$time$mean, 
                  sd=mod.yridx$summary.random$time$sd, Model='yr.idx')
mod.time = mod.time %>% bind_rows(tmp.time)
# ggplot(st.pred %>% group_by(Year) %>% summarise(time.RE=mean(time.RE))) + 
#   geom_line(aes(x=Year, y=time.RE))
ggplot(mod.time) + geom_hline(yintercept = 0, color='black') + 
  geom_ribbon(aes(x=Year, ymin=mu-sd, ymax=mu+sd, group=Model, fill=Model), alpha=0.2) + 
  geom_line(aes(x=Year, y=mu, group=Model, color=Model))
```

Calculate probability that trend is less than zero.
```{r}
P.neg1 = vector(mode = "numeric", length = length(mod.sep$marginals.random$yr.idx))
for (i in 1:length(P.neg1)) {
  marg.i = inla.tmarginal(function(x) x, 
                          mod.sep$marginals.random$yr.idx[[i]], 
                          n=500)
  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
  P.neg1[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
}
tmp1 = tibble(Model='sep', 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.sep$summary.random$yr.idx$`0.5quant`, 
             P.neg=P.neg1, 
             LB=mod.sep$summary.random$yr.idx$`0.025quant`, 
             UB=mod.sep$summary.random$yr.idx$`0.975quant`) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

P.neg2 = vector(mode = "numeric", length = length(mod.yridx$marginals.random$yr.idx))
for (i in 1:length(P.neg2)) {
  marg.i = inla.tmarginal(function(x) x, 
                          mod.yridx$marginals.random$yr.idx[[i]], 
                          n=500)
  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
  P.neg2[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
}
tmp2 = tibble(Model="yr.idx",
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod.yridx$summary.random$yr.idx$`0.5quant`, 
             P.neg=P.neg2, 
             LB=mod.yridx$summary.random$yr.idx$`0.025quant`, 
             UB=mod.yridx$summary.random$yr.idx$`0.975quant`) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)

neg.tmp = tmp1 %>% bind_rows(tmp2)
```

Map of estimated trends with probability of less than zero
```{r}
trends.comp = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% 
  mutate(LB=NA, UB=NA, P.neg=NA, Model="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(neg.tmp)
```

```{r}
ggplot(trends.comp) + geom_density(aes(x=med, group=Model, color=Model)) + 
  xlim(c(-0.1,0.1))

ggplot(trends.comp, aes(color=med)) + 
  geom_sf() + scale_color_gradient2(limits=c(-0.05, 0.05)) + 
  facet_wrap(vars(Model))

ggplot(trends.comp %>% filter(Model!="data")) +
  geom_sf(aes(color=(UB-LB)/2)) + 
  scale_color_viridis_c(option = "plasma", trans="log") + 
  facet_wrap(vars(Model))

ggplot(trends.comp %>% filter(Model!="data")) +
  geom_sf(aes(color=P.neg)) + 
  scale_color_viridis_c() + 
  facet_wrap(vars(Model))

ggplot(trends.comp %>% filter(Model!="data")) + 
  geom_density(aes(x=P.neg, group=Model, color=Model))
```

### 2022-07-11 notebooks/2mod-solution.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

--- 2022-05-12 notebooks/reports/phil_20220506.Rmd
@@
 ---
-title: "WAIS Accum Modeling in INLA"
-output: 
-  html_notebook:
-    code_folding: hide
+title: "Two Model solution"
+output: html_notebook
 ---
 
-Here is the breakdown so far of what I am trying to do in INLA with modeling annual accumulation across central West Antarctica.
-The main parameter of interest for me is the spatial distribution in the change in accumulation over time (with the uncertainty in this trend across the region) while accounting for the temporal and spatial autocorrelation in the data.
-My main questions are (1) do I understand the model properly and is it doing/estimating what I think it is?
-(2) are there better ways that you know of to implement the model for my targets of interest?
-and (3) I'm seeing some weird behavior in some of the marginal posterior distributions that I would like your thoughts on.
-
-## Model setup
-
-*Here's a overview of the statistical underpinnings of the model as I see them.*
-*Feel free to critique/correct any misconceptions or errors you see here (and any issues with equations/notations, as I am new to this and have also been revising/updating it a fair bit).*
-*Many thanks!*
-
-The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension.
-I model accumulation as a gamma distribution to ensure a zero-bounded function with increased variance at higher expected values.
-$$\mathbf{Y} \sim Gamma(a,b)$$
-Rather than using the shape and rate parameters $a$ and $b$, I express this gamma distribution in terms of the expected y value $E(y_i) = a_i/b_i = \mu_i$ and its associated variance (as expressed in INLA using the precision parameter $\phi$) $Var(y_i) = a_i/b_i^2 = \mu_i^2/\phi$.
-This $\phi$ is the same precision parameter as output from `inla` (there is also a scale parameter $s$ that has a default of 1 where $Var(y_i) = 1/\tau$ and $\tau = (s\phi)/\mu_i^2$).
-With rearranging and substituting, I have $a_i = \phi$ and $b_i = \phi/\mu_i$ and I can therefore express the distribution in accumulation observations in terms of $\phi$ and $\mu_i$.
-$$y_i \sim Gamma(\mu_i, \phi)$$
-
-For each observation $y_{i,t} \in \mathbf{Y}$, I construct a linear predictor $\eta_{i,t}$ for the expectation value $\mu_{i,t}$.
-As I assume a gamma-distribution for accumulation, I associate $\eta_{i,t}$ and $\mu_{i,t}$ via the log link function.
-$$\log(\mu_{i,t}) = \eta_{i,t}$$ 
-I can further model $\eta_{i,t}$ as a generalized linear model consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case modeling a spatial random effect $u(\cdot)$ and a temporal random effect $\omega(\cdot)$).
-$$
-\eta_{i,t} = \alpha + \sum^M_{m=1} \beta_m x_{mi} + u(s_i) + \omega(t)
-$$
-
-These latent random effects $u_i$ and $\omega_i$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
-I can model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of some hyperparameters $\pmb{\psi}$ and an associated prior distribution $p(\pmb{\psi})$.
-This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{jk} = Cov(\theta_j, \theta_k)$.
-
-$$
-\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
-$$
-
-If I assume adherence to Markovian properties, the precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of the process, i.e.
-$Q_{jk} \neq 0 \iff k \in \{j,N(j)\}$.
-In this assumption, I specify a Gaussian-Markov Random Field for both the spatial and temporal structures, permitting vast improvements in computational efficiency.
-I also model the spatial and temporal random effects independently, which again greatly improves computation time, but will fail to capture any co-dependencies between these effects.
-
-I model the temporal random effect $\omega$ as a first-order autoregressive model^[I will eventually want to update this to an AR(2) process due to the known temporal autocorrelation in the data] (representing how the spatial field evolves in time) where
-$$\omega_t = \rho \omega_{t-1}$$
-
-The objective, therefore, is to find the joint posterior distribution of the various parameters of interest, as this will allow me to perform inference and make predictions at unobserved locations.
-$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s \mid \mathbf{Y})$$
-
-### Direct temporal trend analysis
-
-*This is where I've (maybe?) diverged a bit from what I discussed above.*
-
-As I am most interested in estimating the spatially-varying trend in accumulation over time, I've attempted to adapt/implement a model to explicitly do this.
-This rendition of the model follows the areal example for disease mapping found in Section 3.3 of [A tutorial in spatial and spatio-temporal models with R-INLA](https://discovery.ucl.ac.uk/id/eprint/1415919/1/Baio_BlaCamBaiRue.pdf), along with some inspiration from an [INLA discussion group thread](https://groups.google.com/g/r-inla-discussion-group/c/X_g4NuqqHfY).
-The [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in "Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA" also discusses how to implement spatially-varying model coefficients in INLA, but near as I can tell it doesn't explicitly discuss extracting the time coefficient (but maybe I missed it, or it is trivial to add this?).
-
-The idea here (i.e.
-in following the disease mapping example) is to directly model the temporal trend as a global fixed effect, but then model the deviations from the global value as a spatially-correlated random effect.
-I want to do this while also accounting for both spatial and temporal correlations in the expected values of accumulation.
-The linear predictor therefore now looks like:
-$$
-\eta_{i,t} = \alpha + \sum^M_{m=1} \beta_m x_{mi} + (\beta_G + \delta_i) \times time \ + u(s_i) + \omega(t)
-$$
-
-This model therefore accounts for fixed effects, a global fixed effect from time $\beta_G$, a spatially-varying deviation (modeled as a spatial random effect indexed on time) from the global time effect $\delta_i$, a spatial random effect, and a temporal random effect.
-
-*When looking at the model implementation below, is the above model what I am actually doing?*
-
 ## Model encoding
-
-### Imports
 
@@
 
-### Data loading
 ```{r}
@@
 
+### Data filtering
+
 ```{r}
+yr.start = 1985
+yr.end = 2004
 set.seed(777)
-gdf_traces = gdf_traces %>% sample_frac(0.01)
+gdf_traces = gdf_traces %>% sample_n(100)
 data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
-  filter(Year >= 1979) %>% filter(Year < 2010) %>% arrange(trace_ID, Year)
+  filter(Year >= yr.start) %>% filter(Year <= yr.end) %>% 
+  group_by(trace_ID) %>% filter(all(yr.start:yr.end %in% Year)) %>% ungroup() %>% 
+  arrange(trace_ID, Year)
 
@@
 # Center covariates
-yr.min = min(dat$Year)-1
+yr.min = min(data$Year)-1
 dat = dat %>% 
@@
          u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
-         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE))
-         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
+         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)) #%>% 
+  # mutate(Year = Year-mean(Year))
 
@@
 
-### INLA  modeling
+### Model 1
 
-I create the mesh used for defining the neighborhoods (this is important for generating the sparse precision matrix $Q$, enabling the GMRF approximation).
-I also create spatial projector matrices for the spatial random effect `A.spat` and the spatial effect on the deviation from the global time trend `A.spYR`.
-I assign priors here using PC matern covariance.
-*I may also need to revisit the priors specification, as I'm not sure I have the right assumption of units/transformations for all of them.*
-
-*These next sections are where I'm not sure I am telling INLA to do what I think I am.*
 ```{r}
@@
 
-spde.yr = inla.spde2.pcmatern(mesh = mesh, 
-  prior.range = c(5000, 0.01), # P(range < 5 km) = 0.01
-  prior.sigma = c(1, 0.01), 
-  constr = TRUE) #Adds an integrate-to-zero constraint (this seemed appropriate as this will model deviations from the global mean effect for time)
-
 # Index for SPDE model and projector matrix
@@
 A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
-
-# Index for SPDE model and projector matrix
-spYR.idx = inla.spde.make.index(name = "spYR.idx", n.spde = spde.yr$n.spde)
-A.spYR = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
 ```
@@
   data = list(y = dat.train$accum), 
-  A = list(1, 1, 1, A.spat, A.spYR),
+  A = list(1, 1, 1, A.spat),
   effects = list(list(Intercept = rep(1, nrow(dat.train))), 
@@
                         u10m=dat.train$u10m, v10m=dat.train$v10m, 
-                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10, 
-                        Year=dat.train$Year), 
+                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10),
                  list(time = dat.train$Year),
-                 spat.idx, 
-                 spYR.idx),
+                 spat.idx),
   tag = 'dat')
@@
   f(spat.idx, model = spde) + #spatial random effect
-  Year + f(spYR.idx, Year, model = spde.yr) + #Global effect from year and deviation from global (modeled as a spatial random effect with with weights given by Year)
   f(time, model = 'ar1', hyper = time.spec) #Temporal random effect modeled as AR1
 ```
-*The above model formula is a main place I'm not sure I am specifying what I think I am for INLA. Is this doing what I think I'm doing i.e. the formula shown in "Direct temporal trend analysis?"*
-
-## Run the model
 
 ```{r}
-mod = inla(form, 
+mod.1 = inla(form, 
                data = inla.stack.data(dat.stack), 
@@
 
-## Evalulate model performance
-
-```{r}
-summary(mod)
-```
-
-Compare results to a standard log-link gamma GLM to ensure correct interpretation of the fixed effects.
-```{r}
-# Formula without random variables
-f.glm = y ~ -1 + Intercept + Year + dem + aspect
-
-# Fitting a frequentist GLM to data
-mod.glm = glm(f.glm, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1), 
-               family = Gamma(link = "log"))
-
-summary(mod.glm)
-```
-
-### Posterior plots
+#### Posterior plots
 
@@
 ```{r}
-marg.fixed = t.marg(mod$marginals.fixed, function(x) exp(x))
+marg.fixed = t.marg(mod.1$marginals.fixed, function(x) exp(x))
 HP.names = c("Gamma precision", "Range for spat.idx", "Stdev for spat.idx", 
-             "Range for spYR.idx", "Stdev for spYR.idx", "Time precision", "Rho for time")
-marg.hyper = t.marg(mod$marginals.hyperpar, function(x) (x), param.names = HP.names)
+             "Time precision", "Rho for time")
+marg.hyper = t.marg(mod.1$marginals.hyperpar, function(x) (x), param.names = HP.names)
 p.selection = c("Intercept", "Year", "dem", "aspect", "Gamma precision", "Range for spat.idx", 
@@
 
-If I understand correctly (and used the right transformations for the links), the following posteriors should show:
+Calculate the model residuals
+```{r}
+# Get indices of observed data
+fitted.idx = inla.stack.index(dat.stack, 'dat')$data
 
-- Intercept: the mean accumulation when all other effects are at their mean values (this will be in the original units of mm w.e. per year) *this value of ~430 mm/a seems reasonable*
-- Gamma precision: the expected residual variability in accumulation from random noise not accounted for in the expected value $\mu_{i,t}$ *(for the average accumulation, this would be a sd of ~86 mm/a, which seems reasonable)*
-- fixed effect coefficients: should be multiplicative e.g.
-a one $\sigma$ increase in elevation (`dem` in the model) results in ~14% reduction in accumulation ($0.86-1= -0.14$) *these values also seem fairly reasonable*
-- range: the area-of-effect for spatial correlation (in meters) *again seems reasonable (~80 km) for spatial correlation in accumulation*
-- rho: the temporal autocorrelation *we assume this to be pretty high, so ~0.9 again makes sense*
-- `stdev for spat.idx`: the standard deviation in accumulation values (in original mm/a?) due to spatial autocorrelation i.e.
-how much variability we expect due to spatial correlation?
-*if that's true, this seems rather small at ~0.15 mm/a, especially given the value for precision.*
-  - *After additional consideration, I think this might be on an untransformed log scale, in which case this represents a $\pm16$% impact on the expected accumulation, which seems reasonable to me.
-Any clarification/confirmation you can offer would be wonderful though.*
-- `stdev for spYR.idx`: the standard deviation in the magnitude of deviation from the global accumulation trend with time i.e.
-how much variability we expect in the `Year` fixed coefficient due to spatial correlation.
-*I don't really have any idea about the units here though, is this perhaps a % change as well?
-If that's true, then the temporal trends STRONGLY vary based on location, given the narrow range in the global coefficient for Year of ~0.5% loss per year and that a decent amount of probability in the deviations shows 25+%.
-Could use some insight on this for sure.*
-  - *Or might this be a % change to the trend value itself? In which case it would represent say a 25% increase/decrease relative to the -0.5% baseline?*
+dat.train = dat.train %>% mutate(accum.mod=mod.1$summary.fitted.values$mean[fitted.idx]) %>% 
+  mutate(accum.res = accum-accum.mod)
 
-For the most part, these posteriors seem reasonable and fit with what I would broadly expect.
-I'm having some trouble interpreting/assessing a couple of them though (as noted above), and would greatly appreciate your thoughts/feedback on it.
-
-I'm also noticing some other weird things that I would like your thoughts on.
-Regarding the deviations from the global time trend, I did a bit more digging and found that the mean values across the region for these deviations are minuscule (shown in the plot below), on the order of $10^{-17}$, while the sd's in these values are much much higher ($10^{-1}$).
-The deviations also do not sum to zero (as I think they should as I give this constraint when defining the SPDE model in Line 154), but maybe this is just an artifact of the values being so close to zero anyways?
-```{r}
-Bi.sp = tibble(East = mesh$loc[,1], North = mesh$loc[,2], 
-               delta.i = mod$summary.random$spYR.idx$mean, 
-               delta.sd = mod$summary.random$spYR.idx$sd) %>% 
-  mutate(B.G = delta.i+mod$summary.fixed["Year",]$mean)
-# ggplot(Bi.sp) + geom_density(aes(x=B.G))
-ggplot(Bi.sp) + geom_density(aes(x=delta.i))
-
-ggplot(Bi.sp) + geom_density(aes(x=delta.sd))
+Is=sample(unique(dat.train$trace_ID),3)
+ggplot(dat.train %>% filter(trace_ID %in% Is), aes(Year, accum.res, group=trace_ID)) + 
+  geom_line(aes(color=as.factor(trace_ID))) + geom_smooth(method = 'lm')
 ```
 
-This is further highlighted by the below plot of a representative marginal distribution in $\beta_G + d_i$ for a single location (after applying an inverse link transformation, so this now shows the % change in accumulation per year due to time for this location).
+Setup of residuals model
 ```{r}
-B.ex = t.marg(mod$summary.fixed$mean[4]+mod$marginals.random$spYR.idx$index.7, 
-              function(x) exp(x)) %>% 
-  mutate(x = x-1)
-ggplot(B.ex) + geom_line(aes(x=x,y=y))
+# Make data stack
+stack.res <- inla.stack(
+  data = list(y = dat.train$accum.res), 
+  A = list(1, 1, A.spat),
+  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
+                 tibble(Year=dat.train$Year), 
+                 spat.idx),
+  tag = 'dat')
+
+# Model formula
+form = y ~ -1 + Intercept + #modeling the intercept
+  f(spat.idx,Year, model = spde) #spatial random effect
 ```
 
-So the modal deviation is negligible relative to the global time trend (-0.5%), but the error on this effect is massive with very fat tails, resulting in a plausible effect anywhere from a ~50% loss to a 100% gain in accumulation per year for this particular location.
-All the other locations have similarly massive uncertainties.
-This seems like far too wide a range to be useful.
-I'm now trying to figure out if this results from a mis-specfication of the model, or if INLA itself is just struggling with these data and this particular parameter for whatever reason.
-
-## Additional plots and comparisons (in case you find them helpful)
-
-Plots of the fitted mean values and the gamma precision uncertainites (both in original units) for Years 1, 11, 21, and 31 (1979, 1989, 1999, 2009).
 ```{r}
-fitted.idx = inla.stack.index(dat.stack, 'dat')$data
-
-val.fitted = tibble(mu = mod$summary.fitted.values$mean[fitted.idx]) %>% 
-  mutate(sd = sqrt(mu^2/mod$summary.hyperpar$mean[1])) %>% 
-  bind_cols(dat.train %>% select(Year, East, North)) %>% 
-  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
-
-ggplot(val.fitted %>% filter(Year %in% c(1,11,21,31))) + 
-  geom_sf(aes(color=mu)) + scale_color_viridis_c() + facet_wrap(~ Year)
-
-ggplot(val.fitted %>% filter(Year %in% c(1,11,21,31))) + 
-  geom_sf(aes(color=sd)) + scale_color_viridis_c(option="plasma") + facet_wrap(~ Year)
+mod.2 = inla(form, 
+               data = inla.stack.data(stack.res), 
+               family = 'Gaussian',
+               quantiles = NULL, 
+               control.predictor = list(compute = TRUE, A = inla.stack.A(stack.res)), 
+               # control.fixed = prior.fixed, 
+               control.family = list(hyper = list(prec = prec.prior)), 
+               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))
 ```
-
-Plots of the mean spatial random effect for accumulation and for the temporal trend in accumulation.
-I am using the assumption that these values are on the log scale, and am therefore representing them transformed to a % increase/decrease in the expected accumulation.
-```{r}
-test.rand = tibble(trend.delta = exp(mod$summary.random$spYR.idx$mean) - 1, 
-                   spat.mu = exp(mod$summary.random$spat.idx$mean) - 1, 
-                   East = mesh$loc[,1], North = mesh$loc[,2]) %>% 
-  st_as_sf(coords=c("East", "North"), crs=st_crs(gdf_traces))
-ggplot(test.rand) + geom_sf(aes(color=spat.mu)) + 
-  scale_color_gradientn(colours = cet_pal(5, name = "d1"))
-ggplot(test.rand) + geom_sf(aes(color=trend.delta)) + 
-  scale_color_continuous()
-```
-
-Plots of representative marginal distributions for the spatial random effect of individual select locations.
-The selected locations represent the min/mid/max effects across the region.
-These are again transformed to represent the % change in expected accumulation value due to the spatial random effect.
-```{r}
-marg.spat = t.marg(mod$marginals.random$spat.idx$index.205, function(x) exp(x)) %>% 
-  mutate(Param = "idx.205")
-marg.spat = marg.spat %>% bind_rows(t.marg(mod$marginals.random$spat.idx$index.1147, function(x) exp(x)) 
-                                    %>% mutate(Param = "idx.1147")) %>% 
-  bind_rows(t.marg(mod$marginals.random$spat.idx$index.214, function(x) exp(x)) 
-            %>% mutate(Param = "idx.214"))
-marg.spat = marg.spat %>% mutate(x = x-1)
-
-ggplot(marg.spat, aes(x=x,y=y)) + geom_line() + 
-  facet_wrap(vars(Param))  
-```
-
-And similar plots for selected years of the temporal random effect (1979, 1989, 1999, 2009).
-```{r}
-marg.time = t.marg(mod$marginals.random$time$index.1, function(x) exp(x)) %>% mutate(Param = "1979")
-marg.time = marg.time %>% bind_rows(t.marg(mod$marginals.random$time$index.11, function(x) exp(x)) 
-                                    %>% mutate(Param = "1989")) %>% 
-  bind_rows(t.marg(mod$marginals.random$time$index.21, function(x) exp(x)) 
-            %>% mutate(Param = "1999")) %>% 
-  bind_rows(t.marg(mod$marginals.random$time$index.31, function(x) exp(x)) 
-            %>% mutate(Param = "2009"))
-marg.time = marg.time %>% mutate(x=x-1)
-
-ggplot(marg.time, aes(x=x,y=y)) + geom_line() + 
-  facet_wrap(vars(Param))  
-```
-
-Below is a plot of the temporal random effect and its estimated uncertainty ($\pm \sigma$) across the full time series.
-```{r}
-time.tbl = tibble(Year=yr.min+(1:nrow(mod$summary.random$time)), 
-                  time.mu = mod$summary.random$time$mean, 
-                   time.sd = mod$summary.random$time$sd)
-
-ggplot(time.tbl, aes(x=Year, y=time.mu)) + 
-  geom_line() + 
-  geom_ribbon(aes(ymin=time.mu-time.sd,ymax=time.mu+time.sd), alpha=0.3) + 
-  ylab("Magnitude of temporal random effect")
-```
-
-## Summary of questions
-
-The following is just a list of the main questions I still have and would appreciate any info/feedback you might have on them.
-
-1. Does the model I have devised seem reasonable, or are there (easy/quick) ways to better get out the information of interest?
-2. Does the model as specified (particularly in regards to the spatial variation in time trends) actually calculate/return what I think it does?
-3. Are my interpretations of model output/transformations correct?
-4. Thoughts/suggestions regarding the very small mean values and comparatively very large uncertainties of $\delta_i$ (the spatial deviations in time trend coefficients) seen in the data?

### 2022-07-11 notebooks/draft-modeling.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

@@
 library(INLA)
-library(INLAutils)
+# library(INLAutils)
 library(spdep)
@@
 data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
-  filter(Year >= 1979) %>% filter(Year < 2014) %>% arrange(trace_ID, Year)
+  filter(Year >= 1979) %>% filter(Year < 2010) %>% arrange(trace_ID, Year)
 ```
@@
 ```
+
 ```{r fig, fig.width=16}

### 2022-07-11 notebooks/exploratory.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

---
title: "WAIS analysis using INLA"
output: html_notebook
---

This is a working version of the general workflow notebook for processing WAIS radar results using INLA.
This includes the general process from raw data import to final processing.
This also includes some instructions for various pre-processing in Python.

## MISC USEFUL LINKS

A few other useful tutorials and such regarding R-INLA and INLA in general:

- [Official R-INLA Website](https://www.r-inla.org/)
- [Spatial Modeling with R-INLA](https://ourcodingclub.github.io/tutorials/spatial-modelling-inla/): A nice walkthrough tutorial of spatial modeling and geostatistics using r-inla
- [Book: Spatiotemporal Bayesian Models using INLA](https://sites.google.com/a/r-inla.org/stbook/home)
- [A gentle INLA tutorial](https://www.precision-analytics.ca/articles/a-gentle-inla-tutorial/)
- [Intro in INLA for geospatial modeling](https://punama.github.io/BDI_INLA/)
- [Digital Book: Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/index.html)
- [Brief intro/tutorial on INLA](https://www.flutterbys.com.au/stats/tut/tut12.9.html)
- [Great mini tutorials using R-INLA](https://haakonbakkagit.github.io/organisedtopics.html)
- [Book: Geospatial Health Data: Modeling and Visualization with R-INLA and Shiny](https://www.paulamoraga.com/book-geospatial/): Walkthroughs using health data, but the explanations make more intuitive sense to me. Section 8 is particularly useful for my work (sadly Section 10 appears to be blank)

## Python preprocessing

To get data in the proper format prior to these analyses, some Python processing is required.
The necessary data are produced and saved to disk by running `preprocessing.py` in the `src` directory.

## INLA modeling

The remainder of this notebook shows the workflow for processing PAIPR WAIS data using rINLA.
I first detail some of the theory and principles behind this, and how specifically I plan to utilize them in this research specifically.

### Background

The accumulation data modeled in this study represents a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension.
We model accumulation as a gamma distribution to ensure a zero-bounded function with increased variance at higher expected values.
$$\mathbf{Y} \sim Gamma(a,b)$$
Rather than using the shape and rate parameters $a$ and $b$, we can express this gamma distribution in terms of the expected y value $E(y_i) = a_i/b_i = \mu_i$ and its associated variance (as expressed using the precision parameter $\phi$) $Var(y_i) = a_i/b_i^2 = \mu_i^2/\phi$.
This $\phi$ is the same precision parameter as output from `inla` (there is also a scale parameter $s$ that has a default of 1 where $Var(y_i) = 1/\tau$ and $\tau = (s\phi)/\mu_i^2$).
With rearranging and substituting, we have $a_i = \phi$ and $b_i = \phi/\mu_i$ and we can therefore express the distribution in accumulation observations in terms of $\phi$ and $\mu_i$.
$$y_i \sim Gamma(\mu_i, \phi)$$

For each observation $y_i \in \mathbf{Y}$, we can construct a linear predictor $\eta_i$ for the expectation value $\mu_i$.
As we assume a gamma-distribution for accumulation, we associate $\eta_i$ and $\mu_i$ via the log link function.
$$\log(\mu_i) = \eta_i$$ 
We can further model $\eta_i$ as a generalized linear model consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case we are modeling a spatial random effect $u_i$ and a temporal random effect $\omega_i$).
$$
\eta_{i} = \alpha +  \sum^M_{m=1} \beta_m x_{mi} + u_i + \omega_{i}
$$

These latent random effects $u_i$ and $\omega_i$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
We can model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian Fields, using a function of some hyperparameters $\psi$ and an associated prior distribution $p(\psi)$.
This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{ij} = Cov(\theta_i, \theta_j)$.

$$
\theta \sim \mathcal{N}(\mu, \Sigma)
$$

If we assume adherence to Markovian properties, our precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of our process, i.e. $Q_{ij} \neq 0 \iff j \in \{i,N(i)\}$.
In this assumption, we specify a Gaussian-Markov Random Field for both our spatial and temporal structures, permitting vast improvements in computational efficiency.
We also model the spatial and temporal random effects independently, which again greatly improves computation time, but will fail to capture any co-dependencies between these effects.

The spatial GMRF $u_i$ is derived from a stochastic partial differential equation...

We model the temporal random effect $\omega_i$ as a first-order autoregressive model^[I will eventually want to update this to an AR(2) process due to the known temporal autocorrelation in our data] (representing how the spatial field evolves in time) where
$$\omega_t = \rho \omega_{t-1}$$

Our objective, therefore, is to find the joint posterior distribution
$$P(\phi, \alpha, \pmb{\beta}, \rho, \mid \mathbf{Y})$$

### INLA installation

The latest version of R (`4.1.1` as of this writing) states it is currently incompatible with R-INLA.
This seems to be some sort of bug (perhaps limited to RStudio?) as it is possible to do this.
It does require installing from a local source instead of following the [standard install instructions](https://www.r-inla.org/download-install) for R-INLA.
Binaries for R-INLA can be found at [this ftp](http://inla.r-inla-download.org/R/testing/src/contrib/).
Then simply follow [these instructions](https://riptutorial.com/r/example/5556/install-package-from-local-source) to do install from local source in RStudio.
This will also require installing some of the dependencies and misc. required R packages as well (e.g. `sp`, `foreach`, etc.).

### Imports

```{r}
library(here)
library(tidyverse)
library(cetcolor)
library(sf)
library(terra)
library(INLA)
library(INLAutils)
library(spdep)
```

### Data loading and pre-processing

```{r}
data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1)
gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()

# Remove points in far northern extent
accum_bnds = st_coordinates(gdf_traces)
gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))

# Remove unrealistically high elevations
trace_drop = gdf_traces %>% filter(elev > 3000)

# Filter out overly large accum sites
trace_drop = trace_drop %>% filter(accum > 1000) %>% select(trace_ID)

# Add East/North coordinates and drop large accum sites
pts_tmp = st_coordinates(gdf_traces)
coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
```

For testing/development purposes, only use a small sample of total
```{r}
# xy.bnds = st_bbox(gdf_traces)
# xy.range = list(x=(xy.bnds$xmax - xy.bnds$xmin), y=(xy.bnds$ymax - xy.bnds$ymin))
# gdf_traces = gdf_traces %>% st_crop(xmin=as.numeric(xy.bnds$xmin+0.25*xy.range$x), 
#                                     xmax=as.numeric(xy.bnds$xmin+0.50*xy.range$x),
#                                     ymin=as.numeric(xy.bnds$ymin), 
#                                     ymax=as.numeric(xy.bnds$ymin+0.25*xy.range$y))

set.seed(777)
gdf_traces = gdf_traces %>% sample_frac(0.01)
data = data %>% filter(trace_ID %in% gdf_traces$trace_ID) %>% 
  filter(Year >= 1980) %>% filter(Year <= 2000) %>% arrange(trace_ID, Year)
```

```{r}
# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, std, elev)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% mutate(Year = Year-yr.min, elev = elev-mean(elev))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

### INLA  modeling

We first create the mesh used for defining the neighborhoods (this is important for generating the sparse precision matrix $Q$, enabling our GMRF approximation).
```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(20000, 50000), cutoff = 1000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")
# mesh = inla.mesh.2d(loc = st_coordinates(gdf_traces), 
#                     max.edge = c(20000, 50000), cutoff = 1000)
# plot(mesh)
# points(st_coordinates(gdf_traces), col = "red")
```

<!-- ## Form 1 -->

<!-- This follows examples as shown in Section 7.1 of [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/ch-spacetime.html#discrete-time-domain). -->

<!-- ### -->

<!-- We then define our SPDE object and create the identifier matrix $A$. -->
<!-- ```{r} -->
<!-- spde = inla.spde2.pcmatern(mesh = mesh,  -->
<!--   prior.range = c(5000, 0.01), # P(range < 5 km) = 0.01 -->
<!--   prior.sigma = c(250, 0.1)) # P(sigma > 250 mm w.e.) = 0.10 -->
<!-- # spde = inla.spde2.matern(mesh=mesh, alpha=2) -->

<!-- iset = inla.spde.make.index('i', n.spde = spde$n.spde, -->
<!--   n.group = max(dat$Year)) -->
<!-- A = inla.spde.make.A(mesh = mesh, -->
<!--   loc = cbind(dat$East, dat$North), group = dat$Year)  -->
<!-- # A_matrix = inla.spde.make.A(mesh, loc=st_coordinates(gdf_traces)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- dat.stack = inla.stack( -->
<!--   data = list(y = dat$accum),  -->
<!--   A = list(A, 1),  -->
<!--   effects = list(iset, data.frame(Intercept = 0, elev = dat$elev)), -->
<!--   tag = 'dat')  -->
<!-- ``` -->

<!-- I also define a PC-prior for the autoregressive temporal autocorrelation. -->
<!-- In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data. -->
<!-- ```{r} -->
<!-- # Prior on autoregressive rho parameter -->
<!-- h.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9))) -->

<!-- # Prior on fixed effects intercept -->
<!-- # prior.fixed <- list(mean.intercept = , prec.intercept = ) -->

<!-- # Model formula -->
<!-- formula = y ~ 0 + Intercept + elev + f(i, model = spde, group = i.group,  -->
<!--   control.group = list(model = 'ar1', hyper = h.spec)) -->

<!-- # PC prior on the autoregressive param -->
<!-- prec.prior <- list(prior = 'pc.prec', param = c(50, 0.9)) -->

<!-- # Model fitting -->
<!-- system.time( -->
<!--   results <- inla(formula,  -->
<!--                family = 'Gamma', -->
<!--                data = inla.stack.data(dat.stack),  -->
<!--                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack)),  -->
<!--                # control.fixed = prior.fixed,  -->
<!--                control.family = list(hyper = list(prec = prec.prior)),  -->
<!--                control.compute = list(dic = TRUE)) -->
<!-- ) -->
<!-- ``` -->

<!-- ## Form 2 -->

<!-- This appears to be the same end result as in Form 1, but a slightly different implementation as followed by Section 8.6.2 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models). -->

<!-- ### -->

<!-- We then define our SPDE object and create the identifier matrix $A$. -->
<!-- ```{r} -->
<!-- spde = inla.spde2.pcmatern(mesh = mesh,  -->
<!--   prior.range = c(5000, 0.01), # P(range < 5 km) = 0.01 -->
<!--   prior.sigma = c(250, 0.1)) # P(sigma > 250 mm w.e.) = 0.10 -->
<!-- # spde = inla.spde2.matern(mesh=mesh, alpha=2) -->

<!-- iset = inla.spde.make.index('i', n.spde = spde$n.spde, -->
<!--   n.group = max(dat$Year)) -->
<!-- A = inla.spde.make.A(mesh = mesh, -->
<!--   loc = cbind(dat$East, dat$North), group = dat$Year)  -->
<!-- # A_matrix = inla.spde.make.A(mesh, loc=st_coordinates(gdf_traces)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- dat.stack = inla.stack( -->
<!--   data = list(y = dat$accum),  -->
<!--   A = list(A, 1),  -->
<!--   effects = list(iset, data.frame(Intercept = 0, elev = dat$elev)), -->
<!--   tag = 'dat')  -->
<!-- ``` -->

<!-- I also define a PC-prior for the autoregressive temporal autocorrelation. -->
<!-- In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data. -->
<!-- ```{r} -->
<!-- # Prior on autoregressive rho parameter -->
<!-- h.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9))) -->

<!-- # Prior on fixed effects intercept -->
<!-- # prior.fixed <- list(mean.intercept = , prec.intercept = ) -->

<!-- # Model formula -->
<!-- formula = y ~ 0 + Intercept + elev + f(i, model = spde, group = i.group,  -->
<!--   control.group = list(model = 'ar1', hyper = h.spec)) -->

<!-- # PC prior on the autoregressive param -->
<!-- prec.prior <- list(prior = 'pc.prec', param = c(50, 0.9)) -->

<!-- # Model fitting -->
<!-- system.time( -->
<!--   results <- inla(formula,  -->
<!--                family = 'Gamma', -->
<!--                data = inla.stack.data(dat.stack),  -->
<!--                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack)),  -->
<!--                # control.fixed = prior.fixed,  -->
<!--                control.family = list(hyper = list(prec = prec.prior)),  -->
<!--                control.compute = list(dic = TRUE)) -->
<!-- ) -->
<!-- ``` -->

## Form 3

This follows examples from Section 8.6.1 in [Bayesian inference with INLA](https://becarioprecario.bitbucket.io/inla-gitbook/ch-temporal.html#separable-models), which details the spatio-temporal component as a separable model more explicitly (two random effects in the formula, one for the spatial model and one for the group-indexed time model).

**THIS IS SO MUCH FASTER!!!**

###

We then define our SPDE object and create the identifier matrix $A$.
```{r}
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(10000, 0.01), # P(range < 10 km) = 0.01
  prior.sigma = c(250, 0.10)) # P(sigma > 250 mm w.e.) = 0.10
# spde = inla.spde2.matern(mesh=mesh, alpha=2)

# iset = inla.spde.make.index('i', n.spde = spde$n.spde,
#   n.group = max(dat$Year))

# Create projector matrix for SPDE mesh
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))
# A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat$East, dat$North))

# Assign index vectors for SPDE model
spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)
```

```{r}
# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev), 
                 list(time = dat.train$Year), 
                 spat.idx),
  tag = 'dat') 
```

I also define a PC-prior for the autoregressive temporal autocorrelation.
In particular, I assume that $P(corr > 0) = 0.9$, indicating a high degree of confidence that some temporal autocorrelation exists in our data.
```{r}
## Define other priors (spatial field priors defined earlier)

# Prior on autoregressive rho parameter
time.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Prior on fixed effects intercept
prior.fixed <- list()

# Prior on precision parameter for Gamma observations
prec.prior <- list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

## PC prior implementations

# # Prior on autoregressive rho parameter
# time.spec <- list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))
# 
# # Prior on fixed effects intercept
# prior.fixed <- list()
# 
# # Prior on precision parameter for Gamma observations
# prec.prior <- list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# Model formula
formula = y ~ -1 + Intercept + elev + 
  f(time, model = 'ar1', hyper = time.spec) + 
  f(spat.field, model = spde)
# formula = y ~ elev + f(spat.field, model = spde) + f(time.field, model = 'ar1', hyper = h.spec)
```

```{r}
# Model fitting
system.time(
  results <- inla(formula, 
                  data = inla.stack.data(dat.stack), 
                  family = 'Gamma',
                  control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1), 
                  # control.fixed = prior.fixed, 
                  control.family = list(hyper = list(prec = prec.prior)), 
                  control.compute = list(dic = TRUE))
)
```

## Continuation

```{r}
# summary(results)
round(results$summary.fixed, 4)
round(results$summary.hyperpar, 3)
results$dic$dic
```

Extract fitted values from `inla` model.
```{r}
dat.idx = inla.stack.index(dat.stack, 'dat')$data
G.params = tibble(mu = results$summary.fitted.values$mean[dat.idx], mu.sd = results$summary.fitted.values$sd[dat.idx])

res.tbl = tibble(trace_ID = dat.train$trace_ID, East = dat.train$East, North = dat.train$North, 
              Year = dat.train$Year, elev = dat.train$elev, accum = dat.train$accum, y.hat = G.params$mu) %>% 
  mutate(y.res = y.hat-accum) %>% 
  st_as_sf(coords = c("East", "North"), crs=st_crs(gdf_traces))

```

## Test data results

```{r}
# Create projector matrix for test data
A.test = inla.spde.make.A(mesh = mesh, loc = cbind(dat.test$East, dat.test$North))

# Create test data stack and join to data stack
test.stack <- inla.stack(
  data = list(y = NA), 
  A = list(1, 1, 1, A.test),
  effects = list(list(Intercept = rep(1, nrow(dat.test))), 
                 tibble(elev = dat.test$elev), 
                 list(time = dat.test$Year), 
                 spat.idx),
  tag = 'test')
val.stack <- inla.stack(dat.stack, test.stack)
```

```{r}
# Model fitting to test data
system.time(
  mod.test <- inla(formula, 
                  data = inla.stack.data(val.stack), 
                  family = 'Gamma',
                  quantiles = NULL, 
                  control.predictor = list(compute = TRUE, A = inla.stack.A(val.stack), link = 1), 
                  control.family = list(hyper = list(prec = prec.prior)), 
                  control.inla = list(strategy = 'adaptive'), 
                  control.mode = list(theta = results$mode$theta, restart = FALSE))
)
```

Below is a function that, given an input of mean estimates, sd on mean estimates, precision parameter for Gamma observations, and sd on precision parameter, returns $n$ samples from the posterior predictive distribution for the given inputs.
```{r}
pp.dist = function(data, prec, prec.sd, n=5000, return_sample=FALSE) {
  if (return_sample==TRUE) {
    ppd = matrix(data=NA, nrow=nrow(data), ncol=n)
  } else {
    ppd = tibble(mu=rep(0,nrow(data)), map=rep(0,nrow(data)), 
                 CI.low=rep(0,nrow(data)), CI.high=rep(0,nrow(data)))
  }

  for (i in 1:nrow(data)) {
    dat.i = data[i,]
    mu = rnorm(n, mean=dat.i$mu, sd=dat.i$mu.sd)
    a = rnorm(n, mean=prec, sd=prec.sd)
    b = a/mu
    dist = rgamma(n=n, shape = a, rate = b)

    if (return_sample==TRUE) {
      ppd[i,] = dist
    } else {
      dist.mu = mean(dist)
      d = density(dist)
      dist.map = d$x[which.max(d$y)]
      dist.hdi = HDInterval::hdi(dist, credMass=0.89)
      ppd[i,] = tibble_row(mu=dist.mu, map=dist.map, 
                           CI.low=as.numeric(dist.hdi[1]), 
                           CI.high=as.numeric(dist.hdi[2]))
    }
  }
  return(ppd)
}
```

```{r}
# Get inla grid indices where test data resides
test.idx = inla.stack.index(val.stack, 'test')$data

test.fitted = tibble(mu = mod.test$summary.fitted.values$mean[test.idx], 
                                  mu.sd = mod.test$summary.fitted.values$sd[test.idx])

system.time(
  post.pred_test <- pp.dist(data=test.fitted, prec=mod.test$summary.hyperpar$mean[1], 
                       prec.sd=mod.test$summary.hyperpar$sd[1])
)
```

```{r}
res.test = dat.test %>% 
  mutate(mu=post.pred_test$mu, CI.low=post.pred_test$CI.low, 
                               CI.hi=post.pred_test$CI.high) %>% 
  mutate(y.res = mu-accum)

ggplot(data=res.test, aes(accum, mu)) + 
  geom_point() + 
  geom_abline(slope=1, intercept=0, color='red') + 
  # geom_smooth(method = 'lm') + 
  lims(x=c(200,600), y=c(200,600))

d = density(res.test$y.res)
d.mode = d$x[which.max(d$y)]
ggplot(res.test) + geom_density(aes(y.res)) + 
  geom_vline(xintercept = d.mode, linetype=2, color='red') + 
  annotate(geom="text", x=d.mode-100, y=max(d$y), 
           label=paste("MAP =", format(d.mode, digits=3)), 
           color="red")
```

###

Here's some experimental stuff for looking at the variability in our estimates.
Basically, I think this might represent the expected value (the `fitted.values` from the model) along with the model variance.

We have data drawn from a Gamma distribution
$$
y_i \mid \mathbf{X_i},\alpha,s_i,\theta \sim Gamma(a_i,b_i)
$$
where the expected value is $E(y_i) = a_i/b_i = \mu_i$ and the variance is $Var(y_i) = a_i/b_i^2 = \mu_i^2/\phi$ where $\phi$ is the precision parameter (this is the same precision parameter as output from `inla` - there is also a scale parameter that has a default of 1 where $Var(y_i) = 1/\tau$ and $\tau = (s\phi)/\mu_i^2$).
This in turn should mean that ??? $a_i = \phi$ and $b_i = \phi/\mu_i$ ???.

We can therefore report the calculated expected values $\mu_i \in \pmb{\mu}$ along with the modeled random variance $\pmb{\mu}^2/\phi$.
I'm not sure how this modeled random variance is supposed to differ from the sd reported in `fitted.values` though (I suppose sd must be the standard deviation on the mean parameter, while the modeled variance is the assumed variability about the true mean).

### 2022-07-11 notebooks/reports/phil_20220506.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

@@
 ---
+
+## Notes
+
+- Try simulating data to test if we can recover values
+  - Ensure spatial variations in simulated data are smooth (sample from true GP?)
+  - Try with normal model first?
+- Prior on $\delta_i$?
+- Try fixing the range?
+- Exploratory analysis on the log scale
+- Ask questions on r-INLA discussion board?
+- Try with a larger data set?
+- Run model excluding time trend to account for other effects, and run additional model on residuals of first model.
 
@@
          u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
-         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE))
+         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)) %>% 
+  mutate(Year = Year-mean(Year))
          # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
@@
                         mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10, 
+                 #        Year=dat.train$Year-mean(dat.train$Year)), 
+                 # list(time = dat.train$Year-mean(dat.train$Year)), 
                         Year=dat.train$Year),
@@
 *The above model formula is a main place I'm not sure I am specifying what I think I am for INLA. Is this doing what I think I'm doing i.e. the formula shown in "Direct temporal trend analysis?"*
+
+```{r}
+# # Make data stack
+# dat.stack <- inla.stack(
+#   data = list(y = dat.train$accum), 
+#   A = list(1,1, A.spYR),
+#   effects = list(list(Intercept = rep(1, nrow(dat.train))), 
+#                  tibble(Year=dat.train$Year),
+#                  spYR.idx),
+#   tag = 'dat')
+# 
+# # Priors for temporal autocorrelation, fixed effects (currently empty), and gamma precision
+# prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%
+# 
+# # Model formula
+# form = y ~ -1 + Intercept + #modeling the intercept
+#   Year + f(spYR.idx, Year, model = spde.yr) #Global effect from year and deviation from global (modeled as a spatial random effect with with weights given by Year)
+```
 

### 2022-07-11 notebooks/wais-tests.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

---
title: "WAIS testing"
output: html_notebook
---

Just some tests of INLA modeling actually using my WAIS data.

```{r}
library(here)
library(tidyverse)
library(cetcolor)
library(sf)
library(tmap)
library(terra)
library(INLA)
library(INLAutils)
library(spdep)
```

```{r}
data_long = read_csv(here('data/traces_long.csv')) %>% select(-...1)
gdf_traces = st_read(here('data/geo_traces.geojson')) %>% as_tibble() %>% st_as_sf()

# Remove points in far northern extent
accum_bnds = st_coordinates(gdf_traces)
gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]), 
                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))
```

```{r}
# ## Find which REMA DEM 8-meter tiles intersect our dataset and download any tiles not
# ## already present in the selected directory "output.dir"
# 
# # Select directory containing REMA mosaic tiles (based on currently-running OS)
# os = Sys.info()['sysname']
# if (os == 'Linux') {
#   REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
# } else {
#   REMA.dir = file.path("G:/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
# }
# # REMA.dir = tk_choose.dir(caption = "Select directory in which to save REMA mosaic tiles")
# 
# # Find all mosaics that intersect the accumulation dataset
# index.file = file.path(dirname(REMA.dir), "REMA_Tile_Index_Rel1.1/REMA_Tile_Index_Rel1.1.shp")
# dem.index = st_read(index.file) %>% st_transform(crs = st_crs(gdf_traces))
# # ggplot() + geom_sf(data = Ant.outline, color = 'grey') + geom_sf(data = dem.index)
# keep.idx = unique(unlist(st_intersects(gdf_traces, dem.index)))
# mosaic.download = dem.index[keep.idx,] %>% mutate(fileurl = as.character(fileurl))
# 
# # Download files from REMA ftp server
# for (file in 1:nrow(mosaic.download)) {
#   tmp.file = tempfile()
#   dir.name = basename(dirname(mosaic.download$fileurl[file]))
#   
#   # If not already downloaded, download tile and extract
#   if (!dir.exists(file.path(REMA.dir, dir.name))) {
#     download.file(mosaic.download$fileurl[file], tmp.file)
#     untar(tmp.file, exdir = file.path(REMA.dir, dir.name))
#   }
#   unlink(tmp.file)
# }
```

```{r}
##----
# ## Generate aspect and slope tiffs for any required REMA tiles, extract slope/aspect
# ## values for all accum points, and add data to data tibbles
# 
# # Function to find which accum points are within a given REMA 8-meter mosaic tile 
# # and extract elevation, slope, and aspect from the REMA raster data
# get_topo_vals = function(tile, accum.pts, tile.dir) {
#   
#   # Logical to filter accum data to those members within the tile extent
#   pts.log = st_intersects(accum.pts, tile, sparse = FALSE)[,1]
#   accum.i = accum.pts %>% filter(pts.log)
#   
#   # Generate file paths to DEM, slope raster and aspect raster
#   dem.path = file.path(tile.dir, as.character(tile$tile), 
#                        paste(as.character(tile$tile), "_8m_dem.tif", sep = ""))
#   slope.path = file.path(tile.dir, as.character(tile$tile), 
#                          paste(as.character(tile$tile), "_8m_slope.tif", sep = ""))
#   aspect.path = file.path(tile.dir, as.character(tile$tile), 
#                           paste(as.character(tile$tile), "_8m_aspect.tif", sep = ""))
#   
#   # Check if slope/aspect rasters exist. If they don't, calculate them from DEM 
#   # and save to REMA directory
#   if (!file.exists(slope.path)) {
#     tile.dem = raster(dem.path)
#     tile.slope = terrain(tile.dem, opt='slope', unit='degrees')
#     writeRaster(tile.slope, filename = slope.path, format = "GTiff")
#     tile.aspect = terrain(tile.dem, opt='aspect', unit='degrees')
#     writeRaster(tile.aspect, filename = aspect.path, format = "GTiff")
#   }
#   
#   # Generate raster stack of DEM, slope, and aspect 
#   # and extract raster values at accum points
#   tile.stack = stack(list(dem.path, slope.path, aspect.path))
#   stack.vals = extract(tile.stack, accum.i)
#   
#   # Get index values of accum data subset within the global accum dataset
#   pts.idx = (1:length(pts.log))[pts.log]
#   
#   # Return index positions and raster stack values as list object
#   return(list(pts.idx, stack.vals))
# }
```

```{r}
# # Add variables for REMA elevation, slope, and aspect
# accum.PS = accum.PS %>% mutate(elev.REMA = NA, slope = NA, aspect = NA)
# 
# for (i in 1:nrow(mosaic.download)) {
#   
#   # Get current tile with coverage in dataset
#   tile = mosaic.download[i,]
#   
#   # Calculate and extract elevation, slope, and aspect
#   vals.list = get_topo_vals(tile, accum.PS, REMA.dir)
#   topo.idx = vals.list[[1]]
#   topo.vals = vals.list[[2]]
#   
#   # Assign topo values to accum tbl
#   accum.PS$elev.REMA[topo.idx] = topo.vals[,1]
#   accum.PS$slope[topo.idx] = topo.vals[,2]
#   accum.PS$aspect[topo.idx] = topo.vals[,3]
# }
```

```{r}
# REMA.200m = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m/REMA_200m_dem_filled.tif"
# elev = rast(REMA.200m)
# crs(r) = st_crs(gdf_traces)$wkt
# 
# slope.200m = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m/REMA_200m_slope.tif"
# slope = terrain(elev, v='slope', filename=slope.200m)
# 
# aspect.200m = "/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m/REMA_200m_aspect.tif"
# slope = terrain(elev, v='aspect', filename=aspect.200m)

REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
crs(dem) = st_crs(gdf_traces)$wkt

slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
crs(slp) = crs(dem)

aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
crs(aspct) = crs(dem)

topo.stk = c(dem, slp, aspct)

topo.pts = extract(topo.stk, vect(gdf_traces$geometry))
gdf_traces = gdf_traces %>% mutate(dem=topo.pts$REMA_200m_dem, 
                               slope=topo.pts$slope, 
                               aspect=topo.pts$aspect)

```

```{r}
RACMO.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data", 
                      "/RACMO/2.3p2_yearly_ANT27_1979_2016/interim")
files = list.files(path = RACMO.dir, full.names = TRUE, 
                   pattern = "\\.tif$")

r.smb = rast(files[1])
names(r.smb) = 1979:2016
# smb.pts = extract(r.smb, vect(gdf_traces$geometry))
r.u10m = rast(files[2])
names(r.u10m) = 1979:2016
# u10m.pts = extract(r.u10m, vect(gdf_traces$geometry))
r.v10m = rast(files[3])
names(r.v10m) = 1979:2016
# v10m.pts = extract(r.v10m, vect(gdf_traces$geometry))

racmo.stk = c(mean(r.smb), mean(r.u10m), mean(r.v10m))
names(racmo.stk) = c("smb", "u10m", "v10m")

racmo.pts = extract(racmo.stk, vect(gdf_traces$geometry))
gdf_traces = gdf_traces %>% mutate(smb.racmo=racmo.pts$smb, 
                               u10m=racmo.pts$u10m, 
                               v10m=racmo.pts$v10m)

gdf_traces = gdf_traces %>% mutate(S_x = abs(slope)*cos(aspect*pi/180), 
                                   S_y = abs(slope)*sin(aspect*pi/180)) %>% 
  mutate(MSWD = S_x*u10m + S_y*v10m)
```

## Data exploration

```{r}
# tmap_mode("view")
# tm_shape(gdf_traces["accum"] %>% filter(accum < 750)) + 
#   tm_dots(col="accum", palette="viridis", style="cont")
# 
# tm_shape(gdf_traces %>% filter(accum < 750) %>% select(std)) + 
#   tm_dots(col="std", palette="plasma", style="cont")

ggplot(gdf_traces) + geom_sf(aes(color=accum)) + 
  scale_color_viridis_c(limits=c(200,500))
# ggplot(gdf_traces) + geom_sf(aes(color=std)) + 
#   scale_color_viridis_c(option="plasma", limits=c(0, 750))
ggplot(gdf_traces) + geom_sf(aes(color=(100*std/accum))) + 
  scale_color_viridis_c(option="plasma", limits=c(0,50))
ggplot(gdf_traces) + geom_sf(aes(color=smb.racmo-accum)) + 
  scale_color_gradientn(colours = cet_pal(5, name = "d1a"), limits=c(-250,250))

ggplot(gdf_traces) + geom_sf(aes(color=dem)) + 
  scale_color_viridis_c(limits=c(750,2000), option="magma") + 
  labs(color = "% error")
ggplot(gdf_traces) + geom_sf(aes(color=elev)) + 
  scale_color_viridis_c(limits=c(1250,2500), option="magma")
ggplot(gdf_traces) + geom_sf(aes(color=slope)) + lims(color=c(0,1))
ggplot(gdf_traces) + geom_sf(aes(color=aspect)) + 
  scale_color_gradientn(colours = cet_pal(5, name = "c2s"))
ggplot(gdf_traces) + geom_sf(aes(color=MSWD)) + scale_color_viridis_c()
```

```{r}
ggplot(gdf_traces, aes(x=accum)) + geom_histogram()
ggplot(gdf_traces, aes(x=std)) + geom_histogram()
ggplot(gdf_traces, aes(x=accum, y=std)) + geom_point()
ggplot(gdf_traces, aes(sample=accum)) + stat_qq() + stat_qq_line()
# vals = log1p(st_drop_geometry(gdf_traces["accum"]))
# ggplot(vals, aes(x=accum)) + geom_histogram()
# ggplot(vals, aes(sample=accum)) + stat_qq() + stat_qq_line()

ggplot(gdf_traces, aes(x=elev, y=dem)) + geom_point()
ggplot(gdf_traces) + geom_density(aes(x=elev), fill='red', alpha=0.5) + 
  geom_density(aes(x=dem), fill='blue', alpha=0.5)
ggplot(gdf_traces) + geom_sf(aes(color=(elev-dem))) + 
  scale_color_gradientn(colours = cet_pal(5, name = "l13"))

ggplot(gdf_traces) + geom_point(aes(x=smb.racmo, y=accum)) + 
  geom_abline(slope = 1, intercept = 0)
ggplot(gdf_traces) + geom_point(aes(x=dem, y=accum))
ggplot(gdf_traces) + geom_point(aes(x=slope, y=accum))
ggplot(gdf_traces) + geom_point(aes(x=aspect, y=accum))

# pairs(gdf_traces %>% select(-trace_ID, -collect_time) %>% st_drop_geometry())
# ggplot(gdf_traces %>% select(-trace_ID, -collect_time) %>% st_drop_geometry(), 
       # aes(x=id, y=value)) + geom_point() + facet_grid(.~variable)

```

## Scaling and centering data

```{r}
# Log transform positive variables
# data = gdf_traces %>% mutate(accum = log(accum), std = log(std), elev = log(elev),
#                              dem = log(dem), smb.racmo = log(smb.racmo)) %>%
#   select(geometry, accum, std, elev, dem, MSWD, smb.racmo)

dat_train = gdf_traces %>% slice_sample(prop = 0.10) %>% arrange(trace_ID)
dat_test = gdf_traces %>% filter(!trace_ID %in% dat_train$trace_ID)
```

## INLA modeling

```{r}

mesh = inla.mesh.2d(loc = st_coordinates(dat_train), 
                    # crs = st_crs(dat_train)$Wkt, 
                    max.edge = c(25000, 100000), cutoff = 1000)
plot(mesh)
points(st_coordinates(dat_train), col = "red")
# Mesh = inla.mesh.2d(st_coordinates(tmp),
#                       max.edge = c(900, 2000),
#                       cutoff = 190)
# 
# plot(Mesh,asp = 1, main = "")
# points(st_coordinates(tmp), col = 'red', pch = 16, cex = 0.1)
```

```{r}

A_matrix = inla.spde.make.A(mesh, loc = st_coordinates(dat_train))
spde.0 = inla.spde2.matern(mesh = mesh, alpha = 2)

# Define penalized-complexity (PC) prior
pcprec <- list(prior = 'pcprec', param = c(1, 0.01))
# Define covariance function
spde_pc <- inla.spde2.pcmatern(mesh = mesh,
  prior.range = c(0.05, 0.01), # P(practic.range < 0.05) = 0.01
  prior.sigma = c(1, 0.01)) # P(sigma > 1) = 0.01

# Make data stack
stk.dat <- inla.stack(
  data = list(y = dat_train$accum), 
  A = list(A_matrix,1),
  effects = list(list(s = 1:spde.0$n.spde), 
                 tibble(Intercept = 1, 
                        gElev = inla.group(dat_train$dem), 
                        gMSWD = inla.group(dat_train$MSWD), 
                        elev = dat_train$dem, 
                        mswd = dat_train$MSWD)), 
  tag = 'dat_train')

f.0 = y ~ 0 + Intercept + 
  f(spatial.field, model = spde.0)
f.fixed = y ~ 0 + Intercept + elev + mswd + 
  f(s, model = spde.0)
f.rw1 = y ~ 0 + Intercept + 
  f(gElev, model = 'rw1', scale.model = TRUE, 
    hyper = list(prec = pcprec)) + 
  f(gMSWD, model = 'rw1', scale.model = TRUE, 
    hyper = list(prec = pcprec)) + 
  f(s, model = spde_pc)
```

```{r}
# # Now we can fit the proper model using the inla() function
# Mod.out_0 = inla(f.0, family = 'Gamma', 
#                 data = list( y = dat_train$accum,         # response variable
#                             Intercept = rep(1,spde.0$n.spde),   # intercept (manually specified)
#                             spatial.field = 1:spde.0$n.spde),  # the spatial random effect (specified with the matern autocorrelation structure from spde)   
#                 control.predictor = list( A = A_matrix,
#                                           link = 1, 
#                                           compute = T),       # this tells the model to compute the posterior marginals for the linear predictor
#                 control.compute = list(cpo = T))

```

```{r}
mod.fixed = inla(f.fixed, family = 'Gamma', 
                 control.compute = list(cpo = TRUE), 
                 data = inla.stack.data(stk.dat), 
                 control.predictor = list(A = inla.stack.A(stk.dat), link = 1))
```

```{r}
# mod.rw1 = inla(f.rw1, family = 'Gamma', 
#                control.compute = list(cpo = TRUE), 
#                  data = inla.stack.data(stk.dat), 
#                  control.predictor = list(A = inla.stack.A(stk.dat), link = 1))
```

```{r}
slcpo <- function(m, na.rm = TRUE) {
  - sum(log(m$cpo$cpo), na.rm = na.rm)
}

# c(mod.0 = slcpo(Mod.out_0), mod.fixed = slcpo(mod.fixed),
#   mod.rw1 = slcpo(mod.rw1))

mod.best = mod.fixed
```

```{r}
spat.tbl = tibble(E = mesh$loc[,1], N = mesh$loc[,2], 
                  spat.mu = mod.best$summary.random$s$mean, 
                   spat.sd = mod.best$summary.random$s$sd) %>% 
  st_as_sf(coords=c("E", "N"), crs=st_crs(dat_train))

ggplot(spat.tbl) + geom_sf(aes(color=spat.mu)) + 
  scale_color_viridis_c()
ggplot(spat.tbl) + geom_sf(aes(color=spat.sd)) + 
  scale_color_viridis_c(option='plasma')
```

```{r}
pred.bbox = st_bbox(gdf_traces)

stepsize = 1000
buffer = 30*stepsize
E.range = round(c(pred.bbox[1]-buffer, pred.bbox[3]+buffer))
N.range = round(c(pred.bbox[2]-buffer, pred.bbox[4]+buffer))

nxy <- round(c(diff(E.range), diff(N.range)) / stepsize) # Calculate the number of cells in the x and y ranges

# Project the spatial field on the mesh vertices using the inla.mesh.projector() function
projgrid <- inla.mesh.projector(mesh,
                                xlim = E.range,
                                ylim = N.range,
                                dims = nxy)

```

```{r}

# xmean = inla.mesh.project(projgrid,
#                            mod.best$summary.random$s$mean)
# xsd = inla.mesh.project(projgrid,
#                          mod.best$summary.random$s$sd)
# 
# r1 <- r2 <- rast(xmin=E.range[1], xmax=E.range[2]-stepsize, 
#               ymin=N.range[1], ymax=N.range[2], 
#               crs=st_crs(dat_train)$proj4string, 
#               resolution=stepsize)
# values(r1) = c(xmean)
# names(r1) = 'spat.mu'
# values(r2) = c(xsd)
# names(r2) = 'spat.sd'
# s.pred = c(flip(r1), flip(r2))
```

```{r}

# Find index of grid points within the country border
grid.pts = tibble(x = projgrid$lattice$loc[,1],
                  y = projgrid$lattice$loc[,2],
                  idx = 1:nrow(projgrid$lattice$loc))
grid.sf = st_as_sf(grid.pts, coords=c('x','y'))
# tmp = st_join(border.sf, grid.sf, join = st_intersects)
# grid.idx = tmp$idx

# Extract topo covariates from rasters for each grid point
topo.pts = extract(topo.stk, vect(grid.sf$geometry))
grid.sf = grid.sf %>% mutate(dem=topo.pts$REMA_200m_dem, 
                               slope=topo.pts$slope, 
                               aspect=topo.pts$aspect)

# Extract clim covariates from rasters for each grid point
racmo.pts = extract(racmo.stk, vect(grid.sf$geometry))
grid.sf = grid.sf %>% mutate(smb.racmo=racmo.pts$smb, 
                               u10m=racmo.pts$u10m, 
                               v10m=racmo.pts$v10m)
grid.sf = grid.sf %>% mutate(S_x = abs(slope)*cos(aspect*pi/180), 
                                   S_y = abs(slope)*sin(aspect*pi/180)) %>% 
  mutate(MSWD = S_x*u10m + S_y*v10m)
# grid.sf = drop_na(grid.sf) %>% st_as_sf(crs = st_crs(data))

```

```{r}

# Define projector matrix for predictions
A.pred = projgrid$proj$A

# Extract prediction coordinates
pred.coords = st_coordinates(grid.sf)

# Generate prediction data stack and join to estimation stack
stk.pred <- inla.stack(
  data = list(y = NA),
  A = list(A.pred, 1), 
  effects = list(list(s = 1:spde.0$n.spde), 
                 tibble(Intercept = 1, 
                        gElev = inla.group(grid.sf$dem), 
                        gMSWD = inla.group(grid.sf$MSWD), 
                        elev = grid.sf$dem, 
                        mswd = grid.sf$MSWD)),
  tag = 'pred') 
stk.all <- inla.stack(stk.dat, stk.pred)
```

```{r}
mod.preds <- inla(f.fixed, family = 'Gamma', 
                  data = inla.stack.data(stk.all), 
                  control.predictor = list(A = inla.stack.A(stk.all),
                                           compute = TRUE, link = 1), 
                  quantiles = NULL, 
                  control.inla = list(strategy = 'adaptive'), 
                  # control.results = list(return.marginals.random = FALSE,
                  #                        return.marginals.predictor = FALSE), 
                  control.mode = list(theta = mod.best$mode$theta,
                                      restart = FALSE))
```

```{r}
# Get inla grid indices where data exists
idx.pred <- inla.stack.index(stk.all, 'pred')$data

# Preallocate matrices for mean and sd predictions
# (using the same grid as generated for spatial rasters above)
sd.pred <- mu.pred <- matrix(NA, nxy[1], nxy[2])
mu.pred = mod.preds$summary.fitted.values$mean[idx.pred]
sd.pred = mod.preds$summary.fitted.values$sd[idx.pred]

bad.idx = mu.pred > 1500
# mu.pred[bad.idx] = NA
# sd.pred[bad.idx] = NA

spat.mu = inla.mesh.project(projgrid,
                           mod.preds$summary.random$s$mean)
spat.sd = inla.mesh.project(projgrid,
                         mod.preds$summary.random$s$sd)
```

```{r}

r1 <- r2 <- r3 <- r4 <- rast(xmin=round(E.range[1]), xmax=round(E.range[2]-stepsize), 
                 ymin=round(N.range[1]), ymax=round(N.range[2]), 
                 resolution=stepsize)
values(r1) = c(mu.pred)
names(r1) = 'pred.mu'
values(r2) = c(sd.pred)
names(r2) = 'pred.sd'
values(r3) = c(spat.mu)
names(r3) = 'spat.mu'
values(r4) = c(spat.sd)
names(r4) = 'spat.sd'

rast.pred = c(flip(r1), flip(r2), flip(r3), flip(r4))
crs(rast.pred) = st_crs(gdf_traces)$wkt

# Extract accum estimates at locations of test data
test.pts = extract(rast.pred, vect(dat_test$geometry))
dat_test = dat_test %>% mutate(pred.mu=test.pts$pred.mu, 
                               pre.sd=test.pts$pred.sd)

racmo.stk = racmo.stk %>% crop(rast.pred)
topo.stk = topo.stk %>% crop(rast.pred)
# plot(rast.pred)
```

```{r}

writeRaster(rast.pred, here('data/rast-pred_20211025.tif'), overwrite = TRUE, filetype = "GTIFF")
writeRaster(racmo.stk, here('data/racmo_20211025.tif'), overwrite = TRUE, filetype = "GTIFF")
writeRaster(topo.stk, here('data/topo_20211025.tif'), overwrite = TRUE, filetype = "GTIFF")
save(dat_test, dat_train, gdf_traces, mod.best, mod.preds, grid.sf, file = here('data/mod.fitted_20211025.RData'))
```

```{r}
rast.pred = rast(here('data/rast-pred_20211025.tif'))
racmo.stk = rast(here('data/racmo_20211025.tif'))
topo.stk = rast(here('data/topo_20211025.tif'))
load(here('data/mod.fitted_20211025.RData'))
```

```{r}
plot(rast.pred)

ggplot(dat_test) + geom_abline(slope = 1, intercept = 0) + 
  geom_point(aes(x=accum, y=pred.mu), color='red') + 
  geom_point(aes(x=accum, y=smb.racmo), color='blue') + 
  xlab('PAIPR accum') + ylab('Modeled accum')
```

```{r}
ggplot(gdf_traces) + geom_sf(aes(color=smb.racmo)) + scale_color_viridis_c(limits = c(150, 650))
ggplot(gdf_traces) + geom_sf(aes(color=accum)) + scale_color_viridis_c(limits = c(150, 650))
```

### 2022-07-11 scratch/simple-mods.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

---
title: "Experimentation with Gaussians and GLM results"
output: html_notebook
---

## Imports

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(cetcolor)
library(sf)
library(terra)
library(INLA)
library(INLAutils)
library(spdep)
```

## Data preprocessing

```{r message=FALSE}
data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1) %>% 
  mutate(Year = as.integer(Year))
gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()

# Remove points in far northern extent
accum_bnds = st_coordinates(gdf_traces)
gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))

# Remove unrealistically high elevations and overly large accum sites
trace_drop = gdf_traces %>% filter(elev > 3000 | accum > 1000)

# Add East/North coordinates and drop filtered sites
pts_tmp = st_coordinates(gdf_traces)
coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
gdf_traces = gdf_traces %>% filter(!trace_ID %in% trace_drop$trace_ID)
```

```{r}
REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
crs(dem) = st_crs(gdf_traces)$wkt

slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
crs(slp) = crs(dem)

aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
crs(aspct) = crs(dem)

topo.stk = c(dem, slp, aspct)

topo.pts = extract(topo.stk, vect(gdf_traces$geometry)) %>% as_tibble() %>% 
  rename(dem=REMA_200m_dem) %>% 
  mutate(trace.ID=gdf_traces$trace_ID)
topo.pts = topo.pts %>% select(-ID)

RACMO.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data", 
                      "/RACMO/2.3p2_yearly_ANT27_1979_2016/interim")
files = list.files(path = RACMO.dir, full.names = TRUE, 
                   pattern = "\\.tif$")

r.smb = rast(files[1])
names(r.smb) = 1979:2016
smb.pts = extract(r.smb, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% mutate(trace.ID=gdf_traces$trace_ID) %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'SMB') %>% 
  mutate(Year = as.integer(Year))

r.u10m = rast(files[2])
names(r.u10m) = 1979:2016
u10m.pts = extract(r.u10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'u10m') %>% 
  mutate(Year = as.integer(Year))

r.v10m = rast(files[3])
names(r.v10m) = 1979:2016
v10m.pts = extract(r.v10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% pivot_longer(cols = num_range(prefix='', range=1979:2016), 
                               names_to = 'Year', values_to = 'v10m') %>% 
  mutate(Year = as.integer(Year))

racmo.pts = smb.pts %>% 
  mutate(u10m=u10m.pts$u10m, v10m=v10m.pts$v10m) %>% 
  select(-ID)

data = data %>% 
  left_join(racmo.pts, by = c("trace_ID" = "trace.ID", "Year")) %>% 
  left_join(topo.pts, by = c("trace_ID" = "trace.ID"))

# data = data %>% mutate(S_x = abs(slope)*cos(asin(aspect)), 
#                        S_y = abs(slope)*aspect) %>% 
#   mutate(MSWD = S_x*u10m + S_y*v10m)
```

```{r}
set.seed(777)
gdf.tmp = gdf_traces %>% sample_frac(0.01)
data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year <= 2004) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m)
# dat = data %>% select(trace_ID, East, North, Year, accum, 
#                       std, elev, dem, slope, aspect, u10m, v10m)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = scale(elev)[,1], 
         dem=scale(dem)[,1], slope=scale(slope)[,1], 
         aspect=sin(pi/180*aspect), 
         u10m=scale(u10m)[,1], v10m=scale(v10m)[,1])

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

## INLA modeling

```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(20000, 50000), cutoff = 1000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")
```

```{r}
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(5000, 0.05), # P(range < 5 km) = 0.01
  prior.sigma = c(500, 0.05)) # P(sigma > 500 mm w.e.) = 0.05

# Create projector matrix for SPDE mesh
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))

# Assign index vectors for SPDE model
spat.idx = inla.spde.make.index(name = "spat.field", n.spde = spde$n.spde)

# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m), 
                 list(time = dat.train$Year), 
                 spat.idx),
  tag = 'dat') 
```

Base model (no random effects)
```{r}
# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# f.base = y ~ -1 + Intercept
f.base = y ~ -1 + Intercept + dem + slope + aspect + slope*aspect

# Model fitting
mod.gauss <- inla(f.base, 
                 data = inla.stack.data(dat.stack), 
                 family = 'Gaussian', 
                 control.family = list(
                                       # hyper = list(prec=prec.prior), 
                                       control.link=list(model="log")),
                 control.predictor = list(compute = TRUE, 
                                          A = inla.stack.A(dat.stack), 
                                          link = 1),
                 control.compute = list(waic = TRUE), 
                 quantiles = NULL)
# Model fitting
mod.gamma <- inla(f.base, 
                 data = inla.stack.data(dat.stack), 
                 family = 'Gamma',
                 control.family = list(hyper = list(prec = prec.prior)), 
                 control.predictor = list(compute = TRUE, 
                                          A = inla.stack.A(dat.stack), 
                                          link = 1), 
                 control.compute = list(waic = TRUE), 
                 quantiles = NULL)

# Fitting a frequentist GLM to data
mod.glm <- glm(f.base, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1), 
               family = gaussian(link = "log"))

summary(mod.glm)
```

Compare marginal distributions of precision (transformed into original units as sqrt(Var)) between Gaussian and gamma models.
```{r}
compare_marginal = function(gauss.mod, gamma.mod, N=100) {

  gauss.var = inla.tmarginal(function(x) 1/x, gauss.mod$marginals.hyperpar$`Precision for the Gaussian observations`, n=N)
  gamma.var = inla.tmarginal(function(x) exp(gamma.mod$summary.fixed$mean[1])^2/x, gamma.mod$marginals.hyperpar$`Precision parameter for the Gamma observations`, n=N)

  df1 = tibble(var.x=gauss.var[,1], var.y=gauss.var[,2], model = "Gaussian")
  df2 = tibble(var.x=gamma.var[,1], var.y=gamma.var[,2], model = "Gamma")

  for (i in 1:length(gauss.mod$names.fixed)) {
    fixed.gauss = inla.tmarginal(function(x) x, gauss.mod$marginals.fixed[[i]], n=N)
    fixed.gamma = inla.tmarginal(function(x) exp(x), gamma.mod$marginals.fixed[[i]], n=N)
    df1 = df1 %>% mutate("{gauss.mod$names.fixed[i]}.x" := fixed.gauss[,1], 
                         "{gauss.mod$names.fixed[i]}.y" := fixed.gauss[,2])
    df2 = df2 %>% mutate("{gamma.mod$names.fixed[i]}.x" := fixed.gamma[,1], 
                         "{gamma.mod$names.fixed[i]}.y" := fixed.gamma[,2])
  }
  # gauss.alpha = inla.tmarginal(function(x) x, gauss.mod$marginals.fixed$Intercept, n=N)
  # gamma.alpha = inla.tmarginal(function(x) exp(x), gamma.mod$marginals.fixed$Intercept, n=N)

  if (any(str_detect(names(gauss.mod$marginals.hyperpar), "Range for"))) {
    tmp1 = inla.tmarginal(function(x) x, 
                         gauss.mod$marginals.hyperpar$`Range for spat.field`, 
                         n=N)
    tmp2 = inla.tmarginal(function(x) x, 
                         gauss.mod$marginals.hyperpar$`Stdev for spat.field`, 
                         n=N)
    df1 = df1 %>% mutate(spat.range.x=tmp1[,1], spat.range.y=tmp1[,2], 
                         spat.std.x=tmp2[,1], spat.std.y=tmp2[,2])
  }
  if (any(str_detect(names(gamma.mod$marginals.hyperpar), "Range for"))) {
    tmp1 = inla.tmarginal(function(x) x, 
                         gamma.mod$marginals.hyperpar$`Range for spat.field`, 
                         n=N)
    tmp2 = inla.tmarginal(function(x) x, 
                         gamma.mod$marginals.hyperpar$`Stdev for spat.field`, 
                         n=N)
    df2 = df2 %>% mutate(spat.range.x=tmp1[,1], spat.range.y=tmp1[,2], 
                         spat.std.x=tmp2[,1], spat.std.y=tmp2[,2])
  }

  df = df1 %>% bind_rows(df2)
  return(df)
}
```

```{r}
df.base = compare_marginal(gauss.mod=mod.gauss, gamma.mod=mod.gamma)

ggplot(df.base, aes(x=Intercept.x, y=Intercept.y, group=model, color=model)) + 
  geom_line() + xlab('Intercept') + ylab('Density')
ggplot(df.base, aes(x=dem.x, y=dem.y, group=model, color=model)) + 
  geom_line() + xlab('Elevation effect') + ylab('Density')
ggplot(df.base, aes(x=slope.x, y=slope.y, group=model, color=model)) + 
  geom_line() + xlab('Slope effect') + ylab('Density')
ggplot(df.base, aes(x=aspect.x, y=aspect.y, group=model, color=model)) + 
  geom_line() + xlab('Aspect effect') + ylab('Density')
ggplot(df.base, aes(x=sqrt(var.x), y=sqrt(var.y), group=model, color=model)) + 
  geom_line() + xlab('St. dev') + ylab('Density')
```

Base model + spatial random effect
```{r}
# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(500, 0.05)) #P(sigma > 250) = 10%

f.sp = y ~ -1 + Intercept + 
  dem + slope + aspect + slope*aspect + 
  f(spat.field, model = spde) #spatial random effect
mod.gauss_sp <- inla(f.sp, 
                 data = inla.stack.data(dat.stack), 
                 control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack)),
                 family = 'Gaussian', 
                 control.family = list(hyper = list(prec = prec.prior)),
                 quantiles = NULL, 
                 control.compute = list(waic = TRUE))
mod.gamma_sp <- inla(f.sp, 
                 data = inla.stack.data(dat.stack), 
                 family = 'Gamma',
                 control.family = list(hyper = list(prec = prec.prior)),
                 control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1),
                 quantiles = NULL, 
                 control.compute = list(waic = TRUE))
```

```{r}
df.sp = compare_marginal(gauss.mod=mod.gauss_sp, gamma.mod=mod.gamma_sp)

ggplot(df.sp, aes(x=a.x, y=a.y, group=model, color=model)) + 
  geom_line() + xlab('Intercept') + ylab('Density')
ggplot(df.sp, aes(x=sqrt(var.x), y=sqrt(var.y), group=model, color=model)) + 
  geom_line() + xlab('St. dev') + ylab('Density')
ggplot(df.sp, aes(x=spat.range.x/1000, y=spat.range.y, group=model, color=model)) + 
  geom_line() + xlab('Range of spatial effect (km)') + ylab('Density')
ggplot(df.sp, aes(x=spat.std.x, y=spat.std.y, group=model, color=model)) + 
  geom_line() + xlab('st. dev of spatial effect') + ylab('Density')
```

```{r}
## Define other priors (spatial field priors defined earlier)

# Prior on autoregressive rho parameter
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Prior on fixed effects intercept (not currently implemented)
prior.fixed = list()
```

### 2022-07-11 scratch/sythnetic-tests.Rmd
INLA (#14)
* Clears old R files in preparation for new imports

* Merges INLA work from r-WAIS into this repo

* Breaks out pre-processing into separate scripts

* Updates INLA and ensures all packages are installed

* Adds explicit document explaining the installation of INLA

* Adds report detailing INLA model and questions for Phil

* Adds notebook comparing SMB estimates in wAIS for RACMO and PAIPR

* Adds notebook for testing spacetime modeling examples with Kronecker sums for effects

* Adds notebook for experimenting with kronecker products for st modeling

* Adds notebooks for comparis

---
title: "Synthetic tests of spatially varying time coefficient"
output: html_notebook
---

This notebook is just to experiment some with some synthetic data to see if I can reproduce the methods.
These tests are based off of examples in an [R-inla discussion post](https://groups.google.com/g/r-inla-discussion-group/c/X_g4NuqqHfY) and the [section on dynamic space-time regression](https://becarioprecario.bitbucket.io/spde-gitbook/ch-stapp.html#dynamic-space-time-regression) in [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/index.html).

```{r}
library(tidyverse)
library(INLA)
library(INLAutils)
library(spdep)
```

## Data generation

We need a function to add spatial correlation to our data.
The below function, provided with the [Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA](https://becarioprecario.bitbucket.io/spde-gitbook/index.html) RBook, does just that.
```{r}
book.rMatern <- function(n, coords, sigma=1, range, kappa = sqrt(8*nu)/range, variance = sigma^2, nu=1) {
    m <- as.matrix(dist(coords))
    m <- exp((1-nu)*log(2) + nu*log(kappa*m)-
             lgamma(nu))*besselK(m*kappa, nu)
    diag(m) <- 1
    return(drop(crossprod(chol(variance*m),
                          matrix(rnorm(nrow(coords)*n), ncol=n))))
}
```

We then define the model parameters/hyperparameters that we wish to recover with our model.
These consist of a global intercept, a global coefficient for the temporal trend, the effective range of our spatial correlation, the variability of our spatial correlation, the strength of the temporal correlation, and the iid Gaussian noise in our observations.
```{r}
alpha = 5.7
beta.G = -0.10
sp.range = 3300
sp.sigma = 0.05
rho = 0.8
tau.e = 20
```

We next generate $n=100$ sample locations and draw $k=10$ independent samples at each location from a Matern GRF (using the above function).
```{r}
n <- 100
k = 10
set.seed(1)
coo <- round(10000 * matrix(runif(2 * n), n))

set.seed(2)
beta.i <- book.rMatern(k, coo, range = sp.range, sigma = sp.sigma)
```

We then add temporal correlation to each time series using an $AR1$ process
```{r}
beta.i[, 1] <- beta.i[, 1] / (1 - rho^2)

for (j in 2:k) {
  beta.i[, j] <- beta.i[, j-1]*rho + beta.i[, j]*(1 - rho^2)
}
```

We then assign a year index, generate iid Gaussian noise, and combine our components to generate our observed values.
```{r}
# Define year index
Year = matrix(1:k, nrow=nrow(coo), ncol=k, byrow=TRUE)

set.seed(3)
# Error in the observation
error <- rnorm(n * k, 0, sqrt(1 / tau.e))

# Dynamic regression part
y <- alpha + (beta.G + beta.i) * Year + error

loc = tibble(idx=as.integer(1:nrow(coo)), East=coo[,1], North=coo[,2])
data = tibble(y=as.vector(y), idx=rep(loc$idx,k), 
              East=rep(loc$East,k), North=rep(loc$North, k), 
              Year=as.vector(Year), 
              B.i = as.vector(beta.i))
```

Here's the variability and range in trends for these data.
```{r}
ggplot(data, aes(x=Year, y=y, group=idx)) + geom_line( alpha=0.1) + geom_smooth(method='lm', alpha=0.2)
```

## Create model

```{r}
mesh <- inla.mesh.2d(coo, max.edge = c(1000, 5000), cutoff = 100)
plot(mesh)
points(loc %>% select(-idx), col = "red")

spde <- inla.spde2.pcmatern(mesh = mesh,
  prior.range = c(500, 0.01), # P(practic.range < 0.05) = 0.01
  prior.sigma = c(0.1, 0.01)) # P(sigma > 1) = 0.01
```

```{r}
yrdiff.idx <- inla.spde.make.index('yrdiff.idx', spde$n.spde, 
                                     n.group = max(unique(data$Year)))
A.yrdiff <- inla.spde.make.A(mesh, cbind(data$East, data$North), 
                             group = data$Year)#, weights = hh)

# A.spat <- inla.spde.make.A(mesh, cbind(data$East, data$North), 
#                            group = data$Year)
# spat.idx <- inla.spde.make.index('spat.field', spde$n.spde, 
#                                  n.group = max(unique(data$Year)))

# # Make data stack
# dat.stack <- inla.stack(
#   data = list(y = data$y),
#   A = list(1, 1, A.yrdiff, A.spat, 1),
#   effects = list(list(Intercept = rep(1, nrow(data))),
#                  tibble(Year=data$Year),
#                  yrdiff.idx, 
#                  spat.idx, 
#                  list(time=data$Year)),
#   tag = 'dat')

# Make data stack
dat.stack <- inla.stack(
  data = list(y = data$y),
  A = list(1, 1, A.yrdiff),
  effects = list(list(Intercept = rep(1, nrow(data))),
                 tibble(Year=data$Year),
                 yrdiff.idx),
  tag = 'dat')
```

```{r}
# # Still experimenting with with one
# form <- y ~ -1 + Intercept + Year + 
#   f(yrdiff.idx, data$Year, model = spde, group = yrdiff.idx.group, 
#     control.group = list(model = 'ar1'))

# This one I know works well!
form <- y ~ -1 + Intercept + Year + 
  f(yrdiff.idx, model = spde, group = yrdiff.idx.group, 
    control.group = list(model = 'ar1'))
```

## Fit model

```{r}
mod <- inla(form, family = 'gaussian',
  data = inla.stack.data(dat.stack), 
  control.predictor = list(A = inla.stack.A(dat.stack)), 
  control.compute = list(config =TRUE))
```

## Evalulate model performance

```{r}
summary(mod)
```

```{r}
dist.Bi = tibble(B.i = mod$summary.random$yrdiff.idx$mean, class = "Mod")
tmp = data %>% select(B.i) %>% mutate(class = "Obs")
dist.Bi = dist.Bi %>% bind_rows(tmp)

mod.marg = tibble(x = mod$marginals.fixed$Intercept[,1], 
                  y = mod$marginals.fixed$Intercept[,2], 
                  Param = "Intercept")
tmp = tibble(x = mod$marginals.fixed$Year[,1], 
                  y = mod$marginals.fixed$Year[,2], 
                  Param = "Year")
mod.marg = mod.marg %>% bind_rows(tmp)

hyper.names = c("Precision", "Range", "Stdev", "Rho")
for (i in 1:length(hyper.names)) {
  tmp = tibble(x=mod$marginals.hyperpar[[i]][,1], y=mod$marginals.hyperpar[[i]][,2], 
               Param=hyper.names[i])
  mod.marg = mod.marg %>% bind_rows(tmp)
}

param.vals = tibble(x = c(alpha, beta.G, tau.e, sp.range, sp.sigma, rho), 
                    Param = c("Intercept", "Year", hyper.names))

ggplot(mod.marg, aes(x=x,y=y)) + geom_line() + 
  geom_vline(data=param.vals, aes(xintercept = x), color='red') + 
  facet_wrap(vars(Param), scales = "free")

# ggplot(mod.marg %>% filter(Param=="Year")) + geom_line(aes(x=x,y=y)) + 
#   geom_vline(xintercept = mu.beta[2], color='red') + xlab("Global year coefficient")

ggplot(dist.Bi, aes(x=B.i, group=class, fill=class)) + 
  geom_density(alpha=0.3) + xlab("Deviations from fixed year coeff")

```

<!-- ### Experimental model -->

<!-- ```{r} -->
<!-- mesh <- inla.mesh.2d(coo, max.edge = c(1000, 5000), cutoff = 100) -->
<!-- plot(mesh) -->
<!-- points(loc %>% select(-idx), col = "red") -->

<!-- spde <- inla.spde2.pcmatern(mesh = mesh, -->
<!--   prior.range = c(500, 0.01), # P(practic.range < 0.05) = 0.01 -->
<!--   prior.sigma = c(0.1, 0.01)) # P(sigma > 1) = 0.01 -->

<!-- # Create projector matrix for SPDE mesh -->
<!-- A.spat = inla.spde.make.A(mesh = mesh, -->
<!--                           loc = cbind(data$East, data$North), -->
<!--                           group = data$Year) -->
<!-- A.yrdiff = inla.spde.make.A(mesh = mesh,  -->
<!--                             loc = cbind(data$East, data$North)) -->

<!-- # Assign index vectors for SPDE model. -->
<!-- yrdiff.idx = inla.spde.make.index(name = "yr.diff",  -->
<!--                                 n.spde = spde$n.spde) -->

<!-- spat.idx = inla.spde.make.index(name = "spat.field", -->
<!--                                 n.spde = spde$n.spde, -->
<!--                                 n.group = max(data$Year)) -->

<!-- # Make data stack -->
<!-- dat.stack <- inla.stack( -->
<!--   data = list(y = data$y), -->
<!--   A = list(1, 1, A.yrdiff, A.spat, 1), -->
<!--   effects = list(list(Intercept = rep(1, nrow(data))), -->
<!--                  tibble(Year=data$Year), -->
<!--                  yrdiff.idx,  -->
<!--                  spat.idx,  -->
<!--                  list(time=data$Year)), -->
<!--   tag = 'dat') -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # form <- y ~ -1 + Intercept + -->
<!-- #   Year -->

<!-- # form <- y ~ -1 + Intercept + -->
<!-- #   Year + f(yr.diff, Year, model = spde) -->

<!-- form <- y ~ -1 + Intercept + -->
<!--   Year + f(yr.diff, Year, model = spde) + -->
<!--   f(time, model = 'ar1') -->

<!-- ``` -->

<!-- ```{r} -->
<!-- res.dev <- inla(form, family = 'gaussian', -->
<!--   data = inla.stack.data(dat.stack),  -->
<!--   control.predictor = list(A = inla.stack.A(dat.stack)),  -->
<!--   control.compute = list(config =TRUE)) -->
<!-- ``` -->

### 2022-07-11 scratch/inla.errors.R

@@ -1,0 +1,60 @@
+# Script to reproduce weird instability errors in inla while using Gaussian model with log link
+
+## Import libraries
+library(tidyverse)
+library(INLA)
+
+# Generate synthetic exponential data
+x = seq(-25,25, length=100)
+A_0 = 300
+k = 0.035
+set.seed(0)
+e = rnorm(length(x), sd=2*k)
+y = A_0*exp(k*x+e)
+df1 = tibble(x=x, y=y, class="OBS")
+
+# Define formula
+formula = y ~ x
+
+# Fit inla model to data
+# prior.fixed = list(mean.intercept=250, prec=)
+inla.bad = inla(formula, data = df1 %>% mutate(Intercept=1), 
+                family = "Gaussian", 
+                control.family = list(control.link=list(model="log")), 
+                control.predictor = list(compute=TRUE, link=1))
+
+# Fit GLM to data (for comparison)
+mod.glm1 = glm(formula, data = df1 %>% mutate(Intercept=1), 
+               family = gaussian(link = "log"))
+
+# Add results to df and plot
+df1 = df1 %>% 
+  bind_rows(tibble(x=x, y=mod.glm1$fitted.values, class="GLM")) %>% 
+  bind_rows(tibble(x=x, y=inla.bad$summary.fitted.values$mean, class="INLA"))
+ggplot(df1) + 
+  geom_line(data=df1 %>% filter(class == "GLM"), aes(x=x, y=y), color='red', alpha=0.33) + 
+  geom_point(aes(x=x,y=y, group=class, color=class), alpha=0.33)
+
+## Rerun with different seed
+
+# Regenerate synthetic data
+set.seed(2)
+e = rnorm(length(x), sd=2*k)
+y = A_0*exp(k*x+e)
+df2 = tibble(x=x, y=y, class="OBS")
+
+# Refit models
+inla.good = inla(formula, data = df2 %>% mutate(Intercept=1), 
+                 family = "Gaussian", 
+                 control.family = list(control.link=list(model="log")), 
+                 control.predictor = list(compute=TRUE, link=1))
+mod.glm2 = glm(formula, data = df2 %>% mutate(Intercept=1), 
+               family = gaussian(link = "log"))
+
+# Add results to df and plot
+df2 = df2 %>% 
+  bind_rows(tibble(x=x, y=mod.glm2$fitted.values, class="GLM")) %>% 
[…]

### 2022-07-11 scripts/topo.calc.R

@@ -1,0 +1,110 @@
+# Script to perform raster calculations and extraction for REMA topographic values
+
+# Import libraries
+library(here)
+library(tidyverse)
+library(sf)
+library(terra)
+
+## Find which REMA DEM 8-meter tiles intersect our dataset and download any tiles not
+## already present in the selected directory "output.dir"
+
+# Select directory containing REMA mosaic tiles (based on currently-running OS)
+os = Sys.info()['sysname']
+if (os == 'Linux') {
+  REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
+} else {
+  REMA.dir = file.path("G:/Research/Antarctica/Data/DEMs/REMA/tiles_8m_v1.1")
+}
+# REMA.dir = tk_choose.dir(caption = "Select directory in which to save REMA mosaic tiles")
+
+# Find all mosaics that intersect the accumulation dataset
+index.file = file.path(dirname(REMA.dir), "REMA_Tile_Index_Rel1.1/REMA_Tile_Index_Rel1.1.shp")
+dem.index = st_read(index.file) %>% st_transform(crs = st_crs(gdf_traces))
+# ggplot() + geom_sf(data = Ant.outline, color = 'grey') + geom_sf(data = dem.index)
+keep.idx = unique(unlist(st_intersects(gdf_traces, dem.index)))
+mosaic.download = dem.index[keep.idx,] %>% mutate(fileurl = as.character(fileurl))
+
+# Download files from REMA ftp server
+for (file in 1:nrow(mosaic.download)) {
+  tmp.file = tempfile()
+  dir.name = basename(dirname(mosaic.download$fileurl[file]))
+
+  # If not already downloaded, download tile and extract
+  if (!dir.exists(file.path(REMA.dir, dir.name))) {
+    download.file(mosaic.download$fileurl[file], tmp.file)
+    untar(tmp.file, exdir = file.path(REMA.dir, dir.name))
+  }
+  unlink(tmp.file)
+}
+
+
+## Generate aspect and slope tiffs for any required REMA tiles, extract slope/aspect
+## values for all accum points, and add data to data tibbles
+# 
+# Function to find which accum points are within a given REMA 8-meter mosaic tile 
+# and extract elevation, slope, and aspect from the REMA raster data
[…]

### 2022-07-11 scripts/article-notebook.Rmd
Adds notebook for paper draft

---
title: "Publication notebook"
output: html_notebook
---

This includes the analysis and figures that will go into the draft publication of the central WAIS trends article.
This will also be a space for drafting and iterating on the wording for methods/analysis/interpretation description.

## METHODS

### Data imports and preprocessing

Load libraries:
```{r}
library(here)
library(INLA)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(colorspace)
library(spdep)
```

```{r}
data.all = readRDS(here('data/Rdata-clean.rds'))
gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))

# # Clip bounds
# bbox = st_bbox(gdf_traces)
# clipper = c(
#   bbox[1]+0.1*(bbox[3]-bbox[1]),
#   bbox[2]+0.5*(bbox[4]-bbox[2]),
#   0.6*(bbox[3]-bbox[1])+bbox[1], 
#   bbox[4]
# )
# gdf_traces = st_crop(gdf_traces, clipper)

# Remove locations with less than 10 years of data
trace.keep = data.all %>% count(trace_ID) %>% filter(n>=10)
data.all = data.all %>% filter(trace_ID %in% trace.keep$trace_ID)
gdf_traces = gdf_traces %>% filter(trace_ID %in% trace.keep$trace_ID)

# Subset data to every 5th location (coarsens data to ~1 km)
skip.int = 5
gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
gdf_traces = gdf_traces[gdf.idx,]

# Filter data to subsetted period
data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
  filter(Year >= 1975) %>% filter(Year < 2015) %>% arrange(trace_ID, Year)

# Remove gdf rows where all data have been filtered out
gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)

# Select variables of interest
dat = data %>% select(trace_ID, East, North, Year, accum, 
                      std, SMB, elev, dem, slope, aspect, 
                      # u10m, v10m, mu.u10, mu.v10
                      )

# Center covariates
dat = dat %>% 
  mutate(elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
         ) %>% 
  mutate(Year.mod = Year-mean(Year), Year.idx = (Year-min(Year)+1))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), 
         # u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), mu.u10=sd(mu.u10), mu.v10=sd(mu.v10)
         )

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

Directly calculate linear coefficients of time from raw data.
```{r}
yr.trends = dat.train %>% 
  group_by(trace_ID) %>% 
  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
  filter(term=='Year.mod') %>% select(-term)
dat.mu = dat.train %>% group_by(trace_ID) %>% 
  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
  mutate(log.est = log(1+(estimate/accum)))
```

### Model setup

The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension (e.g. annually resolved time series).
We therefore choose to model accumulation as a Gamma distribution to ensure a zero-bounded function with increased variance at higher expected values as also observed in the raw data.
This furthermore matches established literature where precipitation values are frequently assumed to follow a Gamma distribution **(citations)**.
We express this distribution in terms of the expected accumulation for an arbitrary space location and time point $\mu_{s,t}$ and a precision parameter at the corresponding time $\phi_t$ 
$$\mathbf{Y} \sim Gamma(\mu_{s,t},\phi_t)$$
where the variance at an arbitrary spacetime point is
$$Var(s_t) = \frac{\mu_{s,t}^2}{\phi_t}$$

For each observation $y_{s,t} \in \mathbf{Y}$, we construct a linear predictor $\eta_{s,t}$ associated with the expectation value $\mu_{s,t}$ via a log link function.
$$\log(\mu_{s,t}) = \eta_{s,t}$$

We further model $\eta_{s,t}$ as a generalized linear equation consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case modeling a spatial random effect $u(\cdot)$ and a temporal random effect $\omega(\cdot)$).
To explicitly model $u(\cdot)$ as a trend in accumulation over time, we further index this random variable by year.
The equation for our generalized linear model is therefore:
$$
\eta_{s,t} = \alpha + \sum^M_{m=1} \beta_m x_{m,s} + u(s)\times t + \omega(t)
$$

This model therefore accounts for fixed global effects, a spatially-varying latent effect indexed on time (i.e. providing explicit estimates of trends in annual acculation), and a temporally-varying latent effect.

The latent random effects $u$ and $\omega$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
We model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of the hyperparameters $\pmb{\psi}$ and an associated prior distribution $p(\pmb{\psi})$.
This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{jk} = Cov(\theta_j, \theta_k)$.

$$
\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
$$

By assuming adherence to Markovian properties, the precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of the process, i.e. $Q_{jk} \neq 0 \iff k \in \{j,N(j)\}$.
The specification of a Gaussian-Markov Random Field for both the spatial and temporal structures permits vast improvements in computational efficiency.
The choice of a separable space/time model (where the spatial and temporal random effects are modeled independently) also greatly improves computation time but will fail to capture any co-dependencies between these effects.

We model the spatial random effect as a stochastic partial differential equation...

We model the temporal random effect $\omega$ (representing how the mean spatial field evolves in time) as a second-order autoregressive model where
$$\omega_t = \rho \omega_{t-1}$$
An AR(2) process was selected due to the observed temporal autocorrelation in the data, which indicates that accumulation retains autocorrelation up to a 2-year lag time.

The modeling objective, therefore, is to find the joint posterior distribution of the various parameters of interest, permitting inference over the processes governing accumulation in the region and making predictions at unobserved locations possible.
$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s \mid \mathbf{Y})$$

```{r}
# More coarse mesh for development purposes
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
                    max.edge = c(25000, 100000), cutoff = 1000)
# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
#                     max.edge = c(30000, 90000), cutoff = 2000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

# 2D spatial Matern GMRF via SPDE (with assigned priors)
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(25000, 0.01), # P(range < 20 km) = 0.01
  prior.sigma = c(1, 0.01)) #P(sd > 1) = 0.05

# Locator index for spatial random effect
spat.idx <- inla.spde.make.index('spat.idx', spde$n.spde)

# Projector matrix for spatial locations (and indexed on Year for trends)
A.spat <- inla.spde.make.A(mesh, 
                         loc=cbind(dat.train$East, dat.train$North), 
                         weights = dat.train$Year.mod)

# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect),
                 list(time = dat.train$Year.mod),
                 spat.idx),
  tag = 'dat')

# Prior for temporal autocorrelation
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))

# Model formula
f.mod = y ~ -1 + Intercept + #Global intercept
  dem + #Fixed effects
  # f(dem, model='clinear', range=c(-Inf,0)) + #Fixed effects
  f(time, model = 'ar1', hyper = time.spec) +  #Temporal random effect modeled as AR1
  f(spat.idx, model = spde) #Spatial randomed effect (indexed by Year)
```

Run and summarize model
```{r}
# Model formula
mod = inla(f.mod,
           data = inla.stack.data(dat.stack),
           family = 'Gamma',
           control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
           # control.inla = list(int.strategy='eb'), #Improves comp time at cost of accuracy/precision
           control.compute = list(waic=TRUE, config=TRUE))

n.iter = 1
while (mod$mode$mode.status>0 && n.iter<=3) {
  print("Issues with Hessian. Re-running to solve negative eigenalues in the Hessian")
  n.iter = n.iter + 1
  print(paste("Starting Iteration", n.iter, "of INLA..."))
  mod = inla.rerun(mod)
}

if (mod$mode$mode.status > 0) {
  print("Issues with Hessian eigenvalues persist. Treat model results with suspicion!")
}
```

```{r}
summary(mod)
```

## Model results

```{r}
# Helper function to extract and transform posteriors
t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {

  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }

  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }

    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }

  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))

  marg.names = names(mod.marg)

  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod$marginals.fixed, function(x) exp(x))
HP.names = c("Gamma precision", 
             "Time precision", "Rho for time", 
             "Range for spat.idx", "Stdev for spat.idx")
marg.hyper = t.marg(mod$marginals.hyperpar, function(x) (x), param.names = HP.names)
p.selection = c("Intercept", "dem", "Gamma precision", 
                "Time precision", "Rho for time", 
                "Range for spat.idx", "Stdev for spat.idx")

posts = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
```

```{r posteriors, fig.width=16}
ggplot(posts, aes(x=x,y=y)) + geom_line() + 
  facet_wrap(vars(Param), scales = "free") + 
  theme(text = element_text(size = 20))  
```

### Components of prediction

```{r}
# Define points of interest to break out components of prediction
pred.trace = gdf_traces %>% mutate(color=as.factor(trace_ID))

# Get raw trend calculations and variable values at locations of interest
dat.pred = dat %>% filter(trace_ID %in% pred.trace$trace_ID) %>% mutate(color=as.factor(trace_ID))
Years = sort(unique(dat.train$Year))
loc.pred = st_coordinates(pred.trace) %x% rep(1, length(Years))
covar.pred = dat.pred %>% select(trace_ID, dem) %>%
  group_by(trace_ID) %>% filter(row_number()==1)
yr.mod = rep(Years - mean(Years), nrow(pred.trace))
st.pred = tibble(X=loc.pred[,1], Y=loc.pred[,2], Year = rep(Years, nrow(pred.trace)), 
                 dem=rep(covar.pred$dem, each=length(Years)))

# Project prediction locations onto mesh
proj.pred = inla.mesh.projector(mesh=mesh, loc = loc.pred)

# Extract spatial RE of trend and time RE at mesh nodes
spat.trend = inla.mesh.project(proj.pred, mod$summary.random$spat.idx$mean)
time.RE = rep(mod$summary.random$time$mean, nrow(pred.trace))

# Construct linear predictor for each spacetime point
lin.est = mod$summary.fixed$mean[1] + 
  mod$summary.fixed$mean[2]*st.pred$dem + spat.trend*yr.mod + time.RE

# Get model components for each st point
st.pred = st.pred %>% 
  mutate(fx.RE=mod$summary.fixed$mean[2]*st.pred$dem, spat.trend=spat.trend*yr.mod, 
         time.RE=time.RE, pred=exp(lin.est)) %>% 
  mutate(pred.sd=sqrt(pred^2/mod$summary.hyperpar$mean[1]), Source="model", 
         color=rep(pred.trace$color, each=length(Years)))

# Average data for each spatial location and add trend estimates
spat.pred = st.pred %>% select(-Year, -spat.trend, -Source, -time.RE) %>% 
  group_by(color) %>% summarize_all(mean, rm.na=TRUE)
spat.pred['trend'] = spat.trend[seq(1,length(spat.trend), 
                                    by=(length(spat.trend)/nrow(spat.pred)))]
```

```{r}
ggplot(spat.pred, aes(x=X,y=Y, color=pred)) + geom_point() + scale_color_viridis_c()
ggplot(spat.pred, aes(x=X,y=Y, color=fx.RE)) + geom_point() + scale_color_gradient2()
ggplot(spat.pred, aes(x=X,y=Y, color=trend)) + geom_point() + scale_color_gradient2()
```

Direct comparisons for individual trace time series between observed and predicted
```{r}
pred.comp = dat.pred %>% select(color, Year, accum, std, SMB) %>% 
  right_join(st.pred, by=c("color", "Year")) %>% 
  mutate(log.diff=log(pred/accum))

ggplot(pred.comp) + 
  geom_line(aes(x=Year, y=log.diff, group=color), size=0.25, alpha=0.1) + 
  theme(legend.position="none")
ggplot(pred.comp) + geom_abline(slope=1, color='black') + 
  geom_point(aes(x=accum, y=SMB), color='red', alpha=0.1) + 
  geom_point(aes(x=accum, y=pred), color='blue', alpha=0.1)
ggplot(pred.comp) + geom_density(aes(x=log(pred/accum)), fill='red', alpha=0.3)
```

```{r}
mod.time = tibble(Year=Years, mu=mod$summary.random$time$mean, 
                  sd=mod$summary.random$time$sd)
tmp = pred.comp %>% group_by(Year) %>% summarize(accum=mean(accum, na.rm=TRUE), pred=mean(pred))
mod.time = mod.time %>% mutate(obs.diff=log(tmp$accum/mean(tmp$accum)), 
                               pred.diff=log(tmp$pred/mean(tmp$pred)))
ggplot(mod.time) + geom_hline(yintercept = 0, color='black', linetype="dashed") + 
  geom_ribbon(aes(x=Year, ymin=mu-sd, ymax=mu+sd), fill='red', alpha=0.2) + 
  geom_line(aes(x=Year, y=mu), color='red') +
  geom_line(aes(x=Year, y=obs.diff), color='blue', alpha=0.5) + 
  geom_line(aes(x=Year, y=pred.diff), color='purple', alpha=0.5)
```

Calculate probability that trend is less than zero.
```{r}
P.neg = vector(mode = "numeric", length = length(mod$marginals.random$spat.idx))
for (i in 1:length(P.neg)) {
  marg.i = inla.tmarginal(function(x) x, 
                          mod$marginals.random$spat.idx[[i]], 
                          n=500)
  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
}
tmp = tibble(Source='model', 
             East=mesh$loc[,1], North=mesh$loc[,2], 
             med=mod$summary.random$spat.idx$`0.5quant`, 
             P.neg=P.neg, 
             LB=mod$summary.random$spat.idx$`0.025quant`, 
             UB=mod$summary.random$spat.idx$`0.975quant`) %>% 
  st_as_sf(coords=c("East", "North"), crs=3031)
```

Map of estimated trends with probability of less than zero
```{r}
trends.comp = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
  rename(med=log.est) %>% 
  mutate(LB=NA, UB=NA, P.neg=NA, Source="data") %>% 
  st_as_sf(coords=c("East", "North"), crs=3031) %>%
  bind_rows(tmp)
```

```{r}
ggplot(trends.comp) + geom_density(aes(x=med, group=Source, color=Source)) + 
  xlim(c(-0.07,0.07))

plt_bnds = st_bbox(gdf_traces)
ggplot(trends.comp, aes(color=med)) + 
  geom_sf() + scale_color_gradient2() + 
  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
  facet_wrap(vars(Source))

ggplot(trends.comp %>% filter(Source!="data")) +
  geom_sf(aes(color=(UB-LB)/2)) + 
  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
  scale_color_viridis_c(option = "plasma", limits=c(0.003, 0.013))
  # scale_color_viridis_c(option = "plasma", trans="log")

ggplot(trends.comp %>% filter(Source!="data")) +
  geom_sf(aes(color=P.neg)) + 
  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
  scale_color_continuous_diverging(palette="Tropic", mid=0.50)

ggplot(trends.comp %>% filter(Source!="data")) + geom_density(aes(x=P.neg), fill='red', alpha=0.3)
```

```{r}
mod.trends = trends.comp %>% filter(Source=="model")
Neg.threshold = 0.95
frac.Neg = nrow(mod.trends %>% filter(P.neg>=Neg.threshold))/nrow(mod.trends)
print(paste(
  format(100*frac.Neg, digits=2), 
  "% of the region has at least ", 
  format(100*Neg.threshold, digits=2), 
  "% probability of a negative accumulation trend", 
  sep=''))

frac.Pos = nrow(mod.trends %>% filter(P.neg<=(1-Neg.threshold)))/nrow(mod.trends)
print(paste(
  format(100*frac.Pos, digits=2), 
  "% of the region has at least ", 
  format(100*Neg.threshold, digits=2), 
  "% probability of a positive accumulation trend", 
  sep=''))
```

### 2022-07-20 scripts/article-notebook.Rmd
Adds updates/modifications to notebook

@@
 output: html_notebook
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
+## INTRODUCTION
+
+The West Antarctic Ice Sheet (WAIS) is highly relevant to our understanding of Southern Hemisphere climate dynamics and to projections of future sea level.
+Already the WAIS is contributing to global sea level changes, having added at least an additional $6.9 \pm 0.6$ mm since 1979 \cite{rignot_four_2019}.
+An understanding of snow accumulation patterns and processes is an essential component of sea level rise calculations, both directly as a contribution to overall mass balance and indirectly, as the surface mass balance (SMB) affects estimates of ice sheet net mass change from other methods---e.g.
+altimetry and gravimetry measurements \cite{velicogna_timevariable_2013, shepherd_trends_2019}.
+Such effects on the net mass change are nontrivial and changes in accumulation can even dominate the integrated height change signal in at least some circumstances \cite{kuipersmunneke_elevation_2015}.
+Additionally, understanding the recent spatio-temporal trends and patterns in precipitation over recent decades provides insight into how atmospheric and oceanic dynamics are changing in the Southern Hemisphere and globally.
+
+Despite the critical need for increased understanding, our knowledge of West Antarctic SMB is hampered by a paucity of data and data coverage \citep{koenig_remote_2014}.
+The extremely remote and harsh environment of the region make \textit{in-situ} measurements challenging and expensive to collect.
+SMB is often highly spatially variable \cite{spikes_variability_2004, banta_spatial_2008, dattler_significant_2019}, leading to questions of how representative small samples of measurements are of the ice sheet as a whole.
+Although satellite-based radar altimeters provide widespread spatial coverage, they typically have coarse resolutions of tens of kilometers and the degree of signal penetration beneath the surface complicates interpretations of height changes \cite{lenaerts_observing_2019}.
+More recent lidar altimeters address some of these limitations, but have a limited temporal range \cite{smith_pervasive_2020}.
+Both radar and lidar altimetry results are further complicated by the need to disentangle signals between changes in SMB, dynamical ice flux, firn compaction, and isostatic uplift.
+Modeling results have seen enormous strides forward, but validation with independent estimates of SMB are still important, especially in data-sparse regions like West Antarctica \citep{lenaerts_observing_2019}.
+Additionally, models still have too coarse of resolution (often 10's of kilometers) to resolve small-scale (\textless5 km) spatial variability \cite{lenaerts_climate_2018} and still have limitations on how well they can capture interannual variability in Antarctic SMB \cite{wang_comparison_2016}.
+Radar has been used for decades to estimate SMB \cite{spikes_variability_2004, anschutz_smallscale_2008, eisen_groundbased_2008}, but still suffers from limitations of homogeneous coverage and processing bottlenecks usually involving labor-intensive manual work, especially for reconstructing annually-resolved SMB time series \cite{medley_airborneradar_2013, lenaerts_observing_2019}.
+
+In spite of these difficulties, substantial progress has been made towards improved understanding of SMB over recent decades in the WAIS.
+Numerous ice core studies suggest accumulation rates in and near the Antarctic Peninsula have recently increased \cite{thomas_doubling_2008, thomas_308_2013, thomas_twentieth_2015, hosking_accumulation_2017, medley_increased_2019}.
+This observation is supported by results from a number of climate models and reanalyses \cite{wang_comparison_2016}.
+Evidence also exists for decreased accumulation over similar time periods for the western portions of WAIS, near the Ross Sea \cite{wang_comparison_2016, wang_snow_2017, medley_increased_2019}, although the \textit{in-situ} observations in this region are more sparse and largely out of date \cite{thomas_regional_2017}.
+As a whole, the central region of WAIS shows smaller changes in snow accumulation, but such results vary considerably both spatially across central WAIS and between studies \cite{banta_spatial_2008, medley_airborneradar_2013, burgener_observed_2013, wang_comparison_2016, thomas_regional_2017, wang_snow_2017}.
+
+Important questions therefore remain regarding the spatial distribution of accumulation trends across WAIS and particularly regarding the nature and location of the transition in central WAIS between the predominantly increasing accumulation of eastern WAIS and the Antarctic Peninsula and the decreasing accumulation in western WAIS.
+Understanding these transitions are important to properly interpret the climatic elements driving the observed changes and to facilitate improved ice sheet-wide estimates of trends in annual SMB.
+
+In this study we construct annually-resolved SMB time series for a region of central WAIS from airborne radar at high spatial resolution and investigate the subsequent spatio-temporal patterns and trends in the data.
+The method for estimating SMB time series from radar is fully automated, includes uncertainty estimation for individual years and locations, and includes a quality-control routine that can automatically remove data poorly suited to the method.
+We use these data to inform a Bayesian statistical model that can leverage the data uncertainties and extend estimates to unobserved spacetime points across the region.
+This modeling therefore permits a more spatially detailed and staistically robust investigation of recent changes in SMB across central WAIS.
+
 ## METHODS
 
-### Data imports and preprocessing
+### Data and study site
+
+This study focuses on a \mytilde250,000 km\textsuperscript{2} region of central West Antarctica (\figref{data-map}).
+The region spans a range of accumulation rates (\textless200 mm w.e.
+to \textgreater750 mm w.e.) and topographic settings, encompassing a portion of the ice divide between the Amundsen Sea and the Ross Sea sectors.
+Although some ice velocities in the coastal areas of this region can exceed 200 m/a, typical \emph{in-situ} velocities for our data coverage are \mytilde20 m/a or less \citep{rignot_ice_2011}.
+One reason this region is significant is because it encompasses portions of the Pine Island Glacier and Thwaites Glacier drainage basins.
+These outlet glaciers had a net mass loss of 1,700 gigatonnes of ice for the period 1979-2017, with the annual contribution also accelerating over that time period \cite{rignot_four_2019}.
+This region additionally has an acceptably dense network of firn density measurements (a total of 16 cores) to properly constrain depth-density modeling (discussed in Section \ref{sec:density}) with increased confidence.
+Finally, this region is between the Antarctic Peninsula---where strong evidence exists of increased accumulation rates over recent decades---and the region near the Ross Sea---where evidence exists for decreased accumulation rates over recent decades \citep[e.g.]{medley_increased_2019, thomas_regional_2017}.
+It is therefore relevant to understanding the extent, boundaries, and patterns of changes in accumulation rates for the entire WAIS.
+
+The main data used in this study are radar echograms from NASA's Operation IceBridge (OIB) Snow Radar.
+This frequency-modulated continuous-wave radar operates in a frequency range (2-8 GHz) with a range resolution (\mytilde5 cm in firn) that permits imaging of annual resolution in Antarctic firn \cite{panzer_ultrawideband_2013}.
+We use portions of 9 OIB flights over the region of interest with collection times spanning 2010-2016.
+
+Estimating annual accumulation from radar echograms relies on the ability of microwave electromagnetic radiation to penetrate the ice sheet surface and propagate through the snow and firn.
+Seasonal variations in the dielectric properties of the preserved firn result in internal reflection horizons (IRHs) in the radar returns.
+When continuous over many kilometers, these IRHs are most often interpreted as isochronous seasonal layers that can be used to determine the age of deposition \cite{eisen_groundbased_2008}.
+To determine accumulation rates, the echogram time-domain must be converted to depth.
+This requires estimates of firn density with depth to determine radar propagation velocity.
+The firn depth-density profile is also used when converting layer thickness to cumulative mass with depth.
+The cumulative mass profile combined with the calculated ages of the IRHs yields annual estimates of accumulation.
+
+In addition to the radar data used, this study utilizes a suite of ice and firn cores, both for informing the the depth-density model and for comparisons of accumulation rates and trends generated with the radar data.
+We use density data from the Satellite Era Accumulation Traverse \cite{burgener_observed_2013}, the WAISCORES program \cite{lamorey_waiscores}, the International Trans-Antarctic Scientific Expedition \cite{mayewski_international_2005}, and additional cores included in the Surface Mass Balance and Snow on Sea Ice Working Group data set \cite{montgomery_sumup_2018}.
+For comparisons of annual accumulation rates and trends we additionally use cores included in the Past Global Changes Antarctica 2k compilation \cite{thomas_regional_2017}.
+
+### Depth-density modeling
+
+The calculation of radar accumulation time series requires estimates of depth-density profiles to convert radar echograms from the time domain into the depth domain and to estimate annual accumulation from layer thickness.
+Various methods have been proposed for estimating the variation of density with depth in Antarctic firn, typically either interpolating from measured ice core or stake farm data \cite{medley_airborneradar_2013} or based on the physical modeling and evolution of firn properties \cite{ligtenberg_improved_2011}.
+
+In this study, we use a statistical estimation and interpolation approach informed by \emph{in-situ} observations as described in \citet{white_hierarchical_2021}.
+This model uses a positive-valued Bayesian semi-parametric model over space and depth below the ice sheet as its building block.
+To impose the monotone increasing shape of the depth-density curve, we integrate the positive-valued function over the depth domain.
+The integrated function is transformed to be bounded below by 0 g/cm$^3$ and 0.917 g/cm$^3$ to constrain the model's mean function to appropriate values for firn/ice density.
+This model imposes smoothness of firn density estimates in the space and depth domains, constrains the shape of firn density curves, and enables simple interpolation (see \citet{white_hierarchical_2021} for more details).
+There are several advantages afforded by this method, most importantly its ability to leverage space-depth autocorrelation to interpolate the shape-constrained depth-density profiles to new locations with relatively few model assumptions.
+Moreover, because we fit the model in a Bayesian framework, our method provides rigorous uncertainty quantification of the depth-density profiles at arbitrary sites.
+This allows us to effectively propagate uncertainty in firn density estimates to SMB estimates.
+
+### SMB estimation with PAIPR
+
+In this study, we estimate annually-resolved accumulation rates from OIB Snow radar using an automated process, the Probabilistic Annual Isochrone Picking Routine---i.e.
+PAIPR \cite{keeler_probabilistic_2020}.
+This algorithm picks continuous IRHs using localized Radon transforms and assigns each a likelihood of representing an annual isochrone based on the relative strength of the radar return and the continuous length of the horizon.
+These likelihoods, along with errors in the depth-density modeling (discussed above), are then used in Monte Carlo sampling to generate uncertainty estimates in the the final annual accumulation estimates.
+The algorithm processes radar data in 30-km chunks, generating results with 25-meter along-track spacing which we aggregate to 200-meter, annually-resolved accumulation distributions.
+We limit analysis to the upper \mytilde25 meters of the firn to ensure no issues with loss of layers due to excessive attenuation of IRH signals with depth.
+The complete period of data coverage therefore varies spatially due to varying accumulation rates, spanning from ten years to fifty-seven years of total record duration.
+Further details of the method are available in \citet{keeler_probabilistic_2020} and \citet{keeler_repeatability_2021}.
+
+The primary advantages of the PAIPR method over more traditional techniques are the fully automated nature of the method and the more complete uncertainty assessment.
+The automated nature of PAIPR allows a wide and rapid application to larger OIB data sets.
+The addition of the quality control routine (see \citet{keeler_repeatability_2021} for details on its implementation) permits this application without careful vetting of input data, as poorly-suited data is flagged and can be removed from additional analyses.
+The inclusion of uncertainties (resulting from both input data and modeling errors) allows a more rigorous assessment of the relevance and significance of observed magnitudes, patterns, and trends.
+These uncertainies are further utilized in our Bayesian modeling framework (see Section ?).
+
+The PAIPR results still have important limitations worth noting.
+The degradation in signal with depth limits the time period over which reliable results are generated, with the temporal duration controlled by the \emph{in-situ} accumulation rate.
+PAIPR cannot produce time series of long duration in high accumulation regions, limiting interpretations of temporal variability and trends in these regions.
+Additionally, the method is reliant on nearby depth-density measurements or modeling, affecting results both in the depth migration steps and in the cumulative mass estimates.
+Reliable SMB estimates therefore rely on sufficient firn/ice core coverage in the region of interest, or else reliable firn density modeling where \emph{in-situ} measurements are not available.
+The method also does not account for dynamical thinning of annual layers.
+The shallow firn column used in PAIPR (\mytilde25 m) limits the impact of this issue, but such effects could still be relevant in regions of high ice velocity (e.g.
+\citet{mouginot_continentwide_2019} show some regions of WAIS near the coast can have \textgreater 1 km a\textsuperscript{-1} ice velocities).
+\citet{konrad_observations_2019} however, found that even for the region of Pine Island Glacier, \mytilde80% of strain rate correction magnitudes were lower than the uncertainty in radar-derived SMB estimates in the region.
+
+The layer-likelihood parameters in PAIPR were trained using manually-traced layers and validated against firn core data.
+The results could therefore be biased towards the specific regions used to train the likelihood parameters.
+The training data were selected to contain a spatially-distributed sample of central WAIS with a range of accumulation rates and conditions.
+This variability in training data, combined with the MC simulations on the uncertainty in the tuned parameters, should make this method applicable to a wide range of conditions, but would be less suited to regions of Antarctica significantly different than the range of conditions represented in the training data.
+Finally, the SMB results are still limited by data availability.
+NASA's Operation IceBridge focused on specific regions and research goals, affecting the selection of flight plans and regions of study.
+This results in some areas with a high density of data coverage (both spatially and temporally) with much more limited or no coverage in other regions (e.g.
+interior portions of the Antarctic ice sheets).
+Even with such limitations, the spatial coverage afforded by OIB SMB results is orders of magnitude greater than traditional firn/ice core, automatic weather stations, or stake farm measurements, regions poorly suited to PAIPR application are automatically filtered from results, and rigorous uncertainties are generated for SMB estimates.
+
+### Data processing and modeling
+
+The final output from PAIPR includes estimates of the modal value and standard deviation of accumulation for each year of a given trace (25-meter along-track spacing, aggregated into 200-meter bins to increase estimate sample size) and a quality rating for whether the results are usable based on the suitability of PAIPR to the particular echogram.
+We remove time series with less than 10 years of estimates, as we frequently observed inaccurate results for regions with few time series elements.
+To avoid biases in early year estimates (i.e. the earliest years are only available for regions of low accumulation rates), we focus on study on the time period 1975--2015.
+This 40-year time period exhibits good spatial coverage across the full region.
+After excluding data flagged as poor and filtering the remaining data with the above criteria, the final dataset consists of 683,676 individual estimates of annual accumulation at 21,887 spatial locations.
+<!-- After excluding data flagged as poor and locations with limited temporal coverage, the final dataset consists of 703,215 individual estimates of annual accumulation at 21,887 spatial locations spanning some portion of the period 1959--2015.
+-->
+
+<!-- Due to the decreased temporal coverage with increasing accumulation rates, the exclusion criteria effectively limit our analysis to regions with annual accumulation less than \mytilde650 mm w.e.
+-->
+
+<!-- Although the remaining time series vary in total duration from 10--57 years, the estimated mean accumulation for the full records change negligibly between using the full time period and using a consistent 40-year time period of 1975--2015 (see \figref{accum-map} for additional details on this comparison).
+-->
+
+<!-- This matches the findings of \citet{wang_comparison_2016}, who found the effect of changing accumulation rates over several decades was small compared to spatial variability. -->
+
+<!-- We therefore ignore the temporal heterogeneity in our mean accumulation investigations in order to improve spatial coverage and extent. -->
+
+<!-- This results in 4,008 unique time series corresponding to different 1-km grid cells. -->
+
+<!-- For trend analyses, we subset the grid cells to time series with complete coverage in the 31-year period 1979-2010. -->
+
+<!-- This time period was selected as a compromise between longer temporal and greater spatial coverage. -->
+
+<!-- The final data set used in trend analyses consists of 2,091 individual time series for 1-km grid cells. -->
+
+<!-- Accumulation time series exhibit high interannual variability. -->
+
+<!-- This, combined with the uncertainty in the estimates of individual years and the 31-year time period, could produce spurious trends in the results \cite{rupper_effects_2015}. -->
+
+<!-- To minimize this impact, we calculate linear trends in accumulation time series using bootstrapping with replacement and weights based on the estimated uncertainty of individual annual estimates.
+-->
+
+<!-- We perform 1,000 iterations to approximate a continuous distribution of trends to generate more robust estimates of the mean trend result and uncertainty in those estimates.  -->
+
+#### Bayesian model
+
+We use a Bayesian approximation method to investigate the spatial variability in the temporal trend in accumulation rates across the study region.
+This choice confers a number of benefits.
+The explicit and joint-probability uncertainty in a Bayesian approach allows more full utilization of the PAIPR-generated uncertainties and provides uncertainty for every parameter and estimate within the model.
+A Bayesian approach also incorporates domain-specific a priori knowledge into model estimates.
+The use of an approximation method vastly improves computation time and makes analyses of large datasets (e.g. times series generated from PAIPR) computationally tractable.
+We utilize Integrated Nested Laplace Approximations INLA) to leverage the noted advantages.
+This method has recently seen widespread adoption in modeling spatiotemporal physical phenomena in a variety of fields \cite{}.
+
+*Will move most of the following to a supplement, with only a brief summary included in the main document.*
+
+The PAIPR-derived accumulation time series represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension (i.e.
+individual years).
+We therefore choose to model accumulation as a Gamma distribution to ensure a zero-bounded function with increased variance at higher expected values as also observed in the raw data.
+This furthermore matches established literature where precipitation values are frequently assumed to follow a Gamma distribution **(citations)**.
+We express this distribution in terms of the expected accumulation for an arbitrary space location and time point $\mu_{s,t}$ and a precision parameter at the corresponding time $\phi_t$ $$\mathbf{Y} \sim Gamma(\mu_{s,t},\phi_t)$$ where the variance at an arbitrary spacetime point is $$Var(s_t) = \frac{\mu_{s,t}^2}{\phi_t}$$
+
+For each observation $y_{s,t} \in \mathbf{Y}$, we construct a linear predictor $\eta_{s,t}$ associated with the expectation value $\mu_{s,t}$ via a log link function.
+$$\log(\mu_{s,t}) = \eta_{s,t}$$
+
+We further model $\eta_{s,t}$ as a generalized linear equation consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case modeling a spatial random effect $u(\cdot)$ and a temporal random effect $\omega(\cdot)$).
+To explicitly model $u(\cdot)$ as a trend in accumulation over time, we further index this random variable by year.
+The equation for our generalized linear model is therefore: $$
+\eta_{s,t} = \alpha + \sum^M_{m=1} \beta_m x_{m,s} + u(s)\times t + \omega(t)
+$$
+
+This model therefore accounts for fixed global effects, a spatially-varying latent effect indexed on time (i.e.
+providing explicit estimates of trends in annual acculation), and a temporally-varying latent effect.
+
+The latent random effects $u$ and $\omega$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
+We model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of the hyperparameters $\pmb{\psi}$ and an associated prior distribution $p(\pmb{\psi})$.
+This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{jk} = Cov(\theta_j, \theta_k)$.
+
+$$
+\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
+$$
+
+By assuming adherence to Markovian properties, the precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of the process, i.e.
+$Q_{jk} \neq 0 \iff k \in \{j,N(j)\}$.
+The specification of a Gaussian-Markov Random Field for both the spatial and temporal structures permits vast improvements in computational efficiency.
+The choice of a separable space/time model (where the spatial and temporal random effects are modeled independently) also greatly improves computation time but will fail to capture any co-dependencies between these effects.
+
+We model the spatial random effect as a stochastic partial differential equation...
+
+We model the temporal random effect $\omega$ (representing how the mean spatial field evolves in time) as a second-order autoregressive model where $$\omega_t = \rho \omega_{t-1}$$ An AR(2) process was selected due to the observed temporal autocorrelation in the data, which indicates that accumulation retains autocorrelation up to a 2-year lag time.
+
+The modeling objective, therefore, is to find the joint posterior distribution of the various parameters of interest, permitting inference over the processes governing accumulation in the region and making predictions at unobserved locations possible.
+$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s \mid \mathbf{Y})$$
+
+## RESULTS
+
+Estimates of mean annual accumulation (shown in \figref{accum-map}) are similar to those found by \citet{medley_airborneradar_2013} using manually traced IRHs from OIB Snow Radar echograms.
+Accumulation is highest (and exhibits the highest variability) close to coastal West Antarctica and progressively decreases towards the interior.
+%, as observed in numerous prior studies \cite{}.
+Considerable spatial variability, however, overlays this broader signal.
+The most likely source of this higher-order spatial variability is accumulation and wind re-distribution governed by surface topography and near-surface winds \cite{dattler_significant_2019}.
+
+The region overall shows generally insignificant trends in annual SMB for the period 1975-2015 (\figref{trend-map}).
+This is in agreement with a number of previous studies that found insignificant changes in accumulation during the latter portion of the 20th century.
+\citet{medley_airborneradar_2013} found no significant trends in SMB for the period 1980-2009 when compositing all OIB Snow Radar time series in their study for an overlapping region.
+\citet{wang_snow_2017} found little to no trend (0.18% per decade) in snow accumulation in central WAIS based on composite ice cores in the region for the period 1900-2010 and weakly positive trends (2.08% per decade) for the same region and time period using ERA-20C reanalysis data.
+\citet{wang_comparison_2016} compared the results of several climate models and reanalyses for the period 1979-2012, including ERA-Interim, JRA-55, MERRA, PMM5, RACMO2.1, and RACMO2.3.
+They found considerable spatial variability in the linear trends in SMB for the central WAIS, including opposing signs of the trend, but none showed significant trends over this period in the region.
+
+Although the majority of our time series show insignificant trends, large spatial heterogeneity exists in the results, with 34% of the total results showing significant negative trends (95% confidence interval) for 1975-2015.
+\textless1% of the trend results show statistically significant positive trends in this time interval.
+More negative trends tend to occur in the western portions of the region, with less negative and insignificant trends towards the east, but with important variability in these general observations (\figref{trend-map}).
+95% of accumulation changes in the region vary between -65 mm decade\textsuperscript{-1} and +7.2 mm decade\textsuperscript{-1} (-17% decade\textsuperscript{-1} and 2.1% decade\textsuperscript{-1} relative to the multi-decadal mean accumulation).
+<!-- The east-west pattern of accumulation trends persists after correcting for mean accumulation rate---i.e.
+more negative accumulation trends as a percent of mean accumulation towards the west (\textbf{b} in \figref{trend-map}).
+--> The observed trend magnitudes are similar to those observed in ice and firn cores across the region and for a wider area of WAIS (\textbf{d} in \figref{trend-map}), but provide important context of the spatial variability and patterns in those trends.
+Although individual cores suggest a similar pattern, they alone are insufficient to resolve the statistical significance of the overall results or to properly demonstrate the spatial nature and heterogeneity of the patterns.
+
+\citet{medley_increased_2019} used ice core compilations combined with reanalysis-derived spatial coherence patterns to estimate Antarctic-wide snow accumulation.
+Their results also show a strong east-west dipole in SMB over WAIS (thought to be driven by the Southern Annular Mode) in the period 1957-2000, with enhanced (decreased) precipitation towards the Antarctic Peninsula (Ross Sea).
+\citet{wang_new_2019} observed a similar dipole in SMB trends for the period 1950-2010 using a kriging reconstruction from ERA-Interim outputs informed by ice core records.
+Our results agree broadly with these conclusions, but indicate greater variability in the trend direction for central WAIS compared to either study, with significant negative trends for the period 1979-2010 extending farther east.
+These results are specifically in agreement with the findings of \citet{burgener_observed_2013}, who found decreasing accumulation since 1975 from several firn cores near WAIS Divide.
+
+\citet{thomas_regional_2017}, using composite ice core records and SMB from RACMO2.3p2, found that for the portion of WAIS between 90\mydegree--150\mydegree W, 50-year running linear trends varied mostly within $\pm1$ mm a\textsuperscript{-1}, with the period 1961-2010 showing a trend of \mytilde0.65 mm a\textsuperscript{-1}.
+Our results---representing a region comprising \~1/3 of the WAIS section in \citet{thomas_regional_2017}---show a spatially integrated trend of -1.8 mm a\textsuperscript{-1} (-0.49% a\textsuperscript{-1}) for the whole region of study, but larger variability spatially of -6.5 mm a\textsuperscript{-1} to +0.72 mm a\textsuperscript{-1}.
+Importantly, the WAIS regions between 90\mydegree--150\mydegree W not included in our results are mostly from the Ross Sea region.
+This region exhibits decreasing accumulation rates in recent decades \cite{wang_snow_2017, bertler_ross_2018, medley_increased_2019, wang_new_2019}, so its exclusion in our study is unlikely to explain the discrepancy between our results and \citet{thomas_regional_2017}.
+
+## Discussion and conclusions
+
+The observation of statistically significant negative trends in this region is somewhat anomalous compared to many previous studies.
+Several mechanisms/rationales could potentially explain the observed patterns of negative trends, including (1) systemic issues/errors in the PAIPR method itself, (2) spurious trends in the data, (3) multi-decadal internal climate variability, and/or (4) significant changes in the atmospheric or climate system of central WAIS.
+
+As with any method, PAIPR could be susceptible to biases or errors not immediately obvious in the results.
+Inherent biases could exist in the depth-density modeling, although such effects should be minimal as our region of study is well-represented with firn cores and modeling error is incorporated into the final uncertainties used in PAIPR \cite{white_hierarchical_2021}.
+Increased difficulty in picking IRHs at depth could lead to artificial negative trends, either by picking too few layers at depth or by picking too many near-surface IRHs.
+Our exclusion of deeper results (\textgreater \mytilde25 m depth) should minimize such errors and our use of bootstrapping to calculate trends will limit the influence of leveraging from both deep and shallow outliers.
+Additionally, \citet{keeler_repeatability_2021} compared SMB time series from repeat OIB flights using PAIPR and showed minimal bias at depth or in overall trends using the PAIPR method.
+Finally, the close agreement with our results with \emph{in-situ} records where available suggest that such widespread and systemic issues with PAIPR, if present, are unlikely to exert a large influence on the results in this study.
+
+If the PAIPR method itself lacks a systemic bias, several other mechanisms could be responsible for the observed trends.
+It's likely some of the observed trends are statically spurious, as is especially common in SMB time series of shorter duration \cite{rupper_effects_2015, wang_comparison_2016}.
+The bootstrapping techniques employed in our trend analyses minimizes the impact of small numbers of erroneous or anomalous points on the computed trends and should therefore limit the introduction of spurious statistically significant trends.
+Furthermore, the widespread and spatially coherent nature of many of these trends, combined with similar agreement in available core time series, supports the idea of limited influence from purely statistical spurious effects.
+
+The observed trends could also be an expression of multi-decadal internal climate variability.
+West Antarctic climate has high variability, with substantial annual, decadal, and centennial scale variability \cite{turner_amundsen_2013, jones_assessing_2016, king_antarctic_2020}.
+The shorter record of much of our investigation (31 years) makes such a possibility more likely.
+A number of these significant trends, however, persist when expanding the results to longer time periods and greater spatial averaging, suggesting they could result from an underlying forcing or change in the climate system.
+Given the high climate variability in the region even at centennial time scales, longer-term trends such as these could occur in the absence of more noteworthy changes in climate.
+
+The observed trends could also result from broader systemic changes in Antarctic climate.
+The observed trends are unlikely to be purely the result of thermodynamic changes, as the high degree of warming observed in the region over recent decades \cite{bromwich_central_2013, jones_assessing_2016} is at odds with the observed losses in accumulation.
+This would suggest changes in atmospheric dynamics and moisture transport as the most likely cause of the observed accumulation trends.
+Several observed changes in the region (discussed below) could explain decreased accumulation over central WAIS.
+These changes are not mutually exclusive, with many of them coupled together or otherwise interrelated.
+
+Sea ice extent in the Ross-Amundsen Sea sector has steadily increased since 1979 \cite{jones_assessing_2016}, which could result in decreased accumulation for the interior of central WAIS.
+Furthermore, sea ice extent has declined in the Amundsen-Bellingshausen Sea sector over the same time period, which could help explain the increased precipitation closer to the Antarctic Peninsula.
+The Southern Annular Mode (SAM), the dominant mode of climate variability in the Southern Hemisphere high latitudes \cite{thompson_annular_2000}, exerts an influence on West Antarctic precipitation, particularly through its effects on extratropical storm tracks \cite{wang_trends_2013, marshall_signatures_2016}.
+The observed trend towards the SAM positive phase since the 1970's could therefore influence accumulation trends in our region of study \cite{marshall_trends_2003a, turner_atmosphere_2017}.
+Several studies have suggested teleconnections between tropical sea surface temperatures and West Antarctic air temperatures \cite{ding_winter_2011, ding_influence_2012, steig_recent_2013} which could have additional influences on SMB in the region.
+
+Much of the climate variability in WAIS specifically results from the presence of the Amundsen Sea Low (ASL) \cite{raphael_amundsen_2015, turner_atmosphere_2017}.
+The zonal and meridional position of the ASL has a well-defined annual cycle, with an extreme southern (northern) and western (eastern) position near the Ross Sea (Antarctic Peninsula) in winter (summer).
+The ASL produces a strong effect on storm tracks in the region, causing a dipole in moisture and warm air advection in West Antarctica, increasing precipitation in eastern West Antarctica near the Peninsula, and decreasing precipitation to the west near the Ross Sea \cite{emanuelsson_role_2018}.
+Although the ASL varies seasonally (with a minimum in winter and maximum in summer \cite{hosking_influence_2013}), the mean depth of the system over recent decades has increased both seasonally and annually \cite{turner_amundsen_2013}.
+This deepening could strengthen the east-west accumulation dipole across WAIS, explaining the first-order spatial pattern of accumulation trends seen in our results.
+
+On a more local scale, the results suggest linear trends in accumulation are spatially variable.
+This includes significant reductions in accumulation rates for central WAIS farther east than found previously.
+The results further show high variability in significant trend magnitudes over distances of a few kilometers, with even opposing signs of trend within 10-km distances.
+Such small scale differences are not discernible with the resolution of most current modeling products over the region \cite{lenaerts_climate_2018}.
+More dense data networks (such as those provided by ice-penetrating radar and automated processes like PAIPR) therefore provide important insights regarding the nature and pattern of these spatiotemporal changes in SMB that are difficult to obtain with other methods.
+The exact cause of this high spatial variability is currently unknown.
+Several studies show surface topography and wind vectors exert strong influences on mean annual accumulation rates in WAIS \cite{dattler_significant_2019}.
+Such factors could theoretically play an important role in the spatial patterns of accumulation trends as well.
+
+In this study, we reconstructed annually-resolved accumulation time series for 1-km aggregated grids from Operation IceBridge Snow Radar echograms using the fully-automated PAIPR method.
+The results showed high spatial variability on the scale of current climate model resolutions, both for mean annual accumulation and for trends in accumulation rates over recent decades.
+The results furthermore demonstrate an east-west dipole in accumulation over central WAIS (as also found by prior investigations) and suggest significant losses in accumulation extend farther east in WAIS than shown in previous studies.
+The results highlight important advantages (fine spatial resolution, broad spatial coverage, annual temporal resolution, etc.) afforded by a dense network of observations as provided by radar echograms compared to either \textit{in-situ} accumulation measurements or modeling efforts.
+Such advantages improve our ability to investigate the spatiotemporal patterns and trends of WAIS SMB over recent decades, both in the context of broader synoptic-scale and finer local-scale mechanisms.
+In this way, we can better ascertain the dominant source of the observed changes (whether natural climate varability or anthropogenic forcings) occurring now and in the future.
 
 Load libraries:
+
 ```{r}
@@
 library(INLA)
+# library(inlabru)
 library(dplyr)
@@
 library(ggplot2)
+library(terra)
 library(colorspace)
@@
 # Subset data to every 5th location (coarsens data to ~1 km)
-skip.int = 5
+skip.int = 3
 gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
@@
 data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
-  filter(Year >= 1975) %>% filter(Year < 2015) %>% arrange(trace_ID, Year)
+  filter(Year >= 1975) %>%
+  filter(Year < 2015) %>% arrange(trace_ID, Year)
 
@@
 Directly calculate linear coefficients of time from raw data.
+
 ```{r}
@@
 
-### Model setup
-
-The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension (e.g.
-annually resolved time series).
-We therefore choose to model accumulation as a Gamma distribution to ensure a zero-bounded function with increased variance at higher expected values as also observed in the raw data.
-This furthermore matches established literature where precipitation values are frequently assumed to follow a Gamma distribution **(citations)**.
-We express this distribution in terms of the expected accumulation for an arbitrary space location and time point $\mu_{s,t}$ and a precision parameter at the corresponding time $\phi_t$ 
-$$\mathbf{Y} \sim Gamma(\mu_{s,t},\phi_t)$$
-where the variance at an arbitrary spacetime point is
-$$Var(s_t) = \frac{\mu_{s,t}^2}{\phi_t}$$
-
-For each observation $y_{s,t} \in \mathbf{Y}$, we construct a linear predictor $\eta_{s,t}$ associated with the expectation value $\mu_{s,t}$ via a log link function.
-$$\log(\mu_{s,t}) = \eta_{s,t}$$
-
-We further model $\eta_{s,t}$ as a generalized linear equation consisting of fixed effect variables $\mathbf{X}$ and smoothed random latent effects (in this case modeling a spatial random effect $u(\cdot)$ and a temporal random effect $\omega(\cdot)$).
-To explicitly model $u(\cdot)$ as a trend in accumulation over time, we further index this random variable by year.
-The equation for our generalized linear model is therefore:
-$$
-\eta_{s,t} = \alpha + \sum^M_{m=1} \beta_m x_{m,s} + u(s)\times t + \omega(t)
-$$
-
-This model therefore accounts for fixed global effects, a spatially-varying latent effect indexed on time (i.e.
-providing explicit estimates of trends in annual acculation), and a temporally-varying latent effect.
-
-The latent random effects $u$ and $\omega$ are indexed by a set of parameters $\theta_s$ and $\theta_t$ that account for the spatial and temporal correlation in the data.
-We model these correlation structures (as defined by the parameters $\pmb{\theta}$) as latent stationary Gaussian fields, using a function of the hyperparameters $\pmb{\psi}$ and an associated prior distribution $p(\pmb{\psi})$.
-This is equivalent to assuming that $\pmb{\theta}$ is associated with some multivariate Normal distribution with mean $\mu = (\mu_1,...,\mu_n)'$ and covariance structure $\Sigma$, where $\Sigma_{jk} = Cov(\theta_j, \theta_k)$.
-
-$$
-\pmb{\theta} \sim \mathcal{N}(\mu_\theta, \Sigma_\theta)
-$$
-
-By assuming adherence to Markovian properties, the precision matrix $Q = \Sigma^{-1}$ is sparse, with the non-zero components of $Q$ completely given by the neighborhood structure ($N$) of the process, i.e.
-$Q_{jk} \neq 0 \iff k \in \{j,N(j)\}$.
-The specification of a Gaussian-Markov Random Field for both the spatial and temporal structures permits vast improvements in computational efficiency.
-The choice of a separable space/time model (where the spatial and temporal random effects are modeled independently) also greatly improves computation time but will fail to capture any co-dependencies between these effects.
-
-We model the spatial random effect as a stochastic partial differential equation...
-
-We model the temporal random effect $\omega$ (representing how the mean spatial field evolves in time) as a second-order autoregressive model where
-$$\omega_t = \rho \omega_{t-1}$$
-An AR(2) process was selected due to the observed temporal autocorrelation in the data, which indicates that accumulation retains autocorrelation up to a 2-year lag time.
-
-The modeling objective, therefore, is to find the joint posterior distribution of the various parameters of interest, permitting inference over the processes governing accumulation in the region and making predictions at unobserved locations possible.
-$$P(\phi, \alpha, \pmb{\beta}, \rho, R_s, \sigma_s \mid \mathbf{Y})$$
-
 ```{r}
 # More coarse mesh for development purposes
+# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
+#                     max.edge = c(17000, 100000), cutoff = 500)
 mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
@@
 Run and summarize model
+
 ```{r}
-# Model formula
-mod = inla(f.mod,
-           data = inla.stack.data(dat.stack),
-           family = 'Gamma',
-           control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
-           # control.inla = list(int.strategy='eb'), #Improves comp time at cost of accuracy/precision
-           control.compute = list(waic=TRUE, config=TRUE))
+# # Model formula
+# mod = inla(f.mod,
+#            data = inla.stack.data(dat.stack),
+#            family = 'Gamma',
+#            control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
+#            # control.inla = list(int.strategy='eb'), #Improves comp time at cost of accuracy/precision
+#            control.compute = list(waic=TRUE, config=TRUE))
+# 
+# n.iter = 1
+# while (mod$mode$mode.status>0 && n.iter<=3) {
+#   print("Issues with Hessian. Re-running to solve negative eigenalues in the Hessian")
+#   n.iter = n.iter + 1
+#   print(paste("Starting Iteration", n.iter, "of INLA..."))
+#   mod = inla.rerun(mod)
+# }
+# 
+# if (mod$mode$mode.status > 0) {
+#   print("Issues with Hessian eigenvalues persist. Treat model results with suspicion!")
+# }
 
-n.iter = 1
-while (mod$mode$mode.status>0 && n.iter<=3) {
-  print("Issues with Hessian. Re-running to solve negative eigenalues in the Hessian")
-  n.iter = n.iter + 1
-  print(paste("Starting Iteration", n.iter, "of INLA..."))
-  mod = inla.rerun(mod)
-}
+mod = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/interim-models/mod.sep.skip3.cut1k.rds')
 
-if (mod$mode$mode.status > 0) {
-  print("Issues with Hessian eigenvalues persist. Treat model results with suspicion!")
-}
+# bru.form = accum ~ Intercept + dem + 
+# bru.test = bru(components = f.mod, )
 ```
@@
 
+### Validate model using hold-out test data
+
+```{r}
+# Projector matrix for spatial locations (and indexed on Year for trends)
+A.valid <- inla.spde.make.A(mesh, 
+                         loc=cbind(dat.test$East, dat.test$North), 
+                         weights = dat.test$Year.mod)
+
+# Make data stack
+valid.stack <- inla.stack(
+  data = list(y = dat.test$accum), 
+  A = list(1, 1, 1, A.valid),
+  effects = list(list(Intercept = rep(1, nrow(dat.test))), 
+                 tibble(elev = dat.test$elev, dem=dat.test$dem, 
+                        slope=dat.test$slope, aspect=dat.test$aspect),
+                 list(time = dat.test$Year.mod),
+                 spat.idx),
+  tag = 'valid')
+stack.full = inla.stack(dat.stack, valid.stack)
+
+# Generate estimates at new locations using fitted model.
+mod.valid = inla(f.mod,
+           data = inla.stack.data(stack.full),
+           family = 'Gamma',
+           control.predictor = list(compute = TRUE, A = inla.stack.A(stack.full), link=1),
+           control.compute = list(return.marginals.predictor=TRUE), 
+           # control.compute = list(waic=TRUE, config=TRUE),
+           control.mode = list(theta = mod$mode$theta, restart=FALSE))
+```
+
+Extract predicted variables/values at validation points
+
+```{r}
+idx.valid <- inla.stack.index(stack.full, tag='valid')$data
+
+valid.df = dat.test %>% select(trace_ID, East, North, Year, accum, std, SMB) %>% 
+  mutate(mu.pred=mod.valid$summary.fitted.values$mean[idx.valid], 
+         mu.sd=mod.valid$summary.fitted.values$sd[idx.valid]) %>% 
+  mutate(y.sd=sqrt(mu.pred^2/mod.valid$summary.hyperpar$mean[1]), 
+         log.diff=log(mu.pred/accum))
+
+ggplot(valid.df, aes(x=accum)) + geom_point(aes(y=mu.pred), alpha=0.1) + 
+  geom_point(aes(y=SMB), color='blue', alpha=0.1) + 
+  geom_abline(intercept=0, slope=1, color='red', linetype='dashed')
+
+# Not really the right metric due to heteroskadasticity
+rmse = sqrt(mean((valid.df$accum-valid.df$mu.pred)^2))
+print(paste("RMSE =", format(rmse, digits=3), "mm w.e."))
+```
+
+Direct comparisons for individual trace time series between observed and predicted
+
+```{r}
+
+idx.obs = inla.stack.index(stack.full, tag='dat')$data
+pred.comp = dat.train %>% select(trace_ID, East, North, Year, accum, std, SMB) %>% 
+  mutate(mu.pred=mod.valid$summary.fitted.values$mean[idx.obs]) %>% 
+  mutate(log.diff=log(mu.pred/accum))
+
+ggplot(valid.df, aes(x=Year, y=log.diff, group=trace_ID)) + 
+  geom_line(size=0.1, alpha=0.1) + 
+  # geom_line(data=pred.comp, size=0.1, alpha=0.1, color='red') + 
+  theme(legend.position="none")
+
+ggplot(valid.df, aes(x=log(mu.pred/accum))) + 
+  geom_density(fill='red', alpha=0.1) + 
+  geom_density(data=pred.comp, fill='blue', alpha=0.1)
+```
+
+### Predict at grid locations
+
+```{r}
+pred.bbox = st_bbox(gdf_traces)
+
+stepsize = 1000
+buffer = 5*stepsize
+E.range = round(c(pred.bbox[1]-buffer, pred.bbox[3]+buffer))
+N.range = round(c(pred.bbox[2]-buffer, pred.bbox[4]+buffer))
+
+nxy <- round(c(diff(E.range), diff(N.range)) / stepsize) # Calculate the number of cells in the x and y ranges
+
+# Project the spatial field on the mesh vertices using the inla.mesh.projector() function
+proj.grid <- inla.mesh.projector(mesh,
+                                xlim = E.range,
+                                ylim = N.range,
+                                dims = nxy)
+
+# Find index of grid points within ROI
+grid.pts = tibble(x = proj.grid$lattice$loc[,1],
+                  y = proj.grid$lattice$loc[,2],
+                  idx = 1:nrow(proj.grid$lattice$loc))
+grid.sf = st_as_sf(grid.pts, coords=c('x','y'), crs=st_crs(gdf_traces))
+
+# Import topo data from REMA tiles
+REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
+dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
+crs(dem) = st_crs(gdf_traces)$wkt
+slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
+crs(slp) = crs(dem)
+aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
+crs(aspct) = crs(dem)
+topo.stk = c(dem, slp, aspct)
+
+# Extract topo covariates from rasters for each grid point
+topo.pts = extract(topo.stk, vect(grid.sf$geometry)) %>% as_tibble() %>%
+  mutate(dem=REMA_200m_dem) %>% select(-REMA_200m_dem)
+
+grid.sf = grid.sf %>% mutate(dem=topo.pts$dem,
+                               slope=topo.pts$slope,
+                               aspect=topo.pts$aspect)
+
+# # Extract clim covariates from rasters for each grid point
+# racmo.pts = extract(racmo.stk, vect(grid.sf$geometry))
+# grid.sf = grid.sf %>% mutate(smb.racmo=racmo.pts$smb,
+#                                u10m=racmo.pts$u10m,
+#                                v10m=racmo.pts$v10m)
+# grid.sf = grid.sf %>% mutate(S_x = abs(slope)*cos(aspect*pi/180),
+#                                    S_y = abs(slope)*sin(aspect*pi/180)) %>%
+#   mutate(MSWD = S_x*u10m + S_y*v10m)
+
+trend.proj = inla.mesh.project(proj.grid, mod$summary.random$spat.idx$mean)
+trend.sd = inla.mesh.project(proj.grid, mod$summary.random$spat.idx$sd)
+
+grid.sf = grid.sf %>% mutate(trend=c(trend.proj), trend.sd=c(trend.sd))
+
+```
+
+```{r}
+# pred.bbox = st_bbox(gdf_traces)
+# 
+# stepsize = 15000
+# buffer = 5*stepsize
+# E.range = round(c(pred.bbox[1]-buffer, pred.bbox[3]+buffer))
+# N.range = round(c(pred.bbox[2]-buffer, pred.bbox[4]+buffer))
+# 
+# nxy <- round(c(diff(E.range), diff(N.range)) / stepsize) # Calculate the number of cells in the x and y ranges
+# 
+# # Project the spatial field on the mesh vertices using the inla.mesh.projector() function
+# proj.grid <- inla.mesh.projector(mesh,
+#                                 xlim = E.range,
+#                                 ylim = N.range,
+#                                 dims = nxy)
+# 
+# # Find index of grid points within ROI
+# grid.pts = tibble(x = proj.grid$lattice$loc[,1],
+#                   y = proj.grid$lattice$loc[,2],
+#                   idx = 1:nrow(proj.grid$lattice$loc))
+# grid.sf = st_as_sf(grid.pts, coords=c('x','y'), crs=st_crs(gdf_traces))
+# 
+# # Import topo data from REMA tiles
+# REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
+# dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
+# crs(dem) = st_crs(gdf_traces)$wkt
+# slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
+# crs(slp) = crs(dem)
+# aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
+# crs(aspct) = crs(dem)
+# topo.stk = c(dem, slp, aspct)
+# 
+# # Extract topo covariates from rasters for each grid point
+# topo.pts = extract(topo.stk, vect(grid.sf$geometry)) %>% as_tibble() %>% 
+#   mutate(dem=REMA_200m_dem) %>% select(-REMA_200m_dem)
+# 
+# # Center and scale topo data
+# topo.pts = topo.pts %>% 
+#   mutate(dem=dem-mean(data$dem, na.rm=TRUE), slope=slope-mean(data$slope, na.rm=TRUE)) %>%
+#   mutate(dem = dem/sd(data$dem, na.rm=TRUE), slope = slope/sd(data$slope, na.rm=TRUE))
+# 
+# grid.sf = grid.sf %>% mutate(dem=topo.pts$dem, 
+#                                slope=topo.pts$slope, 
+#                                aspect=topo.pts$aspect)
+# 
+# # # Extract clim covariates from rasters for each grid point
+# # racmo.pts = extract(racmo.stk, vect(grid.sf$geometry))
+# # grid.sf = grid.sf %>% mutate(smb.racmo=racmo.pts$smb, 
+# #                                u10m=racmo.pts$u10m, 
+# #                                v10m=racmo.pts$v10m)
+# # grid.sf = grid.sf %>% mutate(S_x = abs(slope)*cos(aspect*pi/180), 
+# #                                    S_y = abs(slope)*sin(aspect*pi/180)) %>% 
+# #   mutate(MSWD = S_x*u10m + S_y*v10m)
+# 
+# 
+# Years = 1975:2015
+# loc.pred = st_coordinates(grid.sf) %x% rep(1, length(Years))
+# yr.mod = rep(Years - mean(data$Year), nrow(grid.sf))
+# st.pred = tibble(X=loc.pred[,1], Y=loc.pred[,2], Year = rep(Years, nrow(grid.sf)), 
+#                  Year.mod = yr.mod, 
+#                  dem=rep(grid.sf$dem, each=length(Years)))
+# 
+# # Define projector matrix for predictions
+# A.pred =  inla.spde.make.A(mesh=mesh,
+#                          loc=cbind(st.pred$X, st.pred$Y),
+#                          weights = st.pred$Year.mod)
+# 
+# # Generate prediction data stack and join to estimation stack
+# pred.stack <- inla.stack(
+#   data = list(y = NA),
+#   A = list(1, 1, 1, A.pred),
+#   effects = list(list(Intercept = rep(1, nrow(st.pred))),
+#                  tibble(dem=st.pred$dem,
+#                         # slope=st.pred$slope, aspect=st.pred$aspect
+#                         ),
+#                  list(time = st.pred$Year.mod),
+#                  spat.idx),
+#   tag = 'pred')
+# stack.all <- inla.stack(dat.stack, pred.stack)
+# 
+# # Generate estimates at new locations using fitted model.
+# mod.pred = inla(f.mod,
+#            data = inla.stack.data(stack.all),
+#            family = 'Gamma',
+#            control.predictor = list(compute = TRUE, A = inla.stack.A(stack.all), link=1),
+#            # control.compute = list(waic=TRUE, config=TRUE), 
+#            control.mode = list(theta = mod$mode$theta, restart=FALSE))
+```
+
 ### Components of prediction
@@
 
-Direct comparisons for individual trace time series between observed and predicted
-```{r}
-pred.comp = dat.pred %>% select(color, Year, accum, std, SMB) %>% 
-  right_join(st.pred, by=c("color", "Year")) %>% 
-  mutate(log.diff=log(pred/accum))
-
-ggplot(pred.comp) + 
-  geom_line(aes(x=Year, y=log.diff, group=color), size=0.25, alpha=0.1) + 
-  theme(legend.position="none")
-ggplot(pred.comp) + geom_abline(slope=1, color='black') + 
-  geom_point(aes(x=accum, y=SMB), color='red', alpha=0.1) + 
-  geom_point(aes(x=accum, y=pred), color='blue', alpha=0.1)
-ggplot(pred.comp) + geom_density(aes(x=log(pred/accum)), fill='red', alpha=0.3)
-```
-
 ```{r}
@@
                   sd=mod$summary.random$time$sd)
-tmp = pred.comp %>% group_by(Year) %>% summarize(accum=mean(accum, na.rm=TRUE), pred=mean(pred))
+tmp = pred.comp %>% group_by(Year) %>% summarize(accum=mean(accum, na.rm=TRUE), mu.pred=mean(mu.pred))
 mod.time = mod.time %>% mutate(obs.diff=log(tmp$accum/mean(tmp$accum)), 
-                               pred.diff=log(tmp$pred/mean(tmp$pred)))
+                               pred.diff=log(tmp$mu.pred/mean(tmp$mu.pred)))
 ggplot(mod.time) + geom_hline(yintercept = 0, color='black', linetype="dashed") + 
@@
 
+Plots of trend posteriors for random subsets of locations
+
+```{r}
+# Get df of transformed trend marginals
+marg.num = 100
+marg.trend = tibble(X=numeric(), Y=numeric(), IDX=character())
+idx.name = names(mod$marginals.random$spat.idx)
+for (i in 1:length(idx.name)) {
+  marg.i = inla.tmarginal(function(x) x, 
+                          mod$marginals.random$spat.idx[[i]], 
+                          n=marg.num)
+  marg.tbl = tibble(X=marg.i[,1], Y=marg.i[,2], IDX=idx.name[i])
+  marg.trend = marg.trend %>% bind_rows(marg.tbl)
+}
+
+IDX.subset = sample(idx.name, 12)
+ggplot(marg.trend %>% filter(IDX %in% IDX.subset)) +
+  geom_line(aes(x=X, y=Y, group=IDX, color=IDX), alpha=0.5)
+
+IDX.subset = sample_frac(as_tibble(idx.name), 0.1)
+ggplot(marg.trend %>% filter(IDX %in% IDX.subset$value)) +
+  geom_line(aes(x=X, y=Y, group=IDX), alpha=0.1)
+```
+
 Calculate probability that trend is less than zero.
+
 ```{r}
@@
 for (i in 1:length(P.neg)) {
-  marg.i = inla.tmarginal(function(x) x, 
-                          mod$marginals.random$spat.idx[[i]], 
-                          n=500)
-  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
-  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
+  P.neg[i] = inla.pmarginal(0, mod$marginals.random$spat.idx[[i]])
 }
+# P.neg = vector(mode = "numeric", length = length(mod$marginals.random$spat.idx))
+# for (i in 1:length(P.neg)) {
+#   marg.i = inla.tmarginal(function(x) x, 
+#                           mod$marginals.random$spat.idx[[i]], 
+#                           n=500)
+#   P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
+#   P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
+# }
+
 tmp = tibble(Source='model', 
@@
 Map of estimated trends with probability of less than zero
+
 ```{r}
@@
 mod.trends = trends.comp %>% filter(Source=="model")
-Neg.threshold = 0.95
+Neg.threshold = 0.89
 frac.Neg = nrow(mod.trends %>% filter(P.neg>=Neg.threshold))/nrow(mod.trends)

### 2022-07-20 notebooks/deprecated/article_draft.Rmd
Cleans up old notebook files

---
title: "Ch. 2 Dissertation draft"
csl: ../docs/american-geophysical-union.csl
link-citations: yes
output:
  html_document:
    df_print: paged
bibliography: ../docs/citations.bib
---

## README

This is a draft version of what will eventually become the paper I submit as the second chapter of my dissertation. 
Eventually it would be good for this to be a fully reproducible paper that I can submit to a journal, but that might require more $\LaTeX$ knowledge than I currently possess.

## Introduction

This is where the introduction will go.

## Methods

This is where the methods will go.

```{r, echo=FALSE, message=FALSE}

# Attach required packages
library(tidyverse)
library(here)
library(sf)
library(tcltk)
library(readxl)
library(tictoc)

tic("Total runtime")

# Import accumulation data results
accum.PS = read_rds(here("data/interim_results/accum_data.rds"))
cores.PS = read_rds(here("data/interim_results/core_data.rds"))

# # Code to export data as .csv for use in Python
# accum.exp = accum.PS %>% 
#   mutate(trace_ID = 0:(nrow(accum.PS)-1), Easting=st_coordinates(.)[,1], 
#          Northing=st_coordinates(.)[,2]) %>%
#   st_drop_geometry() %>% unnest(cols = data) %>% 
#   select(trace_ID, Easting, Northing, everything())
# cores.exp = cores.PS %>%
#   mutate(Easting=st_coordinates(.)[,1], Northing=st_coordinates(.)[,2]) %>%
#   st_drop_geometry() %>% unnest(cols = Data) %>% select(Site, Easting, Northing, everything())
# write_csv(accum.exp, '/home/durbank/Documents/Research/Antarctica/WAIS-wind/data/R-trends/accum.csv')
# write_csv(cores.exp, '/home/durbank/Documents/Research/Antarctica/WAIS-wind/data/R-trends/cores.csv')

# Filter accumulation data
accum.PS = accum.PS %>% filter(elev.REMA > 0) %>% 
  filter(slope < 2)

# Import basin shapefiles (in Antarctic Polar Sterographic EPSG:3031)
basins = st_read(here("data/ANT_Basins_IMBIE2_v1.6/ANT_Basins_IMBIE2_v1.6.shp"))

# Import Antarctic outline
Ant.outline = st_read(here("data/Ant_basemap/Coastline_medium_res_polygon.shp"))
```

Below is a reference map showing the location of our data within the greater context of Antarctica.

```{r, fig_inset, echo=FALSE}

# Plot of current Antarctic cores (blue) and soon-to-be-added (red)
cores.plt = ggplot(cores.PS) + geom_sf(data = Ant.outline, color = 'grey') +
  geom_sf(data = sample_n(accum.PS, size = 1000), color = 'red') + 
  geom_sf(aes(text=Site), color = 'blue') + coord_sf(datum = st_crs(cores.PS)) + 
  theme_void()
print(cores.plt)
```

## Results

This is where the results will go.

```{r, echo=FALSE, message=FALSE}

# Create non-referenced tibble of accumulation by location
accum.loc_PS = as_tibble(st_coordinates(accum.PS)) %>%
  rename(Easting = X, Northing = Y) %>% bind_cols(st_set_geometry(accum.PS, NULL))

# Create non-referenced tibble of core accumulation by location
cores.loc_PS = as_tibble(st_coordinates(cores.PS)) %>%
  rename(Easting = X, Northing = Y) %>% bind_cols(st_set_geometry(cores.PS, NULL))
```

```{r, accum_map, echo=FALSE}
# Plot mean accum on spatial grid
accum.plt1 = ggplot(accum.loc_PS, aes(x = Easting, y = Northing, color = accum_mu)) + 
  geom_point() + 
  geom_point(data = cores.loc_PS, aes(x = Easting, y = Northing, color = accum_mu), 
             shape = 12) + 
  scale_color_viridis_c() + 
  # ggtitle("Mean accum in space") + 
  xlab("Easting") + ylab("Northing") + theme_bw()
print(accum.plt1)
```

Above we see how mean accumulation rates vary over the region of interest.

```{r, trend_map, echo=FALSE}
# Plot scaled trends significant at 95% CL
trend.map = ggplot(accum.loc_PS, aes(x = Easting, y = Northing)) + 
  geom_point(aes(color = 100*coeff_perc)) + 
  geom_point(data = filter(accum.loc_PS, p.val_yr < 0.05) %>% sample_n(2000), 
             aes(x=Easting, y=Northing), shape = 4, color = 'black', size = 0.05) + 
  geom_point(data = cores.loc_PS, 
             aes(x = Easting, y = Northing, color = 100*coeff_perc), shape = 8, size = 3) + 
  geom_point(data = filter(cores.loc_PS, p.val_yr < 0.05), 
             aes(x=Easting, y=Northing), shape = 4, color = 'black', size = 0.05) + 
  lims(x = c(-1.5e6,-9e5), y = c(-6e5,0.35e5)) +
  scale_colour_gradient2() +
  labs(x = "Easting", y = "Northing", color = "%/yr") + 
  theme_bw()
print(trend.map)
```

Above we see how the linear time series trend in annual accumulation rate varies spatially across the region of interest.

```{r}
# test = filter(accum.loc_PS, Northing < -4.5E5 & Easting < -1.2E6)
# 
# test.smp = test %>% sample_n(size = 250)
# test.lm = map(test.smp$data, ~ lm(accum ~ Year, data = .x, weights = 1/sd^2))
# 
# test.data = dplyr::select(test.smp, -data, -accum_mu) %>% 
#   mutate(data = map(test.lm, 
#                     ~ dplyr::select(augment(.), Year, accum, .fitted, .se.fit, .resid, .std.resid)))
# 
# tmp.tidy = map(test.lm, ~ tidy(.x))
# test.smp = test.smp %>% 
#   mutate(coeff_yr = simplify(map(tmp.tidy, "estimate") %>% map(2))) %>% 
#   mutate(std.err = simplify(map(tmp.tidy, "std.error") %>% map(2))) %>% 
#   mutate(p.val_yr = simplify(map(tmp.tidy, "p.value") %>% map(2)))
# 
# 
# test2 = test.smp %>% sample_n(size = 10) %>% unnest(cols = c(data))
# 
# ggplot(test2, aes(Year, accum)) + geom_line(aes(group = Easting, color = factor(Easting)))

```

## Comparison of mean accumulation and slope

This section uses 8-meter mosaic tiles from the Reference Elevation Model of Antarctica (REMA) to compare mean accumulation rate to surface slope [@howat_reference_2019].

### Exploratory analyses

```{r, message=FALSE}
library(GGally)

ggpairs(select(accum.loc_PS, Easting, Northing, elev.REMA, slope, aspect, accum_mu))

accum_lm = lm(accum_mu ~ Easting + Northing + elev.REMA + slope + aspect, data = accum.loc_PS)
summary(accum_lm)

core.accum_lm = lm(accum_mu ~ Easting + Northing + Elevation, data = cores.loc_PS)

```

Although all of the included predictors show significant effects on `accum_mu`, the dominant influences on the region-wide trends are concentrated in the spatial components.
Interestingly, elevation shows residual influence on the predictions after accounting it's covariance in x-y space.
The following model summarizes these results for the dominant influences.

```{r}
summary(lm(accum_mu ~ Easting + Northing + elev.REMA, data = accum.loc_PS))
```

We can see that by excluding `slope` and `aspect` from the analysis the model remains virtually the same.
The coefficients for the remaining predictors are very similar as are the p-values.
The new $r^2 = 0.6453$ a neglibible difference from the first model.

The model suggests higher Easting and Northing values (in this case moving south and east respectively) leads to lower accumulation rates, which fits our intuition.
The model further suggests that, after accounting for the covariance of elevation with position, higher elevation sites experience higher accumulation rates. **[What does this mean?]**

### Spatial models

Just because Easting, Northing, and elevation dominate the regional trends does not mean slope and aspect are insignificant.
We therefore utilize spatial regression modeling with slope and aspect to investigate the more localized effects of these two predictors.

```{r, message=FALSE}

# test.data = st_crop(accum.PS, c(xmin=-1.4E6, xmax=(-1.4E6 + 50000), ymin=-5E5, ymax=-4.5E5))
test.data = filter(accum.loc_PS, Easting >= -1.3E6 & Easting <= (-1.3E6+30000)) %>% 
  filter(Northing >= -5E5 & Northing <= -4.5E5)
data.lm = lm(accum_mu ~ Easting + Northing + elev.REMA + slope + aspect, data=test.data)

summary(data.lm)

ggpairs(select(test.data, Easting, Northing, elev.REMA, slope, aspect, accum_mu))
ggplot(test.data, aes(x=Easting, y=Northing)) + 
  geom_point((aes(color=accum_mu))) + scale_color_viridis_c()
```

If we look at the residuals, most don't show additional signal, expect for the East/West spatial components. 
These still show cyclic variations in mean accumulation rates (I would assume lining up with the undulations expressed in the echograms).
The below plot shows residuals plotted against Easting as an example of this.

```{r}
plot(test.data$Easting, residuals(data.lm))
```

### Spatial analysis on individual years

In addition to looking at the mean aggregate accumulation relationships, I should also look at those relationships on the range of individual years (to determine the degree of noise injected by annual fluctuations).
Bear in mind that I will not have predictors at annual resolution.
This may not matter for some predictors (Easting, Northing, maybe elevation) but could be significant for others (slope and aspect?).

```{r}
toc()
```

# REFERENCES

### 2022-07-20 notebooks/deprecated/kronecker-tests.Rmd
Cleans up old notebook files

--- 2022-06-01 notebooks/kronecker-tests.Rmd
@@
 
-set.seed(777)
-gdf_traces = gdf_traces %>% sample_frac(0.01)
-
 # # Subset to smaller region
 # tmp = st_bbox(gdf_traces)
-# East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[1] + (tmp[3]-tmp[1])/2)
-# North = c(tmp[2] + (tmp[4]-tmp[2])/2, tmp[4] - (tmp[4]-tmp[2])/8)
+# # East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[1] + (tmp[3]-tmp[1])/2)
+# # North = c(tmp[2] + (tmp[4]-tmp[2])/2, tmp[4] - (tmp[4]-tmp[2])/8)
+# East = c(tmp[1] + (tmp[3]-tmp[1])/7, tmp[3])
+# North = c(tmp[2], tmp[2] + (tmp[4]-tmp[2])/5)
 # tmp.df = tibble(East=East, North=North)
@@
 #   st_as_sfc()
+# ggplot() + geom_sf(data=poly) + geom_sf(data=gdf_traces)
 # gdf.idx = sapply(st_intersects(gdf_traces, poly),function(x){length(x)==0})
 # gdf_traces = gdf_traces[!gdf.idx,]
+
+set.seed(777)
+# gdf_traces = gdf_traces %>% sample_frac(0.05)
+skip.int = 5
+gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
+gdf_traces = gdf_traces[gdf.idx,]
 
@@
 
+Directly calculate linear coefficients of time from raw data.
+```{r}
+yr.trends = dat.train %>% 
+  group_by(trace_ID) %>% 
+  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
+  filter(term=='Year.mod') %>% select(-term)
+dat.mu = dat.train %>% group_by(trace_ID) %>% 
+  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
+  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
+  mutate(log.est = log(1+(estimate/accum)))
+```
+
 ```{r}
@@
                     max.edge = c(60000, 120000), cutoff = 12000)
-# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
-#                     max.edge = c(30000, 90000), cutoff = 1000)
+mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
+                    max.edge = c(30000, 90000), cutoff = 2000)
 plot(mesh)
@@
 
+Indexing the spacetime function by year directly, rather than having two seperate terms for it (I'm not sure this will work because we don't have unique year values for each spacetime point).
+```{r}
+A.st <- inla.spde.make.A(mesh, 
+                         loc=cbind(dat.train$East, dat.train$North), 
+                         group = dat.train$Year.idx, 
+                         weights = dat.train$Year.mod)
+
+dat.stack <- inla.stack(
+  data = list(y = dat.train$accum), 
+  A = list(1, 1, A.st),
+  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
+                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
+                        # slope=dat.train$slope, aspect=dat.train$aspect, 
+                        Year=dat.train$Year.mod),
+                 st.idx),
+  tag = 'dat')
+
+form = y ~ -1 + Intercept + #modeling the intercept
+  f(st.idx, model = spde, group=st.idx.group, 
+    control.group=list(model='ar1'))
+```
+
 ## Gamma modeling
@@
 
-# mod.gamma = inla(form,
-#                data = inla.stack.data(dat.stack),
-#                family = 'Gamma',
-#                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
-#                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
-#                control.compute = list(waic=TRUE, config=TRUE))
-mod.gamma = readRDS(here('data/interim_results/mod-gamma.rds'))
+mod.gamma = inla(form,
+               data = inla.stack.data(dat.stack),
+               family = 'Gamma',
+               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
+               # control.inla = list(strategy = 'simplified.laplace', int.strategy='ccd'), # Options improve computation time at cost of accuracy/precision
+               control.inla = list(strategy = 'gaussian', int.strategy='eb'), 
+               control.compute = list(waic=TRUE, config=TRUE))
 
+# mod.gamma = readRDS(here('data/interim_results/mod-gamma.rds'))
 summary(mod.gamma)
@@
 ### Posterior checks
-
-Directly calculate linear coefficients of time from raw data.
-```{r}
-yr.trends = dat.train %>% 
-  group_by(trace_ID) %>% 
-  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
-  filter(term=='Year.mod') %>% select(-term)
-dat.mu = dat.train %>% group_by(trace_ID) %>% 
-  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
-  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
-  mutate(log.est = log(1+(estimate/accum)))
-```
 
@@
 It's possible that I am removing temporal signal from my estimates when I model the spacetime auto-correlations.
-Here I run the model without this random effect for comparison (so only using a spatially varying temporal coefficient).
+Here I run the model without this random effect for comparison (so only using a spatially varying temporal coefficient and a separate temporal random effect).
 
 ```{r}
+# Make data stack
+stack.sep <- inla.stack(
+  data = list(y = dat.train$accum), 
+  A = list(1, 1, 1, A.yr),
+  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
+                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
+                        slope=dat.train$slope, aspect=dat.train$aspect, 
+                        Year=dat.train$Year),
+                 list(time = dat.train$Year),
+                 yr.idx),
+  tag = 'dat')
+
+# Priors for temporal autocorrelation
+time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
+
 # Model formula
-f.noST = y ~ -1 + Intercept + 
+f.sep = y ~ -1 + Intercept + 
   # f(st.idx, model = spde, group=st.idx.group, 
   #   control.group=list(model='ar1')) + 
-  f(yr.idx, model = spde)
+  f(yr.idx, model = spde) + 
+  f(time, model = 'ar1', hyper = time.spec) #Temporal random effect modeled as AR1
 
-mod.noST = inla(f.noST,
-                data = inla.stack.data(dat.stack),
+mod.sep = inla(f.sep,
+                data = inla.stack.data(stack.sep),
                 family = 'Gamma',
-                control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
-                control.inla = list(int.strategy='eb', strategy = 'gaussian'), # Options improve computation time at cost of accuracy/precision
+                control.predictor = list(compute = TRUE, A = inla.stack.A(stack.sep), link=1),
+                control.inla = list(int.strategy='eb'), # Option improve computation time at cost of accuracy/precision
                 control.compute = list(waic=TRUE, config=TRUE))
 
-summary(mod.noST)
+summary(mod.sep)
 ```
@@
 
+Extract marginal distributions for subset of coefficients (for plotting purposes)
+```{r}
+n=50
+idx.set = as.integer(seq(1,spde$n.spde, length.out=n))
+marg.set = mod.sep$marginals.random$yr.idx[idx.set]
+
+# Create tbl of marginals
+marg.tbl = tibble()
+for (i in 1:length(idx.set)) {
+  set.i = marg.set[[i]]
+  tbl.i = tibble(Index = idx.set[i], X=set.i[,1], Y=set.i[,2])
+  marg.tbl = marg.tbl %>% bind_rows(tbl.i)
+}
+
+ggplot(marg.tbl) + geom_line(aes(x=X, y=Y, group=Index), color='black', alpha=0.1)
+```
+
+Calculate probability that trend is less than zero.
+```{r}
+P.neg = vector(mode = "numeric", length = length(mod.sep$marginals.random$yr.idx))
+for (i in 1:length(P.neg)) {
+  marg.i = inla.tmarginal(function(x) x, 
+                          mod.sep$marginals.random$yr.idx[[i]], 
+                          n=500)
+  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
+  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
+}
+```
+
+Map of estimated trends with probability of less than zero
+```{r}
+tmp = tibble(Source="model", 
+             East=mesh$loc[,1], North=mesh$loc[,2], 
+             med=mod.sep$summary.random$yr.idx$`0.5quant`, 
+             P.neg=P.neg, 
+             LB=mod.sep$summary.random$yr.idx$`0.025quant`, 
+             UB=mod.sep$summary.random$yr.idx$`0.975quant`) %>% 
+  # mutate(weak=(LB<0 && UB>0)) %>% 
+  st_as_sf(coords=c("East", "North"), crs=3031)
+
+trends.sep = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
+  rename(med=log.est) %>% mutate(LB=NA, UB=NA, P.neg=NA, Source="data") %>% 
+  st_as_sf(coords=c("East", "North"), crs=3031) %>%
+  bind_rows(tmp)
+
+ggplot(trends.sep %>% filter(Source=="data")) + 
+  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
+# ggplot(trends.sep %>% filter(Source=="model")) + 
+#   geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.05, 0.05))
+ggplot(trends.sep %>% filter(Source=="model")) + 
+  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.05, 0.05))
+# ggplot(trends.sep %>% filter(Source=="model")) + 
+#   geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.05, 0.05))
+ggplot(trends.sep %>% filter(Source=='model')) + 
+  geom_sf(aes(color=P.neg)) + scale_color_viridis_c()
+```
+
 ```{r}
 # Posterior draws for year coefficients
-draws=100
+draws=10
 n.locs = spde$n.spde
-samples.noST = inla.posterior.sample(draws, result = mod.noST)
-tmp = inla.posterior.sample.eval(c("yr.idx"), samples.noST)
-ppc.noST = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))
+samples.sep = inla.posterior.sample(draws, result = mod.sep)
+tmp = inla.posterior.sample.eval(c("yr.idx"), samples.sep)
+ppc.sep = tibble(Sample=rep(1:draws, each=n.locs), coeff=as.vector(tmp))
 
-ggplot(ppc.noST %>% group_by(Sample)) + 
+ggplot(ppc.sep %>% group_by(Sample)) + 
   geom_line(aes(x=coeff, group=Sample), stat="density", alpha=0.1) + 
   geom_density(data=dat.mu, aes(x=log.est), color='red') + 
-  xlim(c(-0.03, .03))
+  xlim(c(-0.06, .06))
+
 ```
@@
 ### Spatial comparisons
-
-```{r}
-tmp = tibble(Source="model", 
-             East=mesh$loc[,1], North=mesh$loc[,2], 
-             med=mod.noST$summary.random$yr.idx$`0.5quant`, 
-             LB=mod.noST$summary.random$yr.idx$`0.025quant`, 
-             UB=mod.noST$summary.random$yr.idx$`0.975quant`) %>% 
-  # mutate(weak=(LB<0 && UB>0)) %>% 
-  st_as_sf(coords=c("East", "North"), crs=3031)
-
-trends.noST = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
-  rename(med=log.est) %>% mutate(LB=NA, UB=NA, Source="data") %>% 
-  st_as_sf(coords=c("East", "North"), crs=3031) %>%
-  bind_rows(tmp)
-
-ggplot(trends.noST %>% filter(Source=="data")) + 
-  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
-ggplot(trends.noST %>% filter(Source=="model")) + 
-  geom_sf(aes(color=LB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
-ggplot(trends.noST %>% filter(Source=="model")) + 
-  geom_sf(aes(color=med)) + scale_color_gradient2(limits=c(-0.03, 0.03))
-ggplot(trends.noST %>% filter(Source=="model")) + 
-  geom_sf(aes(color=UB)) + scale_color_gradient2(limits=c(-0.03, 0.03))
-```
 

### 2022-07-20 notebooks/deprecated/wais-from-synth.Rmd
Cleans up old notebook files

---
title: "Testing based off of synthetic data tests"
output: html_notebook
---

## Imports

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(cetcolor)
library(sf)
library(terra)
library(INLA)
library(INLAutils)
library(spdep)
```

## Data loading and pre-processing

```{r data-load, message=FALSE}
data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1) %>% 
  mutate(Year = as.integer(Year))
gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()

# Remove points in far northern extent
accum_bnds = st_coordinates(gdf_traces)
gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))

# # Small-scale testing bounds
# gdf_traces = gdf_traces %>% st_crop(xmin=-1.4e6, xmax=-1.25e6,
#                                     ymin=min(accum_bnds[,2]), ymax=-3e5)

# Remove unrealistically high elevations and overly large accum sites
trace_drop = gdf_traces %>% filter(elev > 3000 | accum > 1000)

# Add East/North coordinates and drop filtered sites
pts_tmp = st_coordinates(gdf_traces)
coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
gdf_traces = gdf_traces %>% filter(!trace_ID %in% trace_drop$trace_ID)
```

We need to import the other covariate data for the region.
This consists of topographic features from REMA and climatic variables from RACMO.
```{r}
REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
crs(dem) = st_crs(gdf_traces)$wkt

slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
crs(slp) = crs(dem)

aspct = rast(file.path(REMA.dir, "REMA_200m_aspect.tif"))
crs(aspct) = crs(dem)

topo.stk = c(dem, slp, aspct)

topo.pts = extract(topo.stk, vect(gdf_traces$geometry)) %>% as_tibble() %>% 
  rename(dem=REMA_200m_dem) %>% 
  mutate(trace.ID=gdf_traces$trace_ID, aspect=sin(pi/180*aspect))
topo.pts = topo.pts %>% select(-ID)

RACMO.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data", 
                      "/RACMO/2.3p2_yearly_ANT27_1979_2016/interim")
files = list.files(path = RACMO.dir, full.names = TRUE, 
                   pattern = "\\.tif$")

r.smb = rast(files[1])
names(r.smb) = 1979:2016
smb.pts = extract(r.smb, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% mutate(trace.ID=gdf_traces$trace_ID) %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'SMB') %>% 
  mutate(Year = as.integer(Year))

r.u10m = rast(files[2])
names(r.u10m) = 1979:2016
u10m.pts = extract(r.u10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = num_range(prefix='', range=1979:2016), 
               names_to = 'Year', values_to = 'u10m') %>% 
  mutate(Year = as.integer(Year))

r.v10m = rast(files[3])
names(r.v10m) = 1979:2016
v10m.pts = extract(r.v10m, vect(gdf_traces$geometry)) %>% 
  as_tibble() %>% pivot_longer(cols = num_range(prefix='', range=1979:2016), 
                               names_to = 'Year', values_to = 'v10m') %>% 
  mutate(Year = as.integer(Year))

racmo.pts = smb.pts %>% 
  mutate(u10m=u10m.pts$u10m, v10m=v10m.pts$v10m) %>% 
  select(-ID)

data = data %>% 
  left_join(racmo.pts, by = c("trace_ID" = "trace.ID", "Year")) %>% 
  left_join(topo.pts, by = c("trace_ID" = "trace.ID"))

# data = data %>% mutate(S_x = abs(slope)*cos(asin(aspect)), 
#                        S_y = abs(slope)*aspect) %>% 
#   mutate(MSWD = S_x*u10m + S_y*v10m)

# Add columns for mean u10m and v10m speeds
mu.wind = data %>% group_by(trace_ID) %>% 
  summarize(mu.u10 = mean(u10m), mu.v10 = mean(v10m))
data = data %>% left_join(mu.wind, by = "trace_ID")
```

While I am testing/developing things, I'm only using a small subset of the data (around 1/2 a percent).
This still gives a few thousand observations to work with though.
Also for initial testing of everything, I am only including a single covariate (elevation).
Once I have a fully fleshed-out workflow though, I intend to add a few more (slope, wind vectors, etc.).
I center this covariate about its mean and also normalize the discrete time dimension, setting the min year as Year 1.

```{r}
set.seed(777)
gdf.tmp = gdf_traces %>% sample_frac(0.01)
data.tmp = data %>% filter(trace_ID %in% gdf.tmp$trace_ID) %>% 
  filter(Year >= 1979) %>% filter(Year < 2014) %>% arrange(trace_ID, Year)

# Select variables of interest
dat = data.tmp %>% select(trace_ID, East, North, Year, accum, 
                      std, elev, dem, slope, aspect, u10m, v10m, 
                      mu.u10, mu.v10)
# dat = data %>% select(trace_ID, East, North, Year, accum, 
#                       std, elev, dem, slope, aspect, u10m, v10m)

# Center covariates
yr.min = min(dat$Year)-1
dat = dat %>% 
  mutate(Year = Year-yr.min, elev = elev-mean(elev, na.rm=TRUE), 
         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
         u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
         mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE))
         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))

# Scale covariates by sd
dat = dat %>% 
  mutate(Year = Year, elev = elev/sd(elev), dem=dem/sd(dem), 
         slope=slope/sd(slope), u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), 
         mu.u10=sd(mu.u10), mu.v10=sd(mu.v10))

# Split into training and testing sets
dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
```

## INLA  modeling

I create the mesh used for defining the neighborhoods (this is important for generating the sparse precision matrix $Q$, enabling the GMRF approximation).
```{r}
mesh = inla.mesh.2d(loc = dat.train %>% select(East, North), 
                    max.edge = c(25000, 75000), cutoff = 5000)
plot(mesh)
points(dat.train %>% select(East, North), col = "red")

spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(10000, 0.05), # P(range < 10 km) = 0.05
  prior.sigma = c(400, 0.05)) # P(sigma > 400 mm w.e.) = 0.05
```

```{r}
sptime.idx <- inla.spde.make.index('sptime.idx', spde$n.spde, 
                                     n.group = max(dat.train$Year))
A.sptime <- inla.spde.make.A(mesh, cbind(dat.train$East, dat.train$North), 
                             group = dat.train$Year)#, weights = dat.train$Year)

# Make data stack
dat.stack <- inla.stack(
  data = list(y = dat.train$accum),
  A = list(1, 1, A.sptime),
  # effects = list(list(Intercept = 1),
  effects = list(list(Intercept = rep(1, nrow(dat.train))),
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        Year=dat.train$Year), 
                 sptime.idx),
  tag = 'dat')
```

```{r}
# Prior on autoregressive rho parameter
time.spec = list(rho = list(prior = 'pc.cor1', param = c(0, 0.9)))

# Prior on fixed effects intercept (not currently implemented)
prior.fixed = list()

# Prior on precision parameter for Gamma observations
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

formula = y ~ -1 + Intercept + Year + 
  f(sptime.idx, 
    Year, 
    # as.numeric(sptime.idx$sptime.idx.group), 
    model = spde, group = sptime.idx.group, 
    control.group = list(model = 'ar1', hyper=time.spec))
```

## Fit model

```{r}
# mod = inla(formula,
#                     data = inla.stack.data(dat.stack),
#                     family = 'Gamma',
#                     quantiles = NULL,
#                     control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link = 1),
#                     # control.fixed = prior.fixed,
#                     control.family = list(hyper = list(prec = prec.prior)),
#                     control.compute = list(waic=TRUE, return.marginals.predictor=FALSE, config=TRUE))
```

### Alternative specification

This is the same model, but with the space and time random effects separated out to be wholly independent.
It still assumes that the temporal trends, however, vary spatially around a global fixed value.
```{r}
spde = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(15000, 0.01), # P(range < 10 km) = 0.01
  prior.sigma = c(10, 0.05))

spde.yr = inla.spde2.pcmatern(mesh = mesh, 
  prior.range = c(5000, 0.01), # P(range < 10 km) = 0.01
  prior.sigma = c(1, 0.01), 
  constr = TRUE)

# Index for SPDE model and projector matrix
spat.idx = inla.spde.make.index(name = "spat.idx", n.spde = spde$n.spde)
A.spat = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))

# Index for SPDE model and projector matrix
spYR.idx = inla.spde.make.index(name = "spYR.idx", n.spde = spde.yr$n.spde)
A.spYR = inla.spde.make.A(mesh = mesh, loc = cbind(dat.train$East, dat.train$North))

# Make data stack
dat.stack.iid <- inla.stack(
  data = list(y = dat.train$accum), 
  A = list(1, 1, 1, A.spat, A.spYR),
  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
                        slope=dat.train$slope, aspect=dat.train$aspect, 
                        u10m=dat.train$u10m, v10m=dat.train$v10m, 
                        mu.u10=dat.train$mu.u10, mu.v10=dat.train$mu.v10, 
                        Year=dat.train$Year), 
                 list(time = dat.train$Year), 
                 spat.idx, 
                 spYR.idx),
  tag = 'dat')

time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
prior.fixed = list()
prec.prior = list(prior = 'pc.prec', param = c(250, 0.1)) #P(sigma > 250) = 10%

# form.iid = y ~ -1 + Intercept + dem + slope + 
#   # f(spat.idx, model = spde) + 
#   Year + f(spat.idx, Year, model = spde.yr) +
#   f(time, model = 'ar1', hyper = time.spec)

form.iid = y ~ -1 + Intercept + dem + 
  # slope +
  aspect + 
  # mu.u10 + mu.v10 + 
  # u10m + v10m + 
  f(spat.idx, model = spde) +
  Year + f(spYR.idx, Year, model = spde.yr) +
  # f(spat.idx, Year, model = spde.yr) +
  f(time, model = 'ar1', hyper = time.spec)
```

```{r}
mod.iid = inla(form.iid, 
               data = inla.stack.data(dat.stack.iid), 
               family = 'Gamma',
               quantiles = NULL, 
               control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack.iid), link = 1), 
               # control.fixed = prior.fixed, 
               control.family = list(hyper = list(prec = prec.prior)), 
               control.compute = list(waic=TRUE, return.marginals.predictor=TRUE, config=TRUE))

```

## Evalulate model performance

```{r}
summary(mod.iid)
```

```{r}

t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {

  if (any(class(mod.marg) == "matrix")) {
    mod.marg = list(mod.marg)
    if (!is.null(param.names)) {
      names(mod.marg = param.names)
    } else {
      names(mod.marg) = "UNNAMED"
    }
  }

  if (is.null(param.names)) {
    param.names = names(mod.marg)
  }

    if (is.list(t.func) != TRUE) {
    t.func = list(t.func)
  }
  if (length(t.func) == 1) {
    t.func = rep(t.func, length(param.names))
  }

  # Preallocate output tbl
  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))

  marg.names = names(mod.marg)

  for (i in 1:length(marg.names)) {
    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[param.names[i]]], n=n)
    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
  }
  return(out)
}
```

```{r}
marg.fixed = t.marg(mod.iid$marginals.fixed, function(x) exp(x))
ggplot(marg.fixed, aes(x=x,y=y)) + geom_line() + facet_wrap(vars(Param), scales = "free")

marg.hyper = t.marg(mod.iid$marginals.hyperpar, function(x) (x))
ggplot(marg.hyper, aes(x=x,y=y)) + geom_line() + facet_wrap(vars(Param), scales = "free")

```

```{r}
Bi.sp = tibble(East = mesh$loc[,1], North = mesh$loc[,2], 
                 B.i = mod.iid$summary.random$spYR.idx$mean) %>% 
  mutate(B.tot = B.i+mod.iid$summary.fixed["Year",]$mean)
ggplot(Bi.sp) + geom_density(aes(x=B.tot))
# ggplot(Bi.sp) + geom_density(aes(x=B.i))

```

Compare results to a normal GLM to ensure correct interpretation
```{r}
# Formula without random variables
f.base = y ~ -1 + Intercept + Year + dem + aspect

# Fitting a frequentist GLM to data
mod.glm = glm(f.base, data = dat.train %>% rename(y=accum) %>% mutate(Intercept=1), 
               family = Gamma(link = "log"))

summary(mod.glm)
```

All of this looks good except for when looking at the individual summaries spatial random effects on the temporal trend.
In this case, the sd for individual estimates is some 4-5 orders of magnitude larger than the mean, making them virtually worthless.
This is visually shown in the below plot (and recall that this is the untransformed values, so after exponentiation this just becomes ridiculous).
```{r}
plot(mod.iid$marginals.random$spYR.idx$index.7, type='l')
```

### 2023-05-23 docs/article-draft.qmd
New quarto draft article

--- 2022-07-11 scripts/article-notebook.Rmd
@@
 ---
-title: "Publication notebook"
-output: html_notebook
+title: "Modeling Article Draft"
+format: html
+  # embed-resources: true
+editor: source
+execute: 
+  echo: false
+  warning: false
+  error: false
+number-sections: true
+bibliography: WAIS-trends.bib
 ---
 
-This includes the analysis and figures that will go into the draft publication of the central WAIS trends article.
-This will also be a space for drafting and iterating on the wording for methods/analysis/interpretation description.
+## INTRODUCTION {#sec-intro}
 
-## METHODS
+The West Antarctic Ice Sheet (WAIS) is highly relevant to our understanding of Southern Hemisphere climate dynamics and to projections of future sea level.
+Already the WAIS is contributing to global sea level changes, having added at least an additional $6.9 \pm 0.6$ mm since 1979 [@rignot_four_2019].
+An understanding of snow accumulation patterns and processes is an essential component of sea level rise calculations, both directly as a contribution to overall mass balance and indirectly, as the surface mass balance (SMB) affects estimates of ice sheet net mass change from other methods---e.g.
+altimetry and gravimetry measurements [@velicogna_timevariable_2013, shepherd_trends_2019].
+Such effects on the net mass change are nontrivial and changes in accumulation can even dominate the integrated height change signal in at least some circumstances [@kuipersmunneke_elevation_2015].
+Additionally, understanding the recent spatio-temporal trends and patterns in precipitation over recent decades provides insight into how atmospheric and oceanic dynamics are changing in the Southern Hemisphere and globally.
 
-### Data imports and preprocessing
+Despite the critical need for increased understanding, our knowledge of West Antarctic SMB is hampered by a paucity of data and data coverage [@koenig_remote_2014].
+The extremely remote and harsh environment of the region make *in-situ* measurements challenging and expensive to collect.
+SMB is often highly spatially variable [@spikes_variability_2004, @banta_spatial_2008, @dattler_significant_2019], leading to questions of how representative small samples of measurements are of the ice sheet as a whole.
+Although satellite-based radar altimeters provide widespread spatial coverage, they typically have coarse resolutions of tens of kilometers and the degree of signal penetration beneath the surface complicates interpretations of height changes[ \cite{]@lenaerts_observing_2019].
+More recent lidar altimeters address some of these limitations, but have a limited temporal range [@smith_pervasive_2020].
+Both radar and lidar altimetry results are further complicated by the need to disentangle signals between changes in SMB, dynamical ice flux, firn compaction, and isostatic uplift.
+Modeling results have seen enormous strides forward, but validation with independent estimates of SMB are still important, especially in data-sparse regions like West Antarctica [@lenaerts_observing_2019].
+Additionally, models still have too coarse of resolution (often 10's of kilometers) to resolve small-scale (\textless5 km) spatial variability [@lenaerts_climate_2018] and still have limitations on how well they can capture interannual variability in Antarctic SMB [@wang_comparison_2016.
+Radar has been used for decades to estimate SMB [@spikes_variability_2004, @anschutz_smallscale_2008, @eisen_groundbased_2008], but still suffers from limitations of homogeneous coverage and processing bottlenecks usually involving labor-intensive manual work, especially for reconstructing annually-resolved SMB time series [@medley_airborneradar_2013, @lenaerts_observing_2019.
 
-Load libraries:
-```{r}
-library(here)
-library(INLA)
-library(dplyr)
-library(tidyr)
-library(broom)
-library(ggplot2)
-library(colorspace)
-library(spdep)
-```
+In spite of these difficulties, substantial progress has been made towards improved understanding of SMB over recent decades in the WAIS.
+Numerous ice core studies suggest accumulation rates in and near the Antarctic Peninsula have recently increased [@thomas_doubling_2008, @thomas_308_2013, @thomas_twentieth_2015, @hosking_accumulation_2017, @medley_increased_2019].
+This observation is supported by results from a number of climate models and reanalyses \cite{wang_comparison_2016}.
+Evidence also exists for decreased accumulation over similar time periods for the western portions of WAIS, near the Ross Sea \cite{wang_comparison_2016, wang_snow_2017, medley_increased_2019}, although the \textit{in-situ} observations in this region are more sparse and largely out of date \cite{thomas_regional_2017}.
+As a whole, the central region of WAIS shows smaller changes in snow accumulation, but such results vary considerably both spatially across central WAIS and between studies \cite{banta_spatial_2008, medley_airborneradar_2013, burgener_observed_2013, wang_comparison_2016, thomas_regional_2017, wang_snow_2017}.
 
-```{r}
-data.all = readRDS(here('data/Rdata-clean.rds'))
-gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
+Important questions therefore remain regarding the spatial distribution of accumulation trends across WAIS and particularly regarding the nature and location of the transition in central WAIS between the predominantly increasing accumulation of eastern WAIS and the Antarctic Peninsula and the decreasing accumulation in western WAIS.
+Understanding these transitions are important to properly interpret the climatic elements driving the observed changes and to facilitate improved ice sheet-wide estimates of trends in annual SMB.
 
-# # Clip bounds
-# bbox = st_bbox(gdf_traces)
-# clipper = c(
-#   bbox[1]+0.1*(bbox[3]-bbox[1]),
-#   bbox[2]+0.5*(bbox[4]-bbox[2]),
-#   0.6*(bbox[3]-bbox[1])+bbox[1], 
-#   bbox[4]
-# )
-# gdf_traces = st_crop(gdf_traces, clipper)
+In this study we construct annually-resolved SMB time series for a region of central WAIS from airborne radar at high spatial resolution and investigate the subsequent spatio-temporal patterns and trends in the data.
+The method for estimating SMB time series from radar is fully automated, includes uncertainty estimation for individual years and locations, and includes a quality-control routine that can automatically remove data poorly suited to the method.
+We use these data to inform a Bayesian statistical model that can leverage the data uncertainties and extend estimates to unobserved spacetime points across the region.
+This modeling therefore permits a more spatially detailed and statistically robust investigation of recent changes in SMB across central WAIS.
 
-# Remove locations with less than 10 years of data
-trace.keep = data.all %>% count(trace_ID) %>% filter(n>=10)
-data.all = data.all %>% filter(trace_ID %in% trace.keep$trace_ID)
-gdf_traces = gdf_traces %>% filter(trace_ID %in% trace.keep$trace_ID)
+## METHODS {#sec-methods}
 
-# Subset data to every 5th location (coarsens data to ~1 km)
-skip.int = 5
-gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
-gdf_traces = gdf_traces[gdf.idx,]
+### Data and study site {#sec-data}
 
-# Filter data to subsetted period
-data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
-  filter(Year >= 1975) %>% filter(Year < 2015) %>% arrange(trace_ID, Year)
+![The region of interest in central West Antarctica and the data sets used for this study.
+Background image denotes 1-km resolution surface topography as a hillshade, derived from the DEM in \citet{slater_new_2018}.
+Colored lines show elevation contours (150 m intervals).
+NASA OIB Snow radar results from PAIPR are shown in red and firn/ice core locations are denoted with blue triangles.
+The figure also denotes the location of WAIS Divide Camp.](Figures/data_panel.png){#fig-datamap}
 
-# Remove gdf rows where all data have been filtered out
-gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)
+This study focuses on a ~250,000 km^2^ region of central West Antarctica (@fig-datamap).
+The region spans a range of accumulation rates (>200 mm w.e. to <750 mm w.e.) and topographic settings, encompassing a portion of the ice divide between the Amundsen Sea and the Ross Sea sectors.
+Although some ice velocities in the coastal areas of this region can exceed 200 m/a, typical *in-situ* velocities for our data coverage are ~20 m/a or less \citep{rignot_ice_2011}.
+One reason this region is significant is because it encompasses portions of the Pine Island Glacier and Thwaites Glacier drainage basins.
+These outlet glaciers had a net mass loss of 1,700 gigatonnes of ice for the period 1979-2017, with the annual contribution also accelerating over that time period \cite{rignot_four_2019}.
+This region additionally has an acceptably dense network of firn density measurements (a total of 16 cores) to properly constrain depth-density modeling (discussed in @sec-density) with increased confidence.
+Finally, this region is between the Antarctic Peninsula---where strong evidence exists of increased accumulation rates over recent decades---and the region near the Ross Sea---where evidence exists for decreased accumulation rates over recent decades \citep[e.g.]{medley_increased_2019, thomas_regional_2017}.
+It is therefore relevant to understanding the extent, boundaries, and patterns of changes in accumulation rates for the entire WAIS.
 
-# Select variables of interest
-dat = data %>% select(trace_ID, East, North, Year, accum, 
-                      std, SMB, elev, dem, slope, aspect, 
-                      # u10m, v10m, mu.u10, mu.v10
-                      )
+The main data used in this study are radar echograms from NASA's Operation IceBridge (OIB) Snow Radar.
+This frequency-modulated continuous-wave radar operates in a frequency range (2-8 GHz) with a range resolution (~5 cm in firn) that permits imaging of annual resolution in Antarctic firn \cite{panzer_ultrawideband_2013}.
+We use portions of 9 OIB flights over the region of interest with collection times spanning 2010-2016.
 
-# Center covariates
-dat = dat %>% 
-  mutate(elev = elev-mean(elev, na.rm=TRUE), 
-         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
-         # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
-         # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
-         ) %>% 
-  mutate(Year.mod = Year-mean(Year), Year.idx = (Year-min(Year)+1))
-         # u10m=mean(u10m, na.rm=TRUE), v10m=mean(v10m, na.rm=TRUE))
+Estimating annual accumulation from radar echograms relies on the ability of microwave electromagnetic radiation to penetrate the ice sheet surface and propagate through the snow and firn.
+Seasonal variations in the dielectric properties of the preserved firn result in internal reflection horizons (IRHs) in the radar returns.
+When continuous over many kilometers, these IRHs are most often interpreted as isochronous seasonal layers that can be used to determine the age of deposition \cite{eisen_groundbased_2008}.
+To determine accumulation rates, the echogram time-domain must be converted to depth.
+This requires estimates of firn density with depth to determine radar propagation velocity.
+The firn depth-density profile is also used when converting layer thickness to cumulative mass with depth.
+The cumulative mass profile combined with the calculated ages of the IRHs yields annual estimates of accumulation.
 
-# Scale covariates by sd
-dat = dat %>% 
-  mutate(elev = elev/sd(elev), dem=dem/sd(dem), 
-         slope=slope/sd(slope), 
-         # u10m=u10m/sd(u10m), v10m=v10m/sd(v10m), mu.u10=sd(mu.u10), mu.v10=sd(mu.v10)
-         )
+In addition to the radar data used, this study utilizes a suite of ice and firn cores, both for informing the the depth-density model and for comparisons of accumulation rates and trends generated with the radar data.
+We use density data from the Satellite Era Accumulation Traverse \cite{burgener_observed_2013}, the WAISCORES program \cite{lamorey_waiscores}, the International Trans-Antarctic Scientific Expedition \cite{mayewski_international_2005}, and additional cores included in the Surface Mass Balance and Snow on Sea Ice Working Group data set \cite{montgomery_sumup_2018}.
+For comparisons of annual accumulation rates and trends we additionally use cores included in the Past Global Changes Antarctica 2k compilation \cite{thomas_regional_2017}.
 
-# Split into training and testing sets
-dat = dat %>% mutate(row.ID = 1:nrow(dat)) %>% relocate(row.ID)
-dat.train = dat %>% slice_sample(prop = 0.80) %>% arrange(row.ID)
-dat.test = dat %>% filter(!row.ID %in% dat.train$row.ID)
-```
+### Depth-density modeling {#sec-density}
 
-Directly calculate linear coefficients of time from raw data.
-```{r}
-yr.trends = dat.train %>% 
-  group_by(trace_ID) %>% 
-  do(tidy(lm(accum ~ Year.mod, data = .))) %>% 
-  filter(term=='Year.mod') %>% select(-term)
-dat.mu = dat.train %>% group_by(trace_ID) %>% 
-  summarize(East=mean(East), North=mean(North), accum=mean(accum)) %>% 
-  left_join(yr.trends %>% select(trace_ID, estimate)) %>% 
-  mutate(log.est = log(1+(estimate/accum)))
-```
+The calculation of radar accumulation time series requires estimates of depth-density profiles to convert radar echograms from the time domain into the depth domain and to estimate annual accumulation from layer thickness.
+Various methods have been proposed for estimating the variation of density with depth in Antarctic firn, typically either interpolating from measured ice core or stake farm data \cite{medley_airborneradar_2013} or based on the physical modeling and evolution of firn properties \cite{ligtenberg_improved_2011}.
 
-### Model setup
+In this study, we use a statistical estimation and interpolation approach informed by *in-situ* observations as described in \citet{white_hierarchical_2021}.
+This model uses a positive-valued Bayesian semi-parametric model over space and depth below the ice sheet as its building block.
+To impose the monotone increasing shape of the depth-density curve, we integrate the positive-valued function over the depth domain.
+The integrated function is transformed to be bounded below by 0 g/cm$^3$ and 0.917 g/cm$^3$ to constrain the model's mean function to appropriate values for firn/ice density.
+This model imposes smoothness of firn density estimates in the space and depth domains, constrains the shape of firn density curves, and enables simple interpolation (see \citet{white_hierarchical_2021} for more details).
+There are several advantages afforded by this method, most importantly its ability to leverage space-depth autocorrelation to interpolate the shape-constrained depth-density profiles to new locations with relatively few model assumptions.
+Moreover, because we fit the model in a Bayesian framework, our method provides rigorous uncertainty quantification of the depth-density profiles at arbitrary sites.
+This allows us to effectively propagate uncertainty in firn density estimates to SMB estimates.
 
-The accumulation data represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension (e.g.
-annually resolved time series).
+### SMB estimation with PAIPR
+
+In this study, we estimate annually-resolved accumulation rates from OIB Snow radar using an automated process, the Probabilistic Annual Isochrone Picking Routine---i.e.
+PAIPR \cite{keeler_probabilistic_2020}.
+This algorithm picks continuous IRHs using localized Radon transforms and assigns each a likelihood of representing an annual isochrone based on the relative strength of the radar return and the continuous length of the horizon.
+These likelihoods, along with errors in the depth-density modeling (discussed above), are then used in Monte Carlo sampling to generate uncertainty estimates in the the final annual accumulation estimates.
+The algorithm processes radar data in 30-km chunks, generating results with 25-meter along-track spacing which we aggregate to 200-meter, annually-resolved accumulation distributions.
+We limit analysis to the upper ~25 meters of the firn to ensure no issues with loss of layers due to excessive attenuation of IRH signals with depth.
+The complete period of data coverage therefore varies spatially due to varying accumulation rates, spanning from ten years to fifty-seven years of total record duration.
+Further details of the method are available in \citet{keeler_probabilistic_2020} and \citet{keeler_repeatability_2021}.
+
+The primary advantages of the PAIPR method over more traditional techniques are the fully automated nature of the method and the more complete uncertainty assessment.
+The automated nature of PAIPR allows a wide and rapid application to larger OIB data sets.
+The addition of the quality control routine (see \citet{keeler_repeatability_2021} for details on its implementation) permits this application without careful vetting of input data, as poorly-suited data is flagged and can be removed from additional analyses.
+The inclusion of uncertainties (resulting from both input data and modeling errors) allows a more rigorous assessment of the relevance and significance of observed magnitudes, patterns, and trends.
+These uncertainies are further utilized in our Bayesian modeling framework (see @sec-bayesian).
+
+The PAIPR results still have important limitations worth noting.
+The degradation in signal with depth limits the time period over which reliable results are generated, with the temporal duration controlled by the *in-situ* accumulation rate.
+PAIPR cannot produce time series of long duration in high accumulation regions, limiting interpretations of temporal variability and trends in these regions.
+Additionally, the method is reliant on nearby depth-density measurements or modeling, affecting results both in the depth migration steps and in the cumulative mass estimates.
+Reliable SMB estimates therefore rely on sufficient firn/ice core coverage in the region of interest, or else reliable firn density modeling where *n-situ* measurements are not available.
+The method also does not account for dynamical thinning of annual layers.
+The shallow firn column used in PAIPR (~25 m) limits the impact of this issue, but such effects could still be relevant in regions of high ice velocity (e.g.
+\citet{mouginot_continentwide_2019} show some regions of WAIS near the coast can have >1 km a^-1^ ice velocities).
+\citet{konrad_observations_2019} however, found that even for the region of Pine Island Glacier, ~80% of strain rate correction magnitudes were lower than the uncertainty in radar-derived SMB estimates in the region.
+
+The layer-likelihood parameters in PAIPR were trained using manually-traced layers and validated against firn core data.
+The results could therefore be biased towards the specific regions used to train the likelihood parameters.
+The training data were selected to contain a spatially-distributed sample of central WAIS with a range of accumulation rates and conditions.
+This variability in training data, combined with the MC simulations on the uncertainty in the tuned parameters, should make this method applicable to a wide range of conditions, but would be less suited to regions of Antarctica significantly different than the range of conditions represented in the training data.
+Finally, the SMB results are still limited by data availability.
+NASA's Operation IceBridge focused on specific regions and research goals, affecting the selection of flight plans and regions of study.
+This results in some areas with a high density of data coverage (both spatially and temporally) with much more limited or no coverage in other regions (e.g.
+interior portions of the Antarctic ice sheets).
+Even with such limitations, the spatial coverage afforded by OIB SMB results is orders of magnitude greater than traditional firn/ice core, automatic weather stations, or stake farm measurements, regions poorly suited to PAIPR application are automatically filtered from results, and rigorous uncertainties are generated for SMB estimates.
+
+### Data processing and modeling {#sec-modeling}
+
+The final output from PAIPR includes estimates of the modal value and standard deviation of accumulation for each year of a given trace (25-meter along-track spacing, aggregated into 200-meter bins to increase estimate sample size) and a quality rating for whether the results are usable based on the suitability of PAIPR to the particular echogram.
+We remove time series with less than 10 years of estimates, as we frequently observed inaccurate results for regions with few time series elements.
+To avoid biases in early year estimates (i.e. the earliest years are only available for regions of low accumulation rates), we focus on study on the time period 1975--2015.
+This 40-year time period exhibits good spatial coverage across the full region.
+After excluding data flagged as poor and filtering the remaining data with the above criteria, the final dataset consists of 683,676 individual estimates of annual accumulation at 21,887 spatial locations.
+
+#### Bayesian model {#sec-bayesian}
+
+We use a Bayesian approximation method to investigate the spatial variability in the temporal trend in accumulation rates across the study region.
+This choice confers a number of benefits.
+The explicit and joint-probability uncertainty in a Bayesian approach allows more full utilization of the PAIPR-generated uncertainties and provides uncertainty for every parameter and estimate within the model.
+A Bayesian approach also incorporates domain-specific a priori knowledge into model estimates.
+The use of an approximation method vastly improves computation time and makes analyses of large datasets (e.g. times series generated from PAIPR) computationally tractable.
+We utilize Integrated Nested Laplace Approximations INLA) to leverage the noted advantages.
+This method has recently seen widespread adoption in modeling spatiotemporal physical phenomena in a variety of fields \cite{}.
+
+*Will move most of the following to a supplement, with only a brief summary included in the main document.*
+
+The PAIPR-derived accumulation time series represent a continuous distribution of a random variable over two continuous spatial dimensions and indexed over an additional discrete temporal dimension (i.e.
+individual years).
 We therefore choose to model accumulation as a Gamma distribution to ensure a zero-bounded function with increased variance at higher expected values as also observed in the raw data.
 This furthermore matches established literature where precipitation values are frequently assumed to follow a Gamma distribution **(citations)**.
-We express this distribution in terms of the expected accumulation for an arbitrary space location and time point $\mu_{s,t}$ and a precision parameter at the corresponding time $\phi_t$ 
-$$\mathbf{Y} \sim Gamma(\mu_{s,t},\phi_t)$$
-where the variance at an arbitrary spacetime point is
-$$Var(s_t) = \frac{\mu_{s,t}^2}{\phi_t}$$
+We express this distribution in terms of the expected accumulation for an arbitrary space location and time point $\mu_{s,t}$ and a precision parameter at the corresponding time $\phi_t$ $$\mathbf{Y} \sim Gamma(\mu_{s,t},\phi_t)$$ where the variance at an arbitrary spacetime point is $$Var(s_t) = \frac{\mu_{s,t}^2}{\phi_t}$$
 
@@
 
-We model the spatial random effect as a stochastic partial differential equation...
-
-We model the temporal random effect $\omega$ (representing how the mean spatial field evolves in time) as a second-order autoregressive model where
-$$\omega_t = \rho \omega_{t-1}$$
-An AR(2) process was selected due to the observed temporal autocorrelation in the data, which indicates that accumulation retains autocorrelation up to a 2-year lag time.
+We model the temporal random effect $\omega$ (representing how the mean spatial field evolves in time) as a second-order autoregressive model where $$\omega_t = \rho \omega_{t-1}$$ An AR(2) process was selected due to the observed temporal autocorrelation in the data, which indicates that accumulation retains autocorrelation up to a 2-year lag time.
 
@@
 
+## RESULTS
+
 ```{r}
-# More coarse mesh for development purposes
-mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
-                    max.edge = c(25000, 100000), cutoff = 1000)
-# mesh = inla.mesh.2d(loc = dat.train %>% select(East, North),
-#                     max.edge = c(30000, 90000), cutoff = 2000)
-plot(mesh)
-points(dat.train %>% select(East, North), col = "red")
+# Import requisite libraries
+library(here)
+library(INLA)
+library(dplyr)
+library(ggplot2)
+library(terra)
+library(colorspace)
 
-# 2D spatial Matern GMRF via SPDE (with assigned priors)
-spde = inla.spde2.pcmatern(mesh = mesh, 
-  prior.range = c(25000, 0.01), # P(range < 20 km) = 0.01
-  prior.sigma = c(1, 0.01)) #P(sd > 1) = 0.05
+# Load necessary data for analysis and plotting (generated using `scripts/article-script.R)
+mod.full = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/interim-models/mod-full.skip10.cut5.rds')
+valid.df = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/dfs/valid.df.rds')
+results.df = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/dfs/results.df.rds')
+grid.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/grid.tif')
+mu.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/mu.tif')
+sd.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/sd.tif')
 
-# Locator index for spatial random effect
-spat.idx <- inla.spde.make.index('spat.idx', spde$n.spde)
-
-# Projector matrix for spatial locations (and indexed on Year for trends)
-A.spat <- inla.spde.make.A(mesh, 
-                         loc=cbind(dat.train$East, dat.train$North), 
-                         weights = dat.train$Year.mod)
-
-# Make data stack
-dat.stack <- inla.stack(
-  data = list(y = dat.train$accum), 
-  A = list(1, 1, 1, A.spat),
-  effects = list(list(Intercept = rep(1, nrow(dat.train))), 
-                 tibble(elev = dat.train$elev, dem=dat.train$dem, 
-                        slope=dat.train$slope, aspect=dat.train$aspect),
-                 list(time = dat.train$Year.mod),
-                 spat.idx),
-  tag = 'dat')
-
-# Prior for temporal autocorrelation
-time.spec = list(rho = list(prior = 'pc.cor1', param = c(0.3, 0.95)))
-
-# Model formula
-f.mod = y ~ -1 + Intercept + #Global intercept
-  dem + #Fixed effects
-  # f(dem, model='clinear', range=c(-Inf,0)) + #Fixed effects
-  f(time, model = 'ar1', hyper = time.spec) +  #Temporal random effect modeled as AR1
-  f(spat.idx, model = spde) #Spatial randomed effect (indexed by Year)
 ```
 
-Run and summarize model
-```{r}
-# Model formula
-mod = inla(f.mod,
-           data = inla.stack.data(dat.stack),
-           family = 'Gamma',
-           control.predictor = list(compute = TRUE, A = inla.stack.A(dat.stack), link=1),
-           # control.inla = list(int.strategy='eb'), #Improves comp time at cost of accuracy/precision
-           control.compute = list(waic=TRUE, config=TRUE))
-
-n.iter = 1
-while (mod$mode$mode.status>0 && n.iter<=3) {
-  print("Issues with Hessian. Re-running to solve negative eigenalues in the Hessian")
-  n.iter = n.iter + 1
-  print(paste("Starting Iteration", n.iter, "of INLA..."))
-  mod = inla.rerun(mod)
-}
-
-if (mod$mode$mode.status > 0) {
-  print("Issues with Hessian eigenvalues persist. Treat model results with suspicion!")
-}
-```
-
-```{r}
-summary(mod)
-```
-
-## Model results
+Estimates of mean annual accumulation (shown in \figref{accum-map}) are similar to those found by \citet{medley_airborneradar_2013} using manually traced IRHs from OIB Snow Radar echograms.
+Accumulation is highest (and exhibits the highest variability) close to coastal West Antarctica and progressively decreases towards the interior.
+Considerable spatial variability, however, overlays this broader signal.
+The most likely source of this higher-order spatial variability is accumulation and wind re-distribution governed by surface topography and near-surface winds \cite{dattler_significant_2019}.
 
@@
 }
+
+# Extract parameter posteriors
+marg.fixed = t.marg(mod.full$marginals.fixed, function(x) exp(x))
+HP.names = c("Gamma precision", 
+             "Time precision", "Rho for time", 
+             "Range for spat.idx", "Stdev for spat.idx")
+marg.hyper = t.marg(mod.full$marginals.hyperpar, function(x) (x), param.names = HP.names)
+p.selection = c("Intercept", "dem", "Gamma precision", 
+                "Time precision", "Rho for time", 
+                "Range for spat.idx", "Stdev for spat.idx")
+posts = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
 ```
@@
 ```{r}
-marg.fixed = t.marg(mod$marginals.fixed, function(x) exp(x))
-HP.names = c("Gamma precision", 
-             "Time precision", "Rho for time", 
-             "Range for spat.idx", "Stdev for spat.idx")
-marg.hyper = t.marg(mod$marginals.hyperpar, function(x) (x), param.names = HP.names)
-p.selection = c("Intercept", "dem", "Gamma precision", 
-                "Time precision", "Rho for time", 
-                "Range for spat.idx", "Stdev for spat.idx")
-
-posts = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
-```
-
-```{r posteriors, fig.width=16}
+#| label: fig-ParamPost
+#| fig-cap: "Posterior distributions of model parameters. Blah blah blah."
 ggplot(posts, aes(x=x,y=y)) + geom_line() + 
@@
 
-### Components of prediction
+The region overall shows generally insignificant trends in annual SMB for the period 1975-2015 (\figref{trend-map}).
+This is in agreement with a number of previous studies that found insignificant changes in accumulation during the latter portion of the 20th century.
+\citet{medley_airborneradar_2013} found no significant trends in SMB for the period 1980-2009 when compositing all OIB Snow Radar time series in their study for an overlapping region.
+\citet{wang_snow_2017} found little to no trend (0.18% per decade) in snow accumulation in central WAIS based on composite ice cores in the region for the period 1900-2010 and weakly positive trends (2.08% per decade) for the same region and time period using ERA-20C reanalysis data.
+\citet{wang_comparison_2016} compared the results of several climate models and reanalyses for the period 1979-2012, including ERA-Interim, JRA-55, MERRA, PMM5, RACMO2.1, and RACMO2.3.
+They found considerable spatial variability in the linear trends in SMB for the central WAIS, including opposing signs of the trend, but none showed significant trends over this period in the region.
 
 ```{r}
-# Define points of interest to break out components of prediction
-pred.trace = gdf_traces %>% mutate(color=as.factor(trace_ID))
-
-# Get raw trend calculations and variable values at locations of interest
-dat.pred = dat %>% filter(trace_ID %in% pred.trace$trace_ID) %>% mutate(color=as.factor(trace_ID))
-Years = sort(unique(dat.train$Year))
-loc.pred = st_coordinates(pred.trace) %x% rep(1, length(Years))
-covar.pred = dat.pred %>% select(trace_ID, dem) %>%
-  group_by(trace_ID) %>% filter(row_number()==1)
-yr.mod = rep(Years - mean(Years), nrow(pred.trace))
-st.pred = tibble(X=loc.pred[,1], Y=loc.pred[,2], Year = rep(Years, nrow(pred.trace)), 
-                 dem=rep(covar.pred$dem, each=length(Years)))
-
-# Project prediction locations onto mesh
-proj.pred = inla.mesh.projector(mesh=mesh, loc = loc.pred)
-
-# Extract spatial RE of trend and time RE at mesh nodes
-spat.trend = inla.mesh.project(proj.pred, mod$summary.random$spat.idx$mean)
-time.RE = rep(mod$summary.random$time$mean, nrow(pred.trace))
-
-# Construct linear predictor for each spacetime point
-lin.est = mod$summary.fixed$mean[1] + 
-  mod$summary.fixed$mean[2]*st.pred$dem + spat.trend*yr.mod + time.RE
-
-# Get model components for each st point
-st.pred = st.pred %>% 
-  mutate(fx.RE=mod$summary.fixed$mean[2]*st.pred$dem, spat.trend=spat.trend*yr.mod, 
-         time.RE=time.RE, pred=exp(lin.est)) %>% 
-  mutate(pred.sd=sqrt(pred^2/mod$summary.hyperpar$mean[1]), Source="model", 
-         color=rep(pred.trace$color, each=length(Years)))
-
-# Average data for each spatial location and add trend estimates
-spat.pred = st.pred %>% select(-Year, -spat.trend, -Source, -time.RE) %>% 
-  group_by(color) %>% summarize_all(mean, rm.na=TRUE)
-spat.pred['trend'] = spat.trend[seq(1,length(spat.trend), 
-                                    by=(length(spat.trend)/nrow(spat.pred)))]
+#| label: fig-trends
+#| fig-cap: "Mean posterior trends across the region with associated errors."
+plot(grid.rast$trend)
+plot(grid.rast$trend.sd)
 ```
 
-```{r}
-ggplot(spat.pred, aes(x=X,y=Y, color=pred)) + geom_point() + scale_color_viridis_c()
-ggplot(spat.pred, aes(x=X,y=Y, color=fx.RE)) + geom_point() + scale_color_gradient2()
-ggplot(spat.pred, aes(x=X,y=Y, color=trend)) + geom_point() + scale_color_gradient2()
-```
+Although the majority of our time series show insignificant trends, large spatial heterogeneity exists in the results, with 34% of the total results showing significant negative trends (95% confidence interval) for 1975-2015.
+<1% of the trend results show statistically significant positive trends in this time interval.
+More negative trends tend to occur in the western portions of the region, with less negative and insignificant trends towards the east, but with important variability in these general observations (\figref{trend-map}).
+95% of accumulation changes in the region vary between -65 mm decade^-1^ and +7.2 mm decade^-1^ (-17% decade^-1^ and 2.1% decade^-1^ relative to the multi-decadal mean accumulation).
+The observed trend magnitudes are similar to those observed in ice and firn cores across the region and for a wider area of WAIS (\textbf{d} in \figref{trend-map}), but provide important context of the spatial variability and patterns in those trends.
+Although individual cores suggest a similar pattern, they alone are insufficient to resolve the statistical significance of the overall results or to properly demonstrate the spatial nature and heterogeneity of the patterns.
 
-Direct comparisons for individual trace time series between observed and predicted
-```{r}
-pred.comp = dat.pred %>% select(color, Year, accum, std, SMB) %>% 
-  right_join(st.pred, by=c("color", "Year")) %>% 
-  mutate(log.diff=log(pred/accum))
+\citet{medley_increased_2019} used ice core compilations combined with reanalysis-derived spatial coherence patterns to estimate Antarctic-wide snow accumulation.
+Their results also show a strong east-west dipole in SMB over WAIS (thought to be driven by the Southern Annular Mode) in the period 1957-2000, with enhanced (decreased) precipitation towards the Antarctic Peninsula (Ross Sea).
+\citet{wang_new_2019} observed a similar dipole in SMB trends for the period 1950-2010 using a kriging reconstruction from ERA-Interim outputs informed by ice core records.
+Our results agree broadly with these conclusions, but indicate greater variability in the trend direction for central WAIS compared to either study, with significant negative trends for the period 1979-2010 extending farther east.
+These results are specifically in agreement with the findings of \citet{burgener_observed_2013}, who found decreasing accumulation since 1975 from several firn cores near WAIS Divide.
 
-ggplot(pred.comp) + 
-  geom_line(aes(x=Year, y=log.diff, group=color), size=0.25, alpha=0.1) + 
-  theme(legend.position="none")
-ggplot(pred.comp) + geom_abline(slope=1, color='black') + 
-  geom_point(aes(x=accum, y=SMB), color='red', alpha=0.1) + 
-  geom_point(aes(x=accum, y=pred), color='blue', alpha=0.1)
-ggplot(pred.comp) + geom_density(aes(x=log(pred/accum)), fill='red', alpha=0.3)
-```
+\citet{thomas_regional_2017}, using composite ice core records and SMB from RACMO2.3p2, found that for the portion of WAIS between 90&deg;--150&deg; W, 50-year running linear trends varied mostly within $\pm1$ mm a^-1^, with the period 1961-2010 showing a trend of ~0.65 mm a^-1^.
+Our results---representing a region comprising ~1/3 of the WAIS section in \citet{thomas_regional_2017}---show a spatially integrated trend of -1.8 mm a^-1^ (-0.49% a^-1^) for the whole region of study, but larger variability spatially of -6.5 mm a^-1^ to +0.72 mm a^-1^.
+Importantly, the WAIS regions between 90&deg;--150&deg; W not included in our results are mostly from the Ross Sea region.
+This region exhibits decreasing accumulation rates in recent decades \cite{wang_snow_2017, bertler_ross_2018, medley_increased_2019, wang_new_2019}, so its exclusion in our study is unlikely to explain the discrepancy between our results and \citet{thomas_regional_2017}.
 
-```{r}
-mod.time = tibble(Year=Years, mu=mod$summary.random$time$mean, 
-                  sd=mod$summary.random$time$sd)
-tmp = pred.comp %>% group_by(Year) %>% summarize(accum=mean(accum, na.rm=TRUE), pred=mean(pred))
-mod.time = mod.time %>% mutate(obs.diff=log(tmp$accum/mean(tmp$accum)), 
-                               pred.diff=log(tmp$pred/mean(tmp$pred)))
-ggplot(mod.time) + geom_hline(yintercept = 0, color='black', linetype="dashed") + 
-  geom_ribbon(aes(x=Year, ymin=mu-sd, ymax=mu+sd), fill='red', alpha=0.2) + 
-  geom_line(aes(x=Year, y=mu), color='red') +
-  geom_line(aes(x=Year, y=obs.diff), color='blue', alpha=0.5) + 
-  geom_line(aes(x=Year, y=pred.diff), color='purple', alpha=0.5)
-```
+## References
 
-Calculate probability that trend is less than zero.
-```{r}
-P.neg = vector(mode = "numeric", length = length(mod$marginals.random$spat.idx))
-for (i in 1:length(P.neg)) {
-  marg.i = inla.tmarginal(function(x) x, 
-                          mod$marginals.random$spat.idx[[i]], 
-                          n=500)
-  P.i = diff(marg.i[,1]) * marg.i[1:(nrow(marg.i)-1),2]
-  P.neg[i] = sum(P.i[which(marg.i[1:(nrow(marg.i)-1),1]<0)])
-}
-tmp = tibble(Source='model', 
-             East=mesh$loc[,1], North=mesh$loc[,2], 
-             med=mod$summary.random$spat.idx$`0.5quant`, 
-             P.neg=P.neg, 
-             LB=mod$summary.random$spat.idx$`0.025quant`, 
-             UB=mod$summary.random$spat.idx$`0.975quant`) %>% 
-  st_as_sf(coords=c("East", "North"), crs=3031)
-```
-
-Map of estimated trends with probability of less than zero
-```{r}
-trends.comp = dat.mu %>% select(-trace_ID, -accum, -estimate)  %>% 
-  rename(med=log.est) %>% 
-  mutate(LB=NA, UB=NA, P.neg=NA, Source="data") %>% 
-  st_as_sf(coords=c("East", "North"), crs=3031) %>%
-  bind_rows(tmp)
-```
-
-```{r}
-ggplot(trends.comp) + geom_density(aes(x=med, group=Source, color=Source)) + 
-  xlim(c(-0.07,0.07))
-
-plt_bnds = st_bbox(gdf_traces)
-ggplot(trends.comp, aes(color=med)) + 
-  geom_sf() + scale_color_gradient2() + 
-  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
-  facet_wrap(vars(Source))
-
-ggplot(trends.comp %>% filter(Source!="data")) +
-  geom_sf(aes(color=(UB-LB)/2)) + 
-  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
-  scale_color_viridis_c(option = "plasma", limits=c(0.003, 0.013))
-  # scale_color_viridis_c(option = "plasma", trans="log")
-
-ggplot(trends.comp %>% filter(Source!="data")) +
-  geom_sf(aes(color=P.neg)) + 
-  xlim(c(plt_bnds[1], plt_bnds[3])) + ylim(c(plt_bnds[2], plt_bnds[4])) + 
-  scale_color_continuous_diverging(palette="Tropic", mid=0.50)
-
-ggplot(trends.comp %>% filter(Source!="data")) + geom_density(aes(x=P.neg), fill='red', alpha=0.3)
-```
-
-```{r}
-mod.trends = trends.comp %>% filter(Source=="model")
-Neg.threshold = 0.95
-frac.Neg = nrow(mod.trends %>% filter(P.neg>=Neg.threshold))/nrow(mod.trends)
-print(paste(
-  format(100*frac.Neg, digits=2), 
-  "% of the region has at least ", 
-  format(100*Neg.threshold, digits=2), 
-  "% probability of a negative accumulation trend", 
-  sep=''))
-
-frac.Pos = nrow(mod.trends %>% filter(P.neg<=(1-Neg.threshold)))/nrow(mod.trends)
-print(paste(
-  format(100*frac.Pos, digits=2), 
-  "% of the region has at least ", 
-  format(100*Neg.threshold, digits=2), 
-  "% probability of a positive accumulation trend", 
-  sep=''))
-```
+::: {#refs}
+:::

### 2023-05-23 scripts/article-script.R

@@ -1,0 +1,305 @@
+# Script to generate data and perform analyses for article on 
+# Bayesian inference of WAIS-central accumulation trends
+
+# Import libraries
+library(here)
+library(INLA)
+library(dplyr)
+library(tidyr)
+library(broom)
+library(terra)
+library(spdep)
+
+## DATA IMPORT AND PREPROCESSING
+
+# Import raw(ish) data
+data.all = readRDS(here('data/Rdata-clean.rds'))
+gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
+
+# # Clip bounds
+# bbox = st_bbox(gdf_traces)
+# clipper = c(
+#   bbox[1]+0.1*(bbox[3]-bbox[1]),
+#   bbox[2]+0.5*(bbox[4]-bbox[2]),
+#   0.6*(bbox[3]-bbox[1])+bbox[1], 
+#   bbox[4]
+# )
+# gdf_traces = st_crop(gdf_traces, clipper)
+
+# Remove locations with less than 10 years of data
+trace.keep = data.all %>% count(trace_ID) %>% filter(n>=10)
+data.all = data.all %>% filter(trace_ID %in% trace.keep$trace_ID)
+gdf_traces = gdf_traces %>% filter(trace_ID %in% trace.keep$trace_ID)
+
+# Subset data to every 5th location (coarsens data to ~2 km)
+skip.int = 10
+gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
+gdf_traces = gdf_traces[gdf.idx,]
+
+# Filter data to subsetted period
+data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
+  filter(Year >= 1975) %>%
+  filter(Year < 2015) %>% arrange(trace_ID, Year)
+
+# Remove gdf rows where all data have been filtered out
+gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)
+
+# Select variables of interest
+dat = data %>% select(trace_ID, East, North, Year, accum, 
+                      std, SMB, elev, dem, slope, aspect, 
+                      # u10m, v10m, mu.u10, mu.v10
+)
+
+# Center covariates
+dat = dat %>% 
+  mutate(elev = elev-mean(elev, na.rm=TRUE), 
+         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
[…]

### 2023-06-01 docs/article-draft.qmd
Updates to fix citations; better figure rendering

@@
 title: "Modeling Article Draft"
-format: html
-  # embed-resources: true
+format: 
+  html:
+    self-contained: true
 editor: source
@@
 An understanding of snow accumulation patterns and processes is an essential component of sea level rise calculations, both directly as a contribution to overall mass balance and indirectly, as the surface mass balance (SMB) affects estimates of ice sheet net mass change from other methods---e.g.
-altimetry and gravimetry measurements [@velicogna_timevariable_2013, shepherd_trends_2019].
+altimetry and gravimetry measurements [@velicogna_time-variable_2013; @shepherd_trends_2019].
 Such effects on the net mass change are nontrivial and changes in accumulation can even dominate the integrated height change signal in at least some circumstances [@kuipersmunneke_elevation_2015].
@@
 The extremely remote and harsh environment of the region make *in-situ* measurements challenging and expensive to collect.
-SMB is often highly spatially variable [@spikes_variability_2004, @banta_spatial_2008, @dattler_significant_2019], leading to questions of how representative small samples of measurements are of the ice sheet as a whole.
-Although satellite-based radar altimeters provide widespread spatial coverage, they typically have coarse resolutions of tens of kilometers and the degree of signal penetration beneath the surface complicates interpretations of height changes[ \cite{]@lenaerts_observing_2019].
+SMB is often highly spatially variable [@spikes_variability_2004; @banta_spatial_2008; @dattler_significant_2019], leading to questions of how representative small samples of measurements are of the ice sheet as a whole.
+Although satellite-based radar altimeters provide widespread spatial coverage, they typically have coarse resolutions of tens of kilometers and the degree of signal penetration beneath the surface complicates interpretations of height changes [@lenaerts_observing_2019].
 More recent lidar altimeters address some of these limitations, but have a limited temporal range [@smith_pervasive_2020].
@@
 Modeling results have seen enormous strides forward, but validation with independent estimates of SMB are still important, especially in data-sparse regions like West Antarctica [@lenaerts_observing_2019].
-Additionally, models still have too coarse of resolution (often 10's of kilometers) to resolve small-scale (\textless5 km) spatial variability [@lenaerts_climate_2018] and still have limitations on how well they can capture interannual variability in Antarctic SMB [@wang_comparison_2016.
-Radar has been used for decades to estimate SMB [@spikes_variability_2004, @anschutz_smallscale_2008, @eisen_groundbased_2008], but still suffers from limitations of homogeneous coverage and processing bottlenecks usually involving labor-intensive manual work, especially for reconstructing annually-resolved SMB time series [@medley_airborneradar_2013, @lenaerts_observing_2019.
+Additionally, models still have too coarse of resolution (often 10's of kilometers) to resolve small-scale (<5 km) spatial variability [@lenaerts_climate_2018] and still have limitations on how well they can capture interannual variability in Antarctic SMB [@wang_comparison_2016].
+Radar has been used for decades to estimate SMB [@spikes_variability_2004; @anschutz_smallscale_2008; @eisen_groundbased_2008], but still suffers from limitations of homogeneous coverage and processing bottlenecks usually involving labor-intensive manual work, especially for reconstructing annually-resolved SMB time series [@medley_airborneradar_2013; @lenaerts_observing_2019].
 
 In spite of these difficulties, substantial progress has been made towards improved understanding of SMB over recent decades in the WAIS.
-Numerous ice core studies suggest accumulation rates in and near the Antarctic Peninsula have recently increased [@thomas_doubling_2008, @thomas_308_2013, @thomas_twentieth_2015, @hosking_accumulation_2017, @medley_increased_2019].
-This observation is supported by results from a number of climate models and reanalyses \cite{wang_comparison_2016}.
-Evidence also exists for decreased accumulation over similar time periods for the western portions of WAIS, near the Ross Sea \cite{wang_comparison_2016, wang_snow_2017, medley_increased_2019}, although the \textit{in-situ} observations in this region are more sparse and largely out of date \cite{thomas_regional_2017}.
-As a whole, the central region of WAIS shows smaller changes in snow accumulation, but such results vary considerably both spatially across central WAIS and between studies \cite{banta_spatial_2008, medley_airborneradar_2013, burgener_observed_2013, wang_comparison_2016, thomas_regional_2017, wang_snow_2017}.
+Numerous ice core studies suggest accumulation rates in and near the Antarctic Peninsula have recently increased [@thomas_doubling_2008; @thomas_308_2013; @thomas_twentieth_2015; @hosking_accumulation_2017; @medley_increased_2019].
+This observation is supported by results from a number of climate models and reanalyses [@wang_comparison_2016].
+Evidence also exists for decreased accumulation over similar time periods for the western portions of WAIS, near the Ross Sea [@wang_comparison_2016; @wang_snow_2017; @medley_increased_2019], although the *in-situ* observations in this region are more sparse and largely out of date [@thomas_regional_2017].
+As a whole, the central region of WAIS shows smaller changes in snow accumulation, but such results vary considerably both spatially across central WAIS and between studies [@banta_spatial_2008; @medley_airborneradar_2013; @burgener_observed_2013; @wang_comparison_2016; @thomas_regional_2017; @wang_snow_2017].
 
@@
 The region spans a range of accumulation rates (>200 mm w.e. to <750 mm w.e.) and topographic settings, encompassing a portion of the ice divide between the Amundsen Sea and the Ross Sea sectors.
-Although some ice velocities in the coastal areas of this region can exceed 200 m/a, typical *in-situ* velocities for our data coverage are ~20 m/a or less \citep{rignot_ice_2011}.
+Although some ice velocities in the coastal areas of this region can exceed 200 m/a, typical *in-situ* velocities for our data coverage are ~20 m/a or less [@rignot_ice_2011].
 One reason this region is significant is because it encompasses portions of the Pine Island Glacier and Thwaites Glacier drainage basins.
-These outlet glaciers had a net mass loss of 1,700 gigatonnes of ice for the period 1979-2017, with the annual contribution also accelerating over that time period \cite{rignot_four_2019}.
+These outlet glaciers had a net mass loss of 1,700 gigatonnes of ice for the period 1979-2017, with the annual contribution also accelerating over that time period [@rignot_four_2019].
 This region additionally has an acceptably dense network of firn density measurements (a total of 16 cores) to properly constrain depth-density modeling (discussed in @sec-density) with increased confidence.
-Finally, this region is between the Antarctic Peninsula---where strong evidence exists of increased accumulation rates over recent decades---and the region near the Ross Sea---where evidence exists for decreased accumulation rates over recent decades \citep[e.g.]{medley_increased_2019, thomas_regional_2017}.
+Finally, this region is between the Antarctic Peninsula---where strong evidence exists of increased accumulation rates over recent decades---and the region near the Ross Sea---where evidence exists for decreased accumulation rates over recent decades e.g.
+[@medley_increased_2019; @thomas_regional_2017].
 It is therefore relevant to understanding the extent, boundaries, and patterns of changes in accumulation rates for the entire WAIS.
@@
 The main data used in this study are radar echograms from NASA's Operation IceBridge (OIB) Snow Radar.
-This frequency-modulated continuous-wave radar operates in a frequency range (2-8 GHz) with a range resolution (~5 cm in firn) that permits imaging of annual resolution in Antarctic firn \cite{panzer_ultrawideband_2013}.
+This frequency-modulated continuous-wave radar operates in a frequency range (2-8 GHz) with a range resolution (~5 cm in firn) that permits imaging of annual resolution in Antarctic firn [@panzer_ultrawideband_2013].
 We use portions of 9 OIB flights over the region of interest with collection times spanning 2010-2016.
@@
 Seasonal variations in the dielectric properties of the preserved firn result in internal reflection horizons (IRHs) in the radar returns.
-When continuous over many kilometers, these IRHs are most often interpreted as isochronous seasonal layers that can be used to determine the age of deposition \cite{eisen_groundbased_2008}.
+When continuous over many kilometers, these IRHs are most often interpreted as isochronous seasonal layers that can be used to determine the age of deposition [@eisen_groundbased_2008].
 To determine accumulation rates, the echogram time-domain must be converted to depth.
@@
 In addition to the radar data used, this study utilizes a suite of ice and firn cores, both for informing the the depth-density model and for comparisons of accumulation rates and trends generated with the radar data.
-We use density data from the Satellite Era Accumulation Traverse \cite{burgener_observed_2013}, the WAISCORES program \cite{lamorey_waiscores}, the International Trans-Antarctic Scientific Expedition \cite{mayewski_international_2005}, and additional cores included in the Surface Mass Balance and Snow on Sea Ice Working Group data set \cite{montgomery_sumup_2018}.
-For comparisons of annual accumulation rates and trends we additionally use cores included in the Past Global Changes Antarctica 2k compilation \cite{thomas_regional_2017}.
+We use density data from the Satellite Era Accumulation Traverse [@burgener_observed_2013], the WAISCORES program [@lamorey_waiscores], the International Trans-Antarctic Scientific Expedition [@mayewski_international_2005], and additional cores included in the Surface Mass Balance and Snow on Sea Ice Working Group data set [@montgomery_sumup_2018].
+For comparisons of annual accumulation rates and trends we additionally use cores included in the Past Global Changes Antarctica 2k compilation [@thomas_regional_2017].
 
@@
 The calculation of radar accumulation time series requires estimates of depth-density profiles to convert radar echograms from the time domain into the depth domain and to estimate annual accumulation from layer thickness.
-Various methods have been proposed for estimating the variation of density with depth in Antarctic firn, typically either interpolating from measured ice core or stake farm data \cite{medley_airborneradar_2013} or based on the physical modeling and evolution of firn properties \cite{ligtenberg_improved_2011}.
+Various methods have been proposed for estimating the variation of density with depth in Antarctic firn, typically either interpolating from measured ice core or stake farm data [@medley_airborneradar_2013] or based on the physical modeling and evolution of firn properties [@ligtenberg_improved_2011].
 
-In this study, we use a statistical estimation and interpolation approach informed by *in-situ* observations as described in \citet{white_hierarchical_2021}.
+In this study, we use a statistical estimation and interpolation approach informed by *in-situ* observations as described in @white_hierarchical_2021.
 This model uses a positive-valued Bayesian semi-parametric model over space and depth below the ice sheet as its building block.
@@
 The integrated function is transformed to be bounded below by 0 g/cm$^3$ and 0.917 g/cm$^3$ to constrain the model's mean function to appropriate values for firn/ice density.
-This model imposes smoothness of firn density estimates in the space and depth domains, constrains the shape of firn density curves, and enables simple interpolation (see \citet{white_hierarchical_2021} for more details).
+This model imposes smoothness of firn density estimates in the space and depth domains, constrains the shape of firn density curves, and enables simple interpolation (see @white_hierarchical_2021 for more details).
 There are several advantages afforded by this method, most importantly its ability to leverage space-depth autocorrelation to interpolate the shape-constrained depth-density profiles to new locations with relatively few model assumptions.
@@
 In this study, we estimate annually-resolved accumulation rates from OIB Snow radar using an automated process, the Probabilistic Annual Isochrone Picking Routine---i.e.
-PAIPR \cite{keeler_probabilistic_2020}.
+PAIPR [@keeler_probabilistic_2020].
 This algorithm picks continuous IRHs using localized Radon transforms and assigns each a likelihood of representing an annual isochrone based on the relative strength of the radar return and the continuous length of the horizon.
@@
 The complete period of data coverage therefore varies spatially due to varying accumulation rates, spanning from ten years to fifty-seven years of total record duration.
-Further details of the method are available in \citet{keeler_probabilistic_2020} and \citet{keeler_repeatability_2021}.
+Further details of the method are available in @keeler_probabilistic_2020 and @keeler_repeatability_2021.
 
@@
 The automated nature of PAIPR allows a wide and rapid application to larger OIB data sets.
-The addition of the quality control routine (see \citet{keeler_repeatability_2021} for details on its implementation) permits this application without careful vetting of input data, as poorly-suited data is flagged and can be removed from additional analyses.
+The addition of the quality control routine (see @keeler_repeatability_2021 for details on its implementation) permits this application without careful vetting of input data, as poorly-suited data is flagged and can be removed from additional analyses.
 The inclusion of uncertainties (resulting from both input data and modeling errors) allows a more rigorous assessment of the relevance and significance of observed magnitudes, patterns, and trends.
@@
 The shallow firn column used in PAIPR (~25 m) limits the impact of this issue, but such effects could still be relevant in regions of high ice velocity (e.g.
-\citet{mouginot_continentwide_2019} show some regions of WAIS near the coast can have >1 km a^-1^ ice velocities).
-\citet{konrad_observations_2019} however, found that even for the region of Pine Island Glacier, ~80% of strain rate correction magnitudes were lower than the uncertainty in radar-derived SMB estimates in the region.
+@mouginot_continentwide_2019 show some regions of WAIS near the coast can have >1 km a^-1^ ice velocities).
+@konrad_observations_2019 however, found that even for the region of Pine Island Glacier, ~80% of strain rate correction magnitudes were lower than the uncertainty in radar-derived SMB estimates in the region.
 
@@
 We utilize Integrated Nested Laplace Approximations INLA) to leverage the noted advantages.
-This method has recently seen widespread adoption in modeling spatiotemporal physical phenomena in a variety of fields \cite{}.
+This method has recently seen widespread adoption in modeling spatiotemporal physical phenomena in a variety of fields **(citations)**.
 
@@
 
-Estimates of mean annual accumulation (shown in \figref{accum-map}) are similar to those found by \citet{medley_airborneradar_2013} using manually traced IRHs from OIB Snow Radar echograms.
+Estimates of mean annual accumulation (shown in \figref{accum-map}) are similar to those found by @medley_airborneradar_2013 using manually traced IRHs from OIB Snow Radar echograms.
 Accumulation is highest (and exhibits the highest variability) close to coastal West Antarctica and progressively decreases towards the interior.
 Considerable spatial variability, however, overlays this broader signal.
-The most likely source of this higher-order spatial variability is accumulation and wind re-distribution governed by surface topography and near-surface winds \cite{dattler_significant_2019}.
+The most likely source of this higher-order spatial variability is accumulation and wind re-distribution governed by surface topography and near-surface winds [@dattler_significant_2019].
 
@@
   facet_wrap(vars(Param), scales = "free") + 
-  theme(text = element_text(size = 20))  
+  theme(text = element_text(size = 12))  
 ```
@@
 This is in agreement with a number of previous studies that found insignificant changes in accumulation during the latter portion of the 20th century.
-\citet{medley_airborneradar_2013} found no significant trends in SMB for the period 1980-2009 when compositing all OIB Snow Radar time series in their study for an overlapping region.
-\citet{wang_snow_2017} found little to no trend (0.18% per decade) in snow accumulation in central WAIS based on composite ice cores in the region for the period 1900-2010 and weakly positive trends (2.08% per decade) for the same region and time period using ERA-20C reanalysis data.
-\citet{wang_comparison_2016} compared the results of several climate models and reanalyses for the period 1979-2012, including ERA-Interim, JRA-55, MERRA, PMM5, RACMO2.1, and RACMO2.3.
+@medley_airborneradar_2013 found no significant trends in SMB for the period 1980-2009 when compositing all OIB Snow Radar time series in their study for an overlapping region.
+@wang_snow_2017 found little to no trend (0.18% per decade) in snow accumulation in central WAIS based on composite ice cores in the region for the period 1900-2010 and weakly positive trends (2.08% per decade) for the same region and time period using ERA-20C reanalysis data.
+@wang_comparison_2016 compared the results of several climate models and reanalyses for the period 1979-2012, including ERA-Interim, JRA-55, MERRA, PMM5, RACMO2.1, and RACMO2.3.
 They found considerable spatial variability in the linear trends in SMB for the central WAIS, including opposing signs of the trend, but none showed significant trends over this period in the region.
@@
 #| label: fig-trends
-#| fig-cap: "Mean posterior trends across the region with associated errors."
+#| fig-cap: "Posterior results for accumulation trends across the region."
+#| fig-subcap: 
+#|   - "Mean trends."
+#|   - "Standard deviation of trends"
+#| layout-ncol: 2
+
 plot(grid.rast$trend)
@@
 95% of accumulation changes in the region vary between -65 mm decade^-1^ and +7.2 mm decade^-1^ (-17% decade^-1^ and 2.1% decade^-1^ relative to the multi-decadal mean accumulation).
-The observed trend magnitudes are similar to those observed in ice and firn cores across the region and for a wider area of WAIS (\textbf{d} in \figref{trend-map}), but provide important context of the spatial variability and patterns in those trends.
+The observed trend magnitudes are similar to those observed in ice and firn cores across the region and for a wider area of WAIS (**d** in \figref{trend-map}), but provide important context of the spatial variability and patterns in those trends.
 Although individual cores suggest a similar pattern, they alone are insufficient to resolve the statistical significance of the overall results or to properly demonstrate the spatial nature and heterogeneity of the patterns.
 
-\citet{medley_increased_2019} used ice core compilations combined with reanalysis-derived spatial coherence patterns to estimate Antarctic-wide snow accumulation.
+@medley_increased_2019 used ice core compilations combined with reanalysis-derived spatial coherence patterns to estimate Antarctic-wide snow accumulation.
 Their results also show a strong east-west dipole in SMB over WAIS (thought to be driven by the Southern Annular Mode) in the period 1957-2000, with enhanced (decreased) precipitation towards the Antarctic Peninsula (Ross Sea).
-\citet{wang_new_2019} observed a similar dipole in SMB trends for the period 1950-2010 using a kriging reconstruction from ERA-Interim outputs informed by ice core records.
+@wang_new_2019 observed a similar dipole in SMB trends for the period 1950-2010 using a kriging reconstruction from ERA-Interim outputs informed by ice core records.
 Our results agree broadly with these conclusions, but indicate greater variability in the trend direction for central WAIS compared to either study, with significant negative trends for the period 1979-2010 extending farther east.
-These results are specifically in agreement with the findings of \citet{burgener_observed_2013}, who found decreasing accumulation since 1975 from several firn cores near WAIS Divide.
+These results are specifically in agreement with the findings of @burgener_observed_2013, who found decreasing accumulation since 1975 from several firn cores near WAIS Divide.
 
-\citet{thomas_regional_2017}, using composite ice core records and SMB from RACMO2.3p2, found that for the portion of WAIS between 90&deg;--150&deg; W, 50-year running linear trends varied mostly within $\pm1$ mm a^-1^, with the period 1961-2010 showing a trend of ~0.65 mm a^-1^.
-Our results---representing a region comprising ~1/3 of the WAIS section in \citet{thomas_regional_2017}---show a spatially integrated trend of -1.8 mm a^-1^ (-0.49% a^-1^) for the whole region of study, but larger variability spatially of -6.5 mm a^-1^ to +0.72 mm a^-1^.
+@thomas_regional_2017, using composite ice core records and SMB from RACMO2.3p2, found that for the portion of WAIS between 90&deg;--150&deg; W, 50-year running linear trends varied mostly within $\pm1$ mm a^-1^, with the period 1961-2010 showing a trend of ~0.65 mm a^-1^.
+Our results---representing a region comprising ~1/3 of the WAIS section in @thomas_regional_2017---show a spatially integrated trend of -1.8 mm a^-1^ (-0.49% a^-1^) for the whole region of study, but larger variability spatially of -6.5 mm a^-1^ to +0.72 mm a^-1^.
 Importantly, the WAIS regions between 90&deg;--150&deg; W not included in our results are mostly from the Ross Sea region.
-This region exhibits decreasing accumulation rates in recent decades \cite{wang_snow_2017, bertler_ross_2018, medley_increased_2019, wang_new_2019}, so its exclusion in our study is unlikely to explain the discrepancy between our results and \citet{thomas_regional_2017}.
+This region exhibits decreasing accumulation rates in recent decades [@wang_snow_2017; @bertler_ross_2018; @medley_increased_2019; @wang_new_2019], so its exclusion in our study is unlikely to explain the discrepancy between our results and @thomas_regional_2017.
 

### 2023-06-08 manuscript/draft.qmd
Adds new directory for reproducible manuscript

--- 2023-06-01 docs/article-draft.qmd
@@
 ---
-title: "Modeling Article Draft"
+title: "Article Draft"
 format: 
@@
 ![The region of interest in central West Antarctica and the data sets used for this study.
-Background image denotes 1-km resolution surface topography as a hillshade, derived from the DEM in \citet{slater_new_2018}.
+Background image denotes 1-km resolution surface topography as a hillshade, derived from the DEM in @slater_new_2018.
 Colored lines show elevation contours (150 m intervals).
 NASA OIB Snow radar results from PAIPR are shown in red and firn/ice core locations are denoted with blue triangles.
-The figure also denotes the location of WAIS Divide Camp.](Figures/data_panel.png){#fig-datamap}
+The figure also denotes the location of WAIS Divide Camp.](figures/data_panel.png){#fig-datamap}
 

### 2023-06-08 manuscript/processing-NB.qmd
Adds new directory for reproducible manuscript

---
title: "Processing Notebook"
format: 
  html:
    self-contained: true
editor: source
---

This notebook contains the processing workflow and instructions in order to reproduce the data, analyses, and plots included in the manuscript.

## Data pre-processing

Some pre-processing, formatting, and data wrangling are required prior to any analyses.
This includes taking the originally generated data (using PAIPR in MATLAB) and getting it into the proper format for processing in `R` using `INLA`.
Some of this processing occurs using `Python` scripts and some of it uses `R` scripts.
Both are detailed in this section.

1. Process the raw data output from PAIPR (performed in `scripts/preprocess.py`)
2. Compute topographic data necessary for later analyses (performed in `scripts/topo.calc.R`)
3. Clean the data and re-format for `R` processing (performed in `scripts/Rdata-prep.R`)
4. Build, run, and validate the `INLA` model (`scripts/inla-model.R`)
5.

### 2023-06-08 manuscript/scripts/Rdata-prep.R

@@ -1,0 +1,107 @@
+# Script to take python-preprocessed data (generated using `preprocess.py`) and perform further R processing.
+# The main points are to add additional topographic and climate covariates to the dfs of interest.
+
+# Import libraries
+library(here)
+library(tidyverse)
+library(sf)
+library(terra)
+
+# Import python data
+data_long = read_csv(here('data/paipr-long.csv')) %>% select(-...1) %>% 
+  mutate(Year = as.integer(Year))
+gdf_traces = st_read(here('data/traces.geojson')) %>% as_tibble() %>% st_as_sf()
+
+# Remove points in far northern extent
+accum_bnds = st_coordinates(gdf_traces)
+gdf_traces = gdf_traces %>% st_crop(xmin=-1.5e6, xmax=max(accum_bnds[,1]),
+                                    ymin=min(accum_bnds[,2]), ymax=max(accum_bnds[,2]))
+
+# # Small-scale testing bounds
+# gdf_traces = gdf_traces %>% st_crop(xmin=-1.4e6, xmax=-1.25e6,
+#                                     ymin=min(accum_bnds[,2]), ymax=-3e5)
+
+# Remove unrealistically high elevations and overly large accum sites
+trace_drop = gdf_traces %>% filter(elev > 3000 | accum > 1000)
+
+# Add East/North coordinates and drop filtered sites
+pts_tmp = st_coordinates(gdf_traces)
+coord_df = tibble(trace_ID = gdf_traces$trace_ID, East = pts_tmp[,1], North = pts_tmp[,2])
+data = data_long %>% left_join(coord_df) %>% filter(!trace_ID %in% trace_drop$trace_ID)
+gdf_traces = gdf_traces %>% filter(!trace_ID %in% trace_drop$trace_ID)
+
+# Import REMA dem data
+REMA.dir = file.path("/media/durbank/WARP/Research/Antarctica/Data/DEMs/REMA/REMA_200m")
+dem = rast(file.path(REMA.dir, "REMA_200m_dem.tif"))
+crs(dem) = st_crs(gdf_traces)$wkt
+
+# Import REMA slope data (calculated using topo.calc script)
+slp = rast(file.path(REMA.dir, "REMA_200m_slope.tif"))
+crs(slp) = crs(dem)
+
+# Import REMA aspect data (calculated using topo.calc script)
[…]

### 2023-06-08 manuscript/scripts/inla-model.R

@@ -1,0 +1,305 @@
+# Script to generate data and perform analyses for article on 
+# Bayesian inference of WAIS-central accumulation trends
+
+# Import libraries
+library(here)
+library(INLA)
+library(dplyr)
+library(tidyr)
+library(broom)
+library(terra)
+library(spdep)
+
+## DATA IMPORT AND PREPROCESSING
+
+# Import raw(ish) data
+data.all = readRDS(here('data/Rdata-clean.rds'))
+gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
+
+# # Clip bounds
+# bbox = st_bbox(gdf_traces)
+# clipper = c(
+#   bbox[1]+0.1*(bbox[3]-bbox[1]),
+#   bbox[2]+0.5*(bbox[4]-bbox[2]),
+#   0.6*(bbox[3]-bbox[1])+bbox[1], 
+#   bbox[4]
+# )
+# gdf_traces = st_crop(gdf_traces, clipper)
+
+# Remove locations with less than 10 years of data
+trace.keep = data.all %>% count(trace_ID) %>% filter(n>=10)
+data.all = data.all %>% filter(trace_ID %in% trace.keep$trace_ID)
+gdf_traces = gdf_traces %>% filter(trace_ID %in% trace.keep$trace_ID)
+
+# Subset data to every 5th location (coarsens data to ~2 km)
+skip.int = 10
+gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
+gdf_traces = gdf_traces[gdf.idx,]
+
+# Filter data to subsetted period
+data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
+  filter(Year >= 1975) %>%
+  filter(Year < 2015) %>% arrange(trace_ID, Year)
+
+# Remove gdf rows where all data have been filtered out
+gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)
+
+# Select variables of interest
+dat = data %>% select(trace_ID, East, North, Year, accum, 
+                      std, SMB, elev, dem, slope, aspect, 
+                      # u10m, v10m, mu.u10, mu.v10
+)
+
+# Center covariates
+dat = dat %>% 
+  mutate(elev = elev-mean(elev, na.rm=TRUE), 
+         dem=dem-mean(dem, na.rm=TRUE), slope=slope-mean(slope, na.rm=TRUE), 
[…]

### 2023-06-08 scripts/article-script.R

@@ -250,6 +250,7 @@ add(grid.rast) <- r.tmp
 dem.fx = (as.matrix(grid.rast$DEM, wide=TRUE) - mean(data$dem))/sd(data$dem)
 
 # Construct time random effect array
+Years = 1975:2014
 tmp = replicate(dim(grid.rast)[2], 
                 matrix(rep(mod.valid$summary.random$time$mean,
                            each=dim(grid.rast)[1]),
@@ -257,7 +258,6 @@ tmp = replicate(dim(grid.rast)[2],
 time.RE = aperm(tmp, c(2,3,1))
 
 # Construct linear predictor for each spacetime grid cell
-Years = 1975:2014
 yr.mod = Years - mean(Years)
 lin.est = mod.valid$summary.fixed$mean[1] + 
   mod.valid$summary.fixed$mean[2] * replicate(length(Years), dem.fx) + 

### 2023-06-13 manuscript/processing-NB.qmd
Experiments in model fitting

@@
 4. Build, run, and validate the `INLA` model (`scripts/inla-model.R`)
-5.
+
+## Analysis
+
+I first import the requisite libraries
+
+```{r}
+library(here)
+library(INLA)
+library(dplyr)
+library(ggplot2)
+library(terra)
+library(colorspace)
+library(sf)
+```
+
+and the model results generated previously.
+
+```{r}
+# Load necessary data for analysis and plotting (generated using `scripts/article-script.R)
+mod.full = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/interim-models/mod-full.skip10.cut5.rds')
+valid.df = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/dfs/valid.df.rds')
+results.df = readRDS('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/dfs/results.df.rds')
+grid.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/grid.tif')
+mu.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/mu.tif')
+sd.rast = rast('/media/durbank/WARP/Research/Antarctica/WAIS-central/data/rasters/sd.tif')
+```
+
+I define a helper function to transform the posteriors to more intuitive scales for improved plots.
+```{r}
+t.marg = function(mod.marg, t.func, param.names=NULL, n=100) {
+
+  if (any(class(mod.marg) == "matrix")) {
+    mod.marg = list(mod.marg)
+    if (!is.null(param.names)) {
+      names(mod.marg = param.names)
+    } else {
+      names(mod.marg) = "UNNAMED"
+    }
+  }
+
+  if (is.null(param.names)) {
+    param.names = names(mod.marg)
+  }
+
+    if (is.list(t.func) != TRUE) {
+    t.func = list(t.func)
+  }
+  if (length(t.func) == 1) {
+    t.func = rep(t.func, length(param.names))
+  }
+
+  # Preallocate output tbl
+  out = tibble(x=numeric(), y=numeric(), Param=vector(mode="character"))
+
+  marg.names = names(mod.marg)
+
+  for (i in 1:length(marg.names)) {
+    marg.i = inla.tmarginal(t.func[[i]], mod.marg[[marg.names[i]]], n=n)
+    out = out %>% bind_rows(tibble(x=marg.i[,1], y=marg.i[,2], Param=param.names[i]))
+  }
+  return(out)
+}
+```
+
+```{r}
+# Extract parameter posteriors
+marg.fixed = t.marg(mod.full$marginals.fixed, function(x) exp(x))
+HP.names = c("Gamma precision", 
+             "Time precision", "Rho for time", 
+             "Range for spat.idx", "Stdev for spat.idx")
+marg.hyper = t.marg(mod.full$marginals.hyperpar, function(x) (x), param.names = HP.names)
+p.selection = c("Intercept", "dem", "Gamma precision", 
+                "Time precision", "Rho for time", 
+                "Range for spat.idx", "Stdev for spat.idx")
+posts = marg.fixed %>% bind_rows(marg.hyper) %>% filter(Param %in% p.selection)
+```
+
+```{r}
+#| label: fig-ParamPost
+#| fig-cap: "Posterior distributions of model parameters. Blah blah blah."
+ggplot(posts, aes(x=x,y=y)) + geom_line() + 
+  facet_wrap(vars(Param), scales = "free") + 
+  theme(text = element_text(size = 12))  
+```
+
+```{r}
+#| label: fig-trends
+#| fig-cap: "Posterior results for accumulation trends across the region."
+#| fig-subcap: 
+#|   - "Mean trends."
+#|   - "Standard deviation of trends"
+#| layout-ncol: 2
+
+plot(grid.rast$trend)
+plot(grid.rast$trend.sd)
+```

### 2023-06-13 manuscript/scripts/inla-model.R

@@ -26,27 +26,27 @@ gdf_traces = readRDS(here('data/Rdata-gdf_trace.rds'))
 # )
 # gdf_traces = st_crop(gdf_traces, clipper)
 
-# Remove locations with less than 10 years of data
-trace.keep = data.all %>% count(trace_ID) %>% filter(n>=10)
+# Remove locations with less than 12 years of data
+trace.keep = data.all %>% count(trace_ID) %>% filter(n>=12)
 data.all = data.all %>% filter(trace_ID %in% trace.keep$trace_ID)
 gdf_traces = gdf_traces %>% filter(trace_ID %in% trace.keep$trace_ID)
 
-# Subset data to every 5th location (coarsens data to ~2 km)
-skip.int = 10
+# Subset data to every 5th location (coarsens data to ~1 km)
+skip.int = 5
 gdf.idx = seq(1, nrow(gdf_traces), by=skip.int)
 gdf_traces = gdf_traces[gdf.idx,]
 
 # Filter data to subsetted period
-data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) %>%
-  filter(Year >= 1975) %>%
-  filter(Year < 2015) %>% arrange(trace_ID, Year)
+data = data.all %>% filter(trace_ID %in% gdf_traces$trace_ID) #%>%
+  # filter(Year >= 1975) %>%
+  # filter(Year < 2015) %>% arrange(trace_ID, Year)
 
 # Remove gdf rows where all data have been filtered out
 gdf_traces = gdf_traces %>% filter(trace_ID %in% data$trace_ID)
 
 # Select variables of interest
-dat = data %>% select(trace_ID, East, North, Year, accum, 
-                      std, SMB, elev, dem, slope, aspect, 
+dat = data %>% select(trace_ID, East, North, Year, accum,
+                      std, SMB, elev, dem, slope, aspect,
                       # u10m, v10m, mu.u10, mu.v10
 )
 
@@ -57,7 +57,8 @@ dat = dat %>%
          # u10m=u10m-mean(u10m, na.rm=TRUE), v10m=v10m-mean(v10m, na.rm=TRUE), 
          # mu.u10=mu.u10-mean(mu.u10, na.rm=TRUE), mu.v10=mu.v10-mean(mu.v10, na.rm=TRUE)
   ) %>% 
-  mutate(Year.mod = Year-mean(1975:2014), Year.idx = (Year-min(Year)+1))
[…]
